### 2014-11-11 appendix-ec.tex
This is what we'll start with.

\section{Expansion Constant Properties}

I talked to someone who wasn't convinced that all of the following useful lemmas
were true, so I went out and proved them.  Now I've typeset them.

\begin{lemma}
\label{lem:ec_rec}
Given an expansion constant $c$ as defined in Definition \ref{def:int_dim}, for any
ball $B_S(p, 2^i \Delta)$ where $p \in S$, $i \in \mathbb{N}$, $\Delta > 0$,

\begin{equation}
| B_S(p, 2^i \Delta) | \le c^i | B_S(p, \Delta) |.
\end{equation}
\end{lemma}

\begin{proof}
This follows from $i$ recursive applications of the definition of the expansion
constant.
\end{proof}

\begin{lemma}
\label{lem:ec_re}
Given an expansion constant $c$ as defined in Definition \ref{def:int_dim}, for any
ball $B_S(p, \alpha \Delta)$ where $p \in S$, $\alpha \in \Real$, $\alpha \ge 1$,
$\Delta > 0$,

\begin{equation}
| B_S(p, \alpha \Delta) | \le c^{\log_2 \alpha} | B_S(p, \Delta) |.
\end{equation}
\end{lemma}

\begin{proof}
By Definition \ref{def:int_dim} and Lemma \ref{lem:ec_rec}, it is obvious that
the statement holds when $\alpha \in \mathbb{N}$.  However, the definition of
expansion constant makes no particular bound on the number of points added when
the ball is not doubled but simply increased by some real factor.  We will
perform this proof by contradiction.

Assume that there exists some $\alpha \in \mathbb{R}$ where $\alpha > 1$ and

\begin{equation}
| B_S(p, \alpha \Delta) | > c^{\log_2 \alpha} | B_S(p, \Delta) |.
\end{equation}

By Definition \ref{def:int_dim}, we also know that

\begin{equation}
| B_S(p, \alpha \Delta) | \le c | B_S(p, (\alpha / 2) \Delta) |.
\end{equation}

Combining these two equations gives that

\begin{equation}
c | B_S(p, (\alpha / 2) \Delta) | > c^{\log_2 \alpha} | B_S(p, \Delta) |
\end{equation}

\noindent or alternately

\begin{equation}
c^{(1 - \log_2 \alpha)} | B_S(p, (\alpha / 2) \Delta) | > | B_S(p, \Delta) |.
\end{equation}

First, assume that $\alpha < 2$.  Then, $|B_S(p, (\alpha / 2) \Delta)| \le
|B_S(p, \Delta)|$.  Thus,

\begin{equation}
c^{(1 - \log_2 \alpha)} | B_S(p, \Delta) | > | B_S(p, \Delta) |
\end{equation}

\noindent which is a contradiction because $c^{(1 - \log_2 \alpha)} > 1$.

Now, assume that $\alpha > 2$.  Define $\beta = 2^i \alpha$ where $i \in
\mathbb{N}$ is such that $1 < \beta < 2$.  Then,

\begin{equation}
| B_S(p, (2^i \beta / 2) \Delta) | \le c^i | B_S(p, (\beta / 2) \Delta) |.
\end{equation}

Because $\beta < 2$, $|B_S(p, (\beta / 2) \Delta)| \le |B_S(p, \Delta)|$.  Thus,

\begin{equation}
|B_S(p, (2^i\beta / 2) \Delta)| \le c^i | B_S(p, \Delta) |.
\end{equation}

From earlier, this implies that

\begin{equation}
\label{eqn:contr}
(c^{(1 - \log_2 (2^i \beta))}) (c^i) |B_S(p, \Delta) | > |B_S(p, \Delta) |.
\end{equation}

Note that $(c^{(1 - \log_2 (2^i \beta))}) (c^i) = c^{(1 - \log_2 \beta)}$, which
is greater than $1$ because $\beta < 2$.  This means that Equation
\ref{eqn:contr} is a contradiction, and thus, for all $\alpha \in \mathbb{R}$,
the statement of the lemma holds.
\end{proof}

\begin{lemma}
\label{lem:ec_neg}
Given an expansion constant $c$ as defined in Definition \ref{def:int_dim}, for
any ball $B_S(p, 2^{-i} \Delta)$ where $p \in S$, $i \in \mathbb{N}$, $\delta >
0$,

\begin{equation}
|B_S(p, 2^{-i} \Delta)| \ge c^{-i} |B_S(p, \Delta)|.
\end{equation}
\end{lemma}

\begin{proof}
Observe that $|B_S(p, \Delta)| \le c^i |B_S(p, 2^{-i} \Delta)|$.  Simple
algebraic manipulation yields the result.
\end{proof}

\begin{lemma}
Given an expansion constant $c$ as defined in Definition \ref{def:int_dim}, for
any ball $B_S(p, \alpha \Delta)$ where $p \in S$, $\alpha \in \mathbb{R}$,
$0 < \alpha \ge 1$, $\delta > 0$,

\begin{equation}
|B_S(p, \alpha \Delta)| \ge c^{\log_2 \alpha} |B_S(p, \Delta)|.
\end{equation}
\end{lemma}

\begin{proof}
(idea) This is a relatively straightforward combination of the proofs of
Lemmas \ref{lem:ec_neg} and \ref{lem:ec_re}.
\end{proof}

Two important observations:

\begin{itemize}
  \item We can place an upper bound on the number of points added as a ball
grows an arbitrary amount.

  \item We can place a lower bound on the number of points removed as a ball
shrinks an arbitrary amount.
\end{itemize}

### 2014-11-11 bounds-files/conclusion.tex
This is what we'll start with.

We have presented a unified framework for bounding the runtimes of dual-tree
algorithms that use cover trees and the standard cover tree pruning dual-tree
traversal (Algorithm \ref{alg:cover-tree-dual}).  Our main result, Theorem
\ref{thm:ct-runtime}, allows plug-and-play runtime bounding of these algorithms.
We have shown that Theorem \ref{thm:ct-runtime} is useful for bounding the
runtime of nearest neighbor search (Theorem \ref{thm:nns}), approximate kernel
density estimation (Theorem \ref{thm:kde-bound}), exact range count, and exact
range search (Theorem \ref{thm:rs}).  With our contribution, bounding a cover
tree dual-tree algorithm is straightforward and only involves bounding the
maximum size of the reference set, $| R^* |$.

### 2014-11-11 bounds-files/general-runtime-bound.tex
This is what we'll start with.

\begin{thm}
\label{thm:ct-runtime}
%For any dual-tree algorithm using cover trees and the standard cover tree
%dual-tree pruning traversal (Algorithm \ref{alg:cover-tree-dual}) with reference
%set $S_r$ that has expansion constant $c_r$ and query set $S_q$, the runtime is
%bounded above by $O(c_r^4 | R^* | \nu \chi \psi N)$, where $ | R^* | $ is
%the maximum size of the reference set $R$ (line
%\ref{alg:line:ct-dual-input}) at any point in the dual-tree pruning traversal,
%$\chi$ is the longest possible running time of \texttt{BaseCase()}, and $\psi$
%is the longest possible running time of \texttt{Score()}.
Given a reference set $S_r$ of size $N$ with an expansion constant $c_r$ and a
set of queries $S_q$ of size $O(N)$, a standard cover tree based dual-tree
algorithm (Algorithm \ref{alg:cover-tree-dual}) takes $O(c_r^4 | R^* | \nu \chi
\psi N)$, where $ | R^* | $ is the maximum size of the reference set $R$ (line
\ref{alg:line:ct-dual-input}) during the dual-tree recursion, $\chi$ is
the maximum possible runtime of \texttt{BaseCase()}, and $\psi$ is the maximum
possible runtime of \texttt{Score()}.
\end{thm}

\vspace*{-1.23em}
\begin{proof}
Consider a reference recursion (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end}).
The work done in the base case loop from lines
\ref{alg:line:ct-dual-base-case-start}--\ref{alg:line:ct-dual-base-case-end} is
$O(\chi | R |)$.  Define $| R^* |$ to be the largest set $|R|$ for any scale
$s_r^{\max}$ and any query node $\mathscr{N}_q$ during the course of the
algorithm; then, it is true that $| R | \le | R^* |$.
%
Then, lines \ref{alg:line:ct-dual-ref-children} and
\ref{alg:line:ct-dual-ref-score} take $O(c_r^4 \psi | R |) \le O(c_r^4 \psi |
R^* |)$ time, because each reference node has $c_r^4$ children.  So, one
full reference recursion takes $O(c_r^4 \psi | R^* |)$ time.

Now, note that there are $O(N)$ nodes in $\mathscr{T}_q$.  Thus, line
\ref{alg:line:ct-dual-query-recursion} is visited $O(N)$ times.  Each of these
$O(N)$ visits to line \ref{alg:line:ct-dual-query-recursion} implies a
recursion, in which the reference set is descended up to $\nu$ times (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
before the query node is descended or the algorithm terminates.  In addition,
each $O(N)$ recursion implies an $O(\psi |R|) \le O(\psi |R^*|)$ operation for
the calculation of $R'$ (line \ref{alg:line:ct-dual-query-pruning}).  Thus, the
full runtime of the algorithm is bounded as $O(c_r^4 |R^*| \nu \chi \psi N +
\psi |R^*| N) = O(c_r^4 |R^*| \nu \chi \psi N)$.
\end{proof}
\vspace*{-0.8em}

This result holds for any dual-tree algorithm regardless of the problem. Hence,
%for any pairwise statistical problem, the runtime of the dual-tree algorithm (of
the runtime of any dual-tree algorithm
%that problem)
would be at least $O(N)$ using our bound, which matches the intuition that
answering $O(N)$ queries would take at least $O(N)$ time. For a particular
problem and data, if $c_r$, $|R^*|$, $\nu$, $\chi$ and
$\psi$ are bounded by constants independent of $N$ (for large enough $N$), then
the dual-tree algorithm for that problem has a runtime linear in $N$. Our
theoretical result separates out the problem-dependent and the
problem-independent elements of the runtime bound, which allows us to simply
plug in the problem-dependent bounds to get runtime bounds for any dual-tree
algorithm without requiring an analysis from scratch.
% This discussion already happened earlier... sort of.
%The expansion constant is related to a notion of intrinsic dimensionality of a
%dataset (as discussed earlier). A relatively small $c_r$ corresponds to
%a dataset with low intrinsic dimensionality, while a value of $c_r$ independent
%of $N$ implies that the intrinsic dimensionality of the data is independent of
%the number of points in the dataset. The latter is a reasonable assumption in
%many scenarios.
%Our bound has a linear dependence on the inverse constant of bichromaticity
%$\nu$; this results from the worst-case situation where the reference tree is
%recursed entirely after all query tree recursions
% this makes sense, given that $\nu$ represents the maximum number of times
%any set $R$ is descended for any query node $\mathscr{N}_q$ before
%$\mathscr{N}_q$ is recursed.
If $\nu \sim \kappa$, which is a reasonable assumption, this is a better result
than the results of Ram et.~al.  \cite{ram2009} who found a dependence of
$c_q^{4 \kappa}$ in their proofs.

The quantity $|R^*|$ bounds the amount of work that needs to be done for each
recursion. In the worst case, $|R^*|$ can be $N$. However,
dual-tree algorithms rely on branch-and-bound techniques to prune away
work (Lines \ref{alg:line:ct-dual-ref-score} and
\ref{alg:line:ct-dual-query-pruning} in Algorithm \ref{alg:cover-tree-dual}). A
small value of $|R^*|$ will imply that the algorithm is extremely successful in
pruning away work at Line \ref{alg:line:ct-dual-ref-score} in Algorithm
\ref{alg:cover-tree-dual}. An (upper) bound on $|R^*|$ (and the algorithm's
success in pruning work) will depend on the problem and the data.  As we will
show, bounding $|R^*|$ is often possible. % In addition, we will also show that
For many dual-tree algorithms, $\chi \sim \psi \sim O(1)$; often, cached
sufficient statistics~\cite{moore2000anchors} can enable $O(1)$ runtime
implementations of \texttt{BaseCase()} and \texttt{Score()}.

% Space constraints :(

%The runtime bound $\chi$ for the \texttt{BaseCase($p_q$, $p_r$)} function
%corresponds to the time required to process a pair of points. For example, in
%range search, \texttt{BaseCase()} corresponds to the
%time taken to compute $d(p_q, p_r)$ and determine if that falls into the desired
%range.  For general kernel summations, this corresponds to computing the
%pairwise kernel function between $p_q$ and $p_r$ and then summing it to the
%output for $p_q$.  In general, the runtime of \texttt{BaseCase()} is dominated
%by the pairwise function computation which is independent of $N$ and is usually
%$O(1)$. The \texttt{Score()} function is an operation on a pair of nodes and
%usually involves computing a bound which requires a pairwise function
%computation between the node centers (similar to the \texttt{BaseCase()}
%function except that it is for node centres) and certain cached node statistics
%(such as node radius in the metric space). Hence its runtime bound $\psi$ is
%usually a constant independent of $N$ or simply $O(1)$.

### 2014-11-11 bounds-files/sec01-supporting-lemmas.tex
This is what we'll start with.

For simplicity, the algorithms considered in this paper will be presented in a
tree-independent context, as in \cite{curtin2013tree}, but the only type of
tree we will consider is the cover tree \cite{langford2006}, and the only type
of traversal we will consider is the cover tree pruning dual-tree traversal, which
we will describe below.

As we will be making heavy use of trees, we must establish notation (taken from
\cite{curtin2013tree}).  $\mathscr{T}_i$ represents a tree; $\mathscr{N}_i$
represents a node; $\mathscr{C}(\mathscr{N}_i)$ is the set of children of the
node $\mathscr{N}_i$; $\mathscr{P}(\mathscr{N}_i)$ is the set of points held in
a node; and $\mathscr{D}^p(\mathscr{N}_i)$ is the set of all descendant points
of the node $\mathscr{N}_i$.

First, we will introduce the cover tree\footnote{In our discussions, we will
only consider the explicit cover tree, as opposed to the implicit cover tree
\cite{langford2006}.  This has the effect of making the traversal more complex,
but it helps simplify our proofs.  Familiarity with the implicit cover tree
representation should not
be required for understanding any of our work here.}.  The cover tree is a
leveled hierarchical data structure originally proposed for the task of nearest
neighbor search by Beygelzimer et~al.~\cite{langford2006}.  Each cover tree node
holds one point $p_i$ and may have a number of child nodes; if the cover tree
node has children, then one child will be the self-child, which holds the same
point.  In addition, each level in the tree is indexed by an integer scale
$s_i$; as the tree is descended, $s_i$ decreases.  Every descendant of a node
$\mathscr{N}_i$ has a maximum distance of $2^{s_i + 1}$ from $p_i$.

Assuming a cover tree $\mathscr{T}$ is built on a dataset $S_r$, let $C_{s_i}$
denote the set of points held in all nodes in $\mathscr{T}$ with scale $s_i$.
Then, the cover tree satisfies the following invariants for all scales $s_i$:

\vspace*{-0.5em}
\begin{itemize}

\item {\em (Nesting)}.  $C_{s_i} \subset C_{s_i - 1}$.  This means that when a
point $p_i \in S_r$ is held in a node at some scale $s_i$, then each smaller
scale will also have a node containing $p_i$.

\item {\em (Covering tree)}.  For every point $p_i \in C_{s_i - 1}$, there
exists a point $p_j \in C_{s_i}$ such that $d(p_i, p_j) < 2^{s_i}$, and the node
holding $p_j$ with scale $s_i$ is the parent node of the node holding $p_i$ with
scale $s_i - 1$.

\item {\em (Separation)}.  For all distinct points $p_i, p_j \in C_{s_i}$,
$d(p_i, p_j) > 2^{s_i}$.

\end{itemize}
\vspace*{-0.5em}

A batch construction algorithm is given in \cite{langford2006}.  In addition,
the cover tree has a number of other useful properties and lemmas based on the
expansion constant \cite{karger2002finding}; we restate the definition below.

\begin{defn}
\label{def:int_dim}
Let $B_S(p, \Delta)$ be the set of points in $S$ within a closed ball of radius
$\Delta$ around some $p \in S$ with respect to a metric $d$:
%
$B_S(p, \Delta) = \{ r \in S \colon d(p, r) \leq \Delta \}$.
%
Then, the {\bf expansion constant} of $S$ with respect to the metric $d$ is the
smallest $c \ge 2$ such that

\vspace*{-0.8em}
\begin{equation}
| B_S(p, 2 \Delta) | \le c | B_S(p, \Delta) |\ \forall\ p \in S,\
\forall\ \Delta > 0.
\end{equation}
\vspace*{-1.4em}

\end{defn}

The expansion constant is used heavily in the cover tree literature.  It is
related to a notion of instrinic dimensionality, and in many scenarios, $c$ is
independent of the number of points in the dataset \cite{langford2006,
karger2002finding}.
%properties of cover trees in \cite{langford2006}.
One result we will use is that
the number of children of any cover tree node is $c^4$ (Lemma 4.1,
\cite{langford2006}), where $c$ is the expansion constant of the dataset the
cover tree is built on.

%\begin{lemma}
%\label{lem:width}
%(Lemma 4.1, \cite{langford2006}) The number of children of any cover tree node $\mathscr{N}_i$ is bounded by
%$c^4$, where $c$ is the expansion constant of the dataset the cover tree is
%built on, as defined in Definition \ref{def:int_dim}.
%\end{lemma}

%\begin{lemma}
%\label{lem:depth}
%(Lemma 4.3, \cite{langford2006}) The maximum depth of any point $p_r$ in a cover
%tree $\mathscr{T}_r$ is $O(c^2 \log N)$, where $N$ is the number of points in
%the dataset that $\mathscr{T}_r$ is built on.
%\end{lemma}

Next we show the standard cover tree dual-tree pruning traversal, in a
problem-independent form.  This is given in Algorithm \ref{alg:cover-tree-dual}.
The algorithm is adapted from the original cover tree batch nearest neighbors
algorithm \cite{langford2006} and implements the dual-tree recursion for the
cover tree.  This traversal was first presented by Curtin and Ram
\cite{curtin2014dual}.
%
The correctness of this
traversal for various problem-dependent choices of the \texttt{BaseCase()} and
the \texttt{Score()} functions has been proven \cite{curtin2013tree}.
%  Now, we must introduce the inverse degree of bichromaticity.

\begin{algorithm}[htb]
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, set of reference nodes
$R$ \label{alg:line:ct-dual-input}
    \STATE {\bfseries Output:} none

    \medskip
    \STATE $s^{\max}_r \gets \max_{\mathscr{N}_r \in R} s_r$
    \IF{$(s_q < s^{\max}_r)$ {\bf or} $(s^{\max}_r = -\infty)$}
\label{alg:line:ct-dual-ref-recursion-start}
      \FORALL{$\mathscr{N}_r \in R$} \label{alg:line:ct-dual-base-case-start}
        \STATE \texttt{BaseCase($p_q$, $p_r$)}
      \ENDFOR \label{alg:line:ct-dual-base-case-end}
      \STATE $R_r \gets \{ \mathscr{N}_r \in R : s_r = s^{\max}_r \}$
\label{alg:line:ct-dual-ref-set}
      \STATE $R_{r - 1} \gets \{ \mathscr{C}(\mathscr{N}_r) : \mathscr{N}_r \in
R_r \} \cup (R \setminus R_r)$ \label{alg:line:ct-dual-ref-children}
      \STATE $R'_{r - 1} \gets \{ \mathscr{N}_r \in R_{r - 1} :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-ref-score}
      \STATE recurse with $\mathscr{N}_q$ and $R'_{r - 1}$
\label{alg:line:ct-dual-ref-recursion-end}
    \ELSE \label{alg:line:ct-dual-query-recursion-start}
      \FORALL{$\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q)$}
        \STATE $R' \gets \{ \mathscr{N}_r \in R :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-query-pruning}
        \STATE recurse with $\mathscr{N}_{qc}$ and $R'$
\label{alg:line:ct-dual-query-recursion}
      \ENDFOR \label{alg:line:ct-dual-query-recursion-end}
    \ENDIF
  \end{algorithmic}
  \caption{The standard pruning dual-tree traversal for cover trees.}
  \label{alg:cover-tree-dual}
\end{algorithm}

\begin{defn}
\label{def:bichromaticity}
Let $\mathscr{T}_q$ and $\mathscr{T}_r$ be two cover trees built on query set
$S_q$ and reference set $S_r$, respectively.  Now consider a pruning dual-tree
traversal (such as Algorithm \ref{alg:cover-tree-dual}) with the property that
the scales of nodes in $\mathscr{T}_q$ and $\mathscr{T}_r$ are kept as close as
possible---that is, the tree with the larger scale is always descended.  Then,
the {\bf inverse degree of bichromaticity} $\nu$ of the tree pair
$(\mathscr{T}_q, \mathscr{T}_r)$ is the maximum number of recursions in
$\mathscr{T}_r$ following a recursion in $\mathscr{T}_q$ before another
recursion in $\mathscr{T}_q$ or the termination of the algorithm (whichever
happens first).

This quantity is related to the {\bf degree of bichromaticity} $\kappa$
\cite{ram2009}, which is the maximum number of recursions in $\mathscr{T}_q$
between any two recursions in $\mathscr{T}_r$.
\end{defn}

Although $\kappa$ and $\nu$ can be difficult to bound, it is reasonable to
%monochromatic case where $S_q = S_r$, the trees $\mathscr{T}_q$ and
%$\mathscr{T}_r$ are identical, so $\kappa = \nu = 1$.  It is reasonable to
say that $\kappa \sim \nu \sim O(1)$ for datasets with roughly similar
distributions of pairwise distances.

\begin{lemma}
Consider a dataset $S$ with expansion constant $c$ and a subset $C \subseteq S$
such that every point in $C$ is separated by $\delta$.  Then, given any query
point $p \not\in S$ and some radius $\rho \delta$,

\vspace*{-0.7em}
\begin{equation}
| B_S(p, \rho \delta) \cap C | \le c^{2 + \log_2 \rho}.
\end{equation}
\vspace*{-1.1em}
\label{lem:packing}
\end{lemma}

\vspace*{-1em}
\begin{proof}
This is based on the packing argument from Lemma 4.1 in \cite{langford2006}.
Observe that $B_S(p, \rho \delta) \subseteq B_S(p_i, 2 \rho \delta)$ for any
$p_i \in S$, and that $| B_S(p, 2 \rho \delta) | = c^{2 + \log_2 \rho} |
B_S(p, \delta / 2) |$.  Because each point in $C$ is separated by $\delta$, the
number of points in $B_S(p, \rho \delta) \cap C$ is
bounded by the number of disjoint balls of radius $\delta / 2$ that can be
packed into $B_S(p, \rho \delta)$.  In the worst case, this packing is
perfect, and

\vspace*{-0.7em}
\begin{equation}
|B_S(p, \rho \delta)| \le \frac{|B_S(p_i, 2 \rho \delta)|}{|B_S(p_i, \delta
/ 2)|} \le c^{2 + \log_2 \rho}.
\end{equation}\end{proof}
\vspace*{-2.2em}

### 2014-11-11 bounds-files/sec02-nearest-neighbor-search.tex
This is what we'll start with.

Consider nearest neighbor search, given as a \texttt{BaseCase()} function
(Algorithm \ref{alg:nn_base_case}) and a \texttt{Score()} function (Algorithm
\ref{alg:nn_score}).  The goal is, for every query point $p_q$, find the nearest
neighbor $p_r$ in the reference set $S_r$.  This algorithm stores the current
nearest neighbor candidates in the array $N$ and their distances in the array
$D$.\footnote{See Curtin et.~al. \cite{curtin2013tree} for a more complete
discussion of how this algorithm works and a proof of correctness.}  The
\texttt{Score()} function depends on a bound function $B(\mathscr{N}_q)$ which
represents the maximum distance that could possibly improve a nearest neighbor
candidate for any descendant point of the query node $\mathscr{N}_q$.  The
standard bound function $B(\mathscr{N}_q)$ used for cover trees is
%\begin{equation}
$B(\mathscr{N}_q) \le D[p_q] + 2^{s_q + 1}$ (this is adapted from
Beygelzimer et.~al. \cite{langford2006}).
%\end{equation}

%\begin{figure}

%\begin{minipage}{0.4\textwidth}
\begin{algorithm}[tb]
%\begin{subfigure}[b]{0.48\textwidth}
%  \SetAlgoLined
\begin{algorithmic}
    \STATE {\bfseries Input:} query point $p_q$, reference point $p_r$, list of
candidate neighbors $N$ and distances $D$
    \STATE {\bfseries Output:} distance $d$ between $p_q$ and $p_r$

    \medskip
%    \STATE $d \gets \| p_q - p_r \|$
%    \medskip

    \IF{$d(p_q, p_r) < D[p_q]$ \AND \texttt{BaseCase($p_q$, $p_r$)} not yet called}
    \STATE  $D[p_q] \gets d(p_q, p_r)$, and $N[p_q] \gets p_r$
    \ENDIF

    \RETURN $d(p_q, p_r)$
  \end{algorithmic}

  \caption{Nearest neighbor search \texttt{BaseCase()}}
  \label{alg:nn_base_case}
%\end{subfigure}
\end{algorithm}
%\end{minipage}

%\begin{subfigure}[b]{0.48\textwidth}
%\begin{minipage}{0.4\textwidth}
\begin{algorithm}[tb]
  \begin{algorithmic}
    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, reference node
$\mathscr{N}_r$
    \STATE {\bfseries Output:} a score for the node combination $(\mathscr{N}_q,
\mathscr{N}_r)$, or $\infty$ if the combination should be pruned

    \medskip

    \IF{$d_{\min}(\mathscr{N}_q, \mathscr{N}_r) < B(\mathscr{N}_q)$}
      \RETURN $d_{\min}(\mathscr{N}_q, \mathscr{N}_r)$
    \ENDIF

%    \medskip
    \RETURN $\infty$
  \end{algorithmic}

  \caption{Nearest neighbor search \texttt{Score()}}
  \label{alg:nn_score}
\end{algorithm}
%\end{minipage}
%\end{subfigure}
%\end{figure}

%\begin{equation}
%\begin{split}
%B(\mathscr{N}_q) = \min &\left\{ \max \big\{ \max_{p \in \mathscr{P}_q}
%D[p], \max_{\mathscr{N}_c \in \mathscr{C}_q} B(\mathscr{N}_c) \big\},
%\right. \\
% & \left. \min_{p \in \mathscr{P}_q} \big(D[p] + \rho(\mathscr{N}_q) +
%\lambda(\mathscr{N}_q)\big)
%,\right. \\
% & \left. \min_{\mathscr{N}_c \in \mathscr{C}_q} \Big(B(\mathscr{N}_c) + 2
%\big(\lambda(\mathscr{N}_q) - \lambda(\mathscr{N}_c)\big)\Big) \right. \\
% & B(\operatorname{Par}(\mathscr{N}_q)) \bigg\} . \\
%\end{split}
%\end{equation}

%Although that is a particularly complex definition, what we'll be talking about
%is only cover trees, and we can relax the bound a bit:

%\begin{eqnarray}
%B(\mathscr{N}_q) &\le& \min_{p \in \mathscr{P}_q} \big(D[p] +
%\rho(\mathscr{N}_q) + \lambda(\mathscr{N}_q) \big) \\
% &=& \min_{p \in \mathscr{P}_q} \big(D[p] + \lambda(\mathscr{N}_q) \big) \\
% &=& D[p_q] + \lambda(\mathscr{N}_q) \\
% &=& D[p_q] + 2^{s_q + 1} \\
%&=:& B_c(\mathscr{N}_q).
%\end{eqnarray}

In this formulation, the query node $\mathscr{N}_q$ holds the the query point
$p_q$, the quantity $D[p_q]$ is the current nearest neighbor candidate distance
for the query point $p_q$, and $2^{s_q + 1}$ corresponds to the furthest
descendant distance of $\mathscr{N}_q$.  For notational convenience in the
following proof, take $c_{qr} = \max((\max_{p_q \in S_q} c'_r), c_r)$, where
$c'_r$ is the expansion constant of the set $S_r \cup \{ p_q \}$.

%This follows because $\rho(\mathscr{N}_q) = 0$ for all cover tree
%nodes because each cover tree node only holds one point and that point is the
%centroid of the node.  Also, for a cover tree node $\mathscr{N}_i$ with scale
%$s_i$,

%\begin{equation}
%\label{eqn:ct-fdd-bound}
%\lambda(\mathscr{N}_i) \le 2^{s_i + 1}.
%\end{equation}

\begin{thm}
Using cover trees, the standard cover tree pruning dual-tree traversal, and the
nearest neighbor search \texttt{BaseCase()} and \texttt{Score()} as given in
Algorithms \ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively, and also
given a reference set $S_r$ with expansion constant $c_r$, a query set $S_q$,
and an inverse degree of bichromaticity $\nu$, the running time of the
algorithm is bounded by $O(c_r^4 c_{qr}^5 \nu N)$.
\label{thm:nns}
\end{thm}

\begin{proof}
The running time of \texttt{BaseCase()} and \texttt{Score()} are clearly $O(1)$.
Due to Theorem \ref{thm:ct-runtime}, we therefore know that the runtime of the
algorithm is bounded by $O(c_r^4 \nu |R^*| N)$.  Thus, the only thing that
remains is to bound the maximum size of the reference set, $|R^*|$.

Assume that when $R^*$ is encountered, the maximum reference scale is
$s_r^{\max}$ and the query node is $\mathscr{N}_q$.  Every node $\mathscr{N}_r
\in R^*$ satisfies the property enforced in line
\ref{alg:line:ct-dual-ref-score} that
%\begin{equation}
$d_{\min}(\mathscr{N}_q, \mathscr{N}_r) \le B(\mathscr{N}_q)$.
%\end{equation}
%Relaxing this slightly,
%\begin{equation}
%d_{\min}(\mathscr{N}_q, \mathscr{N}_r) \le B_c(\mathscr{N}_q).
%\end{equation}
Using the definition of $d_{\min}(\cdot, \cdot)$ and $B(\cdot)$, we
expand the equation.  Note that $p_q$ is the point held in $\mathscr{N}_q$ and
$p_r$ is the point held in $\mathscr{N}_r$.  Also, take $\hat{p}_r$ to be the
current nearest neighbor candidate for $p_q$; that is, $D[p_q] = d(p_q,
\hat{p}_r)$.  Then,

\vspace*{-1em}
\begin{eqnarray}
%d(p_q, p_r) - \lambda(\mathscr{N}_q) - \lambda(\mathscr{N}_r) &\le& D[p_q] +
%2^{s_q + 1} \\
%d(p_q, p_r) &\le& d(p_q, \hat{p}_r) + \lambda(\mathscr{N}_q) +
%\lambda(\mathscr{N}_r) + 2^{s_q + 1} \\
%d(p_q, p_r) - 2^{s_q + 1} - 2^{s_r + 1} &\le& D[p_q] + 2^{s_q + 1} \\
d(p_q, p_r) &\le& d(p_q, \hat{p}_r) + 2^{s_q + 1} + 2^{s_r + 1} + 2^{s_q + 1}
\label{eqn:pr_dist} \\
 &\le& d(p_q, \hat{p}_r) + 2(2^{s_r^{\max} + 1})
\end{eqnarray}
\vspace*{-0.9em}

\noindent where the last step follows because $s_q + 1 \le s_r^{\max}$ and $s_r
\le s_r^{\max}$.  Define the set of points $P$ as the points held in each node
in $R^*$ (that is, $P = \{ p_r \in \mathscr{P}(\mathscr{N}_r) : \mathscr{N}_r
\in R^* \}$).  Then, we can write

%\begin{equation}
% = \{ p_r : \mathscr{N}_r \in \mathscr{C}(\mathscr{N}_p), \; s_p = s_r^{\max} + 1 \}.
%\end{equation}
%
%  Stated differently, we have that each point $p_r$ for each node
%$\mathscr{N}_r \in R$ is such that

\vspace*{-0.8em}
\begin{equation}
P \subseteq B_{S_r}(p_q, d(p_q, \hat{p}_r) + 2(2^{s_r^{\max} + 1}))
\end{equation}
\vspace*{-0.8em}

%\noindent or that

%\begin{equation}
%R \subseteq B_{S_r}(p_q, d(p_q, \hat{p}_r) + 3(2^{s_r^{\max} + 1})).
%\end{equation}

Suppose that the true nearest neighbor is $p_r^*$ and $d(p_q, p_r^*) >
2^{s_r^{\max} + 1}$.  Then, $p_r^*$ must be held as a descendant point of some
node in $R^*$ which holds some point $\tilde{p}_r$.  Using the triangle
inequality,

%\begin{eqnarray}
\vspace*{-1.0em}
\begin{equation}
d(p_q, \hat{p}_r) \le d(p_q, \tilde{p}_r)
 \le d(p_q, p_r^*) + d(\tilde{p}_r, p_r^*)
 \le d(p_q, p_r^*) + 2^{s_r^{\max} + 1}.
%\end{eqnarray}
\end{equation}
\vspace*{-0.8em}

This gives that
%\begin{equation}
$P \subseteq B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) + 3(2^{s_r^{\max} +
1}))$.
% &\subseteq& B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) + 3(2^{s_r^{\max} + 1}))
%\end{equation}
The previous step is necessary: to apply the definition of the expansion
constant, the ball must be centered at a point in the set; now, the center
($p_q$) is part of the set.

\vspace*{-1.2em}
\begin{eqnarray}
| B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) + 3(2^{s_r^{\max} + 1})) | &\le& |
B_{S_r \cup \{ p_q \}}(p_q, 4 d(p_q, p_r^*)) | \\
% &\le& | B_{S_{qr}}(p_q, 8 d(p_q, p_r^*)) | \\
% &\le& (c'_r)^3 | B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2) | \\
 &\le& c_{qr}^3 | B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2 |
\end{eqnarray}
\vspace*{-0.8em}

\noindent which follows because the expansion constant of the set $S_r \cup \{
p_q \}$ is bounded above by $c_{qr}$.  Next, we know that $p_r^*$ is the closest
point to $p_q$ in $S_r \cup \{ p_q \}$; thus, there cannot exist a point $p'_r
\ne p_q \in S_r \cup \{ p_q \}$ such that
%\begin{equation}
$p'_r \in B_{S_{qr}}(p_q, d(p_q, p_r^*) / 2)$
%\end{equation}
because that would imply that $d(p_q, p'_r) < d(p_q, p_r^*)$, which is
a contradiction.  Thus, the only point in the ball is $p_q$, and we have that $|
B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2) | = 1$, giving the result that
$|R| \le c_{qr}^3$ in this case.

%%%%%%%%

% Changes from below by Pari are applied.

%****************************************************

%Again, I think the proof from Eq 18 onwards is incorrect. If $B_{S_{qr}}(p_q, \cdot)$ is infact the ball around the data, then $|B_{S_{qr}}(p_q, d(p_q, \hat{p}_r) / 2)| \not= 1$. This would hold only for the actual nearest-neighbor of $p_q$, not any neighbor candidate. If you mean that $B_{S_{qr}}(p_q, \cdot)$ is infact the intersection of the ball around the data and the set of all (implicit) nodes at scale $s_r^{\max}$, then the definition of expansion constant (as in Eq 19) cannot be applied to that set. It can only be applied to the actual ball around the data.

%The fix is straightforward though -- you just have to replace $\hat{p}_r$ with $p_r^*$, the true nearest neighbour of $p_q$. Now $\hat{p}_r$ at least has an implicit node at scale $s_r^{\max}$. Let $\tilde{p}_r$ be the node at scale $s_r^{\max}$ which contains $p_r^*$ as a descendant. So $\tilde{p}_r$'s farthest descendant is at most $2^{s_r^{\max} + 1}$ away from it. So it follows that:
%\begin{eqnarray*}
%d(p_q, \hat{p}_r) & \leq & d(p_q, \tilde{p}_r) \mbox{ since $\hat{p}_r$ is the current closest, } \\
% & \leq & d(p_q, p_r^*) + d(\tilde{p}_r, p_r^*) \mbox{ by triangle inequality, } \\
% & \leq & d(p_q, p_r^*) + 2^{s_r^{\max} + 1}.
%\end{eqnarray*}
%Now you can say that
%$$B_{S_{qr}}(p_q, d(p_q, \hat{p}_r) + 3(2^{s_r^{\max} + 1})) \subseteq B_{S_{qr}}(p_q, d(p_q, p_r^*) + 2^{s_r^{\max} + 1} + 3(2^{s_r^{\max} + 1})),$$
%and work with the right hand side ball and use the argument that $|B_{S_{qr}}(p_q, d(p_q, p_r^*) / 2)| = 1$.

%****************************************************

%%%%%%%%

The other case is when $d(p_q, p_r^*) \le 2^{s_r^{\max} + 1}$, which means that
$d(p_q, \hat{p}_r) \le 2^{s_r^{\max} + 2}$. %  We use a packing argument,
%similar to the proof of Lemma~4.1 in \cite{langford2006}.
Note that $P \in C_{s_r^{\max}}$, and therefore

\vspace*{-1em}
\begin{eqnarray}
P &\subseteq& B_{S_r}(p_q, d(p_q, p_r^*) + 3(2^{s_r^{\max} + 1})) \cap
C_{s_r^{\max}} \\
 &\subseteq& B_{S_r}(p_q, 4(2^{s_r^{\max} + 1})) \cap C_{s_r^{\max}}.
\end{eqnarray}
\vspace*{-0.6em}

Every point in $C_{s_r^{\max}}$ is separated by at least $2^{s_r^{\max}}$.
Using Lemma \ref{lem:packing} with $\delta = 2^{s_r^{\max}}$ and $\rho = 8$
yields that $|P| \le c_r^5$.  This gives the result, because $c_r^5 \le
c_{qr}^5$.
\end{proof}

%By assumption (and Equation \ref{eqn:pr_dist}), for any $p_r \in
%P$,
%\begin{equation}
%$d(p_q, p_r) \leq 2^{s_r^{\max} + 3}$.
%\end{equation}
%Therefore, for any such $p_r$,

%\vspace*{-0.5em}
%\begin{eqnarray}
%\begin{equation}
%B_{S_r} \left( p_q, d(p_q, p_r^*) + 3(2^{s_r^{\max} + 1}) \right) \subseteq
%B_{S_r} \left( p_q, 2^{s_r^{\max} + 3} \right)
% \subseteq B_{S_r} \left(p_r, 2^{s_r^{\max} + 4} \right).
% &\subseteq& B_{S_r} \left(p_r, 2^{s_r^{\max} + 4} \right).
%\end{eqnarray}
%\end{equation}

%Applying the definition of $c_r$, we have that
%\begin{equation}
%$| B_{S_r} (p_r, 2^{s_r^{\max} + 4} ) | \leq c_r^5 | B_{S_r} (p_r,
%2^{s_r^{\max}-1} ) |$.
%\label{eqn:packing1}
%\end{equation}

%Any two points corresponding to nodes in $R^*$ are separated by at least
%$2^{s_r^{\max}}$, so each ball
%can  contain only one node in $R^*$. Therefore, the total
%number of nodes in $R^*$ is bounded by the number of disjoint balls of size
%$2^{s_r^{\max} - 1}$ that can be packed into a ball of size $2^{s_r^{\max} +
%4}$.
%
%In the worst case, this packing is perfect, and we have that
%\begin{equation}
%|B_{S_r}(p_q, d(p_q, p_r^*) + 4(2^{s_r^{\max} + 1})) \cap C_{s_r^{\max}}| \;
%\leq \;  \frac{\left| B_{S_r} \left(p_r, 2^{s_r^{\max} + 4} \right) \right|}{
%\left| B_{S_r} \left(p_q, 2^{s_r^{\max} - 1} \right) \right|} \; \leq \; c_r^5
%\end{equation}

%\noindent which gives the result because $c_r^5 \le c_{qr}^5$.

In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 \nu N)$ because
$c = c_r = c_{qr}$.

### 2014-11-11 bounds-files/sec03-approx-kde.tex
This is what we'll start with.

Ram et.~al. present a clever technique for bounding the
running time of approximate kernel density estimation based on the properties of
the kernel, when the kernel is shift-invariant and satisfies a few assumptions
\cite{ram2009}.  We will restate these assumptions and provide an adapted proof
using Theorem \ref{thm:ct-runtime}, which gives a tighter bound.

Approximate kernel density estimation is a common application of dual-tree
algorithms \cite{gray2003nonparametric, nbody}.  Given a query set $S_q$, a
reference set $S_r$ of size $N$, and a kernel function $\mathcal{K}(\cdot,
\cdot)$, the goal is to calculate, for each $p_q \in S_q$, a density estimate
$f(p_q)$ such that $| f(p_q) - f^*(p_q) | < N \epsilon$ for some
parameter $\epsilon$ and

%\vspace*{-0.5em}
\begin{equation}
f^*(p_q) = \sum_{p_r \in S_r} \mathcal{K}(p_q, p_r).
\end{equation}
%\vspace*{-0.5em}

This is related to the well-studied problem of kernel summation, which can
also be solved with dual-tree algorithms \cite{lee2006faster, lee2008fast}.  In
both of those problems, simple geometric observations can be made to accelerate
computation: when $\mathcal{K}(\cdot, \cdot)$ is shift-invariant, faraway points
have very small kernel evaluations.  Thus, trees can be built on $S_q$ and
$S_r$, and node combinations can be pruned when the nodes are far apart while
still obeying the error bound $N \epsilon$.\footnote{Here we consider
only absolute-value approximation.  Our methods can be easily adapted to other
schemes, such as relative-value approximation.  \cite{ram2009} considers
relative-value approximation in far more detail.}

A tree-independent algorithm for solving approximate kernel density estimation
is given as a \texttt{BaseCase()} function in Algorithm \ref{alg:kde_base_case}
and a \texttt{Score()} function in Algorithm \ref{alg:kde_score} (a correctness
proof can be found in \cite{curtin2013tree}).  This function
assumes a shift-invariant kernel $\mathcal{K}(p_i, p_j) = \mathcal{K}(d( p_i,
p_j ))$.  The list $f_p$ holds partial kernel density estimates for
each query point, and the list $f_n$ holds partial kernel density
estimates for each query node.  At the beginning of the dual-tree traversal, the
lists $f_p$ and $f_n$, which are both of size $O(N)$, are each
initialized to 0.

After the dual-tree traversal, the actual kernel density estimates $f$
must be extracted.  This can be done by traversing the query tree and
calculating $f(p_q) = f_p(p_q) + \sum_{\mathscr{N}_i \in T}
f_n(\mathscr{N}_i)$, where $T$ is the set of nodes in $\mathscr{T}_q$ that
have $p_q$ as a descendant.  %It is possible to implement this in a way where
Each query node needs to be visited only once.

Note that this version is far simpler than the algorithm proposed in,
e.g., \cite{gray2003nonparametric}; however, it is sufficient for our runtime
analysis.  Real-world implementations, such as the one found in \textbf{mlpack}
\cite{mlpack2013}, tend to be far more complex.

\begin{algorithm}[tb]
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:} query point $p_q$, reference point $p_r$, list of
  kernel point estimates $\hat{f}_p$
    \STATE {\bfseries Output:} kernel value $\mathcal{K}(p_q, p_r)$

    \medskip
    \STATE $f_p(p_q) \gets f_p(p_q) + \mathcal{K}(p_q, p_r)$
    \RETURN $\mathcal{K}(p_q, p_r)$
  \end{algorithmic}

  \caption{Approximate kernel density estimation \texttt{BaseCase()}}
  \label{alg:kde_base_case}
\end{algorithm}

\begin{algorithm}[tb]
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, reference node
$\mathscr{N}_r$, list of node kernel estimates $\hat{f}_n$
    \STATE {\bfseries Output:} a score for the node combination $(\mathscr{N}_q,
\mathscr{N}_r)$, or $\infty$ if the combination should be pruned

    \medskip

    \IF{$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) < \epsilon$}
      \STATE $f_n(\mathscr{N}_q) \gets f_n(\mathscr{N}_q) + |
\mathscr{D}^p(\mathscr{N}_r) | \left(\mathcal{K}(d_{\min}(\mathscr{N}_q,
\mathscr{N}_r)) + \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))\right) /\;2$
      \RETURN $\infty$
    \ENDIF

    \RETURN $\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))$
  \end{algorithmic}

  \caption{Approximate kernel density estimation \texttt{Score()}}
  \label{alg:kde_score}
\end{algorithm}

\begin{thm}
Assume that $\mathcal{K}(p_i, p_j) = \mathcal{K}\left(d(p_i, p_j)\right)$ is a shift-invariant kernel with some bandwidth $h > 0$, where $\mathcal{K}(x)$ is smooth, non-negative, monotonically decreasing, and concave for $x \in [0, h]$ and convex for $x \in (h, \infty)$. Then, given a query set $S_q$ and a reference set $S_r$ with
expansion constant $c_r$, an inverse degree of bichromaticity $\nu$ and using the
approximate kernel density estimation \texttt{BaseCase()} and \texttt{Score()}
as given in Algorithms \ref{alg:kde_base_case} and \ref{alg:kde_score},
respectively, with the traversal given in Algorithm \ref{alg:cover-tree-dual},
the running time of approximate kernel density estimation for some $\epsilon$
error parameter is bounded by
%\begin{equation}
$O(c_r^{8 + \log_2 \zeta} \nu N)$
%\end{equation}
where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$.

\label{thm:kde-bound}
\end{thm}

\begin{proof}
It is clear that \texttt{BaseCase()} and \texttt{Score()} both take $O(1)$ time,
so Theorem \ref{thm:ct-runtime} implies the total runtime of the dual-tree
algorithm is bounded by $O(c_r^4 \nu |R^*| N)$.  Thus, we will bound $|R^*|$
using techniques related to those used in \cite{ram2009}.  The bounding of
$|R^*|$ is split into two sections: first, show that when the scale $s_r^{\max}$
is small enough, $R^*$ is empty.  Second, bound $R^*$ when $s_r^{\max}$ is larger.

The \texttt{Score()} function is such that any node in $|R^*|$ for a given query
node $\mathscr{N}_q$ obeys

\vspace*{-0.8em}
\begin{equation}
\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))
\ge \epsilon.
\end{equation}

Thus, we are interested in the maximum possible value $\mathcal{K}(a) -
\mathcal{K}(b)$ for a fixed value of $b - a > 0$.  Due to the nature of
$\mathcal{K}(\cdot)$, the maximum value of $\mathcal{K}'(\cdot)$ is
$\mathcal{K}'(h)$; therefore, the maximum possible value of $\mathcal{K}(a) -
\mathcal{K}(b)$ is when the interval $[a, b]$ is centered on $h$.  Further,
$\mathcal{K}(a) - \mathcal{K}(b) \le \epsilon$ when $(b - a) \le (-\epsilon /
\mathcal{K}'(h))$.  Note that

\vspace*{-1.5em}
\begin{equation}
d_{\max}(\mathscr{N}_q, \mathscr{N}_r) - d_{\min}(\mathscr{N}_q, \mathscr{N}_r)
= d(p_q, p_r) + 2^{s_r^{\max} + 1} - d(p_q, p_r) + 2^{s_r^{\max} + 1} = 2^{s_r^{\max} + 2}.
\end{equation}
\vspace*{-0.8em}

Therefore, $|R^*| = \emptyset$ when
%\begin{eqnarray}
$2^{s_r^{\max} + 2} \le -\epsilon / \mathcal{K}'(h)$, or when
%\begin{equation}
$s_r^{\max} \le \log_2( -\epsilon / \mathcal{K}'(h) ) - 2$.
%\end{equation}
%
Consider, then, the case when $s_r^{\max} > \log_2( -\epsilon /
\mathcal{K}'(h) ) - 2$.  It is trivial to see that for any $\mathscr{N}_r \in
R^*$, $\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) > \epsilon$;
alternately stated, this means that
%\begin{equation}
$d(p_q, p_r) < \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}$.
%\end{equation}
Therefore, bounding the number of points in the set
%\begin{equation}
$B_{S_r}(p_q, \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}) \cap
C_{s_r^{\max}}$
%\end{equation}
is sufficient to bound $|R^*|$.  For notational convenience, define $\omega = \mathcal{K}^{-1}(\epsilon) /
2^{s_r^{\max} + 1} + 1$.  Using Lemma \ref{lem:packing} with $\delta =
2^{s_r^{\max}}$ and $\rho = 2 \omega$ gives $|R^*| = c_r^{3 + \log_2 \omega}$.

%The procedure to bound this will be the same as
%for the nearest neighbor search proof---a packing bound.  See that

%\vspace*{-1.0em}
%\begin{eqnarray}
%B_{S_r}\left(p_q, \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}\right) &\subseteq&
%B_{S_r}\left(p_r, 2 (\mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1})\right) \\
% &=& B_{S_r}\left(p_r, 2^{s_r^{\max} + 2} \left(
%\frac{\mathcal{K}^{-1}(\epsilon)}{2^{s_r^{\max} + 1}} + 1\right)\right).
%\label{eqn:kde_ball}
%\end{eqnarray}
%\vspace*{-0.5em}

%For notational convenience, define $\omega = \mathcal{K}^{-1}(\epsilon) /
%2^{s_r^{\max} + 1} + 1$.  Applying the definition of expansion constant,

%\begin{equation}
%\left| B_{S_r}\left(p_r, 2^{s_r^{\max} + 2} \omega \right) \right|
%\le c_r^{3 + \log_2 \omega} \left|
%B_{S_r}\left(p_r, 2^{s_r^{\max} - 1}\right) \right|.
%\end{equation}

%The total number of nodes in $R^*$ is bounded by the number of disjoint balls of
%size $2^{s_r^{\max} - 1}$ that can be packed into the ball in Equation
%\ref{eqn:kde_ball}.  In the worst case, this packing is perfect, and

%\begin{equation}
%\left| B_{S_r}(p_q, \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}) \cap
%C_{s_r^{\max}} \right| \le \frac{\left| B_{S_r}\left(p_r, 2^{s_r^{\max} + 2}
%\omega \right) \right|}{\left| B_{S_r}\left(p_r, 2^{s_r^{\max} - 1}\right) \right|} \le c_r^{3
%+ \log_2 \omega}.
%\end{equation}

The value $\omega$ is maximized when $s_r^{\max}$ is minimized.  Using the lower
bound on $s_r^{\max}$, $\omega$ is bounded as
%
%\begin{equation}
%$\omega = \frac{\mathcal{K}^{-1}(\epsilon)}{2^{s_r^{\max} + 1} + 1} \le
%\frac{\mathcal{K}^{-1}(\epsilon)}{2^{s_r^{\max} + 1}} = -2 \mathcal{K}'(h)
%\mathcal{K}^{-1}(\epsilon) \epsilon^{-1}.
%\end{equation}
$\omega = -2 \mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$.
%
Finally, with $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon)
\epsilon^{-1}$, we are able to conclude that $|R^*| \le c_r^{3 + \log_2 (2
\zeta)} = c_r^{4 + \log_2 \zeta}$.  Therefore, the entire dual-tree traversal
takes $O(c_r^{8 + \log_2 \zeta} \nu N)$ time.

The postprocessing step to extract the estimates $f(\cdot)$ requires one
traversal of the tree $\mathscr{T}_r$; the tree has $O(N)$ nodes, so this takes
only $O(N)$ time.
%However, there is one more component to consider: the extraction of the
%estimates $f(p_q)$ for each $p_q$.  As discussed earlier, this process can be
%completed in one breadth-first or depth-first traversal of the query tree
%$\mathscr{T}_q$; this takes $O(N)$ time because there are $O(N)$ nodes.%: at
%each node $\mathscr{N}_q$, distribute the partial kernel density estimate to
%each of its children; for each $\mathscr{N}_c \in \mathscr{C}(\mathscr{N}_q)$,
%set
%\begin{equation}
%$f_n(\mathscr{N}_c) \gets f_n(\mathscr{N}_c) +
%f_n(\mathscr{N}_q).$
%\end{equation}
%Continue this process until the leaf nodes are reached; then, for each leaf node
%$\mathscr{N}_l$ which holds point $p_l$, take
%\begin{equation}
%$f(p_l) \gets f_p(p_l) + f_n(\mathscr{N}_l)$
%\end{equation}
%and this will produce the correct estimates $f(\cdot)$.  Because
%there are $O(N)$ nodes and each node has up to $c^4$ children, this process can
%take no longer than $O(c^4 N)$ time.
This is less than the runtime of the
dual-tree traversal, so the runtime of the dual-tree traversal dominates the
algorithm's runtime, and the theorem holds.
\end{proof}

%Define the inverse of the kernel function $\mathcal{K}(\cdot)$ as
%$\mathcal{K}^{-1}(\cdot)$, and define the set $P^*$ as the points held in each
%node in $R^*$. The set $P^*$ for a query node $\mathscr{N}_q$ with
%point $p_q$ can be split into three disjoint sets:

%\begin{eqnarray}
%P^*_l &=& \{ p_r \in P^* : d(p_q, p_r) < h - 2^{s_r^{\max} + 1} \}, \\
%P^*_m &=& \{ p_r \in P^* : h - 2^{s_r^{\max} + 1} \le d(p_q, p_r) \le h +
%2^{s_r^{\max} + 1} \}, \\
%P^*_r &=& \{ p_r \in P^* : d(p_q, p_r) > h + 2^{s_r^{\max} + 1} \}.
%\end{eqnarray}

%First, consider the set $P^*_l$.  For any $p_r \in P^*_l$, we can use the
%concavity of $\mathcal{K}(\cdot)$ in the region $[0, h]$ to see that

%\begin{eqnarray}
%\epsilon &<& \mathcal{K}(\max(0, (d(p_q, p_r) - 2^{s_r^{\max} + 1}))) -
%\mathcal{K}(d(p_q, p_r) + 2^{s_r^{\max} + 1}) \\
% &\le& \mathcal{K}(d(p_q, p_r) + 2^{s_r^{\max} + 1}) - 2^{s_r^{\max} + 2}
%\mathcal{K}'(d(p_q, p_r) + 2^{s_r^{\max} + 1}) - \mathcal{K}(d(p_q, p_r) +
%2^{s_r^{\max} + 1}) \\
% &=& -2^{s_r^{\max} + 2} \mathcal{K}'(d(p_q, p_r) + 2^{s_r^{\max} + 1}).
%\end{eqnarray}

%Denoting the inverse function of $\mathcal{K}'(\cdot)$ as
%$\mathcal{K}'^{-1}_{[a, b]}(\cdot)$ where the output value is restricted to be
%in the range $[a, b]$, we can place a lower bound on $d(p_q, p_r)$:

%\begin{equation}
%d(p_q, p_r) > \mathcal{K}'^{-1}_{[0, h]}\left( \frac{-\epsilon}{2^{s_r^{\max} +
%2}} \right) - 2^{s_r^{\max} + 1}.
%\end{equation}

%Now, consider the set $P^*_u$.  We can make a similar argument using the
%convexity of $\mathcal{K}(\cdot)$ in the region $[h, \infty]$ to see that

%\begin{eqnarray}
%\epsilon &<& \mathcal{K}(d(p_r, p_q) - 2^{s_r^{\max} + 1}) -
%\mathcal{K}(d(p_q, p_r) + 2^{s_r^{\max} + 1}) \\
% &\le& \mathcal{K}(d(p_q, p_r) - 2^{s_r^{\max} + 1}) - \left(\mathcal{K}(d(p_q,
%p_r) - 2^{s_r^{\max} + 1}) + 2^{s_r^{\max} + 2} \mathcal{K}'(d(p_q, p_r) -
%2^{s_r^{\max} + 1}) \\
% &=& -2^{s_r^{\max} + 2} \mathcal{K}'(d(p_q, p_r) - 2^{s_r^{\max} + 1}).
%\end{eqnarray}

%This leads to an upper bound on $d(p_q, p_r)$:

%\begin{equation}
%d(p_q, p_r) < \mathcal{K}'^{-1}_{[h, \infty]}\left(
%\frac{-\epsilon}{2^{s_r^{\max} + 2}}\right) + 2^{s_r^{\max} + 1}.
%\end{equation}

%Lastly, we consider the set $P^*_m$.  Here, we have that

%\begin{eqnarray}
%\epsilon &<& \mathcal{K}(d(p_q, p_r) - 2^{s_r^{\max} + 1}) - \mathcal{K}(d(p_q,
%p_r) + 2^{s_r^{\max} + 1}) \\
% &\le& -2^{s_r^{\max} + 1} \mathcal{K}'(h).
%\end{eqnarray}

%To complete the analysis, we must split into two cases for $s_r^{\max}$.  When
%$s_r^{\max} < \lfloor \log_2 \left

%In the monochromatic case, $\nu = 1$ and the bound is easier to
%understand.
The dependence on $\epsilon$ (through $\zeta$) is expected: as $\epsilon \to 0$
and the search becomes exact, $\zeta$ diverges both because $\epsilon^{-1}$
diverges and also because $\mathcal{K}^{-1}(\epsilon)$ diverges, and the runtime
goes to the worst-case $O(N^2)$; exact kernel density estimation means no nodes
can be pruned at all.

%As we will show, many kernels have a property such that $-\mathcal{K}'(h)
%\mathcal{K}^{-1}(\epsilon)$ has zero dependence on the bandwidth of the kernel.
%Although this result may seem counterintuitive, the only quantities of the
%kernel that affect the runtime are the maximum derivative (which is at the
%inflection point) and the radius of the ball for which $\mathcal{K}(\cdot) \ge
%\epsilon$.

%To see an example of the lack of dependence on kernel bandwidth, consider the
%common Gaussian kernel:

For the Gaussian kernel with bandwidth $\sigma$ defined by $\mathcal{K}_g(d) =
\exp(-d^2 / (2 \sigma^2))$, $\zeta$ does not
depend on the kernel bandwidth; only
the approximation parameter $\epsilon$.  For this kernel, $h = \sigma$ and
therefore $-\mathcal{K}'_g(h) = \sigma^{-1} e^{-1 / 2}$.  Additionally,
$\mathcal{K}_g^{-1}(\epsilon) = \sigma \sqrt{2 \ln (1 / \epsilon)}$.  This means
that for the Gaussian kernel, $\zeta = \sqrt{(-2 \ln \epsilon) / (e
\epsilon^2)}$.  Again, as $\epsilon \to 0$, the
runtime diverges; however, note that there is no dependence on the kernel
bandwidth $\sigma$.  For a reasonably chosen $\epsilon = 0.05$, the runtime is
approximately $O(c_r^{8.89} \nu N)$; for $\epsilon = 0.01$, the runtime is
approximately $O(c_r^{11.52} \nu N)$.  For very small $\epsilon = 0.00001$, the
runtime is approximately $O(c_r^{22.15} \nu N)$.

%Next, consider this sigmoidal kernel: $\mathcal{K}_s(d) = (1 + \exp(\alpha (d -
%\beta)))^{-1}$; note that this is not the more common `sigmoid kernel', which is
%actually a hyperbolic tangent.  Instead, this kernel resembles a step function.
%Some elementary calculus gives an inflection point $h = \beta$, $\mathcal{K}'(h)
%= -\alpha / 4$, and $\mathcal{K}^{-1}(\epsilon) = \ln(\epsilon^{-1} - 1) /
%\alpha + \beta$.  This gives the following expression for $\zeta$:

%\vspace*{-0.8em}
%\begin{equation}
%\zeta = (1/4) ( \epsilon^{-1} \ln(\epsilon^{-1} - 1) +
%(\beta \alpha) / \epsilon ).
%\end{equation}
%\vspace*{-0.8em}
%
%Now, consider the problem of range count, where the number of points in the
%range $[0, u]$ for each query point $p_q \in S_q$ is desired.  If the sigmoidal
%kernel is used with $\beta = u$ for approximate kernel density estimation, then
%as $\alpha \to 0$ and $\epsilon \to 0$, the results will converge to the result
%for exact range count.  Therefore, when $\alpha > 0$ or $\epsilon > 0$, the
%result is approximate range count, and we can use the results of Theorem
%\ref{thm:kde-bound} to show a runtime bound.
%
%Suppose, for the sake of simplicity, that $\alpha = O(\epsilon^{-1})$; then,
%$\zeta = O(\epsilon^{-1} \log \epsilon^{-1} + \epsilon^{-2} u) = O(\epsilon^{-2}
%u)$.  Then, the runtime of approximate range count is bounded as
%
%\vspace*{-1em}
%\begin{eqnarray}
%\begin{equation}
%O(c_r^{8 + \log_2(\epsilon^{-2} u)} \nu N) = O(c_r^{8 + 2 \log_2(\epsilon^{-1}) +
%\log_2(u)} \nu N)
% = O(\epsilon^{-1} u c_r^{8 + 3 \log_2 c_r} \nu N).
%\end{equation}
%\end{eqnarray}
%\vspace*{-1.5em}

%Unfortunately, this bounding technique gives a bad relationship to the
%approximation parameter $\epsilon$ (and rederivations with $\alpha \ne \epsilon$
%suffer from the same issue).  Another drawback of this technique is that it
%cannot account for a range $[l, h]$ with $l > 0$.  Therefore, in the next
%section, we develop an alternate methodology for bounding the running time of
%range search and range count.

### 2014-11-11 bounds-files/sec04-range-search.tex
This is what we'll start with.

\begin{algorithm}[tb]
%  \SetAlgoLined
\begin{algorithmic}[1]
    \STATE {\bfseries Input:} query point $p_q$, reference point $p_r$, range
sets $N[p_q]$ and range $[l, u]$
    \STATE {\bfseries Output:} distance $d$ between $p_q$ and $p_r$

%    \medskip
%    \STATE $d \gets \| p_q - p_r \|$
%    \medskip

    \IF{$d(p_q, p_r) \in [r_{\min}, r_{\max}]$ \AND \texttt{BaseCase($p_q$, $p_r$)} not yet called}
    \STATE  $N[p_q] \gets N[p_q] \cup \{ p_r \}$
    \ENDIF

    \RETURN $d$
  \end{algorithmic}
  \caption{Range search \texttt{BaseCase()}}
  \label{alg:rs_bc}
\end{algorithm}

\begin{algorithm}[tb]
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, reference node
$\mathscr{N}_r$
    \STATE {\bfseries Output:} a score for the node combination $(\mathscr{N}_q,
\mathscr{N}_r)$, or $\infty$ if the combination should be pruned

    \medskip

    \IF{$d_{\min}(\mathscr{N}_q, \mathscr{N}_r) \in [l, u]$ or
$d_{\max}(\mathscr{N}_q, \mathscr{N}_r) \in [l, u]$}
      \RETURN $d_{\min}(\mathscr{N}_q, \mathscr{N}_r)$
    \ENDIF

    \RETURN $\infty$
  \end{algorithmic}
  \caption{Range search \texttt{Score()}}
  \label{alg:rs_sc}
\end{algorithm}

In the range search problem, the task is to find the set of reference points
%\begin{equation}
$N[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}$
%\end{equation}
for each query point $p_q$, where $[l, u]$ is the given
range.  The range count problem is practically identical, but only the size of
the set, $|N[p_q]|$, is desired.
Our proof works for both of these algorithms
identically, but we will focus on range search.  A \texttt{BaseCase()} and
\texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
\ref{alg:rs_sc}, respectively (a correctness proof can be found in
\cite{curtin2013tree}).  The sets $N[p_q]$ (for each $p_q$) are
initialized to $\emptyset$ at the beginning of the traversal.

\begin{thm}
Given a reference set $S_r$ of size $N$ with expansion constant $c_r$, and a
query set $S_q$ of size $O(N)$, a search range of $[l, u]$, and using the range
search \texttt{BaseCase()} and \texttt{Score()} as given in Algorithms
\ref{alg:rs_bc} and \ref{alg:rs_sc}, respectively, with the standard cover tree
pruning dual-tree traversal as given in Algorithm \ref{alg:cover-tree-dual}, and
also assuming that for some $\alpha > 0$,

\vspace*{-1em}
\begin{equation}
\left| \{ p_r: p_r \in S_r, (1 - \alpha) l \le d(p_q, p_r) \le (1 + \alpha) u \}
\right| = O(1),
\end{equation}
\vspace*{-1.5em}

\noindent the running time of range search or range count is bounded by
$O(c_r^{8 + \beta} \nu N)$ with
%\begin{equation}
$\beta = \log_2 (1 + \alpha^{-1})$.
%\end{equation}
\label{thm:rs}
\end{thm}

\vspace*{-2.5em}
\begin{proof}
Both \texttt{BaseCase()} (Algorithm \ref{alg:rs_bc}) and \texttt{Score()}
(Algorithm \ref{alg:rs_sc}) take $O(1)$ time.  Therefore, using Lemma
\ref{thm:ct-runtime} we know that the runtime of the algorithm is bounded by
$O(c_r^4 \nu |R^*| N)$. %  Thus, our only task is to bound the maximum size of the
%reference set, $|R^*|$.
By the pruning rule, for a query node
$\mathscr{N}_q$, the reference set $R^*$
is made up of reference nodes $\mathscr{N}_r$ that are within a margin of
$2^{s_q + 1} + 2^{s_r + 1} \le 2^{s_r^{\max} + 2}$ of the range $[l, u]$.  Given
that $p_r$ is the point in $\mathscr{N}_r$,

\vspace*{-1em}
\begin{equation}
%p_r &\in& \left( B_{S_r}(p_q, u + \lambda_q + \lambda_r) \cap
%C_{s_r^{\max}}\right) \setminus \left( B_{S_r}(p_q, l - \lambda_q - \lambda_r)
%\cap C_{s_r^{\max}} \right) \\
p_r \in \left( B_{S_r}(p_q, u + 2^{s_r^{\max} + 2}) \cap
C_{s_r^{\max}}\right)\setminus \left( B_{S_r}(p_q, l - 2^{s_r^{\max} + 2}) \cap
C_{s_r^{\max}} \right). \label{eqn:rsballs}
\end{equation}
\vspace*{-1em}

A bound on the number of elements in this set is a bound on $|R^*|$.  %First,
%consider the case where $u \le 2^{s_r^{\max} + 1}$.  Then,
%\begin{eqnarray}
%|B_{S_r}(p_q, u + 2^{s_r^{\max} + 1})| - |B_{S_r}(p_q, l - 2^{s_r^{\max} + 1})|
%&\le& |B_{S_r}(p_q, 2^{s_r^{\max} + 2})| \\
%&\le& c_{qr}^3 |B_{S_r}(p_q, 2^{s_r^{\max} - 1})| \\
%&\le& c_{qr}^3
%\end{eqnarray}
%\noindent where the last step follows because $|B_{S_r}(p_q, 2^{s_r^{\max} -
%1})| = 1$ due to the separation invariant.
First, consider the case where $u \le \alpha^{-1} 2^{s_r^{\max} + 2}$.  Ignoring
the smaller ball, take $\delta = 2^{s_r^{\max}}$ and $\rho = 4 (1 +
\alpha^{-1})$ and apply Lemma \ref{lem:packing} to produce the bound

\begin{equation}
|R^*| \le c_r^{4 + \log_2(1 + \alpha^{-1})}.
\end{equation}

% , and ignoring
%the smaller inner ball,
% note that for any $p_r \in P$, $d(p_q, p_r) \le (1 +
%\alpha^{-1}) 2^{s_r^{\max} + 2}$.  Therefore,

%\begin{equation}
%|B_{S_r}(p_q, u + 2^{s_r^{\max} + 2}) \cap C_{s_r^{\max}} | \le |B_{S_r}(p_q,
%(1 + \alpha^{-1}) 2^{s_r^{\max} + 2}) \cap C_{s_r^{\max}} |.
%\end{equation}

%In addition, we know that each point in $C_{s_r^{\max}}$ is separated by a
%distance of at least $2^{s_r^{\max}}$.  We use a packing argument similar to the
%one used in Theorem \ref{thm:nns}.

%Given that $P$ is the set of points held in the nodes $R$, it is true that for
%any $p_r \in P$, $d(p_q, p_r) \le (1 + \alpha^{-1}) 2^{s_r^{\max} + 1}$.
%Therefore,

%\vspace*{-1em}
%\begin{eqnarray}
%|B_{S_r}(p_q, (1 + \alpha^{-1}) 2^{s_r^{\max} + 2})| &\le& |B_{S_r}(p_r, (1 +
%\alpha^{-1}) 2^{s_r^{\max} + 3})| \\
%|B_{S_r}(p_q, u + 2^{s_r^{\max} + 1})| - |B_{S_r}(p_q, l - 2^{s_r^{\max} + 1})|
%&\le& |B_{S_r}(p_q, (1 + \alpha^{-1}) 2^{s_r^{\max} + 1})| \nonumber \\
%&\le& |B_{S_r}(p_r, (1 + \alpha^{-1}) 2^{s_r^{\max} + 2})| \nonumber \\
%&\le& |B_{S_r \cup \{ p_q \}}(p_q, (1 + \alpha^{-1}) 2^{s_r^{\max} + 1})
%\nonumber \\
%&\le& c_{qr}^{(2 + \log_2(1 + \alpha^{-1}))} |B_{S_r \cup \{ p_q \}}(p_q, 2^{s_r^{\max} -
%1})| \nonumber \\
%&\le& c_r^{(4 + \log_2(1 + \alpha^{-1}))} | B_{S_r}(p_r, 2^{s_r^{\max} - 1}) |.
%&\le& (1 + \alpha^{-1}) c_r^{3 + \log_2 c_r} | B_{S_r}(p_r, 2^{s_r^{\max} - 1})
%|.
%&\le& (1 + \alpha^{-1}) c_{qr}^{2 + \log_2 c_{qr}}
%\end{eqnarray}
%\vspace*{-1em}

%Because each point in $P \subseteq C_{s_r^{\max}}$ is separated by at least
%$2^{s_r^{\max}}$, $| B_{S_r}(p_r, 2^{s_r^{\max} - 1}) | = 1$ for any $p_r \in
%P$.  Thus, the total number of nodes in $R$ is the number of balls of radius
%$2^{s_r^{\max} - 1}$ that can be packed into a ball of radius $(1 + \alpha^{-1})
%2^{s_r^{\max} + 2}$.  In the worst case, this packing is perfect, giving

%\vspace*{-1em}
%\begin{equation}
%\left| B_{S_r}\left(p_q, (1 + \alpha^{-1}) 2^{s_r^{\max} + 2}\right) \cap C_{s_r^{\max}}
%\right| \le \frac{\left| B_{S_r}\left( p_r, (1 + \alpha^{-1}) 2^{s_r^{\max} +
%3}\right) \right|}{\left| B_{S_r}\left( p_r, (1 + \alpha^{-1}) 2^{s_r^{\max} -
%1}\right) \right|} \le c_r^{4 + \log_2 (1 + \alpha^{-1})}.
%\end{equation}
%\vspace*{-1em}

%Therefore, in this case, $|R^*| \le c_r^{4 + \log_2 (1 + \alpha^{-1})}$.
Now, consider the other case: $u > \alpha^{-1} 2^{s_r^{\max} + 1}$.
%Then,
%\begin{eqnarray}
%B_{S_r}(p_q, u + 2^{s_r^{\max} + 1}) &\subseteq& B_{S_r}(p_q, (1 + \alpha) u), \\
%B_{S_r}(p_q, l - 2^{s_r^{\max} + 1}) &\supseteq& B_{S_r}(p_q, (1 - \alpha) l).
%\end{eqnarray}
This means

\vspace*{-1em}
\begin{equation}
B_{S_r}(p_q, u + 2^{s_r^{\max} + 1}) \setminus B_{S_r}(p_q, l - 2^{s_r^{\max} +
1}) \subseteq B_{S_r}(p_q, (1 + \alpha) u) \setminus B_{S_r}(p_q, (1 - \alpha)
l)
\end{equation}

\noindent and, due to the assumptions, $| B_{S_r}(p_q, (1 + \alpha) u) | - |
B_{S_r}(p_q, (1 - \alpha) l) | = O(1)$, $|R^*| = O(1)$.  Thus, the theorem
holds.
\end{proof}
\vspace*{-0.4em}

%\begin{figure}
%\begin{centering}
%\input{rs-bounds}
%\caption{The shape of geometry to come.}
%\label{fig:rs_shapes}
%\end{centering}
%\end{figure}

%Figure \ref{fig:rs_shapes} details the possibilities when a particular
%combination is visited.

%\begin{enumerate}
%\item If the reference node is entirely contained in the range $[l +
%\lambda_q, u - \lambda_q]$ (as in $\mathscr{N}_{r1}$), then every
%descendant point of the reference node will be in the results for every
%descendant point of the query node; these can be added and then the combination
%can be pruned.\footnote{Because we have assumed that each query point will have
%$O(1)$ results, the total runtime of this step for each query node is $O(1)$, or
%$O(N)$ during the whole algorithm.  We will not consider the runtime
%contribution of this situation further, because other factors will outweigh
%this.}

%\item If the reference node is not contained at all in the range $[l -
%\lambda_q, u + \lambda_q]$ (as in $\mathscr{N}_{r2}$ and
%$\mathscr{N}_{r3}$), then no descendant points of the reference node will be in
%the results for any descendant point of the query node; thus, the combination
%can simply be pruned.

%\item If the reference node overlaps with the ranges $[l - \lambda_q,
%l + \lambda_q]$ or $[u - \lambda_q, u + \lambda_q]$ (as in
%$\mathscr{N}_{r4}$, $\mathscr{N}_{r5}$, and $\mathscr{N}_{r6}$), the node
%combination cannot be pruned by \texttt{Score()}, and must be descended.  This
%is the interesting case that we will focus on hereafter.  Only combinations of
%this type will make up the reference nodes in $|R_r^*|$.
%\end{enumerate}

%In the last situation, a reference node $\mathscr{N}_r$ with center $p_r$ and
%furthest descendant distance $\lambda_r$ satisfies at least one of the following
%two conditions:

%\begin{eqnarray}
%p_r &\in& B_{S_r}(p_q, u + \lambda_q + \lambda_r) \setminus B_{S_r}(p_q, u -
%\lambda_q - \lambda_r), \\
%p_r &\in& B_{S_r}(p_q, l + \lambda_q + \lambda_r) \setminus B_{S_r}(p_q, l -
%\lambda_q - \lambda_r).
%\end{eqnarray}

%Assuming first that $l - \lambda_q - \lambda_r > 0$, the reference node lies in
%the outer shell of a ball.  This is true of any $\mathscr{N}_r \in R_r$ for any
%cover set $R_r$ during the traversal.  If we can bound the size of the set

%\begin{equation}
%\left(B_{S_r}(p_q, u + \lambda_q + \lambda_r) \setminus B_{S_r}(p_q, u -
%\lambda_q - \lambda_r) \right) \cup \left(B_{S_r}(p_q, l + \lambda_q +
%\lambda_r) \setminus B_{S_r}(p_r, l - \lambda_q - \lambda_r) \right)
%\end{equation}

%\noindent then we will have bounded $| R_r^* |$.  First, observe that $\lambda_q
%+ \lambda_r \le 2^{s_r^{\max} + 1}$; then, the following algebraic
%manipulations lead us to our result.

%\begin{cor}
%In the monochromatic case ($S_q = S_r$, and $c_r = c_{qr} = c$), the
%running time of range search or range count is bounded by $O(c^{8 + \beta} N)$.
%\end{cor}

To understand the relation of the running time to the bounding quantity
$\alpha$, consider (for the monochromatic case with $\nu \sim O(1)$) $\alpha =
(1 / 3)$; this yields a running time of $O(c^8 N)$.  $\alpha = (1 / 7)$ yields
$O(c^9 N)$, $\alpha = (1 / 15)$ yields $O(c^{10} N)$, and so forth.  As $\alpha$
gets smaller, the exponent on $c$ gets larger, and diverges as $\alpha \to 0$.

The assumption that on number of points in the range of $[(1 - \alpha)l, (1 +
\alpha)u]$ for a single query is not too restrictive, but must be satisfied for
the result to be meaningful.  After all, if that assumption is not satisfied,
then it is not possible to perform range search in $O(1)$ time for each query
point, making an $O(N)$ bound for the whole query set impossible.

To show where the assumption is valid, consider that when presented with a very
large dataset, a practitioner may perform monochromatic range search on a range
$[l, u]$ for a geometric subset of their data; for instance, a hypersquare
with side length $2u$.  This might be done in order to find a range $[l, u]$
that produces a manageable number of results (if the range is too large, the
number of results returned for each query point may be enormous).  Then, that
practitioner may scale to a larger dataset by simply increasing the size of
their hypersquare.  In this setting, the number of points in the range of
$[(1 - \alpha)l, (1 + \alpha)u]$ for a single query remains $O(1)$ with respect
to $N$.  Alternately, it should be possible to tighten the range $[l, u]$ as $N$
increases in such a way that the number of points in the range remains constant.

%In cases where this assumption is not thought to be valid, it is straightforward
%to prove a bound on the running time of range search or range count by using the
%aspect ratio of the data, as in \cite{curtin2014kernel}.  However, the aspect
%ratio is particularly sensitive to datasets that have arbitrarily close points,
%and it is difficult to make assumptions about the aspect ratio.

Alternately, if the assumption is not valid but it can be shown that the number
of points in the range $[(1 - \alpha)l, (1 + \alpha)u]$ is $O(k)$ where $k$ has
some dependence on $N$, then the running time can be easily bounded as $O(c^{4 +
\beta} kN)$, in the monochromatic case with $\nu \sim O(1)$.

%Also, note that the proof considers points with distance from $p_q$ in
%the range $[l + \lambda_q + \lambda_r, u - \lambda_q - \lambda_r]$; however, the
%pruning rule could also prune points that lie entirely in the range $[l, u]$ for
%additional speedup (Algorithm \ref{alg:rs_sc} does not do this because there is
%no theoretical benefit in our proof).  Unfortunately, the expansion constant
%does not make working with slices of balls easy, and as a result the runtime
%bound as given is likely to be loose when this additional prune is made,
%especially when $u - l$ is large.

### 2014-11-11 bounds-files/sec05-range-search-aspect-ratio.tex
This is what we'll start with.

In this section, we will bound the running time of range search in a different
way, by using the aspect ratio of the reference set $S_r$.

Define $\tau_r$ to be the aspect ratio of the reference set $S_r$, which is the
largest distance between any two points in $S_r$ (denote this quantity as
$\gamma_r$) divided by the smallest nonzero distance between any two points in
$S_r$ (denote this quantity as $\delta_r$).

\begin{thm}
Under the standard assumptions and given a reference set aspect ratio of
$\tau_r$, using the range search \texttt{BaseCase()} and \texttt{Score()} as
given in Algorithms \ref{alg:rs_bc} and \ref{alg:rs_sc}, respectively, with a
search range of $[l, u]$, and also assuming that the length of the list of
results for each query point in $S_q$ is $O(1)$, the running time of range
search or range count is bounded by

\begin{equation}
\begin{split}
  O((1 + \tau_r) c_r^{7 + \log_2 c_r} \nu N),& \text{\ \ \ \ if } u \ge \gamma_r\\
  O((1 + \frac{u}{\gamma_r} \tau_r) c_r^{7 + \log_2 c_r} \nu N),&
\text{\ \ \ \ otherwise}.
\end{split}
\end{equation}

\end{thm}

\begin{proof}
As in the previous proof, we know that the running time of \texttt{BaseCase()}
and \texttt{Score()} are $O(1)$, and that the runtime of the algorithm is
bounded by $O(c_r^4 \nu |R^*| N)$.  Thus, again, our goal is to bound the
maximum size of the reference set $| R^* |$.

From Equation \ref{eqn:rsballs}, we can ignore the inner ball term to see that

\begin{equation}
|R^*| \le \left| B_{S_r}(p_q, u + 2^{s_r^{\max} + 1}) \cap C_{s_r^{\max}}
\right|.
\end{equation}

It is true that $|B_{S_r}(p_r, \gamma_r)| = N$ for all $p_r \in S_r$; also, if
$u \ge \gamma_r$, then $B_{S_r}(p_r, u) = B_{S_r}(p_r, \gamma_r)$.  Thus, define
the auxiliary quantity $\zeta_r = \min (u, \gamma_r)$.  Using this auxiliary
quantity, we can now employ a packing argument, as in previous proofs.  For any
$p_r \in P$,

\begin{eqnarray}
|B_{S_r}(p_q, u + 2^{s_r^{\max} + 1})| &\le& |B_{S_r}(p_q, \zeta_r +
2^{s_r^{\max} + 1})| \\
 &\le& |B_{S_r}(p_q, (1 + \frac{\zeta_r}{2^{s_r^{\max} + 1}}) 2^{s_r^{\max} +
1})| \\
 &\le& |B_{S_r}(p_r, (1 + \frac{\zeta_r}{2^{s_r^{\max} + 1}}) 2^{s_r^{\max} +
2})| \\
 &\le& c_r^{3 \log_2{(1 + \zeta_r / 2^{s_r^{\max} + 1})}} | B_{S_r}(p_r,
2^{s_r^{\max} - 1})|.
\end{eqnarray}

The total number of nodes in $R$ is bounded by the number of disjoint balls of
size $2^{s_r^{\max} - 1}$ that can be packed into a ball of radius $(1 + \zeta_r
/ 2^{s_r^{\max} + 1}) 2^{s_r^{\max} + 2}$.  In the worst case, this packing is
perfect, and we have that

\begin{equation}
\left| B_{S_r}(p_q, \zeta_r + 2^{s_r^{\max} + 1}) \cap C_{s_r^{\max}} \right|
\le
\frac{| B_{S_r}(p_r, (1 + \zeta_r / 2^{s_r^{\max} + 1}) 2^{s_r^{\max} + 2}) |}{|
B_{S_r}(p_r, 2^{s_r^{\max} - 1}) |} \le c_r^{3 \log_2{(1 + \zeta_r /
2^{s_r^{\max} + 1})}}.
\end{equation}

We can simplify the last term:

\begin{equation}
c_r^{3 \log_2(1 + \zeta_r / 2^{s_r^{\max} + 1})} = (1 +
\frac{\zeta_r}{2^{s_r^{\max} + 1}}) c_r^{3 \log_2 c_r}.
\end{equation}

In the worst case, $s_r^{\max}$ is the bottom level of the tree, and
$2^{s_r^{\max}}$ is the distance between the two closest points in $S_r$ (call
this distance $\delta_r$).  Expand $\zeta_r$ to its definition to get the result
that

\begin{equation}
|R^*| =
\begin{cases}
  (1 + \tau_r) c^{3 \log_2 c},& \text{if } u \ge \gamma_r \\
  (1 + \frac{u}{\gamma_r} \tau_r) c_r^{3 \log_2 c_r},& \text{otherwise}.
\end{cases}
\end{equation}

\end{proof}

It is important to note that as $u$ decreases, the running time of the algorithm
also decreases accordingly.

Possibilities for improvement:

\begin{itemize}
  \item Can we use a packing argument to bound the number of balls in a `slice'?

  \item Can we incorporate the lower bound $l$ into the proof to subtract large
parts away?

  \item Can we modify the \texttt{Score()} rules for pruning when a node is
entirely contained within the range, and then only consider slices of size
$2^{s_r^{\max} + 1}$ centered at radius $u$ and $l$?
\end{itemize}

The second two ideas depend on the first idea.

### 2014-11-11 bounds.tex
This is what we'll start with.

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\hypersetup{colorlinks,
            urlcolor=blue,
            citecolor=green,
            linkcolor=red}
\usepackage{url}
%\documentstyle[nips13submit_09,times,art10]{article} % For LaTeX 2.09
\usepackage{mathrsfs}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{color}

\usepackage{pstricks}
\usepackage{tikz}
%\usepackage{subcaption}

\title{Plug-and-play dual-tree algorithm runtime analysis}

%\author{
%David S.~Hippocampus\thanks{ Use footnote for providing further information
%about author (webpage, alternative address)---\emph{not} for acknowledging
%funding agencies.} \\
%Department of Computer Science\\
%Cranberry-Lemon University\\
%Pittsburgh, PA 15213 \\
%\texttt{hippo@cs.cranberry-lemon.edu} \\
%\And
%Coauthor \\
%Affiliation \\
%Address \\
%\texttt{email} \\
%\AND
%Coauthor \\
%Affiliation \\
%Address \\
%\texttt{email} \\
%\And
%Coauthor \\
%Affiliation \\
%Address \\
%\texttt{email} \\
%\And
%Coauthor \\
%Affiliation \\
%Address \\
%\texttt{email} \\
%(if needed)\\
%}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

% Misc. math commands.  I couldn't figure out how to make the Unicode cat
% character, sadly.
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\var}{\operatorname{var}}

\newcommand{\K}{\mathcal{K}}
\newcommand{\hilbert}{\mathcal{H}}
\newcommand{\Real}{\mathcal{R}}

\renewcommand{\algorithmicforall}{\textbf{for each}}
\renewcommand{\algorithmicrepeat}{\textbf{do}}
\renewcommand{\algorithmicuntil}{\textbf{while}}

\newtheorem{defn}{Definition}
\newtheorem{thm}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{cor}{Corollary}

%\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}

\maketitle

\vspace*{-0.8em}
\begin{abstract}
%A large collection of machine learning algorithms contain subproblems that can
%be efficiently solved with trees.  In the situation where a dual-tree algorithm
%with cover trees is used to solve a problem, rigorous worst-case runtime bounds
%can often be shown.  A worst-case runtime bound is proven for all cover tree
%dual-tree algorithms; this bound can be employed to give useful runtime bounds
%for individual problems.  We show a new, tighter bound of $O(c^9 \nu N)$ for
%all-nearest-neighbor search and also provide linear runtime bounds for
%approximate kernel density estimation and exact range count/search under some
%mild assumptions.  These are the first linear runtime bounds we are aware of for
%range count/search.
Numerous machine learning algorithms contain pairwise statistical
problems in their core. Often, tree structures are used
to solve these problems efficiently. Dual-tree
algorithms using cover trees have
rigorous worst-case runtime guarantees for some of these problems. In this paper, we present a {\em
problem-independent} runtime guarantee for {\em any} dual-tree
algorithm using the cover tree, separating out the problem-dependent and
the problem-independent elements. This allows us to just plug in
bounds for the problem-dependent elements to get runtime guarantees
for dual-tree algorithms for any pairwise statistical problem. We
demonstrate this plug-and-play procedure for nearest-neighbor search
and approximate kernel density estimation to get improved runtime
guarantees.  Under some assumptions, we also present the first linear runtime
guarantee for dual-tree based range search.
\end{abstract}

\vspace*{-1em}

%\section{Dual-tree algorithms for pairwise statistical problems (Introduction)}
\section{Dual-tree algorithms}
\vspace*{-0.5em}

%\begin{itemize}
%\item What are pairwise statistical problems? Examples?
%
%A surprising number of machine learning algorithms have bottlenecks which, if
%implemented naively, depend on the calculation of potentially large numbers of
%distances.  A common example is nearest neighbor search: if search
%for the nearest neighbor of a single query point in a reference set of size $N$,
%simple linear scan costs $O(N)$ time.  For many decades, spatial data structures
%have been employed to accelerate this type of computation, beginning with the
%quad tree \cite{finkel1974quad}; building trees on the reference set often
%results in expected runtimes of $O(\log N)$ per query point.
%
A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems.
All-nearest-neighbor search is one such problem, naively requiring $O(N)$ time
to answer in single query in a reference set of size $N$; answering $O(N)$
queries subsequently requiring the prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, naively requiring $O(N^2)$
time to answer $O(N)$ queries. The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\cite{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.

%\item What are dual-tree algorithms? Brief explanation.
%
%More recently, Gray and Moore \cite{nbody} observed that problems often have
%large query {\em sets}, not just a few query points, and introduced dual-tree
%algorithms.  One example is all-nearest-neighbor search: find the nearest
%neighbor of every point in the reference set.  Simple implementations scale
%poorly, but dual-tree algorithms can obtain orders-of-magnitude speedup for a
%multitude of problems.
%
%Following this, numerous dual-tree algorithms have been developed to solve a
%wide variety of problems; these include kernel density estimation
%\cite{gray2003nonparametric}, fast max-kernel search \cite{curtin2013fast},
%approximate nearest neighbor search \cite{ram2009rank}, mean shift
%\cite{wang2007fast}, minimum spanning tree calculation
%\cite{march2010euclidean}, $n$-point correlation function estimation
%\cite{march2012fast}, kernel summation \cite{lee2006faster}, particle
%smoothing \cite{klaas2006fast}, variational inference \cite{amizadeh2012variational} and range search \cite{nbody}.
%
Gray and Moore \cite{nbody} generalized the fast-multipole-method from
computational physics to obtain {\em dual-tree algorithms}. These are extremely
useful when there are large query {\em sets}, not just a few query points.
%\item What (pairwise statistical) problems have they been applied to?
%
Following this, numerous dual-tree algorithms have been developed to solve a
wide variety of problems; these include kernel density estimation
\cite{gray2003nonparametric}, fast max-kernel search \cite{curtin2013fast},
approximate nearest neighbor search \cite{ram2009rank}, mean shift
\cite{wang2007fast}, minimum spanning tree calculation
\cite{march2010euclidean}, $n$-point correlation function estimation
\cite{march2012fast}, kernel summation \cite{lee2006faster}, particle
smoothing \cite{klaas2006fast}, variational inference \cite{amizadeh2012variational} and range search \cite{nbody}.
%
%\item Existing analyses of dual-tree algorithms
Some of these algorithms are derived using the cover tree
\cite{langford2006}, a data structure with beneficial theoretical qualities.
Dual-tree all nearest-neighbor search and approximate kernel density estimation with cover trees have $O(N)$ runtime guarantees for $O(N)$ queries \cite{ram2009}; minimum spanning tree calculation scales as $O(N \log N)$ \cite{march2010euclidean}. Other problems have similar worst-case guarantees \cite{curtin2014dual, march2013multi}.
% Is there a better reference for the npcf runtime bound?

%\item General tree/problem independent dual-tree algorithms (maybe)
%\item What is lacking -- general dual-tree analysis (since the algorithms are so similar in structure)
%A recent framework for understanding and unifying dual-tree algorithms
%observes that all dual-tree algorithms are particularly similar, and generalizes
%them \cite{curtin2013tree}.  A
%nice benefit of this viewpoint is that it allows quick and easy development of
%new dual-tree algorithms.  In this paper, we make the same observation about
%runtime proofs for cover tree dual-tree algorithms, and show that
%runtime analysis of generalized cover tree dual-tree algorithms is
%possible.  Thus, the contributions of this paper are:
%
%\vspace*{-0.5em}
%\begin{itemize}
%  \item A general runtime bound for dual-tree algorithms that use the cover
%tree.
%
%  \item \mbox{A nearest neighbor search runtime bound,
%slightly improving on existing linear-time results.}
%
%  \item An approximate kernel density estimation runtime bound,
%improving on existing linear-time results.
%
%  \item Novel runtime analysis of dual-tree range count and range
%search.
%\end{itemize}
Curtin et.~al. \cite{curtin2013tree} observe that all dual-tree
algorithms are similar and propose a unified framework which allows the
development of dual-tree algorithms in a problem independent (and tree
independent) manner. In this paper, we make the same observation about runtime
proofs for cover tree dual-tree algorithms, and show that runtime analysis
of a generalized dual-tree algorithm is possible. The specific contributions of
this paper are:
\vspace*{-0.3em}
\begin{itemize}
  \item A general runtime bound for dual-tree algorithms that use the cover
tree.
  \item An improved linear runtime guarantee for all nearest-neighbor search.
  \item An improved linear runtime guarantee for approximate kernel density
estimation.
  \item The first linear runtime guarantee for dual-tree range count and range
search.
\end{itemize}

%The following section will detail theory prerequisites, and then introduce the
%general runtime bound.

%\item This paper's contribution:
%  \begin{itemize}
%  \item Problem independent dual-tree algorithm analysis
%  \item Examples of final problem dependent runtime bounds for NNS and KDE which match existing results (from LTAPS) -- implying nothing is lost yet for doing problem independent analysis
%  \item Novel runtime analyses of dual-tree range count (or range search).
%  \end{itemize}
%\item This paper's structure (to be written last)
%\end{itemize}

\vspace*{-1em}
\section{Preliminaries}
\vspace*{-0.6em}

\input{bounds-files/sec01-supporting-lemmas.tex}

\vspace*{-0.6em}
\section{General runtime bound}
\vspace*{-0.4em}

\input{bounds-files/general-runtime-bound.tex}

\vspace*{-0.8em}
\section{Nearest neighbor search}
\vspace*{-0.4em}

\input{bounds-files/sec02-nearest-neighbor-search.tex}

\vspace*{-0.6em}
\section{Approximate kernel density estimation}
\vspace*{-0.6em}

\input{bounds-files/sec03-approx-kde.tex}

\vspace*{-0.8em}
\section{Range Search and Range Count}
\vspace*{-0.4em}

\input{bounds-files/sec04-range-search.tex}

%\section{Range Search with Aspect Ratio}

%\input{bounds-files/sec05-range-search-aspect-ratio.tex}

\section{Conclusion}

\input{bounds-files/conclusion.tex}

\bibliographystyle{plain}
\bibliography{bounds}

% Appendix doesn't have anything useful or non-straightforward.
%\appendix
%\input{appendix-ec}

\end{document}

### 2014-11-11 bounds-files/imbalance.tex
Some thoughts on tree imbalance.

A better understanding of the behavior of cover trees can be acheived partially
by introducing a measure of tree imbalance.  Numerous measures of tree imbalance
have already been established, but we aim to capture a different measure of
imbalance that utilizes the leveled structure of the cover tree.

We already know each node in a cover tree is indexed with an integer level (or
scale).  In the explicit representation of the cover tree, each non-leaf node
has children at a lower level.  But these children need not be strictly one
level lower; see Figure \ref{fig:imbalance}.  In Figure
\ref{fig:imbalance-good}, each cover tree node has children that are strictly
one level lower; we will refer to this as a {\em perfectly balanced cover tree}.
Figure \ref{fig:imbalance-bad}, on the other hand, has many children with far
lower levels than their parents.  We will refer to this as an {\em imbalanced
cover tree}.

\begin{figure}
\caption{Balanced and imbalanced cover trees.}
\label{fig:imbalance}
\end{figure}

An imbalanced cover tree can---and often does---happen in practice.  Consider a
dataset with a single outlier which is very far away from all of the other
points.  Figure \ref{fig:outlier} shows what happens in this situation: the root
node has two children; one of these children has only the outlier as a
descendant, and the other child has the rest of the points in the dataset as a
descendant.  In fact, it is easy to find datasets with multiple outliers which
give rise to a chain-like structure at the top of the tree: see Figure
\ref{fig:outliers} for an illustration\footnote{As a side note, this behavior is
not limited to cover trees, and can happen to mean-split $kd$-trees too,
especially in higher dimensions.}.

\begin{figure}
\caption{Single-outlier cover tree.}
\label{fig:outlier}
\end{figure}

\begin{figure}
\caption{Multiple-outlier cover tree.}
\label{fig:outliers}
\end{figure}

A tree that has this chain-like structure all the way down is going to perform
abysmally; motivated by this observation, we define a measure of tree imbalance.

\begin{defn}
The {\it cover node imbalance} $i(\mathscr{N}_i)$ for a cover tree node
$\mathscr{N}_i$ with scale $s_i$ in the cover tree $\mathscr{T}$ is defined as
the cumulative number of missing levels between the node and its children.  If
the node has a leaf child (which has level $-\infty$), then number of missing
levels is defined as the difference between $s_i$ and $s_{\min} - 1$ where
$s_{\min}$ is the smallest scale of a non-leaf node in $\mathscr{T}$.
Explicitly written, this calculation is

\begin{equation}
i(\mathscr{N}_i) = \left( \sum_{\mathscr{N}_c \in \mathscr{C}(\mathscr{N}_i),
\mathscr{C}(\mathscr{N}_c) \ne \emptyset} s_i - s_c - 1 \right) + \left(
\sum_{\mathscr{N}_c \in \mathscr{C}(\mathscr{N}_i), \mathscr{C}(\mathscr{N}_c) =
\emptyset} \max(s_i - s_{\min} - 1, 0) \right).
\end{equation}
\end{defn}

Given this definition for a single node, then, we can generalize to the full
tree.

\begin{defn}
The {\it cover tree imbalance} $b(\mathscr{T})$ for a cover tree $\mathscr{T}$
is defined as the cumulative number of missing levels in the tree.  This can be
expressed as a function of cover node imbalances easily:

\begin{equation}
b(\mathscr{T}) = \sum_{\mathscr{N}_i \in \mathscr{T}} i(\mathscr{N}_i).
\end{equation}
\end{defn}

### 2014-11-12 bounds-files/general-runtime-bound.tex
Remove negative vspace and remove NIPS formatting.
Update; rewritten intro, imbalance, and bound.

@@
+Perhaps more interesting than measures of tree imbalance is the way cover trees
+are actually used in dual-tree algorithms.  Although cover trees were originally
+intended for nearest neighbor search (see Algorithm \texttt{Find-All-Nearest},
+\cite{langford2006}), they can be adapted to a wide variety of problems: minimum
+spanning tree calculation \cite{march2010fast}, approximate nearest neighbor
+search \cite{ram2010rank}, Gaussian processes posterior calculation
+\cite{moore2014fast}, max-kernel search \cite{curtin2013fast,
+curtin2014dual} are some examples.  Further, through the tree-independent
+dual-tree algorithm abstraction of Curtin et~al. \cite{curtin2013tree}, other
+existing dual-tree algorithms can easily be adapted for use with cover trees.
+
+In the framework of tree-independent dual-tree algorithms
+\cite{curtin2013tree}, all that is necessary to describe a dual-tree algorithm
+is a point-to-point base case function (\texttt{BaseCase()}) and a node-to-node
+pruning rule (\texttt{Score()}).  These functions, which are often very
+straightforward, are then paired with a type of tree and a pruning dual-tree
+traversal to produce a working algorithm.  In later sections, we will consider
+specific examples.
+
+When using cover trees, the typical pruning dual-tree traversal is an adapted
+form of the original nearest neighbor search algorithm
+(\texttt{Find-All-Nearest}, \cite{langford2006}); this traversal is implemented
+in both the cover tree reference implementation and in the more flexible {\bf
+mlpack} library \cite{curtin2013mlpack}.  The algorithm is given in Algorithm
+\ref{alg:cover-tree-dual} and was originally presented by Curtin and Ram
+\cite{curtin2014dual}.  Initially, it is called with the root of the query tree
+and a reference set containing only the root of the reference tree.
+
+\begin{algorithm}[htb]
+  \begin{algorithmic}[1]
+    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, set of reference nodes
+$R$ \label{alg:line:ct-dual-input}
+    \STATE {\bfseries Output:} none
+
+    \medskip
+    \STATE $s^{\max}_r \gets \max_{\mathscr{N}_r \in R} s_r$
+    \IF{$(s_q < s^{\max}_r)$ {\bf or} $(s^{\max}_r = -\infty)$}
+\label{alg:line:ct-dual-ref-recursion-start}
+      \FORALL{$\mathscr{N}_r \in R$} \label{alg:line:ct-dual-base-case-start}
+        \STATE \texttt{BaseCase($p_q$, $p_r$)}
+      \ENDFOR \label{alg:line:ct-dual-base-case-end}
+      \STATE $R_r \gets \{ \mathscr{N}_r \in R : s_r = s^{\max}_r \}$
+\label{alg:line:ct-dual-ref-set}
+      \STATE $R_{r - 1} \gets \{ \mathscr{C}(\mathscr{N}_r) : \mathscr{N}_r \in
+R_r \} \cup (R \setminus R_r)$ \label{alg:line:ct-dual-ref-children}
+      \STATE $R'_{r - 1} \gets \{ \mathscr{N}_r \in R_{r - 1} :
+\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
+\label{alg:line:ct-dual-ref-score}
+      \STATE recurse with $\mathscr{N}_q$ and $R'_{r - 1}$
+\label{alg:line:ct-dual-ref-recursion-end}
+    \ELSE \label{alg:line:ct-dual-query-recursion-start}
+      \FORALL{$\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q)$}
+        \STATE $R' \gets \{ \mathscr{N}_r \in R :
+\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
+\label{alg:line:ct-dual-query-pruning}
+        \STATE recurse with $\mathscr{N}_{qc}$ and $R'$
+\label{alg:line:ct-dual-query-recursion}
+      \ENDFOR \label{alg:line:ct-dual-query-recursion-end}
+    \ENDIF
+  \end{algorithmic}
+  \caption{The standard pruning dual-tree traversal for cover trees.}
+  \label{alg:cover-tree-dual}
+\end{algorithm}
+
+This dual-tree recursion is a depth-first recursion in the query tree and a
+breadth-first recursion in the reference tree.  It descends either the query
+tree or the reference tree (not both) during a single iteration, and the
+conditional in line \ref{alg:line:ct-dual-ref-recursion-start} ensures that the
+combinations of query and reference nodes that are being visited are close in
+scale.
+
+Due to the favorable theoretical properties of the cover tree, we can prove
+general runtime bounds when this traversal is used.
+
 \begin{thm}
@@
 set of queries $S_q$ of size $O(N)$, a standard cover tree based dual-tree
-algorithm (Algorithm \ref{alg:cover-tree-dual}) takes $O(c_r^4 | R^* | \nu \chi
-\psi N)$, where $ | R^* | $ is the maximum size of the reference set $R$ (line
+algorithm (Algorithm \ref{alg:cover-tree-dual}) takes $O(c_r^4 | R^* | \chi
+\psi (N + \theta))$, where $ | R^* | $ is the maximum size of the reference set $R$ (line
 \ref{alg:line:ct-dual-input}) during the dual-tree recursion, $\chi$ is
-the maximum possible runtime of \texttt{BaseCase()}, and $\psi$ is the maximum
-possible runtime of \texttt{Score()}.
+the maximum possible runtime of \texttt{BaseCase()}, $\psi$ is the maximum
+possible runtime of \texttt{Score()}, and $\theta$ is a measure of imbalance
+defined by
+
+\begin{equation}
+\theta = \max(\log_2(\eta_r / \eta_q), 0) + i(\mathscr{T}_q) + N
+\max(\log_2(\delta_q / \delta_r), 0)
+\end{equation}
+
+\noindent with $\eta_q$ as the maximum pairwise distance between points in
+$S_q$, $\delta_q$ as the minimum nonzero pairwise distance between points in
+$S_q$, $\eta_r$ as the maximum pairwise distance between points in $S_r$, and
+$\delta_r$ as the minimum nonzero pairwise distance between points in $S_r$.
 \end{thm}
 
-\vspace*{-1.23em}
 \begin{proof}
+First, split the algorithm into two parts: reference recursions (lines
+\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
+and query recursions (lines
+\ref{alg:line:ct-dual-query-recursion-start}--\ref{alg:line:ct-dual-query-recursion-end}).
+The runtime of the algorithm is bounded as the runtime of a reference recursion
+times the total number of reference recursions plus the total runtime of all
+query recursions.
+
 Consider a reference recursion (lines
@@
 R^* |)$ time, because each reference node has $c_r^4$ children.  So, one
-full reference recursion takes $O(c_r^4 \psi | R^* |)$ time.
+full reference recursion takes $O(c_r^4 \psi \chi | R^* |)$ time.
 
 Now, note that there are $O(N)$ nodes in $\mathscr{T}_q$.  Thus, line
-\ref{alg:line:ct-dual-query-recursion} is visited $O(N)$ times.  Each of these
-$O(N)$ visits to line \ref{alg:line:ct-dual-query-recursion} implies a
-recursion, in which the reference set is descended up to $\nu$ times (lines
-\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
-before the query node is descended or the algorithm terminates.  In addition,
-each $O(N)$ recursion implies an $O(\psi |R|) \le O(\psi |R^*|)$ operation for
-the calculation of $R'$ (line \ref{alg:line:ct-dual-query-pruning}).  Thus, the
-full runtime of the algorithm is bounded as $O(c_r^4 |R^*| \nu \chi \psi N +
-\psi |R^*| N) = O(c_r^4 |R^*| \nu \chi \psi N)$.
+\ref{alg:line:ct-dual-query-recursion} is visited $O(N)$ times.  The amount of
+work in line \ref{alg:line:ct-dual-query-pruning}, like in the reference
+recursion, is bounded as $O(c_r^4 \psi | R^* |)$. Therefore, the total
+runtime of all query recursions is $O(c_r^4 \psi | R^* | N)$.
+
+Lastly, we must bound the total number of reference recursions.  Reference
+recursions happen in three cases: \textit{(1)} $s_r^{\max}$ is greater than the
+scale of the root of the query tree (no query recursions have happened yet);
+\textit{(2)} $s_r^{\max}$ is less than or equal to the scale of the root of the
+query tree, but is greater than the minimum scale of the query tree that is not
+$-\infty$; \textit{(3)} $s_r^{\max}$ is less than the minimum scale of the query
+tree that is not $-\infty$.
+
+First, consider case \textit{(1)}.  If the scale of the root of the query tree
+is $s_q^T$ and the scale of the root of the reference tree is $s_r^T$, then
+$s_r^T - s_q^T$ recursions of this sort occur before the first query recursion.
+
+Next, consider case \textit{(2)}.  Each query recursion implies at least one
+reference recursion; thus there will be at least $O(N)$ recursions.  Any
+extra recursions above and beyond this occur because the query node does not
+have scale which is exactly one less than its parent.  When the query node is
+not a leaf, the number of extra reference recursions is bounded by the
+difference between the scales of the query node and its parent minus one (one is
+subtracted to account for the one reference recursion that must happen).  Each
+query node is visited once, and therefore the total number of extra recursions
+when the query node is not a leaf is bounded by
+
+\begin{equation}
+\sum_{\mathscr{N}_q \in \mathscr{T}_q} \sum_{\mathscr{N}_{qc} \in
+\mathscr{C}(\mathscr{N}_q), s_{qc} \ne -\infty} s_q - s_{qc} - 1.
+\end{equation}
+
+When the query node is a leaf, then the number of recursions in case
+\textit{(2)} for that query node is bounded by the difference between the scale
+of the query node's parent and the smallest scale of a non-leaf node in the
+query tree (call this $s_q^{\min}$) minus one:
+
+\begin{equation}
+\sum_{\mathscr{N}_q \in \mathscr{T}_q} \sum_{\mathscr{N}_{qc} \in
+\mathscr{C}(\mathscr{N}_q), s_{qc} = -\infty} \max(s_q - s_q^{\min} - 1, 0).
+\end{equation}
+
+The $\max$ is necessary to account for the case where the query node's parent
+has scale $s_q^{\min}$; no reference recursions of type \textit{(2)} happen at
+all in this case.  Therefore the total number of extra reference recursions of
+type \textit{(2)} is bounded by $b$, where
+
+\begin{equation}
+b = \sum_{\mathscr{N}_q \in \mathscr{T}_q} \left(\sum_{\mathscr{N}_{qc} \in
+\mathscr{C}(\mathscr{N}_q), s_{qc} \ne -\infty} s_q - s_{qc} - 1 \right) +
+\left(\sum_{\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q), s_{qc} = -\infty}
+\max(s_q - s_q^{\min} - 1, 0) \right).
+\end{equation}
+
+(Note that $b$ can be calculated in a single $O(N)$ traversal of the tree.)
+
+Lastly, case \textit{(3)} occurs when the reference scale is less than the
+minimum query scale $s_q^{\min}$.  Define the minimum reference scale as
+$s_r^{\min}$.  If $s_q^{\min} - s_r^{\min} > 0$, each query node may have up to
+$s_q^{\min} - s_r^{\min}$ reference recursions of type \textit{(3)}, and there
+are $N$ query nodes.  Thus, the total number of reference recursions of type
+\textit{(3)} is bounded by $N \max(s_q^{\min} - s_r^{\min}, 0)$.
+
+Combining these three cases, we get the following expression for an upper bound
+$\theta$ on the total number of reference recursions:
+
+\begin{equation}
+\theta = \max(s_r^T - s_q^T, 0) + b + N (s_q^{\min} - s_r^{\min}).
+\end{equation}
+
+Now, we can express these in terms of more common constants which are related to
+the aspect ratio of the data.  Define $\delta_q$ as the minimum nonzero pairwise
+distance in $S_q$ and $\eta_q$ as the furthest pairwise distance in $S_q$ (and
+define $\delta_r$ and $\eta_r$ accordingly for the set $S_r$).  Then, $s_r^T =
+\lceil \log_2 \eta_r \rceil$, $s_q^T = \lceil \log_2 \eta_q$, $s_r^{\min} =
+\lceil \log_2 \delta_r \rceil$, and $s_q^{\min} = \lceil \log_2 \delta_q
+\rceil$.  This gives a more reasonable expression:
+
+\begin{equation}
+\theta = \max(\log_2(\eta_r / \eta_q), 0) + b + N \max(\log_2(\delta_q /
+\delta_r), 0).
+\end{equation}
+
+Returning to our original line of reasoning, we can combine these results to
+give the result of the theorem:
+
+\begin{equation}
+O(c_r^4 \psi \chi | R^* | \theta) + O(c_r^4 \psi | R^* | N) \sim O(c_r^4 \psi
+\chi |R^*| (N + \theta).
+\end{equation}
 \end{proof}
-\vspace*{-0.8em}
+
+A quick discussion of the numerous constants we have introduced and what they
+mean.  In the monochromatic case, the terms of $\theta$ depending on $\eta_q$,
+$\eta_r$, $\delta_q$, and $\delta_r$ disappear entirely.  However we are left
+with the $b$ term, which is kind of a measure of imbalance of the tree.  The
+best case is a case where the tree is perfectly balanced and there are no
+`missing levels' where a child of a node is more than one level away from its
+parent.  In practice, this quantity is probably much less than $N$ and should
+not scale significantly with $N$ (those points which contribute to $b$ the most
+will probably be outliers).  In the worst case, $b$ is bounded by the maximum
+depth of the tree multiplied by the number of points, which is trivially $O(N
+\log (\eta_q / \eta_r)$.  It is likely that a tighter bound of $O(N \log N)$
+could be established.
+
+There are two corner cases for the non-monochromatic case: all extra reference
+recursions are of type \textit{(1)}, and all extra reference recursions are of
+type \textit{(3)}.  The first case presents the best situation, because all of
+the reference recursions are done before the first query node is recursed in.
+Therefore the bulk of the reference recursion work is done by with the query
+tree root as the query node; this is the best case.  The worst case is the
+latter, where recursion in the reference tree cannot begin until reaching the
+bottom of the query tree.  Thus, no pruning can occur until reaching the leaves
+of the query tree, and the recursion is essentially equivalent to a single-tree
+recursion for each query point plus the added overhead of the recursion of the
+query tree with one reference node.  This second case degenerates to something
+like $O(N \log N)$ or $O(N^2)$ time depending on how you look at it.
 

### 2014-11-12 bounds-files/imbalance.tex
I'll figure out how to do the bound soon.
Update; rewritten intro, imbalance, and bound.

@@
 \end{defn}
+
+A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
+imbalance $i(\mathscr{T}_b) = 0$.  A worst-case cover tree $\mathscr{T}_w$ which
+is entirely a chain-like structure with maximum scale $s_{\max}$ and minimum
+scale $s_{\min}$ will have imbalance $i(\mathscr{T}_w) = N (s_{\max} -
+s_{\min})$.  Because of this chain-like structure, each level has only one node
+and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
+that the imbalance is actually quadratic in $N$!
+
+As we will see later empirically, the notion of cover tree imbalance does a good
+job of capturing the `goodness' of a tree.  We will use this notion in order to
+quantify the worst-case performance of dual-tree algorithms using the cover
+tree.
+
+{\bf TODO: Can any bound be proven on $i(\mathscr{T})$?  I haven't thought of
+anything successfully yet.}

### 2014-11-12 bounds-files/sec01-supporting-lemmas.tex
Remove negative vspace and remove NIPS formatting.
Update; rewritten intro, imbalance, and bound.

@@
 For simplicity, the algorithms considered in this paper will be presented in a
-tree-independent context, as in \cite{curtin2013tree}, but the only type of
-tree we will consider is the cover tree \cite{langford2006}, and the only type
-of traversal we will consider is the cover tree pruning dual-tree traversal, which
+tree-independent context, as in \cite{curtin2013tree}, but the only type of tree
+we will consider is the cover tree \cite{langford2006}, and the only type of
+traversal we will consider is the cover tree pruning dual-tree traversal, which
 we will describe below.
@@
 As we will be making heavy use of trees, we must establish notation (taken from
-\cite{curtin2013tree}).  $\mathscr{T}_i$ represents a tree; $\mathscr{N}_i$
-represents a node; $\mathscr{C}(\mathscr{N}_i)$ is the set of children of the
-node $\mathscr{N}_i$; $\mathscr{P}(\mathscr{N}_i)$ is the set of points held in
-a node; and $\mathscr{D}^p(\mathscr{N}_i)$ is the set of all descendant points
-of the node $\mathscr{N}_i$.
+\cite{curtin2013tree}).  The notation we will be using is defined in Table
+\ref{tab:notation}.
 
-First, we will introduce the cover tree\footnote{In our discussions, we will
-only consider the explicit cover tree, as opposed to the implicit cover tree
-\cite{langford2006}.  This has the effect of making the traversal more complex,
-but it helps simplify our proofs.  Familiarity with the implicit cover tree
-representation should not
-be required for understanding any of our work here.}.  The cover tree is a
-leveled hierarchical data structure originally proposed for the task of nearest
-neighbor search by Beygelzimer et~al.~\cite{langford2006}.  Each cover tree node
-holds one point $p_i$ and may have a number of child nodes; if the cover tree
-node has children, then one child will be the self-child, which holds the same
-point.  In addition, each level in the tree is indexed by an integer scale
-$s_i$; as the tree is descended, $s_i$ decreases.  Every descendant of a node
-$\mathscr{N}_i$ has a maximum distance of $2^{s_i + 1}$ from $p_i$.
+\begin{table}
+{\small
+\begin{center}
+\begin{tabular}{|c|l|}
+\hline
+{\bf Symbol} & {\bf Description} \\ \hline
+$\mathscr{N}$ & A tree node \\ \hline
+$\mathscr{C}_i$ & Set of child nodes of $\mathscr{N}_i$ \\ \hline
+$\mathscr{P}_i$ & Set of points held in $\mathscr{N}_i$ \\ \hline
+$\mathscr{D}_i^n$ & Set of descendant nodes of $\mathscr{N}_i$ \\ \hline
+$\mathscr{D}_i^p$ & Set of points contained in $\mathscr{N}_i$ and
+$\mathscr{D}_i^n$ \\ \hline
+$\mu_i$ & Center of $\mathscr{N}_i$ (for cover trees, $\mu_i = p_i$) \\ \hline
+$\lambda_i$ & Furthest descendant distance \\ \hline
+\end{tabular}
+\end{center}
+}
+\caption{Notation for trees.  See \cite{curtin2013tree} for details.}
+\label{tab:notation}
+\end{table}
+
+The cover tree is a leveled hierarchical data structure originally proposed for
+the task of nearest neighbor search by Beygelzimer
+et~al.~\cite{langford2006}.  The original description of the tree describes both
+an {\it implicit} and {\it explicit} representation, where the explicit
+representation is the actual structure that would be implemented in practice.
+Thus, we will consider only the explicit representation in this paper.
+
+Each node $\mathscr{N}_i$ in the cover tree is indexed by an integer scale $s_i$
+and holds one point $p_i$ which is also the center of the node ($p_i = \mu_i$).
+The cover tree node may have a number of child nodes; if the cover tree node
+does have children, then one child will be the self-child, which holds the same
+point $p_i$ with a smaller scale $s_i$.  Each child of $\mathscr{N}_i$ will have
+scale less than $s_i$, and every descendant point of a node $\mathscr{N}_i$
+(that is, the set $\mathscr{D}_i^p$) has a maximum distance of $2^{s_i + 1}$
+from the point $p_i$.
 
@@
 
-\vspace*{-0.5em}
 \begin{itemize}
@@
 \end{itemize}
-\vspace*{-0.5em}
 
@@
 
-\vspace*{-0.8em}
 \begin{equation}
@@
 \end{equation}
-\vspace*{-1.4em}
 
@@
 karger2002finding}.
-%properties of cover trees in \cite{langford2006}.
-One result we will use is that
-the number of children of any cover tree node is $c^4$ (Lemma 4.1,
-\cite{langford2006}), where $c$ is the expansion constant of the dataset the
-cover tree is built on.
+
+The expansion constant can be used to show a few useful bounds on various
+properties of the cover tree; we restate these results below, given some cover
+tree built on a dataset $S$ with expansion constant $c$ and $|S| = N$:
+
+\begin{itemize}
+  \item {\bf Width bound:} no cover tree node has more than $c^4$ children
+(Lemma 4.1, \cite{langford2006}).
+
+  \item {\bf Depth bound:} the maximum depth of any node is $O(c^2 \log N)$
+(Lemma 4.3, \cite{langford2006}).
+
+  \item {\bf Space bound:} a cover tree has $O(N)$ nodes (Theorem 1,
+\cite{langford2006}).
+\end{itemize}
 
@@
 
-Next we show the standard cover tree dual-tree pruning traversal, in a
-problem-independent form.  This is given in Algorithm \ref{alg:cover-tree-dual}.
-The algorithm is adapted from the original cover tree batch nearest neighbors
-algorithm \cite{langford2006} and implements the dual-tree recursion for the
-cover tree.  This traversal was first presented by Curtin and Ram
-\cite{curtin2014dual}.
-%
-The correctness of this
-traversal for various problem-dependent choices of the \texttt{BaseCase()} and
-the \texttt{Score()} functions has been proven \cite{curtin2013tree}.
-%  Now, we must introduce the inverse degree of bichromaticity.
-
-\begin{algorithm}[htb]
-  \begin{algorithmic}[1]
-    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, set of reference nodes
-$R$ \label{alg:line:ct-dual-input}
-    \STATE {\bfseries Output:} none
-
-    \medskip
-    \STATE $s^{\max}_r \gets \max_{\mathscr{N}_r \in R} s_r$
-    \IF{$(s_q < s^{\max}_r)$ {\bf or} $(s^{\max}_r = -\infty)$}
-\label{alg:line:ct-dual-ref-recursion-start}
-      \FORALL{$\mathscr{N}_r \in R$} \label{alg:line:ct-dual-base-case-start}
-        \STATE \texttt{BaseCase($p_q$, $p_r$)}
-      \ENDFOR \label{alg:line:ct-dual-base-case-end}
-      \STATE $R_r \gets \{ \mathscr{N}_r \in R : s_r = s^{\max}_r \}$
-\label{alg:line:ct-dual-ref-set}
-      \STATE $R_{r - 1} \gets \{ \mathscr{C}(\mathscr{N}_r) : \mathscr{N}_r \in
-R_r \} \cup (R \setminus R_r)$ \label{alg:line:ct-dual-ref-children}
-      \STATE $R'_{r - 1} \gets \{ \mathscr{N}_r \in R_{r - 1} :
-\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
-\label{alg:line:ct-dual-ref-score}
-      \STATE recurse with $\mathscr{N}_q$ and $R'_{r - 1}$
-\label{alg:line:ct-dual-ref-recursion-end}
-    \ELSE \label{alg:line:ct-dual-query-recursion-start}
-      \FORALL{$\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q)$}
-        \STATE $R' \gets \{ \mathscr{N}_r \in R :
-\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
-\label{alg:line:ct-dual-query-pruning}
-        \STATE recurse with $\mathscr{N}_{qc}$ and $R'$
-\label{alg:line:ct-dual-query-recursion}
-      \ENDFOR \label{alg:line:ct-dual-query-recursion-end}
-    \ENDIF
-  \end{algorithmic}
-  \caption{The standard pruning dual-tree traversal for cover trees.}
-  \label{alg:cover-tree-dual}
-\end{algorithm}
-
-\begin{defn}
-\label{def:bichromaticity}
-Let $\mathscr{T}_q$ and $\mathscr{T}_r$ be two cover trees built on query set
-$S_q$ and reference set $S_r$, respectively.  Now consider a pruning dual-tree
-traversal (such as Algorithm \ref{alg:cover-tree-dual}) with the property that
-the scales of nodes in $\mathscr{T}_q$ and $\mathscr{T}_r$ are kept as close as
-possible---that is, the tree with the larger scale is always descended.  Then,
-the {\bf inverse degree of bichromaticity} $\nu$ of the tree pair
-$(\mathscr{T}_q, \mathscr{T}_r)$ is the maximum number of recursions in
-$\mathscr{T}_r$ following a recursion in $\mathscr{T}_q$ before another
-recursion in $\mathscr{T}_q$ or the termination of the algorithm (whichever
-happens first).
-
-This quantity is related to the {\bf degree of bichromaticity} $\kappa$
-\cite{ram2009}, which is the maximum number of recursions in $\mathscr{T}_q$
-between any two recursions in $\mathscr{T}_r$.
-\end{defn}
-
-Although $\kappa$ and $\nu$ can be difficult to bound, it is reasonable to
-%monochromatic case where $S_q = S_r$, the trees $\mathscr{T}_q$ and
-%$\mathscr{T}_r$ are identical, so $\kappa = \nu = 1$.  It is reasonable to
-say that $\kappa \sim \nu \sim O(1)$ for datasets with roughly similar
-distributions of pairwise distances.
-
 \begin{lemma}
@@
 
-\vspace*{-0.7em}
 \begin{equation}
@@
 \end{equation}
-\vspace*{-1.1em}
 \label{lem:packing}
@@
 
-\vspace*{-1em}
 \begin{proof}
@@
 \end{equation}\end{proof}
-\vspace*{-2.2em}

### 2014-11-12 bounds.tex
Remove negative vspace and remove NIPS formatting.
Some thoughts on tree imbalance.
Update; rewritten intro, imbalance, and bound.

@@
 \documentclass{article} % For LaTeX2e
-\usepackage{nips14submit_e,times}
 \usepackage{hyperref}
@@
 \title{Plug-and-play dual-tree algorithm runtime analysis}
+\author{Ryan R. Curtin, Dongryeol Lee, William B. March, Parikshit Ram}
+\date{\today}
 
@@
 
-%\nipsfinalcopy % Uncomment for camera-ready version
-
 \begin{document}
@@
 
-\vspace*{-0.8em}
 \begin{abstract}
@@
 
-\vspace*{-1em}
-
 %\section{Dual-tree algorithms for pairwise statistical problems (Introduction)}
 \section{Dual-tree algorithms}
-\vspace*{-0.5em}
 
@@
 bottlenecks that can be expressed as pairwise statistical problems.
-All-nearest-neighbor search is one such problem, naively requiring $O(N)$ time
+Nearest neighbor search is one such problem, naively requiring $O(N)$ time
 to answer in single query in a reference set of size $N$; answering $O(N)$
-queries subsequently requiring the prohibitive $O(N^2)$ time. Kernel density
+queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
 estimation is also a pairwise statistical problem, naively requiring $O(N^2)$
@@
 
+Gray and Moore \cite{nbody} generalized the fast multipole method from
+computational physics to obtain dual-tree algorithms.  These are extremely
+useful when there are large query sets, not just a few query points.  Dual-tree
+algorithms can be easily understood through the recent framework of Curtin
+et~al. \cite{curtin2013tree}: two trees (a query tree and a reference tree) are
+traversed by a {\it pruning dual-tree traversal}.  This traversal visits
+combinations of nodes from the trees in some sequence (each combination consisting of a query node and a
+reference node), calling a problem-specific \texttt{Score()} function to
+determine if the node combination can be pruned.  If not, then a
+problem-specific \texttt{BaseCase()} function is called for each combination of
+points held in the query node and reference node.  This has significant
+similarity to the more common single-tree branch-and-bound algorithms, except
+that the algroithm must recurse into child nodes of {\it both} the query tree
+and reference tree.
+
+There exist numerous dual-tree algorithms for problems as diverse as kernel
+density estimation \cite{gray2003nonparametric}, mean shift \cite{wang2007fast},
+minimum spanning tree calcul
+
 %\item General tree/problem independent dual-tree algorithms (maybe)
@@
 %
-%\vspace*{-0.5em}
 %\begin{itemize}
@@
 this paper are:
-\vspace*{-0.3em}
 \begin{itemize}
@@
 
-\vspace*{-1em}
 \section{Preliminaries}
-\vspace*{-0.6em}
 
@@
 
-\vspace*{-0.6em}
+\section{Tree Imbalance}
+
+\input{bounds-files/imbalance.tex}
+
 \section{General runtime bound}
-\vspace*{-0.4em}
 
@@
 
-\vspace*{-0.8em}
 \section{Nearest neighbor search}
-\vspace*{-0.4em}
 
@@
 
-\vspace*{-0.6em}
 \section{Approximate kernel density estimation}
-\vspace*{-0.6em}
 
@@
 
-\vspace*{-0.8em}
 \section{Range Search and Range Count}
-\vspace*{-0.4em}
 
@@
 % Appendix doesn't have anything useful or non-straightforward.
-%\appendix
-%\input{appendix-ec}
+\appendix
+\input{appendix-ec}
 

### 2014-11-12 bounds.tex
Refactor introduction.

@@
 %range count/search.
-Numerous machine learning algorithms contain pairwise statistical
-problems in their core. Often, tree structures are used
-to solve these problems efficiently. Dual-tree
-algorithms using cover trees have
-rigorous worst-case runtime guarantees for some of these problems. In this paper, we present a {\em
-problem-independent} runtime guarantee for {\em any} dual-tree
-algorithm using the cover tree, separating out the problem-dependent and
-the problem-independent elements. This allows us to just plug in
-bounds for the problem-dependent elements to get runtime guarantees
-for dual-tree algorithms for any pairwise statistical problem. We
-demonstrate this plug-and-play procedure for nearest-neighbor search
-and approximate kernel density estimation to get improved runtime
-guarantees.  Under some assumptions, we also present the first linear runtime
-guarantee for dual-tree based range search.
+Numerous machine learning algorithms contain pairwise statistical problems in
+their core. Often, tree structures are used to solve these problems efficiently.
+Dual-tree algorithms using cover trees have rigorous worst-case runtime
+guarantees for some of these problems. In this paper, we present a {\em
+problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
+the cover tree, separating out the problem-dependent and the problem-independent
+elements. This allows us to just plug in bounds for the problem-dependent
+elements to get runtime guarantees for dual-tree algorithms for any pairwise
+statistical problem. We demonstrate this plug-and-play procedure for
+nearest-neighbor search and approximate kernel density estimation to get
+improved runtime guarantees.  Under some assumptions, we also present the first
+linear runtime guarantee for dual-tree based range search.
 \end{abstract}
 
-%\section{Dual-tree algorithms for pairwise statistical problems (Introduction)}
 \section{Dual-tree algorithms}
 
-%\begin{itemize}
-%\item What are pairwise statistical problems? Examples?
-%
-%A surprising number of machine learning algorithms have bottlenecks which, if
-%implemented naively, depend on the calculation of potentially large numbers of
-%distances.  A common example is nearest neighbor search: if search
-%for the nearest neighbor of a single query point in a reference set of size $N$,
-%simple linear scan costs $O(N)$ time.  For many decades, spatial data structures
-%have been employed to accelerate this type of computation, beginning with the
-%quad tree \cite{finkel1974quad}; building trees on the reference set often
-%results in expected runtimes of $O(\log N)$ per query point.
-%
 A surprising number of machine learning algorithms have computational
@@
 
-%\item What are dual-tree algorithms? Brief explanation.
-%
-%More recently, Gray and Moore \cite{nbody} observed that problems often have
-%large query {\em sets}, not just a few query points, and introduced dual-tree
-%algorithms.  One example is all-nearest-neighbor search: find the nearest
-%neighbor of every point in the reference set.  Simple implementations scale
-%poorly, but dual-tree algorithms can obtain orders-of-magnitude speedup for a
-%multitude of problems.
-%
-%Following this, numerous dual-tree algorithms have been developed to solve a
-%wide variety of problems; these include kernel density estimation
-%\cite{gray2003nonparametric}, fast max-kernel search \cite{curtin2013fast},
-%approximate nearest neighbor search \cite{ram2009rank}, mean shift
-%\cite{wang2007fast}, minimum spanning tree calculation
-%\cite{march2010euclidean}, $n$-point correlation function estimation
-%\cite{march2012fast}, kernel summation \cite{lee2006faster}, particle
-%smoothing \cite{klaas2006fast}, variational inference \cite{amizadeh2012variational} and range search \cite{nbody}.
-%
-Gray and Moore \cite{nbody} generalized the fast-multipole-method from
-computational physics to obtain {\em dual-tree algorithms}. These are extremely
-useful when there are large query {\em sets}, not just a few query points.
-%\item What (pairwise statistical) problems have they been applied to?
-%
-Following this, numerous dual-tree algorithms have been developed to solve a
-wide variety of problems; these include kernel density estimation
-\cite{gray2003nonparametric}, fast max-kernel search \cite{curtin2013fast},
-approximate nearest neighbor search \cite{ram2009rank}, mean shift
-\cite{wang2007fast}, minimum spanning tree calculation
-\cite{march2010euclidean}, $n$-point correlation function estimation
-\cite{march2012fast}, kernel summation \cite{lee2006faster}, particle
-smoothing \cite{klaas2006fast}, variational inference \cite{amizadeh2012variational} and range search \cite{nbody}.
-%
-%\item Existing analyses of dual-tree algorithms
-Some of these algorithms are derived using the cover tree
-\cite{langford2006}, a data structure with beneficial theoretical qualities.
-Dual-tree all nearest-neighbor search and approximate kernel density estimation with cover trees have $O(N)$ runtime guarantees for $O(N)$ queries \cite{ram2009}; minimum spanning tree calculation scales as $O(N \log N)$ \cite{march2010euclidean}.
-Other problems have similar worst-case guarantees \cite{curtin2014dual, march2013multi}.
-% Is there a better reference for the npcf runtime bound?
-
 Gray and Moore \cite{nbody} generalized the fast multipole method from
 computational physics to obtain dual-tree algorithms.  These are extremely
-useful when there are large query sets, not just a few query points.  Dual-tree
-algorithms can be easily understood through the recent framework of Curtin
-et~al. \cite{curtin2013tree}: two trees (a query tree and a reference tree) are
-traversed by a {\it pruning dual-tree traversal}.  This traversal visits
-combinations of nodes from the trees in some sequence (each combination consisting of a query node and a
-reference node), calling a problem-specific \texttt{Score()} function to
-determine if the node combination can be pruned.  If not, then a
-problem-specific \texttt{BaseCase()} function is called for each combination of
-points held in the query node and reference node.  This has significant
-similarity to the more common single-tree branch-and-bound algorithms, except
-that the algroithm must recurse into child nodes of {\it both} the query tree
-and reference tree.
+useful when there are large query sets, not just a few query points.  Instead of
+building a tree on the reference set and searching with each query point
+separately, Gray and Moore suggest also building a query tree and traversing
+both the query and reference trees simultaneously (a {\it dual-tree traversal},
+from which the class of algorithms takes its name).
+
+Dual-tree algorithms can be easily understood through the recent framework of
+Curtin et~al. \cite{curtin2013tree}: two trees (a query tree and a reference
+tree) are traversed by a {\it pruning dual-tree traversal}.  This traversal
+visits combinations of nodes from the trees in some sequence (each combination
+consisting of a query node and a reference node), calling a problem-specific
+\texttt{Score()} function to determine if the node combination can be pruned.
+If not, then a problem-specific \texttt{BaseCase()} function is called for each
+combination of points held in the query node and reference node.  This has
+significant similarity to the more common single-tree branch-and-bound
+algorithms, except that the algroithm must recurse into child nodes of {\it
+both} the query tree and reference tree.
 
@@
 density estimation \cite{gray2003nonparametric}, mean shift \cite{wang2007fast},
-minimum spanning tree calcul
+minimum spanning tree calculation \cite{march2010euclidean}, $n$-point
+correlation function estimation \cite{march2012fast}, max-kernel search
+\cite{curtin2013fast}, particle smoothing \cite{klaas2006fast}, variational
+inference \cite{amizadeh2012variational}, and range search \cite{nbody}, to name
+a few.
 
-%\item General tree/problem independent dual-tree algorithms (maybe)
-%\item What is lacking -- general dual-tree analysis (since the algorithms are so similar in structure)
-%A recent framework for understanding and unifying dual-tree algorithms
-%observes that all dual-tree algorithms are particularly similar, and generalizes
-%them \cite{curtin2013tree}.  A
-%nice benefit of this viewpoint is that it allows quick and easy development of
-%new dual-tree algorithms.  In this paper, we make the same observation about
-%runtime proofs for cover tree dual-tree algorithms, and show that
-%runtime analysis of generalized cover tree dual-tree algorithms is
-%possible.  Thus, the contributions of this paper are:
-%
-%\begin{itemize}
-%  \item A general runtime bound for dual-tree algorithms that use the cover
-%tree.
-%
-%  \item \mbox{A nearest neighbor search runtime bound,
-%slightly improving on existing linear-time results.}
-%
-%  \item An approximate kernel density estimation runtime bound,
-%improving on existing linear-time results.
-%
-%  \item Novel runtime analysis of dual-tree range count and range
-%search.
-%\end{itemize}
-Curtin et.~al. \cite{curtin2013tree} observe that all dual-tree
-algorithms are similar and propose a unified framework which allows the
-development of dual-tree algorithms in a problem independent (and tree
-independent) manner. In this paper, we make the same observation about runtime
-proofs for cover tree dual-tree algorithms, and show that runtime analysis
-of a generalized dual-tree algorithm is possible. The specific contributions of
-this paper are:
-\begin{itemize}
-  \item A general runtime bound for dual-tree algorithms that use the cover
-tree.
-  \item An improved linear runtime guarantee for all nearest-neighbor search.
-  \item An improved linear runtime guarantee for approximate kernel density
-estimation.
-  \item The first linear runtime guarantee for dual-tree range count and range
-search.
-\end{itemize}
+Some of these algorithms are derived using the cover tree \cite{langford2006}, a
+data structure with beneficial theoretical qualities.  Dual-tree all
+nearest-neighbor search and approximate kernel density estimation with cover
+trees have $O(N)$ runtime guarantees for $O(N)$ queries \cite{ram2009}; minimum
+spanning tree calculation scales as $O(N \log N)$ \cite{march2010euclidean}.
+Other problems have similar worst-case guarantees \cite{curtin2014dual,
+march2013multi}.
 
-%The following section will detail theory prerequisites, and then introduce the
-%general runtime bound.
+In this work we combine the generalization of Curtin et~al.
+\cite{curtin2013tree} with the theoretical results of Beygelzimer et~al.
+\cite{langford2006} (and others) in order to develop a worst-case runtime bound
+for any dual-tree algorithm when the cover tree is used.
 
-%\item This paper's contribution:
-%  \begin{itemize}
-%  \item Problem independent dual-tree algorithm analysis
-%  \item Examples of final problem dependent runtime bounds for NNS and KDE which match existing results (from LTAPS) -- implying nothing is lost yet for doing problem independent analysis
-%  \item Novel runtime analyses of dual-tree range count (or range search).
-%  \end{itemize}
-%\item This paper's structure (to be written last)
-%\end{itemize}
+Section \ref{sec:preliminaries} lays out the required background, notation, and
+introduces the cover tree and its associated theoretical properties.  Readers
+familiar with the cover tree literature and dual-tree algorithms (especially
+\cite{curtin2013tree}) may find that section to be review.  Following that, we
+introduce an intuitive measure of cover tree imbalance in Section
+\ref{sec:imbalance}.  This measure of imbalance is then used to prove the main
+result of the paper in Section \ref{sec:bound}, which is a worst-case runtime
+bound for generalized dual-tree algorithms.  We apply this result to three
+specific problems: nearest neighbor search (Section \ref{sec:nns}), approximate
+kernel density estimation (Section \ref{sec:akde}), and range search / range
+count (Section \ref{sec:rs}), showing linear runtime bounds for each of those
+algorithms.  Each of these bounds is an improvement on the state-of-the-art, and
+in the case of range search, is the first such bound.  We then investigate the
+empirical performance of cover trees for these algorithms, with a focus on
+understanding the dataset-dependent constants that our bounds use.
 
 \section{Preliminaries}
+\label{sec:preliminaries}
 
@@
 \section{Tree Imbalance}
+\label{sec:imbalance}
 
@@
 \section{General runtime bound}
+\label{sec:bound}
 
@@
 \section{Nearest neighbor search}
+\label{sec:nns}
 
@@
 \section{Approximate kernel density estimation}
+\label{sec:akde}
 
@@
 \section{Range Search and Range Count}
+\label{sec:rs}
 

### 2014-11-13 figures/balanced-cover-tree.tex
Four figures I made.  I'm not sure if I like graphviz or not.

\begin{tikzpicture}[>=latex,line join=bevel,scale=0.47]
  \pgfsetlinewidth{1bp}
%%
\begin{scope}
  \pgfsetstrokecolor{black}
  \definecolor{strokecol}{rgb}{1.0,1.0,1.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{1.0,1.0,1.0};
  \pgfsetfillcolor{fillcol}
  \filldraw (0bp,0bp) -- (0bp,180bp) -- (486bp,180bp) -- (486bp,0bp) -- cycle;
\end{scope}
\begin{scope}
  \pgfsetstrokecolor{black}
  \definecolor{strokecol}{rgb}{1.0,1.0,1.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{1.0,1.0,1.0};
  \pgfsetfillcolor{fillcol}
  \filldraw (0bp,0bp) -- (0bp,180bp) -- (486bp,180bp) -- (486bp,0bp) -- cycle;
\end{scope}
\begin{scope}
  \pgfsetstrokecolor{black}
  \definecolor{strokecol}{rgb}{1.0,1.0,1.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{1.0,1.0,1.0};
  \pgfsetfillcolor{fillcol}
  \filldraw (0bp,0bp) -- (0bp,180bp) -- (486bp,180bp) -- (486bp,0bp) -- cycle;
\end{scope}
  \pgfsetcolor{black}
  % Edge: c2 -> c6
  \draw [-] (228.43bp,74.834bp) .. controls (218.25bp,64.938bp) and (204.48bp,51.546bp)  .. (185.8bp,33.385bp);
  % Edge: c3 -> c9
  \draw [-] (387bp,71.697bp) .. controls (387bp,63.983bp) and (387bp,54.712bp)  .. (387bp,36.104bp);
  % Edge: c3 -> c10
  \draw [-] (401.57bp,74.834bp) .. controls (411.75bp,64.938bp) and (425.52bp,51.546bp)  .. (444.2bp,33.385bp);
  % Edge: c2 -> c8
  \draw [-] (257.57bp,74.834bp) .. controls (267.75bp,64.938bp) and (281.52bp,51.546bp)  .. (300.2bp,33.385bp);
  % Edge: c1 -> c4
  \draw [-] (84.43bp,74.834bp) .. controls (74.25bp,64.938bp) and (60.476bp,51.546bp)  .. (41.796bp,33.385bp);
  % Edge: c1 -> c5
  \draw [-] (99bp,71.697bp) .. controls (99bp,63.983bp) and (99bp,54.712bp)  .. (99bp,36.104bp);
  % Edge: root -> c1
  \draw [-] (221.75bp,150.67bp) .. controls (197.4bp,138.83bp) and (157.28bp,119.33bp)  .. (120.33bp,101.37bp);
  % Edge: root -> c2
  \draw [-] (243bp,143.7bp) .. controls (243bp,135.98bp) and (243bp,126.71bp)  .. (243bp,108.1bp);
  % Edge: c2 -> c7
  \draw [-] (243bp,71.697bp) .. controls (243bp,63.983bp) and (243bp,54.712bp)  .. (243bp,36.104bp);
  % Edge: root -> c3
  \draw [-] (264.25bp,150.67bp) .. controls (288.6bp,138.83bp) and (328.72bp,119.33bp)  .. (365.67bp,101.37bp);
  % Node: c7
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (243bp,18bp) ellipse (27bp and 18bp);
  \draw (243bp,18bp) node {$\mathscr{N}_h$};
\end{scope}
  % Node: c9
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (387bp,18bp) ellipse (27bp and 18bp);
  \draw (387bp,18bp) node {$\mathscr{N}_j$};
\end{scope}
  % Node: c8
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (315bp,18bp) ellipse (27bp and 18bp);
  \draw (315bp,18bp) node {$\mathscr{N}_i$};
\end{scope}
  % Node: c3
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (387bp,90bp) ellipse (27bp and 18bp);
  \draw (387bp,90bp) node {$\mathscr{N}_d$};
\end{scope}
  % Node: c2
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (243bp,90bp) ellipse (27bp and 18bp);
  \draw (243bp,90bp) node {$\mathscr{N}_c$};
\end{scope}
  % Node: c1
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (99bp,90bp) ellipse (27bp and 18bp);
  \draw (99bp,90bp) node {$\mathscr{N}_b$};
\end{scope}
  % Node: c10
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (459bp,18bp) ellipse (27bp and 18bp);
  \draw (459bp,18bp) node {$\mathscr{N}_k$};
\end{scope}
  % Node: root
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (243bp,162bp) ellipse (27bp and 18bp);
  \draw (243bp,162bp) node {$\mathscr{N}_a$};
\end{scope}
  % Node: c6
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (171bp,18bp) ellipse (27bp and 18bp);
  \draw (171bp,18bp) node {$\mathscr{N}_g$};
\end{scope}
  % Node: c5
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (99bp,18bp) ellipse (27bp and 18bp);
  \draw (99bp,18bp) node {$\mathscr{N}_f$};
\end{scope}
  % Node: c4
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (27bp,18bp) ellipse (27bp and 18bp);
  \draw (27bp,18bp) node {$\mathscr{N}_e$};
\end{scope}
%
\draw[gray,thin,dashed] (0bp,58bp) -- (500bp,58bp);
\draw[gray,thin,dashed] (0bp,133bp) -- (500bp,133bp);

\draw (470bp,180bp) node[darkgray] {\scriptsize $s_a$};
\draw (470bp,122bp) node[darkgray] {\scriptsize $s_a - 1$};
\draw (470bp,47bp) node[darkgray] {\scriptsize $s_a - 2$};
\end{tikzpicture}

### 2014-11-13 figures/imbalanced-cover-tree.tex
Four figures I made.  I'm not sure if I like graphviz or not.

\begin{tikzpicture}[>=latex,line join=bevel,scale=0.47]
  \pgfsetlinewidth{1bp}
%%
\begin{scope}
  \pgfsetstrokecolor{black}
  \definecolor{strokecol}{rgb}{1.0,1.0,1.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{1.0,1.0,1.0};
  \pgfsetfillcolor{fillcol}
  \filldraw (0bp,0bp) -- (0bp,180bp) -- (342bp,180bp) -- (342bp,0bp) -- cycle;
\end{scope}
  \pgfsetcolor{black}
  % Edge: c2 -> c6
  \draw [-] (156.43bp,74.834bp) .. controls (146.25bp,64.938bp) and (132.48bp,51.546bp)  .. (113.8bp,33.385bp);
  % Edge: root -> c4
  \draw [-] (156.4bp,146.6bp) .. controls (131.06bp,121.62bp) and (78.82bp,70.101bp)  .. (41.639bp,33.435bp);
  % Edge: c2 -> c8
  \draw [-] (185.57bp,74.834bp) .. controls (195.75bp,64.938bp) and (209.52bp,51.546bp)  .. (228.2bp,33.385bp);
  % Edge: root -> c10
  \draw [-] (185.6bp,146.6bp) .. controls (210.94bp,121.62bp) and (263.18bp,70.101bp)  .. (300.36bp,33.435bp);
  % Edge: root -> c2
  \draw [-] (171bp,143.7bp) .. controls (171bp,135.98bp) and (171bp,126.71bp)  .. (171bp,108.1bp);
  % Edge: c2 -> c7
  \draw [-] (171bp,71.697bp) .. controls (171bp,63.983bp) and (171bp,54.712bp)  .. (171bp,36.104bp);
  % Node: c7
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (171bp,18bp) ellipse (27bp and 18bp);
  \draw (171bp,18bp) node {$\mathscr{N}_r$};
\end{scope}
  % Node: c8
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (243bp,18bp) ellipse (27bp and 18bp);
  \draw (243bp,18bp) node {$\mathscr{N}_s$};
\end{scope}
  % Node: c2
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (171bp,90bp) ellipse (27bp and 18bp);
  \draw (171bp,90bp) node {$\mathscr{N}_n$};
\end{scope}
  % Node: c10
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (315bp,18bp) ellipse (27bp and 18bp);
  \draw (315bp,18bp) node {$\mathscr{N}_t$};
\end{scope}
  % Node: root
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (171bp,162bp) ellipse (27bp and 18bp);
  \draw (171bp,162bp) node {$\mathscr{N}_m$};
\end{scope}
  % Node: c6
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (99bp,18bp) ellipse (27bp and 18bp);
  \draw (99bp,18bp) node {$\mathscr{N}_q$};
\end{scope}
  % Node: c4
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (27bp,18bp) ellipse (27bp and 18bp);
  \draw (27bp,18bp) node {$\mathscr{N}_p$};
\end{scope}
%
\draw[darkgray, thin] (-30bp,0bp) -- (-30bp,190bp);
\draw[gray,thin,dashed] (0bp,58bp) -- (350bp,58bp);
\draw[gray,thin,dashed] (0bp,133bp) -- (350bp,133bp);

\draw (320bp,180bp) node[darkgray] {\scriptsize $s_m$};
\draw (320bp,122bp) node[darkgray] {\scriptsize $s_m - 1$};
\draw (320bp,47bp) node[darkgray] {\scriptsize $s_m - 2$};
\end{tikzpicture}

### 2014-11-13 bounds-files/imbalance.tex
Refactor some text; add the figures; fix the citations.

@@
-A better understanding of the behavior of cover trees can be acheived partially
-by introducing a measure of tree imbalance.  Numerous measures of tree imbalance
-have already been established, but we aim to capture a different measure of
-imbalance that utilizes the leveled structure of the cover tree.
+It is well-known that imbalance in trees leads to degradation in performance;
+for instance, a $kd$-tree node with every descendant except one in its left
+child is effectively useless.  A $kd$-tree full of nodes like this will perform
+abysmally for nearest neighbor search, and it is not hard to generate a dataset
+that will cause a $kd$-tree of this sort.
+
+This sort of imbalance applies to all types of trees, not just $kd$-trees.  In
+our situation, we are interested in a better understanding of this imbalance for
+cover trees, and thus endeavor to introduce a more formal measure of imbalance
+which is correlated with tree performance.  Numerous measures of tree
+imbalance have already been established; one example is that proposed by
+\citet{colless1982review}, and another is Sackin's index \citep{sackin1972good},
+but we aim to capture a different measure of imbalance that utilizes the leveled
+structure of the cover tree.
 
@@
 one level lower; we will refer to this as a {\em perfectly balanced cover tree}.
-Figure \ref{fig:imbalance-bad}, on the other hand, has many children with far
-lower levels than their parents.  We will refer to this as an {\em imbalanced
-cover tree}.
+Figure \ref{fig:imbalance-bad}, on the other hand, contains the node
+$\mathscr{N}_m$ which has two children with scale two less than $s_m$.  We will
+refer to this as an {\em imbalanced cover tree}.  Note that in our definition,
+the balance of a cover tree has nothing to do with differing number of
+descendants in each child branch but instead only missing levels.
 
 \begin{figure}
+\begin{subfigure}[b]{0.585\textwidth}
+  \begin{center}
+    \input{figures/balanced-cover-tree.tex}
+  \end{center}
+  \caption{Balanced cover tree.}
+  \label{fig:imbalance-good}
+\end{subfigure}
+\begin{subfigure}[b]{0.415\textwidth}
+  \begin{center}
+    \input{figures/imbalanced-cover-tree.tex}
+  \end{center}
+  \caption{Imbalanced cover tree.}
+  \label{fig:imbalance-bad}
+\end{subfigure}
 \caption{Balanced and imbalanced cover trees.}
@@
 
-An imbalanced cover tree can---and often does---happen in practice.  Consider a
-dataset with a single outlier which is very far away from all of the other
-points.  Figure \ref{fig:outlier} shows what happens in this situation: the root
-node has two children; one of these children has only the outlier as a
-descendant, and the other child has the rest of the points in the dataset as a
-descendant.  In fact, it is easy to find datasets with multiple outliers which
-give rise to a chain-like structure at the top of the tree: see Figure
-\ref{fig:outliers} for an illustration\footnote{As a side note, this behavior is
-not limited to cover trees, and can happen to mean-split $kd$-trees too,
-especially in higher dimensions.}.
+An imbalanced cover tree can---and often does---happen in practice, and the
+imbalance may be far worse than the simple graphs of Figure \ref{fig:imbalance}.
+Consider a dataset with a single outlier which is very far away from all of the
+other points\footnote{Note also that for an outlier sufficiently far away, the
+expansion constant is $N - 1$.}.  Figure \ref{fig:outlier} shows what happens in
+this situation: the root node has two children; one of these children has only
+the outlier as a descendant, and the other child has the rest of the points in
+the dataset as a descendant.  In fact, it is easy to find datasets with multiple
+outliers which give rise to a chain-like structure at the top of the tree: see
+Figure \ref{fig:outliers} for an illustration\footnote{As a side note, this
+behavior is not limited to cover trees, and can happen to mean-split $kd$-trees
+too, especially in higher dimensions.}.
 
 \begin{figure}
+\begin{center}
+  \input{figures/single-outlier-tree.tex}
+\end{center}
 \caption{Single-outlier cover tree.}
@@
 \begin{figure}
-\caption{Multiple-outlier cover tree.}
+\begin{center}
+  \input{figures/multiple-outlier-tree.tex}
+\end{center}
+\caption{A badly-drawn multiple-outlier cover tree.}
 \label{fig:outliers}
@@
 A tree that has this chain-like structure all the way down is going to perform
-abysmally; motivated by this observation, we define a measure of tree imbalance.
+horrendously; motivated by this observation, we define a measure of tree
+imbalance.
 

### 2014-11-13 bounds-files/sec01-supporting-lemmas.tex
Refactor some text; add the figures; fix the citations.

@@
 For simplicity, the algorithms considered in this paper will be presented in a
-tree-independent context, as in \cite{curtin2013tree}, but the only type of tree
-we will consider is the cover tree \cite{langford2006}, and the only type of
-traversal we will consider is the cover tree pruning dual-tree traversal, which
-we will describe below.
+tree-independent context, as in \citet{curtin2013tree}, but the only type of
+tree we will consider is the cover tree \citep{langford2006}, and the only type
+of traversal we will consider is the cover tree pruning dual-tree traversal,
+which we will describe below.
 
-As we will be making heavy use of trees, we must establish notation (taken from
-\cite{curtin2013tree}).  The notation we will be using is defined in Table
+As we will be making heavy use of trees, we must establish notation \citep[taken
+from][]{curtin2013tree}.  The notation we will be using is defined in Table
 \ref{tab:notation}.
@@
 The cover tree is a leveled hierarchical data structure originally proposed for
-the task of nearest neighbor search by Beygelzimer
-et~al.~\cite{langford2006}.  The original description of the tree describes both
-an {\it implicit} and {\it explicit} representation, where the explicit
-representation is the actual structure that would be implemented in practice.
-Thus, we will consider only the explicit representation in this paper.
+the task of nearest neighbor search.  The original description of the tree
+describes both an {\it implicit} and {\it explicit} representation, where the
+explicit representation is the actual structure that would be implemented in
+practice.  Thus, we will consider only the explicit representation in this
+paper.
 
@@
 
-A batch construction algorithm is given in \cite{langford2006}.  In addition,
-the cover tree has a number of other useful properties and lemmas based on the
-expansion constant \cite{karger2002finding}; we restate the definition below.
+A batch construction algorithm is given by \citet{langford2006}, called
+\texttt{Construct}; but the details of construction are not relevant to our
+work, so we only mention this as a reference.  More relevant to our work is the
+number of useful and unique theoretical properties the cover tree possesses,
+based on the expansion constant \citep{karger2002finding}; we restate the
+definition below.
 
@@
 The expansion constant is used heavily in the cover tree literature.  It is
-related to a notion of instrinic dimensionality, and in many scenarios, $c$ is
-independent of the number of points in the dataset \cite{langford2006,
-karger2002finding}.
+related to a notion of instrinic dimensionality, and previous work has shown
+that there are many scenarios where $c$ is independent of the number of points
+in the dataset.
 
@@
 
+Lastly, we introduce a convenience lemma of our own which is a generalization of
+the packing arguments used by \citet{langford2006}.  This is a more flexible
+version of their argument.
+
 \begin{lemma}
@@
 / 2)|} \le c^{2 + \log_2 \rho}.
-\end{equation}\end{proof}
+\end{equation}
+\end{proof}

### 2014-11-13 bounds.tex
Refactor some text; add the figures; fix the citations.

@@
-\documentclass{article} % For LaTeX2e
+\documentclass[twoside,11pt]{article} % For LaTeX2e
+\usepackage{jmlr2e}
 \usepackage{hyperref}
@@
 \usepackage{mathrsfs}
-\usepackage{amsthm}
+%\usepackage{amsthm}
 \usepackage{amsmath}
 \usepackage{amsfonts}
-\usepackage{amssymb}
+%\usepackage{amssymb}
 \usepackage{algorithmic}
@@
 \usepackage{tikz}
-%\usepackage{subcaption}
-
-\title{Plug-and-play dual-tree algorithm runtime analysis}
-\author{Ryan R. Curtin, Dongryeol Lee, William B. March, Parikshit Ram}
-\date{\today}
-
-%\author{
-%David S.~Hippocampus\thanks{ Use footnote for providing further information
-%about author (webpage, alternative address)---\emph{not} for acknowledging
-%funding agencies.} \\
-%Department of Computer Science\\
-%Cranberry-Lemon University\\
-%Pittsburgh, PA 15213 \\
-%\texttt{hippo@cs.cranberry-lemon.edu} \\
-%\And
-%Coauthor \\
-%Affiliation \\
-%Address \\
-%\texttt{email} \\
-%\AND
-%Coauthor \\
-%Affiliation \\
-%Address \\
-%\texttt{email} \\
-%\And
-%Coauthor \\
-%Affiliation \\
-%Address \\
-%\texttt{email} \\
-%\And
-%Coauthor \\
-%Affiliation \\
-%Address \\
-%\texttt{email} \\
-%(if needed)\\
-%}
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to \LaTeX{} to determine where to break
-% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
-% puts 3 of 4 authors names on the first line, and the last on the second
-% line, try using \AND instead of \And before the third author name.
+\usepackage{subcaption}
 
@@
 
+%\newtheorem{defn}{Definition}
+%\newtheorem{thm}{Theorem}
+%\newtheorem{lemma}{Lemma}
+%\newtheorem{cor}{Corollary}
 \newtheorem{defn}{Definition}
 \newtheorem{thm}{Theorem}
-\newtheorem{lemma}{Lemma}
 \newtheorem{cor}{Corollary}
@@
 \begin{document}
+
+\title{Plug-and-play dual-tree algorithm runtime analysis}
+
+\author{\name Ryan R. Curtin \email ryan@ratml.org\\
+  \addr School of Computational Science and Engineering\\
+  Georgia Institute of Technology\\
+  Atlanta, GA 30332-0250, USA
+  \AND
+  \name Dongryeol Lee \email drselee@gmail.com\\
+  \addr GE Research\\
+  Schenedectadadaty, NY\\
+  \AND
+  \name William B. March \email march@ices.utexas.edu\\
+  \addr Austin, TX, USA\\
+  \AND
+  \name Parikshit Ram \email p.ram@gatech.edu\\
+  \addr Skytree
+}
+
+\editor{Unknown}
 
@@
 spatial data structures to accelerate this type of computation
-\cite{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
+\citep{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
 query under favorable conditions.
 
-Gray and Moore \cite{nbody} generalized the fast multipole method from
+\citet{nbody} generalized the fast multipole method from
 computational physics to obtain dual-tree algorithms.  These are extremely
@@
 Dual-tree algorithms can be easily understood through the recent framework of
-Curtin et~al. \cite{curtin2013tree}: two trees (a query tree and a reference
+\citet{curtin2013tree}: two trees (a query tree and a reference
 tree) are traversed by a {\it pruning dual-tree traversal}.  This traversal
@@
 There exist numerous dual-tree algorithms for problems as diverse as kernel
-density estimation \cite{gray2003nonparametric}, mean shift \cite{wang2007fast},
-minimum spanning tree calculation \cite{march2010euclidean}, $n$-point
-correlation function estimation \cite{march2012fast}, max-kernel search
-\cite{curtin2013fast}, particle smoothing \cite{klaas2006fast}, variational
-inference \cite{amizadeh2012variational}, and range search \cite{nbody}, to name
+density estimation \citep{gray2003nonparametric}, mean shift \citep{wang2007fast},
+minimum spanning tree calculation \citep{march2010euclidean}, $n$-point
+correlation function estimation \citep{march2012fast}, max-kernel search
+\citep{curtin2013fast}, particle smoothing \citep{klaas2006fast}, variational
+inference \citep{amizadeh2012variational}, and range search \citep{nbody}, to name
 a few.
 
-Some of these algorithms are derived using the cover tree \cite{langford2006}, a
+Some of these algorithms are derived using the cover tree \citep{langford2006}, a
 data structure with beneficial theoretical qualities.  Dual-tree all
 nearest-neighbor search and approximate kernel density estimation with cover
-trees have $O(N)$ runtime guarantees for $O(N)$ queries \cite{ram2009}; minimum
-spanning tree calculation scales as $O(N \log N)$ \cite{march2010euclidean}.
-Other problems have similar worst-case guarantees \cite{curtin2014dual,
+trees have $O(N)$ runtime guarantees for $O(N)$ queries \citep{ram2009}; minimum
+spanning tree calculation scales as $O(N \log N)$ \citep{march2010euclidean}.
+Other problems have similar worst-case guarantees \citep{curtin2014dual,
 march2013multi}.
 
-In this work we combine the generalization of Curtin et~al.
-\cite{curtin2013tree} with the theoretical results of Beygelzimer et~al.
-\cite{langford2006} (and others) in order to develop a worst-case runtime bound
-for any dual-tree algorithm when the cover tree is used.
+In this work we combine the generalization of \citet{curtin2013tree} with the
+theoretical results of \citet{langford2006} and others in order to develop a
+worst-case runtime bound for any dual-tree algorithm when the cover tree is
+used.
 
@@
 introduces the cover tree and its associated theoretical properties.  Readers
-familiar with the cover tree literature and dual-tree algorithms (especially
-\cite{curtin2013tree}) may find that section to be review.  Following that, we
-introduce an intuitive measure of cover tree imbalance in Section
-\ref{sec:imbalance}.  This measure of imbalance is then used to prove the main
-result of the paper in Section \ref{sec:bound}, which is a worst-case runtime
-bound for generalized dual-tree algorithms.  We apply this result to three
-specific problems: nearest neighbor search (Section \ref{sec:nns}), approximate
-kernel density estimation (Section \ref{sec:akde}), and range search / range
-count (Section \ref{sec:rs}), showing linear runtime bounds for each of those
-algorithms.  Each of these bounds is an improvement on the state-of-the-art, and
-in the case of range search, is the first such bound.  We then investigate the
-empirical performance of cover trees for these algorithms, with a focus on
-understanding the dataset-dependent constants that our bounds use.
+familiar with the cover tree literature and dual-tree algorithms
+\citep[especially][]{curtin2013tree} may find that section to be review.
+Following that, we introduce an intuitive measure of cover tree imbalance in
+Section \ref{sec:imbalance}.  This measure of imbalance is then used to prove
+the main result of the paper in Section \ref{sec:bound}, which is a worst-case
+runtime bound for generalized dual-tree algorithms.  We apply this result to
+three specific problems: nearest neighbor search (Section \ref{sec:nns}),
+approximate kernel density estimation (Section \ref{sec:akde}), and range search
+/ range count (Section \ref{sec:rs}), showing linear runtime bounds for each of
+those algorithms.  Each of these bounds is an improvement on the
+state-of-the-art, and in the case of range search, is the first such bound.  We
+then investigate the empirical performance of cover trees for these algorithms,
+with a focus on understanding the dataset-dependent constants that our bounds
+use.
 
@@
 
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
 \bibliography{bounds}

### 2014-11-13 bounds-files/general-runtime-bound.tex
Refactor some of this, and change the definition of node imbalance to make the general runtime bound much simpler to write about.

@@
 are actually used in dual-tree algorithms.  Although cover trees were originally
-intended for nearest neighbor search (see Algorithm \texttt{Find-All-Nearest},
-\cite{langford2006}), they can be adapted to a wide variety of problems: minimum
-spanning tree calculation \cite{march2010fast}, approximate nearest neighbor
-search \cite{ram2010rank}, Gaussian processes posterior calculation
-\cite{moore2014fast}, max-kernel search \cite{curtin2013fast,
-curtin2014dual} are some examples.  Further, through the tree-independent
-dual-tree algorithm abstraction of Curtin et~al. \cite{curtin2013tree}, other
-existing dual-tree algorithms can easily be adapted for use with cover trees.
+intended for nearest neighbor search \citep*[see Algorithm
+\texttt{Find-All-Nearest},][]{langford2006}, they can be adapted to a wide
+variety of problems: minimum
+spanning tree calculation \citep{march2010euclidean}, approximate nearest neighbor
+search \citep{ram2009rank}, Gaussian processes posterior calculation
+\citep{moore2014fast}, max-kernel search \citep{curtin2014dual} are some
+examples.  Further, through the tree-independent dual-tree algorithm abstraction
+of \citet{curtin2013tree}, other existing dual-tree algorithms can easily be
+adapted for use with cover trees.
 
-In the framework of tree-independent dual-tree algorithms
-\cite{curtin2013tree}, all that is necessary to describe a dual-tree algorithm
-is a point-to-point base case function (\texttt{BaseCase()}) and a node-to-node
-pruning rule (\texttt{Score()}).  These functions, which are often very
-straightforward, are then paired with a type of tree and a pruning dual-tree
-traversal to produce a working algorithm.  In later sections, we will consider
-specific examples.
+In the framework of tree-independent dual-tree algorithms, all that is necessary
+to describe a dual-tree algorithm is a point-to-point base case function
+(\texttt{BaseCase()}) and a node-to-node pruning rule (\texttt{Score()}).  These
+functions, which are often very straightforward, are then paired with a type of
+tree and a pruning dual-tree traversal to produce a working algorithm.  In later
+sections, we will consider specific examples.
 
-When using cover trees, the typical pruning dual-tree traversal is an adapted
-form of the original nearest neighbor search algorithm
-(\texttt{Find-All-Nearest}, \cite{langford2006}); this traversal is implemented
-in both the cover tree reference implementation and in the more flexible {\bf
-mlpack} library \cite{curtin2013mlpack}.  The algorithm is given in Algorithm
-\ref{alg:cover-tree-dual} and was originally presented by Curtin and Ram
-\cite{curtin2014dual}.  Initially, it is called with the root of the query tree
-and a reference set containing only the root of the reference tree.
-
-\begin{algorithm}[htb]
+\begin{algorithm}[tb]
   \begin{algorithmic}[1]
@@
 
+When using cover trees, the typical pruning dual-tree traversal is an adapted
+form of the original nearest neighbor search algorithm \citep[see
+\texttt{Find-All-Nearest},][]{langford2006}; this traversal is implemented in
+both the cover tree reference implementation and in the more flexible {\bf
+mlpack} library \citep{mlpack2013}.  The algorithm is given in Algorithm
+\ref{alg:cover-tree-dual} and was originally presented by
+\citet{curtin2014dual}.  Initially, it is called with the root of the query tree
+and a reference set containing only the root of the reference tree.
+
 This dual-tree recursion is a depth-first recursion in the query tree and a
@@
 set of queries $S_q$ of size $O(N)$, a standard cover tree based dual-tree
-algorithm (Algorithm \ref{alg:cover-tree-dual}) takes $O(c_r^4 | R^* | \chi
-\psi (N + \theta))$, where $ | R^* | $ is the maximum size of the reference set $R$ (line
-\ref{alg:line:ct-dual-input}) during the dual-tree recursion, $\chi$ is
+algorithm (Algorithm \ref{alg:cover-tree-dual}) takes
+
+\begin{equation}
+O\left(c_r^4 | R^* | \chi \psi (N + \theta)\right)
+\end{equation}
+
+\noindent time, where $ | R^* | $ is the maximum size of the reference set $R$
+(line \ref{alg:line:ct-dual-input}) during the dual-tree recursion, $\chi$ is
 the maximum possible runtime of \texttt{BaseCase()}, $\psi$ is the maximum
@@
 
-Next, consider case \textit{(2)}.  Each query recursion implies at least one
-reference recursion; thus there will be at least $O(N)$ recursions.  Any
-extra recursions above and beyond this occur because the query node does not
-have scale which is exactly one less than its parent.  When the query node is
-not a leaf, the number of extra reference recursions is bounded by the
-difference between the scales of the query node and its parent minus one (one is
-subtracted to account for the one reference recursion that must happen).  Each
-query node is visited once, and therefore the total number of extra recursions
-when the query node is not a leaf is bounded by
+Next, consider case \textit{(2)}.  In this situation, each query recursion
+implies at least one reference recursion before another query recursion.  For
+some query node $\mathscr{N}_q$, the
+exact number of reference recursions before the children of $\mathscr{N}_q$ are
+recursed into is bounded above by $i(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has
+imbalance $0$, then it is exactly one level below its parent, and thus there is
+only one reference recursion.  On the other hand, if $\mathscr{N}_q$ is many
+levels below its parent, then it is possible that a reference recursion may
+occur for each level in between; this is a maximum of $i(\mathscr{N}_q) + 1$.
+
+Because each query node in $\mathscr{T}_q$ is recursed into once, the total
+number of reference recursions before each query recursion is
 
 \begin{equation}
-\sum_{\mathscr{N}_q \in \mathscr{T}_q} \sum_{\mathscr{N}_{qc} \in
-\mathscr{C}(\mathscr{N}_q), s_{qc} \ne -\infty} s_q - s_{qc} - 1.
+\sum_{\mathscr{N}_q \in \mathscr{T}_q} i(\mathscr{N}_q) + 1 = b(\mathscr{T}_q) +
+O(N)
 \end{equation}
 
-When the query node is a leaf, then the number of recursions in case
-\textit{(2)} for that query node is bounded by the difference between the scale
-of the query node's parent and the smallest scale of a non-leaf node in the
-query tree (call this $s_q^{\min}$) minus one:
-
-\begin{equation}
-\sum_{\mathscr{N}_q \in \mathscr{T}_q} \sum_{\mathscr{N}_{qc} \in
-\mathscr{C}(\mathscr{N}_q), s_{qc} = -\infty} \max(s_q - s_q^{\min} - 1, 0).
-\end{equation}
-
-The $\max$ is necessary to account for the case where the query node's parent
-has scale $s_q^{\min}$; no reference recursions of type \textit{(2)} happen at
-all in this case.  Therefore the total number of extra reference recursions of
-type \textit{(2)} is bounded by $b$, where
-
-\begin{equation}
-b = \sum_{\mathscr{N}_q \in \mathscr{T}_q} \left(\sum_{\mathscr{N}_{qc} \in
-\mathscr{C}(\mathscr{N}_q), s_{qc} \ne -\infty} s_q - s_{qc} - 1 \right) +
-\left(\sum_{\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q), s_{qc} = -\infty}
-\max(s_q - s_q^{\min} - 1, 0) \right).
-\end{equation}
-
-(Note that $b$ can be calculated in a single $O(N)$ traversal of the tree.)
+\noindent since there are $O(N)$ nodes in the query tree.
 
 Lastly, case \textit{(3)} occurs when the reference scale is less than the
-minimum query scale $s_q^{\min}$.  Define the minimum reference scale as
-$s_r^{\min}$.  If $s_q^{\min} - s_r^{\min} > 0$, each query node may have up to
-$s_q^{\min} - s_r^{\min}$ reference recursions of type \textit{(3)}, and there
-are $N$ query nodes.  Thus, the total number of reference recursions of type
-\textit{(3)} is bounded by $N \max(s_q^{\min} - s_r^{\min}, 0)$.
+minimum query scale $s_q^{\min}$; that is, all of the query recursions are done.
+Define the minimum reference scale as $s_r^{\min}$.  If $s_q^{\min} - s_r^{\min}
+> 0$, each query node may have up to $s_q^{\min} - s_r^{\min}$ reference
+recursions of type \textit{(3)}, and there are $O(N)$ query nodes.  Thus, the
+total number of reference recursions of type \textit{(3)} is bounded by $O(N)
+\max(s_q^{\min} - s_r^{\min}, 0)$.
 
@@
 \begin{equation}
-\theta = \max(s_r^T - s_q^T, 0) + b + N (s_q^{\min} - s_r^{\min}).
+\theta = \max(s_r^T - s_q^T, 0) + b(\mathscr{T}_q) + O(N) + O(N) (s_q^{\min} -
+s_r^{\min}).
 \end{equation}
@@
 define $\delta_r$ and $\eta_r$ accordingly for the set $S_r$).  Then, $s_r^T =
-\lceil \log_2 \eta_r \rceil$, $s_q^T = \lceil \log_2 \eta_q$, $s_r^{\min} =
-\lceil \log_2 \delta_r \rceil$, and $s_q^{\min} = \lceil \log_2 \delta_q
+\lceil \log_2 \eta_r \rceil$, $s_q^T = \lceil \log_2 \eta_q \rceil$, $s_r^{\min}
+= \lceil \log_2 \delta_r \rceil$, and $s_q^{\min} = \lceil \log_2 \delta_q
 \rceil$.  This gives a more reasonable expression:
@@
 \begin{equation}
-\theta = \max(\log_2(\eta_r / \eta_q), 0) + b + N \max(\log_2(\delta_q /
-\delta_r), 0).
+\theta = \max\left(\log_2\left(\frac{\eta_r}{\eta_q}\right), 0\right) +
+b(\mathscr{T}_q) + O(N) + O(N)
+\max\left(\log_2\left(\frac{\delta_q}{\delta_r}\right), 0\right).
 \end{equation}
@@
 O(c_r^4 \psi \chi | R^* | \theta) + O(c_r^4 \psi | R^* | N) \sim O(c_r^4 \psi
-\chi |R^*| (N + \theta).
+\chi |R^*| (N + \theta)).
 \end{equation}
@@
 
-A quick discussion of the numerous constants we have introduced and what they
-mean.  In the monochromatic case, the terms of $\theta$ depending on $\eta_q$,
-$\eta_r$, $\delta_q$, and $\delta_r$ disappear entirely.  However we are left
-with the $b$ term, which is kind of a measure of imbalance of the tree.  The
-best case is a case where the tree is perfectly balanced and there are no
-`missing levels' where a child of a node is more than one level away from its
-parent.  In practice, this quantity is probably much less than $N$ and should
-not scale significantly with $N$ (those points which contribute to $b$ the most
-will probably be outliers).  In the worst case, $b$ is bounded by the maximum
-depth of the tree multiplied by the number of points, which is trivially $O(N
-\log (\eta_q / \eta_r)$.  It is likely that a tighter bound of $O(N \log N)$
-could be established.
+When we consider the monochromatic case (where $S_q = S_r$), the results
+trivially simplify.
 
-There are two corner cases for the non-monochromatic case: all extra reference
-recursions are of type \textit{(1)}, and all extra reference recursions are of
-type \textit{(3)}.  The first case presents the best situation, because all of
-the reference recursions are done before the first query node is recursed in.
+\begin{cor}
+\label{cor:ct-runtime-mono}
+Given the situation of Theorem \ref{thm:ct-runtime} but with $S_q = S_r = S$ so
+that $c_q = c_r = c$ and $\mathscr{T}_q = \mathscr{T}_r = \mathscr{T}$, a
+standard cover tree based dual-tree algorithm (Algorithm
+\ref{alg:cover-tree-dual}) takes
+
+\begin{equation}
+O\left(c^4 |R^*| \chi \psi (N + i(\mathscr{T}))\right)
+\end{equation}
+
+\noindent time, where $ | R^* | $ is the maximum size of the reference set $R$
+(line \ref{alg:line:ct-dual-input}) during the dual-tree recursion, $\chi$ is
+the maximum possible runtime of \texttt{BaseCase()}, and $\psi$ is the maximum
+possible runtime of \texttt{Score()}.
+\end{cor}
+
+An intuitive understanding of these bounds is best achieved by first considering
+the monochromatic case (this case arises, for instance, in all-nearest-neighbor
+search).  The linear dependence on $N$ arises from the fact that all query nodes
+must be visited.  The dependence on the reference tree, however, is encapsulated
+by the term $c^4 |R^*|$, with $|R^*|$ being the maximum size of the reference
+set $R$; this value must be derived for each specific problem.  The bad
+performance of poorly-behaved datasets with large $c$ (or, in the worst case, $c
+\sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
+also have a high cover tree imbalance $i(\mathscr{T})$; the linear dependence of
+runtime on imbalance is thus sensible.
+
+The bichromatic case ($S_q \ne S_r$) is a slightly more complex result which
+deserves a bit more attention.  The intuition for all terms except $\theta$
+remain virtually the same, although it is worth pointing out that there is
+no dependence on the reference set size.
+
+The term $\theta$ captures the effect of query and reference datasets with
+different widths, and has two corner cases: all extra reference recursions are of
+type \textit{(1)}, and all extra reference recursions are of type \textit{(3)}.
+The first case presents the best situation, because all of the reference
+recursions are done before the first query node is recursed into\footnote{In
+fact, re-derivation for this specific case can remove the dependence on
+$i(\mathscr{T}_q)$ entirely.  But our bound, as derived, does not bound this
+behavior tightly.  To do so would introduce a number of complex constants, and
+thus in that respect $i(\mathscr{T}_q)$ is a more suitable bounding quantity}.
 Therefore the bulk of the reference recursion work is done by with the query
@@
 recursion for each query point plus the added overhead of the recursion of the
-query tree with one reference node.  This second case degenerates to something
-like $O(N \log N)$ or $O(N^2)$ time depending on how you look at it.
+query tree with one reference node.
 
-This result holds for any dual-tree algorithm regardless of the problem. Hence,
+{\bf TODO: what does the second case degenerate to in terms of running time?}
+
+The quantity $|R^*|$ bounds the amount of work that needs to be done for each
+recursion. In the worst case, $|R^*|$ can be $N$. However,
+dual-tree algorithms rely on branch-and-bound techniques to prune away
+work (Lines \ref{alg:line:ct-dual-ref-score} and
+\ref{alg:line:ct-dual-query-pruning} in Algorithm \ref{alg:cover-tree-dual}). A
+small value of $|R^*|$ will imply that the algorithm is extremely successful in
+pruning away work at Line \ref{alg:line:ct-dual-ref-score} in Algorithm
+\ref{alg:cover-tree-dual}. An (upper) bound on $|R^*|$ (and the algorithm's
+success in pruning work) will depend on the problem and the data.  As we will
+show, bounding $|R^*|$ is often possible. % In addition, we will also show that
+For many dual-tree algorithms, $\chi \sim \psi \sim O(1)$; often, cached
+sufficient statistics \citep{moore2000anchors} can enable $O(1)$ runtime
+implementations of \texttt{BaseCase()} and \texttt{Score()}.
+
+These results hold for any dual-tree algorithm regardless of the problem. Hence,
 %for any pairwise statistical problem, the runtime of the dual-tree algorithm (of
@@
 answering $O(N)$ queries would take at least $O(N)$ time. For a particular
-problem and data, if $c_r$, $|R^*|$, $\nu$, $\chi$ and
-$\psi$ are bounded by constants independent of $N$ (for large enough $N$), then
-the dual-tree algorithm for that problem has a runtime linear in $N$. Our
-theoretical result separates out the problem-dependent and the
+problem and data, if $c_r$, $|R^*|$, $\chi$, $\psi$ are bounded by constants
+independent of $N$ and $\theta$ is no more than linear in $N$ (for large enough
+$N$), then the dual-tree algorithm for that problem has a runtime linear in $N$.
+Our theoretical result separates out the problem-dependent and the
 problem-independent elements of the runtime bound, which allows us to simply
@@
 algorithm without requiring an analysis from scratch.
+
+Our results are similar to that of \citet{ram2009}, but those results depend on
+a quantity called the {\it constant of bichromaticity}, denoted $\kappa$, which
+has unclear relation to cover tree imbalance.  The dependence on $\kappa$ is
+given as $c_q^{4 \kappa}$, which is not a good bound, especially because
+$\kappa$ may easily be very large.  Further, it is incorrectly claimed that
+$\kappa = 1$ in the monochromatic case.
+
+The more recent results of \citet{curtin2014dual} are more related to these
+results, but they depend on the {\it inverse constant of bichromaticity} $\nu$
+which suffers from the same problem as $\kappa$.  Although the dependence on
+$\nu$ is linear (that is, $O(\nu N)$), bounding $\nu$ is difficult and it is not
+true that $\nu = 1$ in the monochromatic case.
+
+Both $\nu$ and $\kappa$ are difficult to empirically calculate and require an
+entire run of the dual-tree algorithm.  On the other hand, bounding
+$i(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree (assuming
+the tree is already built).  Thus, not only is our bound tighter, it more
+closely reflects the actual behavior of dual-tree algorithms, and the constants
+which it depends upon are straightforward to calculate.
+
+In the following sections, we will apply our results to specific problems and
+show the utility of our bound in simplifying runtime proofs for dual-tree
+algorithms.
+
 % This discussion already happened earlier... sort of.
@@
 %$\mathscr{N}_q$ is recursed.
-If $\nu \sim \kappa$, which is a reasonable assumption, this is a better result
-than the results of Ram et.~al.  \cite{ram2009} who found a dependence of
-$c_q^{4 \kappa}$ in their proofs.
-
-The quantity $|R^*|$ bounds the amount of work that needs to be done for each
-recursion. In the worst case, $|R^*|$ can be $N$. However,
-dual-tree algorithms rely on branch-and-bound techniques to prune away
-work (Lines \ref{alg:line:ct-dual-ref-score} and
-\ref{alg:line:ct-dual-query-pruning} in Algorithm \ref{alg:cover-tree-dual}). A
-small value of $|R^*|$ will imply that the algorithm is extremely successful in
-pruning away work at Line \ref{alg:line:ct-dual-ref-score} in Algorithm
-\ref{alg:cover-tree-dual}. An (upper) bound on $|R^*|$ (and the algorithm's
-success in pruning work) will depend on the problem and the data.  As we will
-show, bounding $|R^*|$ is often possible. % In addition, we will also show that
-For many dual-tree algorithms, $\chi \sim \psi \sim O(1)$; often, cached
-sufficient statistics~\cite{moore2000anchors} can enable $O(1)$ runtime
-implementations of \texttt{BaseCase()} and \texttt{Score()}.
 

### 2014-11-13 bounds-files/imbalance.tex
Refactor some of this, and change the definition of node imbalance to make the general runtime bound much simpler to write about.

@@
 \end{center}
+\label{fig:outlier}
 \caption{Single-outlier cover tree.}
-\label{fig:outlier}
 \end{figure}
@@
 $\mathscr{N}_i$ with scale $s_i$ in the cover tree $\mathscr{T}$ is defined as
-the cumulative number of missing levels between the node and its children.  If
-the node has a leaf child (which has level $-\infty$), then number of missing
-levels is defined as the difference between $s_i$ and $s_{\min} - 1$ where
-$s_{\min}$ is the smallest scale of a non-leaf node in $\mathscr{T}$.
+the cumulative number of missing levels between the node and its parent
+$\mathscr{N}_p$ (which has scale $s_p$).  If
+the node is a leaf child (that is, $s_i = -\infty$), then number of missing
+levels is defined as the difference between $s_p$ and $s_{\min} - 1$ where
+$s_{\min}$ is the smallest scale of a non-leaf node in $\mathscr{T}$.  If
+$\mathscr{N}_i$ is the root of the tree, then the cover node imbalance is 0.
 Explicitly written, this calculation is
@@
 \begin{equation}
-i(\mathscr{N}_i) = \left( \sum_{\mathscr{N}_c \in \mathscr{C}(\mathscr{N}_i),
-\mathscr{C}(\mathscr{N}_c) \ne \emptyset} s_i - s_c - 1 \right) + \left(
-\sum_{\mathscr{N}_c \in \mathscr{C}(\mathscr{N}_i), \mathscr{C}(\mathscr{N}_c) =
-\emptyset} \max(s_i - s_{\min} - 1, 0) \right).
+i(\mathscr{N}_i) = \begin{dcases*}
+  s_p - s_i - 1 & if $\mathscr{N}_i$ is not a leaf and not the root node \\
+  \max(s_p - s_{\min} - 1, \; 0) & if $\mathscr{N}_i$ is a leaf \\
+  0 & if $\mathscr{N}_i$ is the root node.
+  \end{dcases*}
 \end{equation}
@@
 
-Given this definition for a single node, then, we can generalize to the full
-tree.
+This simple definition of cover node imbalance is easy to calculate, and using
+it, we can generalize to a measure of imbalance for the full tree.
 
@@
 A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
-imbalance $i(\mathscr{T}_b) = 0$.  A worst-case cover tree $\mathscr{T}_w$ which
-is entirely a chain-like structure with maximum scale $s_{\max}$ and minimum
-scale $s_{\min}$ will have imbalance $i(\mathscr{T}_w) = N (s_{\max} -
-s_{\min})$.  Because of this chain-like structure, each level has only one node
-and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
-that the imbalance is actually quadratic in $N$!
+imbalance $i(\mathscr{T}_b) = 0$ (for instance, Figure
+\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
+entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
+$s_{\min}$ will have imbalance $i(\mathscr{T}_w) \sim N (s_{\max} - s_{\min})$.
+Because of this chain-like structure, each level has only one node and thus
+there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning that the
+imbalance is actually quadratic in $N$!
 

### 2014-11-14 bounds-files/general-runtime-bound.tex
Rename imbalance to i_t.

@@
 \begin{equation}
-\theta = \max(\log_2(\eta_r / \eta_q), 0) + i(\mathscr{T}_q) + N
+\theta = \max(\log_2(\eta_r / \eta_q), 0) + i_t(\mathscr{T}_q) + N
 \max(\log_2(\delta_q / \delta_r), 0)
@@
 exact number of reference recursions before the children of $\mathscr{N}_q$ are
-recursed into is bounded above by $i(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has
+recursed into is bounded above by $i_t(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has
 imbalance $0$, then it is exactly one level below its parent, and thus there is
@@
 levels below its parent, then it is possible that a reference recursion may
-occur for each level in between; this is a maximum of $i(\mathscr{N}_q) + 1$.
+occur for each level in between; this is a maximum of $i_t(\mathscr{N}_q) + 1$.
 
@@
 \begin{equation}
-\sum_{\mathscr{N}_q \in \mathscr{T}_q} i(\mathscr{N}_q) + 1 = b(\mathscr{T}_q) +
+\sum_{\mathscr{N}_q \in \mathscr{T}_q} i_t(\mathscr{N}_q) + 1 = b(\mathscr{T}_q) +
 O(N)
@@
 \begin{equation}
-O\left(c^4 |R^*| \chi \psi (N + i(\mathscr{T}))\right)
+O\left(c^4 |R^*| \chi \psi (N + i_t(\mathscr{T}))\right)
 \end{equation}
@@
 \sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
-also have a high cover tree imbalance $i(\mathscr{T})$; the linear dependence of
+also have a high cover tree imbalance $i_t(\mathscr{T})$; the linear dependence of
 runtime on imbalance is thus sensible.
@@
 fact, re-derivation for this specific case can remove the dependence on
-$i(\mathscr{T}_q)$ entirely.  But our bound, as derived, does not bound this
+$i_t(\mathscr{T}_q)$ entirely.  But our bound, as derived, does not bound this
 behavior tightly.  To do so would introduce a number of complex constants, and
-thus in that respect $i(\mathscr{T}_q)$ is a more suitable bounding quantity}.
+thus in that respect $i_t(\mathscr{T}_q)$ is a more suitable bounding quantity}.
 Therefore the bulk of the reference recursion work is done by with the query
@@
 entire run of the dual-tree algorithm.  On the other hand, bounding
-$i(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree (assuming
-the tree is already built).  Thus, not only is our bound tighter, it more
-closely reflects the actual behavior of dual-tree algorithms, and the constants
-which it depends upon are straightforward to calculate.
+$i_t(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree
+(assuming the tree is already built).  Thus, not only is our bound tighter, it
+more closely reflects the actual behavior of dual-tree algorithms, and the
+constants which it depends upon are straightforward to calculate.
 

### 2014-11-14 bounds-files/imbalance.tex
Rename imbalance to i_t.

@@
 \begin{defn}
-The {\it cover node imbalance} $i(\mathscr{N}_i)$ for a cover tree node
+The {\it cover node imbalance} $i_n(\mathscr{N}_i)$ for a cover tree node
 $\mathscr{N}_i$ with scale $s_i$ in the cover tree $\mathscr{T}$ is defined as
@@
 \begin{equation}
-i(\mathscr{N}_i) = \begin{dcases*}
+i_n(\mathscr{N}_i) = \begin{dcases*}
   s_p - s_i - 1 & if $\mathscr{N}_i$ is not a leaf and not the root node \\
@@
 \begin{defn}
-The {\it cover tree imbalance} $b(\mathscr{T})$ for a cover tree $\mathscr{T}$
+The {\it cover tree imbalance} $i_t(\mathscr{T})$ for a cover tree $\mathscr{T}$
 is defined as the cumulative number of missing levels in the tree.  This can be
@@
 \begin{equation}
-b(\mathscr{T}) = \sum_{\mathscr{N}_i \in \mathscr{T}} i(\mathscr{N}_i).
+i_t(\mathscr{T}) = \sum_{\mathscr{N}_i \in \mathscr{T}} i_n(\mathscr{N}_i).
 \end{equation}
@@
 A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
-imbalance $i(\mathscr{T}_b) = 0$ (for instance, Figure
+imbalance $i_t(\mathscr{T}_b) = 0$ (for instance, Figure
 \ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
 entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
-$s_{\min}$ will have imbalance $i(\mathscr{T}_w) \sim N (s_{\max} - s_{\min})$.
-Because of this chain-like structure, each level has only one node and thus
-there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning that the
-imbalance is actually quadratic in $N$!
+$s_{\min}$ will have imbalance $i_t(\mathscr{T}_w) \sim N (s_{\max} -
+s_{\min})$.  Because of this chain-like structure, each level has only one node
+and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
+that the imbalance is actually quadratic in $N$!
 
@@
 
-{\bf TODO: Can any bound be proven on $i(\mathscr{T})$?  I haven't thought of
+{\bf TODO: Can any bound be proven on $i_t(\mathscr{T})$?  I haven't thought of
 anything successfully yet.}

### 2014-11-14 bounds-files/sec02-nearest-neighbor-search.tex
Remove negative vspace and remove NIPS formatting.
Start refactoring NNS section.

@@
-Consider nearest neighbor search, given as a \texttt{BaseCase()} function
-(Algorithm \ref{alg:nn_base_case}) and a \texttt{Score()} function (Algorithm
-\ref{alg:nn_score}).  The goal is, for every query point $p_q$, find the nearest
-neighbor $p_r$ in the reference set $S_r$.  This algorithm stores the current
-nearest neighbor candidates in the array $N$ and their distances in the array
-$D$.\footnote{See Curtin et.~al. \cite{curtin2013tree} for a more complete
-discussion of how this algorithm works and a proof of correctness.}  The
-\texttt{Score()} function depends on a bound function $B(\mathscr{N}_q)$ which
-represents the maximum distance that could possibly improve a nearest neighbor
-candidate for any descendant point of the query node $\mathscr{N}_q$.  The
-standard bound function $B(\mathscr{N}_q)$ used for cover trees is
-%\begin{equation}
-$B(\mathscr{N}_q) \le D[p_q] + 2^{s_q + 1}$ (this is adapted from
-Beygelzimer et.~al. \cite{langford2006}).
-%\end{equation}
-
-%\begin{figure}
-
 %\begin{minipage}{0.4\textwidth}
@@
 %\end{figure}
+
+The standard task of nearest neighbor search can be simply described: given a
+query set $S_q$ and a reference set $S_r$, for each query point $p_q \in S_q$,
+find the nearest neighbor $p_r$ in the reference set $S_r$.  The task is
+well-studied and well-known, and there exist numerous approaches for both exact
+and approximate nearest neighbor search, including the cover tree nearest
+neighbor search algorithm due to \citet{langford2006}.  We will consider that
+algorithm, but in a tree-independent sense as given by \citet{curtin2013tree};
+this means that to describe the algorithm, we require only a \texttt{BaseCase()}
+and \texttt{Score()} function; these are given in Algorithms
+\ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively.
+
+Given a type of tree and traversal, these two functions store the current
+nearest neighbor candidates in the array $N$ and their distances in the array
+$D$. \citep[see][for a more complete
+discussion of how this algorithm works and a proof of
+correctness.]{curtin2013tree}  The
+\texttt{Score()} function depends on a bound function $B(\mathscr{N}_q)$ which
+represents the maximum distance that could possibly improve a nearest neighbor
+candidate for any descendant point of the query node $\mathscr{N}_q$.  The
+standard bound function $B(\mathscr{N}_q)$ used for cover trees is
+%\begin{equation}
+$B(\mathscr{N}_q) \le D[p_q] + 2^{s_q + 1}$ \citep[this is adapted from][]{langford2006}.
+%\end{equation}
+
+%\begin{figure}
 
@@
 
-\vspace*{-1em}
 \begin{eqnarray}
@@
 \end{eqnarray}
-\vspace*{-0.9em}
 
@@
 \end{equation}
-\vspace*{-0.8em}
 
@@
 %\begin{eqnarray}
-\vspace*{-1.0em}
 \begin{equation}
@@
 
-\vspace*{-1.2em}
 \begin{eqnarray}
@@
 \end{eqnarray}
-\vspace*{-0.8em}
 
@@
 \end{eqnarray}
-\vspace*{-0.6em}
 
@@
 
-%\vspace*{-0.5em}
 %\begin{eqnarray}

### 2014-11-14 bounds-files/sec02-nearest-neighbor-search.tex
Refactor NNS section a bit.

@@
 given a reference set $S_r$ with expansion constant $c_r$, a query set $S_q$,
-and an inverse degree of bichromaticity $\nu$, the running time of the
-algorithm is bounded by $O(c_r^4 c_{qr}^5 \nu N)$.
+the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N + \theta))$
+with $\theta$ defined as in Theorem \ref{thm:ct-runtime}.
 \label{thm:nns}
@@
 Due to Theorem \ref{thm:ct-runtime}, we therefore know that the runtime of the
-algorithm is bounded by $O(c_r^4 \nu |R^*| N)$.  Thus, the only thing that
-remains is to bound the maximum size of the reference set, $|R^*|$.
+algorithm is bounded by $O(c_r^4 |R^*| (N + \theta))$.  Thus, the only thing
+that remains is to bound the maximum size of the reference set, $|R^*|$.
 
@@
 
-In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 \nu N)$ because
-$c = c_r = c_{qr}$.
+In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 (N +
+i_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = i_t(\mathscr{T})$.

### 2014-11-14 bounds-files/sec03-approx-kde.tex
Remove negative vspace and remove NIPS formatting.
Refactor KDE a bit; we can write about both approximation schemes.

@@
-Ram et.~al. present a clever technique for bounding the
+\citet{ram2009} present a clever technique for bounding the
 running time of approximate kernel density estimation based on the properties of
-the kernel, when the kernel is shift-invariant and satisfies a few assumptions
-\cite{ram2009}.  We will restate these assumptions and provide an adapted proof
-using Theorem \ref{thm:ct-runtime}, which gives a tighter bound.
+the kernel, when the kernel is shift-invariant and satisfies a few assumptions.
+We will restate these assumptions and provide an adapted proof using Theorem
+\ref{thm:ct-runtime}, which gives a tighter bound.
 
 Approximate kernel density estimation is a common application of dual-tree
-algorithms \cite{gray2003nonparametric, nbody}.  Given a query set $S_q$, a
+algorithms \citep{gray2003nonparametric, nbody}.  Given a query set $S_q$, a
 reference set $S_r$ of size $N$, and a kernel function $\mathcal{K}(\cdot,
-\cdot)$, the goal is to calculate, for each $p_q \in S_q$, a density estimate
-$f(p_q)$ such that $| f(p_q) - f^*(p_q) | < N \epsilon$ for some
-parameter $\epsilon$ and
+\cdot)$, the true kernel density estimate for a query point $p_q$ is given as
 
-%\vspace*{-0.5em}
 \begin{equation}
@@
 \end{equation}
-%\vspace*{-0.5em}
 
-This is related to the well-studied problem of kernel summation, which can
-also be solved with dual-tree algorithms \cite{lee2006faster, lee2008fast}.  In
-both of those problems, simple geometric observations can be made to accelerate
-computation: when $\mathcal{K}(\cdot, \cdot)$ is shift-invariant, faraway points
-have very small kernel evaluations.  Thus, trees can be built on $S_q$ and
-$S_r$, and node combinations can be pruned when the nodes are far apart while
-still obeying the error bound $N \epsilon$.\footnote{Here we consider
-only absolute-value approximation.  Our methods can be easily adapted to other
-schemes, such as relative-value approximation.  \cite{ram2009} considers
-relative-value approximation in far more detail.}
+There are two simple ways to approximate $f^*(p_q)$: {\it absolute value
+approximation} and {\it relative value approximation}.  Absolute value
+approximation requires that each density estimate $f(p_q)$ is within $\epsilon$
+of the true estimate $f^*(p_q)$:
+
+\begin{equation}
+| f(p_q) - f^*(p_q) | < N \epsilon.
+\end{equation}
+
+Relative value approximation is a more flexible approximation scheme; given some
+parameter $\epsilon$, the requirement is that each density estimate is within a
+relative tolerance of $f^*(p_q):$
+
+\begin{equation}
+\frac{| f(p_q) - f^*(p_q) |}{| f^*(p_q) |} < N \epsilon.
+\end{equation}
+
+Kernel density estimation is related to the well-studied problem of kernel
+summation, which can also be solved with dual-tree algorithms
+\citep{lee2006faster, lee2008fast}.  In both of those problems, regardless of
+the approximation scheme, simple geometric observations can be made to
+accelerate computation: when $\mathcal{K}(\cdot, \cdot)$ is shift-invariant,
+faraway points have very small kernel evaluations.  Thus, trees can be built on
+$S_q$ and $S_r$, and node combinations can be pruned when the nodes are far
+apart while still obeying the error bounds.
+
+In the following two subsections, we will separately consider both the absolute
+value approximation scheme and the relative value approximation scheme.
+
+\subsection{Absolute value approximation}
 
@@
 
-\vspace*{-1.5em}
 \begin{equation}
@@
 
-%\vspace*{-1.0em}
 %\begin{eqnarray}
@@
 %\end{eqnarray}
-%\vspace*{-0.5em}
 
@@
 
-%\vspace*{-0.8em}
 %\begin{equation}
@@
 %\end{equation}
-%\vspace*{-0.8em}
 %
@@
 %
-%\vspace*{-1em}
 %\begin{eqnarray}
@@
 %\end{eqnarray}
-%\vspace*{-1.5em}
 

### 2014-11-18 bounds-files/sec03-approx-kde.tex
Some further refactoring.

@@
 A tree-independent algorithm for solving approximate kernel density estimation
+with absolute value approximation
 is given as a \texttt{BaseCase()} function in Algorithm \ref{alg:kde_base_case}
-and a \texttt{Score()} function in Algorithm \ref{alg:kde_score} (a correctness
-proof can be found in \cite{curtin2013tree}).  This function
+and a \texttt{Score()} function in Algorithm \ref{alg:kde_score}  \citep[a
+correctness proof can be found in][]{curtin2013tree}.  This function
 assumes a shift-invariant kernel $\mathcal{K}(p_i, p_j) = \mathcal{K}(d( p_i,
@@
 
-Note that this version is far simpler than the algorithm proposed in,
-e.g., \cite{gray2003nonparametric}; however, it is sufficient for our runtime
-analysis.  Real-world implementations, such as the one found in \textbf{mlpack}
-\cite{mlpack2013}, tend to be far more complex.
+Note that this version is far simpler than other dual-tree algorithms that have
+been proposed for approximate kernel density estimation \citep[see, for
+instance][]{gray2003nonparametric}; however, this version is sufficient for our
+runtime analysis.  Real-world implementations, such as the one found in
+\textbf{mlpack} \citep{mlpack2013}, tend to be far more complex.
 
@@
 
-  \caption{Approximate kernel density estimation \texttt{Score()}}
+  \caption{Absolute-value approximate kernel density estimation
+\texttt{Score()}}
   \label{alg:kde_score}
@@
 
+\begin{algorithm}[tb]
+  \begin{algorithmic}[1]
+    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, reference node
+$\mathscr{N}_r$, list of node kernel estimates $\hat{f}_n$
+    \STATE {\bfseries Output:} a score for the node combination $(\mathscr{N}_q,
+\mathscr{N}_r)$, or $\infty$ if the combination should be pruned
+
+    \medskip
+
+    \IF{$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
+\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) /
+\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) < \epsilon$}
+      \STATE $f_n(\mathscr{N}_q) \gets f_n(\mathscr{N}_q) + |
+\mathscr{D}^p(\mathscr{N}_r) | \left(\mathcal{K}(d_{\min}(\mathscr{N}_q,
+\mathscr{N}_r)) + \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))\right) /\;2$
+      \RETURN $\infty$
+    \ENDIF
+
+    \RETURN $\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
+\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))$
+  \end{algorithmic}
+
+  \caption{Relative-value approximate kernel density estimation
+\texttt{Score()}}
+  \label{alg:kde_rva_score}
+\end{algorithm}
+
 \begin{thm}
-Assume that $\mathcal{K}(p_i, p_j) = \mathcal{K}\left(d(p_i, p_j)\right)$ is a shift-invariant kernel with some bandwidth $h > 0$, where $\mathcal{K}(x)$ is smooth, non-negative, monotonically decreasing, and concave for $x \in [0, h]$ and convex for $x \in (h, \infty)$.
-Then, given a query set $S_q$ and a reference set $S_r$ with
-expansion constant $c_r$, an inverse degree of bichromaticity $\nu$ and using the
+Assume that $\mathcal{K}(p_i, p_j) = \mathcal{K}\left(d(p_i, p_j)\right)$ is a
+shift-invariant kernel with some bandwidth $h > 0$, where $\mathcal{K}(x)$ is
+smooth, non-negative, monotonically decreasing, and concave for $x \in [0, h]$
+and convex for $x \in (h, \infty)$. Then, given a query set $S_q$ and a
+reference set $S_r$ with
+expansion constant $c_r$, and using the
 approximate kernel density estimation \texttt{BaseCase()} and \texttt{Score()}
@@
 %\begin{equation}
-$O(c_r^{8 + \log_2 \zeta} \nu N)$
+$O(c_r^{8 + \log_2 \zeta} (N + \theta))$
 %\end{equation}
-where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$.
+where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$ and
+$\theta$ as in Theorem \ref{thm:ct-runtime}.
 
@@
 so Theorem \ref{thm:ct-runtime} implies the total runtime of the dual-tree
-algorithm is bounded by $O(c_r^4 \nu |R^*| N)$.  Thus, we will bound $|R^*|$
-using techniques related to those used in \cite{ram2009}.  The bounding of
-$|R^*|$ is split into two sections: first, show that when the scale $s_r^{\max}$
-is small enough, $R^*$ is empty.  Second, bound $R^*$ when $s_r^{\max}$ is larger.
+algorithm is bounded by $O(c_r^4 |R^*| (N + \theta))$.  Thus, we will bound
+$|R^*|$ using techniques related to those used by \citet{ram2009}.  The bounding
+of $|R^*|$ is split into two sections: first, show that when the scale
+$s_r^{\max}$ is small enough, $R^*$ is empty.  Second, bound $R^*$ when
+$s_r^{\max}$ is larger.
 
@@
 bandwidth $\sigma$.  For a reasonably chosen $\epsilon = 0.05$, the runtime is
-approximately $O(c_r^{8.89} \nu N)$; for $\epsilon = 0.01$, the runtime is
-approximately $O(c_r^{11.52} \nu N)$.  For very small $\epsilon = 0.00001$, the
-runtime is approximately $O(c_r^{22.15} \nu N)$.
+approximately $O(c_r^{8.89} (N + \theta))$; for $\epsilon = 0.01$, the runtime
+is approximately $O(c_r^{11.52} (N + \theta))$.  For very small $\epsilon =
+0.00001$, the runtime is approximately $O(c_r^{22.15} (N + \theta))$.
 
@@
 %range search and range count.
+
+\subsection{Relative Value Approximation}
+
+Approximate kernel density estimation using relative-value approximation may be
+bounded in a similar manner.  A \texttt{Score()} function is given in Algorithm
+\ref{alg:kde_rva_score}.

### 2014-11-18 bounds-files/sec04-range-search.tex
Remove negative vspace and remove NIPS formatting.
Some further refactoring.

@@
 \texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
-\ref{alg:rs_sc}, respectively (a correctness proof can be found in
-\cite{curtin2013tree}).  The sets $N[p_q]$ (for each $p_q$) are
+\ref{alg:rs_sc}, respectively \citep[a correctness proof can be found
+in][]{curtin2013tree}.  The sets $N[p_q]$ (for each $p_q$) are
 initialized to $\emptyset$ at the beginning of the traversal.
@@
 
-\vspace*{-1em}
 \begin{equation}
@@
 \end{equation}
-\vspace*{-1.5em}
 
 \noindent the running time of range search or range count is bounded by
-$O(c_r^{8 + \beta} \nu N)$ with
+$O(c_r^{8 + \beta} (N + \theta))$ with $\theta$ defined as in Theorem
+\ref{thm:ct-runtime} and
 %\begin{equation}
@@
 
-\vspace*{-2.5em}
 \begin{proof}
@@
 \ref{thm:ct-runtime} we know that the runtime of the algorithm is bounded by
-$O(c_r^4 \nu |R^*| N)$. %  Thus, our only task is to bound the maximum size of the
+$O(c_r^4 |R^*| (N + \theta))$. %  Thus, our only task is to bound the maximum size of the
 %reference set, $|R^*|$.
@@
 \end{equation}
-\vspace*{-1em}
 
@@
 
-%\vspace*{-1em}
 %\begin{eqnarray}
@@
 %\end{eqnarray}
-%\vspace*{-1em}
 
@@
 
-%\vspace*{-1em}
 %\begin{equation}
@@
 %\end{equation}
-%\vspace*{-1em}
 
@@
 \end{proof}
-\vspace*{-0.4em}
 
@@
 To understand the relation of the running time to the bounding quantity
-$\alpha$, consider (for the monochromatic case with $\nu \sim O(1)$) $\alpha =
-(1 / 3)$; this yields a running time of $O(c^8 N)$.  $\alpha = (1 / 7)$ yields
-$O(c^9 N)$, $\alpha = (1 / 15)$ yields $O(c^{10} N)$, and so forth.  As $\alpha$
-gets smaller, the exponent on $c$ gets larger, and diverges as $\alpha \to 0$.
+$\alpha$, consider $\alpha =
+(1 / 3)$; this yields a running time of $O(c^8 (N + \theta))$.  $\alpha = (1 /
+7)$ yields $O(c^9 (N + \theta))$, $\alpha = (1 / 15)$ yields $O(c^{10} (N +
+\theta))$, and so forth.  As $\alpha$ gets smaller, the exponent on $c$ gets
+larger, and diverges as $\alpha \to 0$.
 
@@
 some dependence on $N$, then the running time can be easily bounded as $O(c^{4 +
-\beta} kN)$, in the monochromatic case with $\nu \sim O(1)$.
+\beta} k(N + \theta))$.
 

### 2014-11-18 bounds.tex
mathtools for equation alignment.
Some further refactoring.

@@
 \usepackage{color}
+\usepackage{mathtools}
 
@@
 state-of-the-art, and in the case of range search, is the first such bound.  We
-then investigate the empirical performance of cover trees for these algorithms,
-with a focus on understanding the dataset-dependent constants that our bounds
-use.
+then investigate the empirical performance of cover trees for these algorithms
+in Section \ref{sec:emp}, with a focus on understanding the dataset-dependent
+constants that our bounds use.
 
@@
 
-\section{Tree Imbalance}
+\section{Tree imbalance}
 \label{sec:imbalance}
@@
 
-\section{Range Search and Range Count}
+\section{Range search and range count}
 \label{sec:rs}
@@
 %\input{bounds-files/sec05-range-search-aspect-ratio.tex}
+
+\section{Empirical investigations}
+\label{sec:emp}
+
+\input{bounds-files/empirical.tex}
 

### 2014-11-18 bounds-files/empirical.tex
New file for empirical results.

In the previous section we have expressed a number of bounds of roughly the form

\begin{equation}
O(c^a (N + \theta))
\end{equation}

\noindent for some $a$.  Although both $c$ and $\theta$ are well-defined and
straightforward, there is limited practical utility to these bounds without
empirical experimentation.  One can easily say `these algorithms scale linearly
if $c$ has no dependence on $N$ and $\theta$ is not superlinear in $N$'; this is
useful.  But this does not answer the question of `on what datasets are these
assumptions valid, and how do these algorithms actually perform in practice?'

To help answer this oft-overlooked query, we perform an empirical analysis of
$c$, $i_t(\mathscr{T}_q)$, and $\theta$ on a variety of different datasets for
each algorithm we have presented.

\subsection{The expansion constant $c$}

\subsection{Tree imbalance $i_t(\mathscr{T})$}

\subsection{Nearest neighbor search}

\subsection{Kernel density estimation}

\subsection{Range search}

### 2014-11-19 bounds-files/empirical.tex
Add a table with information I am slowly gathering.

@@
 
+\begin{table}
+
+\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+\hline
+{\bf Dataset & $n$ & $d$ & $c$ & $i_t$ & NNS & KDE & RS} \\
+\hline
+cloud & 2048 & 10 & 178.5 &
+winequality & 6497 & 11 & 4133.0 &
+birch3 & 100000 & 2 & 1866 &
+corel & 37749 & 32 & 23952 &
+covertype & 581012 & 54 &
+isolet & 7797 & 617 & 7797 &
+miniboone & 130064 & 50 & 130064 &
+phy & 150000 & 78 & 27731 &
+bio &
+lcdm & 1808174 & 3 &
+yp-msd & 515345 & 90 & 515243 &
+mnist & 70000 & 784 & 70000 &
+randu & 1000000 & 10 &
+power & 2075260 & 7 &
+\end{tabular}
+\end{table}
+
 \subsection{Tree imbalance $i_t(\mathscr{T})$}

### 2014-11-19 bounds-files/sec03-approx-kde.tex
Fixes pointed out by Bill.

@@
 
-\begin{equation}
+\begin{eqnarray}
 d_{\max}(\mathscr{N}_q, \mathscr{N}_r) - d_{\min}(\mathscr{N}_q, \mathscr{N}_r)
-= d(p_q, p_r) + 2^{s_r^{\max} + 1} - d(p_q, p_r) + 2^{s_r^{\max} + 1} = 2^{s_r^{\max} + 2}.
-\end{equation}
+&\le& d(p_q, p_r) + 2^{s_r^{\max} + 1} - d(p_q, p_r) + 2^{s_r^{\max} + 1} \\
+ &\le& 2^{s_r^{\max} + 2}.
+\end{eqnarray}
 
@@
 \mathcal{K}'(h) ) - 2$.  It is trivial to see that for any $\mathscr{N}_r \in
-R^*$, $\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) > \epsilon$;
+R^*$, $\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) > \epsilon$;
 alternately stated, this means that

### 2014-11-20 bounds-files/empirical.tex
Changes for today.

--- 2014-11-18 bounds-files/empirical.tex
@@
 
+\begin{table}
+
+\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+\hline
+{\bf Dataset} & {\bf $n$} & {\bf $d$} & {\bf $c$} & {\bf $i_t$} & {\bf NNS} & {\bf KDE} & {\bf RS} \\
+\hline
+cloud & 2048 & 10 & 178.5 & \\
+winequality & 6497 & 11 & 4133.0 & \\
+birch3 & 100000 & 2 & 1866 & \\
+corel & 37749 & 32 & 23952 & \\
+covertype & 581012 & 54 & \\
+isolet & 7797 & 617 & 7797 & \\
+miniboone & 130064 & 50 & 130064 & \\
+phy & 150000 & 78 & 27731 & \\
+bio & \\
+lcdm & 1808174 & 3 & \\
+yp-msd & 515345 & 90 & 515243 & \\
+mnist & 70000 & 784 & 70000 & \\
+randu & 1000000 & 10 & \\
+power & 2075260 & 7 & \\
+\end{tabular}
+\end{table}
+
 \subsection{Tree imbalance $i_t(\mathscr{T})$}

### 2014-11-20 bounds-files/sec03-approx-kde.tex
Remove legacy \nu.
Changes for today.

@@
 
-There are two simple ways to approximate $f^*(p_q)$: {\it absolute value
-approximation} and {\it relative value approximation}.  Absolute value
-approximation requires that each density estimate $f(p_q)$ is within $\epsilon$
-of the true estimate $f^*(p_q)$:
+In the case of an infinite-tailed kernel $\mathcal{K}(\cdot, \cdot)$, the
+computation cannot be accelerated; thus, attention has turned towards tractable
+approximation schemes.  Two simple schemes for the approximation of $f^*(p_q)$
+are well-known: {\it absolute value approximation} and {\it relative value
+approximation}.  Absolute value approximation requires that each density
+estimate $f(p_q)$ is within $\epsilon$ of the true estimate $f^*(p_q)$:
 
@@
 In the following two subsections, we will separately consider both the absolute
-value approximation scheme and the relative value approximation scheme.
+value approximation scheme and the relative value approximation scheme, under
+the assumption of a shift-invariant kernel $\mathcal{K}(p_q, p_r) =
+\mathcal{K}(\| p_q - p_r \|)$ which is monotonically decreasing and
+non-negative.  In addition, we assume that there exists some bandwidth $h$ such
+that $\mathcal{K}(d)$ must be concave for $d \in [0, h]$ and convex for $d \in
+[h, \infty)$.  This assumption implies that the magnitude of the derivative
+$|\mathcal{K}'(d)|$ is maximized at $d = h$.  These are not restrictive
+assumptions; most standard kernels fall into this class, including the Gaussian,
+Laplacian and Epanechnikov kernels.
 
@@
 A tree-independent algorithm for solving approximate kernel density estimation
-with absolute value approximation
+with absolute value approximation under the previous assumptions on th kernel
 is given as a \texttt{BaseCase()} function in Algorithm \ref{alg:kde_base_case}
 and a \texttt{Score()} function in Algorithm \ref{alg:kde_score}  \citep[a
-correctness proof can be found in][]{curtin2013tree}.  This function
-assumes a shift-invariant kernel $\mathcal{K}(p_i, p_j) = \mathcal{K}(d( p_i,
-p_j ))$.  The list $f_p$ holds partial kernel density estimates for
-each query point, and the list $f_n$ holds partial kernel density
-estimates for each query node.  At the beginning of the dual-tree traversal, the
-lists $f_p$ and $f_n$, which are both of size $O(N)$, are each
-initialized to 0.
+correctness proof can be found in][]{curtin2013tree}.  The list $f_p$ holds
+partial kernel density estimates for each query point, and the list $f_n$ holds
+partial kernel density estimates for each query node.  At the beginning of the
+dual-tree traversal, the lists $f_p$ and $f_n$, which are both of size $O(N)$,
+are each initialized to 0.  As the traversal proceeds, node combinations are
+pruned if the difference between the maximum kernel value
+$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r))$ and the minimum kernel
+value $\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))$ is sufficiently
+small (line \ref{alg:ava-kde-prune}).  If the node combination can be pruned,
+then the partial node estimate is updated (line \ref{alg:ava-kde-update}).  When
+node combinations cannot be pruned, \texttt{BaseCase()} may be called, which
+simply updates the partial point estimate with the exact kernel evaluation (line
+\ref{alg:kde-bc-update}).
 
@@
     \STATE $f_p(p_q) \gets f_p(p_q) + \mathcal{K}(p_q, p_r)$
+\ref{alg:kde-bc-update}
     \RETURN $\mathcal{K}(p_q, p_r)$
@@
 \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) < \epsilon$}
+\label{alg:ava-kde-prune}
       \STATE $f_n(\mathscr{N}_q) \gets f_n(\mathscr{N}_q) + |
 \mathscr{D}^p(\mathscr{N}_r) | \left(\mathcal{K}(d_{\min}(\mathscr{N}_q,
-\mathscr{N}_r)) + \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))\right) /\;2$
+\mathscr{N}_r)) + \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))\right)
+/\;2$ \label{alg:ava-kde-update}
       \RETURN $\infty$
@@
 \begin{thm}
-Assume that $\mathcal{K}(p_i, p_j) = \mathcal{K}\left(d(p_i, p_j)\right)$ is a
-shift-invariant kernel with some bandwidth $h > 0$, where $\mathcal{K}(x)$ is
-smooth, non-negative, monotonically decreasing, and concave for $x \in [0, h]$
-and convex for $x \in (h, \infty)$. Then, given a query set $S_q$ and a
-reference set $S_r$ with
-expansion constant $c_r$, and using the
-approximate kernel density estimation \texttt{BaseCase()} and \texttt{Score()}
-as given in Algorithms \ref{alg:kde_base_case} and \ref{alg:kde_score},
-respectively, with the traversal given in Algorithm \ref{alg:cover-tree-dual},
-the running time of approximate kernel density estimation for some $\epsilon$
-error parameter is bounded by
+Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
+of the previous section.  Then, given a query set $S_q$ and a reference set
+$S_r$ with expansion constant $c_r$, and using the approximate kernel density
+estimation \texttt{BaseCase()} and \texttt{Score()} as given in Algorithms
+\ref{alg:kde_base_case} and \ref{alg:kde_score}, respectively, with the
+traversal given in Algorithm \ref{alg:cover-tree-dual}, the running time of
+approximate kernel density estimation for some $\epsilon$ error parameter is
+bounded by
 %\begin{equation}
@@
 Thus, we are interested in the maximum possible value $\mathcal{K}(a) -
-\mathcal{K}(b)$ for a fixed value of $b - a > 0$.  Due to the nature of
-$\mathcal{K}(\cdot)$, the maximum value of $\mathcal{K}'(\cdot)$ is
+\mathcal{K}(b)$ for a fixed value of $b - a > 0$.  Due to our assumptions, the
+maximum value of $\mathcal{K}'(\cdot)$ is
 $\mathcal{K}'(h)$; therefore, the maximum possible value of $\mathcal{K}(a) -
@@
 
-Therefore, $|R^*| = \emptyset$ when
+Therefore, $R^* = \emptyset$ when
 %\begin{eqnarray}
@@
 \zeta)} = c_r^{4 + \log_2 \zeta}$.  Therefore, the entire dual-tree traversal
-takes $O(c_r^{8 + \log_2 \zeta} \nu N)$ time.
+takes $O(c_r^{8 + \log_2 \zeta} (N + \theta))$ time.
 
@@
 
+To understand the lack of dependence on kernel bandwidth more intuitively,
+consider that as the kernel bandwidth increases, two things happen: {\it (a)}
+the reference set $R$ becomes empty at larger scales, and {\it (b)}
+$\mathcal{K}^{-1}(\epsilon)$ grows, allowing less pruning at higher levels.
+These effects are opposite, and for the Gaussian kernel they cancel each other
+out, giving the same bound regardless of bandwidth.
+
+Other kernels do not exhibit this effect.  Consider the Laplacian kernel:
+
 %Next, consider this sigmoidal kernel: $\mathcal{K}_s(d) = (1 + \exp(\alpha (d -

### 2014-11-26 bounds-files/sec01-supporting-lemmas.tex
Check in discussion of implicit and explicit tree, and discuss the expansion constant a bit more.  I'm still waiting on numbers.

@@
 
+\subsection{The cover tree}
+
 The cover tree is a leveled hierarchical data structure originally proposed for
-the task of nearest neighbor search.  The original description of the tree
-describes both an {\it implicit} and {\it explicit} representation, where the
-explicit representation is the actual structure that would be implemented in
-practice.  Thus, we will consider only the explicit representation in this
-paper.
+the task of nearest neighbor search by \citet{langford2006}.  Each node
+$\mathscr{N}_i$ in the cover tree is associated with a single point $p_i$.  An
+adequate description is given in their work (we have adapted notation slightly):
 
-Each node $\mathscr{N}_i$ in the cover tree is indexed by an integer scale $s_i$
-and holds one point $p_i$ which is also the center of the node ($p_i = \mu_i$).
-The cover tree node may have a number of child nodes; if the cover tree node
-does have children, then one child will be the self-child, which holds the same
-point $p_i$ with a smaller scale $s_i$.  Each child of $\mathscr{N}_i$ will have
-scale less than $s_i$, and every descendant point of a node $\mathscr{N}_i$
-(that is, the set $\mathscr{D}_i^p$) has a maximum distance of $2^{s_i + 1}$
-from the point $p_i$.
-
-Assuming a cover tree $\mathscr{T}$ is built on a dataset $S_r$, let $C_{s_i}$
-denote the set of points held in all nodes in $\mathscr{T}$ with scale $s_i$.
-Then, the cover tree satisfies the following invariants for all scales $s_i$:
+\begin{quote}
+A {\it cover tree} $\mathscr{T}$ on a dataset $S$ is a leveled tree where each
+level is a ``cover'' for the level beneath it.  Each level is indexed by an
+integer scale $s_i$ which decreases as the tree is descended.  Every {\it node}
+in the tree is associated with a point in $S$.  Each {\it point} in $S$ may be
+associated with multiple nodes in the tree; however, we require that any point
+appears at most once in every level.  Let $C_{s_i}$ denote the set of points in
+$S$ associated with the nodes at level $s_i$.  The cover tree obeys the
+following invariants for all $s_i$:
 
 \begin{itemize}
+  \item {\em (Nesting)}. $C_{s_i} \subset C_{s_i - 1}$.  This implies that once a
+point $p \in S$ appears in $C_{s_i}$ then {\it every} lower level in the tree
+has a node associated with $p$.
 
-\item {\em (Nesting)}.  $C_{s_i} \subset C_{s_i - 1}$.  This means that when a
-point $p_i \in S_r$ is held in a node at some scale $s_i$, then each smaller
-scale will also have a node containing $p_i$.
+  \item {\em (Covering tree)}. For every $p_i \in C_{s_i - 1}$, there exists a
+$p_j \in C_{s_i}$ such that $d(p_i, p_j) < 2^{s_i}$ and the node in level $s_i$
+associated with $p_j$ is a parent of the node in level $s_i - 1$ associated with
+$p_i$.
 
-\item {\em (Covering tree)}.  For every point $p_i \in C_{s_i - 1}$, there
-exists a point $p_j \in C_{s_i}$ such that $d(p_i, p_j) < 2^{s_i}$, and the node
-holding $p_j$ with scale $s_i$ is the parent node of the node holding $p_i$ with
-scale $s_i - 1$.
+  \item {\em (Separation)}.  For all distinct $p_i, p_j \in C_{s_i}$, $d(p_i,
+p_j) > 2^{s_i}$.
+\end{itemize}
+\end{quote}
 
-\item {\em (Separation)}.  For all distinct points $p_i, p_j \in C_{s_i}$,
-$d(p_i, p_j) > 2^{s_i}$.
+A consequence of this definition is if there exists a node $\mathscr{N}_i$,
+containing the point $p_i$ at some scale $s_i$, then there will also exist a
+self-child node $\mathscr{N}_{ic}$ containing the point $p_i$ at scale $s_i - 1$
+which is a child of $\mathscr{N}_i$.  In addition, every descendant point of the
+node $\mathscr{N}_i$ is contained within a ball of radius $2^{s_i + 1}$ centered
+at the point $p_i$; therefore, $\lambda_i = 2^{s_i + 1}$ and $\mu_i = p_i$
+(Table \ref{tab:notation}).
 
-\end{itemize}
+Note that the cover tree may be interpreted as an infinite-leveled tree, with
+$C_{\infty}$ containing only the root point, $C_{-\infty} = S$, and all levels
+between defined according to the definitions above.  \citep{langford2006} find
+this representation (which they call the {\it implicit} representation) easier
+for description of their algorithms and some of their proofs.  But naturally,
+this is not suitable for implementation; hence, there is an {\it explicit}
+representation in which all nodes that have only a self-child are coalesced.
+Figure \ref{fig:implicit-explicit} shows an example cover tree and highlights
+implicit and explicit nodes.
 
-A batch construction algorithm is given by \citet{langford2006}, called
-\texttt{Construct}; but the details of construction are not relevant to our
-work, so we only mention this as a reference.  More relevant to our work is the
-number of useful and unique theoretical properties the cover tree possesses,
-based on the expansion constant \citep{karger2002finding}; we restate the
-definition below.
+In our work here, we consider only the explicit representation of a cover tree,
+and do not concern ourselves with the construction of the tree\footnote{A batch
+construction algorithm is given by \citet{langford2006}, called
+\texttt{Construct}.}.
+
+\subsection{Expansion constant}
+
+The explicit representation of a cover tree has a number of useful theoretical
+properties based on the expansion constant \citep{karger2002finding}; we restate
+its definition below.
 
@@
 
-The expansion constant is used heavily in the cover tree literature.  It is
-related to a notion of instrinic dimensionality, and previous work has shown
+The expansion constant is used heavily in the cover tree literature.  It is,
+in some sense, a notion of instrinic dimensionality, and previous work has shown
 that there are many scenarios where $c$ is independent of the number of points
-in the dataset.
+in the dataset \citep{karger2002finding, langford2006,
+krauthgamer2004navigating, ram2009}.  Note also that if points in $S \subset
+\mathcal{H}$ are being drawn according to a stationary distribution $f(x)$, then
+$c$ will converge to some finite value $c_f$ as $|S| \to \infty$.  To see this,
+define $c_f$ as a generalization of the expansion constant for distributions.
+$c_f \ge 2$ is the smallest value such that
+
+\begin{equation}
+\int_{\mathcal{B}_{\mathcal{H}}(p, 2 \Delta)} f(x) dx \le c_{f}
+\int_{\mathcal{B}_{\mathcal{H}}(p, \Delta)} f(x)
+dx
+\end{equation}
+
+\noindent for all $p \in \mathcal{H}$ and $\Delta > 0$ such that
+$\int_{\mathcal{B}_{\mathcal{H}}(p, \Delta)} f(x) dx > 0$, and with
+$\mathcal{B}_{\mathcal{H}}(p, \Delta)$ defined as the closed ball of radius
+$\Delta$ in the space $\mathcal{H}$.
+
+Take $f(x)$ as a uniform spherical distribution in $\mathcal{R}^d$: for any $|x|
+\le 1$, $f(x)$ is a constant; for $|x| > 1$, $f(x) = 0$.  It is easy to see that
+$c_f$ in this situation is $2^d$.  To demonstrate this convergence, Table
+\ref{tab:ec_scaling} shows results for 10 trials of empirically calculated
+expansion constants for datasets drawn from this distribution, for varying $d$.
+
+\begin{table}[htb]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|}
+\hline
+$|S|$ & $d = 2$ & $d = 5$ & $d = 10$ & $d = 20$ \\
+\hline
+10      & 6     & 10    & 10    &         \\
+31      & 12    & 31    & 31    &         \\
+100     & 13    & 64    & 100   &         \\
+316     & 23    & 96    & 286   &         \\
+1000    & 26    & 126   & 673   &         \\
+3162    & 29    & 189   & 1348  &         \\
+10000   & 30    & 259   & 2646  &         \\
+31622   & 39    & 309   & 3858  &         \\
+100000  & 40    & 360   & 7162  &         \\
+316227  & 46    & 385   & 8297  &         \\
+1000000 &       &       &       &         \\
+$\infty$ ($c_f$)   & 4     & 32    & 1024  & 1048576 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Calculated $c$ for uniform hypersphere distribution as
+$|S|$ scales.}
+\label{tab:ec_scaling}
+\end{table}
+
+We are somewhat limited in the sizes of $|S|$ we can scale to, because
+calculating the expansion constant of a dataset is a computationally intensive
+task.  Nonetheless, we can see that the expansion constant moves towards the
+final value of $c_f$.  We postulate that the behavior of more complex real-world
+datasets is similar.
+
+There are some other important observations about the behavior of $c$.  Adding a
+single point to $S$ may increase $c$ arbitrarily: consider a set $S$ distributed
+entirely on the surface of a unit hypersphere.  If one adds a single point at
+the origin, producing the set $S'$, then $c$ explodes to $|S'|$ whereas before
+it may have been much smaller than $|S|$.  Adding a single point may also
+decrease $c$ significantly.  Suppose one adds a point arbitrarily close to the
+origin to $S'$; now, the expansion constant will be $|S'| / 2$.  So, although we
+can bound the behavior of $c$ as $|S| \to \infty$ for $S$ from a stationary
+distribution, we are not able to easily say much about its convergence behavior.
 

### 2014-11-26 bounds-files/imbalance.tex
Refactor slightly and prepare for some empirical results.

@@
 child is effectively useless.  A $kd$-tree full of nodes like this will perform
-abysmally for nearest neighbor search, and it is not hard to generate a dataset
-that will cause a $kd$-tree of this sort.
+abysmally for nearest neighbor search, and it is not hard to generate a
+pathological dataset that will cause a $kd$-tree of this sort.
 
@@
 \end{center}
+\caption{Single-outlier cover tree.}
 \label{fig:outlier}
-\caption{Single-outlier cover tree.}
 \end{figure}
@@
 
-As we will see later empirically, the notion of cover tree imbalance does a good
-job of capturing the `goodness' of a tree.  We will use this notion in order to
-quantify the worst-case performance of dual-tree algorithms using the cover
-tree.
+For most real-world datasets, the tree imbalance is sublinear in the number of
+points.  To demonstrate this, Table \ref{tab:emp_imbalance} shows empirically
+calculated cover tree imbalance for cover trees on a number of datasets built
+with both the cover tree implementation in {\bf mlpack} \citep{mlpack2013} and
+the original reference implementation \citep{langford2006}.
+
+\begin{table}[tb]
+\begin{center}
+\begin{tabular}{|c|c|c|r|r|}
+\hline
+Dataset & $N$ & $d$ & reference & mlpack \\
+\hline
+wine & & & & \\
+isolet & & & & \\
+corel & & & & \\
+mnist & 70000 & 784 & & \\
+bio & & & & \\
+phy & & & & \\
+covertype & & & & \\
+LCDM & & & & \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Empirical tree imbalance calculations.}
+\label{tab:emp_imbalance}
+\end{table}
 

### 2014-11-26 bounds-files/sec01-supporting-lemmas.tex
Some changes and a better range search proof using Pari's assumptions.

@@
 task.  Nonetheless, we can see that the expansion constant moves towards the
-final value of $c_f$.  We postulate that the behavior of more complex real-world
-datasets is similar.
+final value of $c_f$, though potentially from above ({\it The results are not
+done yet.  This paragraph may need rewriting.  But I strongly suspect the
+results will show what I've said.}).  We postulate that the behavior of more
+complex real-world datasets is similar.
 

### 2014-11-26 bounds-files/sec04-range-search.tex
Some changes and a better range search proof using Pari's assumptions.

@@
     \IF{$d(p_q, p_r) \in [r_{\min}, r_{\max}]$ \AND \texttt{BaseCase($p_q$, $p_r$)} not yet called}
-    \STATE  $N[p_q] \gets N[p_q] \cup \{ p_r \}$
+    \STATE  $S[p_q] \gets S[p_q] \cup \{ p_r \}$
     \ENDIF
@@
 %\begin{equation}
-$N[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}$
+$S[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}$
 %\end{equation}
@@
 range.  The range count problem is practically identical, but only the size of
-the set, $|N[p_q]|$, is desired.
+the set, $|S[p_q]|$, is desired.
 Our proof works for both of these algorithms
@@
 initialized to $\emptyset$ at the beginning of the traversal.
+
+In order to bound the running time of dual-tree range search, we require better
+notions for understanding the difficulty of the problem.  Observe that if the
+range is sufficiently large, then for every query point $p_q$, $S[p_q] = S_r$.
+Clearly, for $S_q \sim S_r \sim O(N)$, this cannot be solved in anything less
+than quadratic time simply due to the time required to fill each output array
+$S[p_q]$.  Define the maximum result size for a given query set $S_q$, reference
+set $S_r$, and range $[l, u]$ as
+
+\begin{equation}
+| S_{\max} | = \max_{p_q \in S_q} | S[p_q] |.
+\label{eqn:smax}
+\end{equation}
+
+Small $| S_{\max} |$ implies an easy problem; large $| S_{\max} |$ implies a
+difficult problem.  For bounding the running time of range search, we require
+one more definition.
+
+\begin{defn}
+For a range search problem with query set $S_q$, reference set $S_r$, range $[l,
+u]$, and results $S[p_q]$ for each query point $p_q$ given as
+
+\begin{equation}
+S[p_q] = \{ p_r : p_r \in S_r, l \le d(p_q, p_r) \le u \},
+\end{equation}
+
+\noindent define the {\it $\alpha$-expansion} of the range set $S[p_q]$ as the slightly larger set
+
+\begin{equation}
+S^{\alpha}[p_q] = \{ p_r : p_r \in S_r, (1 - \alpha) l \le d(p_q, p_r) \le (1 +
+\alpha) u \}.
+\end{equation}
+\end{defn}
+
+Using these notions, we may now bound the running time of range search.
 
@@
 \begin{equation}
-\left| \{ p_r: p_r \in S_r, (1 - \alpha) l \le d(p_q, p_r) \le (1 + \alpha) u \}
-\right| = O(1),
+| S^{\alpha}[p_q] \setminus S[p_q] | \le C \; \; \forall \; p_q \in S_q,
 \end{equation}
 
+%\begin{equation}
+%\left| \{ p_r: p_r \in S_r, (1 - \alpha) l \le d(p_q, p_r) \le (1 + \alpha) u \}
+%\right| = O(1),
+%\end{equation}
+
 \noindent the running time of range search or range count is bounded by
-$O(c_r^{8 + \beta} (N + \theta))$ with $\theta$ defined as in Theorem
-\ref{thm:ct-runtime} and
-%\begin{equation}
-$\beta = \log_2 (1 + \alpha^{-1})$.
+
+\begin{equation}
+O\left(c_r^{4} \max\left(c_r^{4 + \beta}, |S_{\max}| + C\right) (N + \theta)
+\right)
+\end{equation}
+
+with $\theta$ defined as in Theorem
+\ref{thm:ct-runtime},
+$\beta = \log_2 (1 + \alpha^{-1})$, and $S_{\max}$ as defined in Equation
+\ref{eqn:smax}.
 %\end{equation}
@@
 \ref{thm:ct-runtime} we know that the runtime of the algorithm is bounded by
-$O(c_r^4 |R^*| (N + \theta))$. %  Thus, our only task is to bound the maximum size of the
-%reference set, $|R^*|$.
-By the pruning rule, for a query node
-$\mathscr{N}_q$, the reference set $R^*$
+$O(c_r^4 |R^*| (N + \theta))$.  As with the previous proofs, then, our only task
+is to bound the maximum size of the reference set, $|R^*|$.
+
+By the pruning rule, for a query node $\mathscr{N}_q$, the reference set $R^*$
 is made up of reference nodes $\mathscr{N}_r$ that are within a margin of
@@
 1}) \subseteq B_{S_r}(p_q, (1 + \alpha) u) \setminus B_{S_r}(p_q, (1 - \alpha)
-l)
+l).
 \end{equation}
 
-\noindent and, due to the assumptions, $| B_{S_r}(p_q, (1 + \alpha) u) | - |
-B_{S_r}(p_q, (1 - \alpha) l) | = O(1)$, $|R^*| = O(1)$.  Thus, the theorem
-holds.
+This set is necessarily a subset of $S^{\alpha}[p_q]$; by
+definition, the number of points in this set is bounded above by $|S_{\max}| +
+C$.  We may then conclude that $|R^*| \le |S_{\max}| + C$.  By taking the
+maximum of the sizes of $|R^*|$ in both cases above, we obtain the statement of
+the theorem.
 \end{proof}
+
+This bound displays both the expected dependence on $c_r$ and $|S_{\max}|$.  As
+the largest range set $S_{\max}$ increases in size (with the worst case being
+$S_{\max} \sim N$), the runtime degenerates to quadratic.  But for adequately
+small $S_{\max}$ the runtime is instead dependent on $c_r$ and the parameter $C$
+of the $\alpha$-expansion of $S_{\max}$.  This situation leads to a
+simplification.
+
+\begin{cor}
+For sufficiently small $|S_{\max}|$ and sufficiently small $C$, the runtime of
+range search under the conditions of Theorem \ref{thm:rs} simplifies to
+
+\begin{equation}
+O(c_r^{8 + \beta} (N + \theta)).
+\end{equation}
+\label{cor:rs}
+\end{cor}
 
@@
 
-To understand the relation of the running time to the bounding quantity
-$\alpha$, consider $\alpha =
-(1 / 3)$; this yields a running time of $O(c^8 (N + \theta))$.  $\alpha = (1 /
-7)$ yields $O(c^9 (N + \theta))$, $\alpha = (1 / 15)$ yields $O(c^{10} (N +
-\theta))$, and so forth.  As $\alpha$ gets smaller, the exponent on $c$ gets
-larger, and diverges as $\alpha \to 0$.
+In this setting we can more easily consider the relation of the running time to
+$\alpha$.  Consider $\alpha = (1 / 3)$; this yields a running time of $O(c^8 (N
++ \theta))$.  $\alpha = (1 / 7)$ yields $O(c^9 (N + \theta))$, $\alpha = (1 /
+15)$ yields $O(c^{10} (N + \theta))$, and so forth.  As $\alpha$ gets smaller,
+the exponent on $c$ gets larger, and diverges as $\alpha \to 0$.
 
-The assumption that on number of points in the range of $[(1 - \alpha)l, (1 +
-\alpha)u]$ for a single query is not too restrictive, but must be satisfied for
-the result to be meaningful.  After all, if that assumption is not satisfied,
-then it is not possible to perform range search in $O(1)$ time for each query
-point, making an $O(N)$ bound for the whole query set impossible.
+For reasonable runtime it is necessary that the $\alpha$-expansion of $S_{\max}$
+be bounded.  This is because the dual-tree recursion must retain reference nodes
+which may contain descendants in the range set $S[p_q]$ for some query $p_q$.
+The parameter $C$ of the $\alpha$-expansion allows us to bound the number of
+reference nodes of this type, and if $\alpha$ increases but $C$ remains small
+enough that Corollary \ref{cor:rs} applies, then we are able to obtain tighter
+running bounds.
 
-To show where the assumption is valid, consider that when presented with a very
-large dataset, a practitioner may perform monochromatic range search on a range
-$[l, u]$ for a geometric subset of their data; for instance, a hypersquare
-with side length $2u$.  This might be done in order to find a range $[l, u]$
-that produces a manageable number of results (if the range is too large, the
-number of results returned for each query point may be enormous).  Then, that
-practitioner may scale to a larger dataset by simply increasing the size of
-their hypersquare.  In this setting, the number of points in the range of
-$[(1 - \alpha)l, (1 + \alpha)u]$ for a single query remains $O(1)$ with respect
-to $N$.  Alternately, it should be possible to tighten the range $[l, u]$ as $N$
-increases in such a way that the number of points in the range remains constant.
+%The assumption that on number of points in the range of $[(1 - \alpha)l, (1 +
+%\alpha)u]$ for a single query is not too restrictive, but must be satisfied for
+%the result to be meaningful.  After all, if that assumption is not satisfied,
+%then it is not possible to perform range search in $O(1)$ time for each query
+%point, making an $O(N)$ bound for the whole query set impossible.
+
+%To show where the assumption is valid, consider that when presented with a very
+%large dataset, a practitioner may perform monochromatic range search on a range
+%$[l, u]$ for a geometric subset of their data; for instance, a hypersquare
+%with side length $2u$.  This might be done in order to find a range $[l, u]$
+%that produces a manageable number of results (if the range is too large, the
+%number of results returned for each query point may be enormous).  Then, that
+%practitioner may scale to a larger dataset by simply increasing the size of
+%their hypersquare.  In this setting, the number of points in the range of
+%$[(1 - \alpha)l, (1 + \alpha)u]$ for a single query remains $O(1)$ with respect
+%to $N$.  Alternately, it should be possible to tighten the range $[l, u]$ as $N$
+%increases in such a way that the number of points in the range remains constant.
 
@@
 
-Alternately, if the assumption is not valid but it can be shown that the number
-of points in the range $[(1 - \alpha)l, (1 + \alpha)u]$ is $O(k)$ where $k$ has
-some dependence on $N$, then the running time can be easily bounded as $O(c^{4 +
-\beta} k(N + \theta))$.
+%Alternately, if the assumption is not valid but it can be shown that the number
+%of points in the range $[(1 - \alpha)l, (1 + \alpha)u]$ is $O(k)$ where $k$ has
+%some dependence on $N$, then the running time can be easily bounded as $O(c^{4 +
+%\beta} k(N + \theta))$.
 

### 2014-12-01 bounds-files/sec01-supporting-lemmas.tex
clarified proof of lemma 1

@@
 Consider a dataset $S$ with expansion constant $c$ and a subset $C \subseteq S$
-such that every point in $C$ is separated by $\delta$.  Then, given any query
-point $p \not\in S$ and some radius $\rho \delta$,
-
+such that every point in $C$ is separated by $\delta$.  Then, for any point 
+$p \not\in S$ and any radius $\rho \delta > 0$:
 \begin{equation}
-| B_S(p, \rho \delta) \cap C | \le c^{2 + \log_2 \rho}.
+| B_S(p, \rho \delta) \cap C | \le c^{2 + \lceil \log_2 \rho \rceil}.
 \end{equation}
@@
 \begin{proof}
-This is based on the packing argument from Lemma 4.1 in \cite{langford2006}.
-Observe that $B_S(p, \rho \delta) \subseteq B_S(p_i, 2 \rho \delta)$ for any
-$p_i \in S$, and that $| B_S(p, 2 \rho \delta) | = c^{2 + \log_2 \rho} |
-B_S(p, \delta / 2) |$.  Because each point in $C$ is separated by $\delta$, the
+The proof is based on the packing argument from Lemma 4.1 in 
+\cite{langford2006}. Consider two cases: first, let $d(p, p_i) > \rho \delta$ 
+for any $p_i \in S$. In this case, $B_S(p, \rho \delta) = \emptyset$ and the 
+lemma holds trivially. 
+Otherwise, let $p_i \in S$ be a point such that $d(p, p_i) \leq \rho \delta$.
+Observe that $B_S(p, \rho \delta) \subseteq B_S(p_i, 2 \rho \delta)$. 
+Also, $| B_S(p_i, 2 \rho \delta) | = c^{2 + \lceil \log_2 \rho 
+\rceil} | B_S(p_i, \delta / 2) |$ by the definition of the expansion constant.
+Because each point in $C$ is separated by $\delta$, the
 number of points in $B_S(p, \rho \delta) \cap C$ is
@@
 |B_S(p, \rho \delta)| \le \frac{|B_S(p_i, 2 \rho \delta)|}{|B_S(p_i, \delta
-/ 2)|} \le c^{2 + \log_2 \rho}.
+/ 2)|} \le c^{2 + \lceil \log_2 \rho \rceil}.
 \end{equation}

### 2014-12-02 bounds-files/sec03-approx-kde.tex
Some changes and a better range search proof using Pari's assumptions.
Propagate changes to Lemma 1 to the document.

@@
     \STATE $f_p(p_q) \gets f_p(p_q) + \mathcal{K}(p_q, p_r)$
-\ref{alg:kde-bc-update}
+\label{alg:kde-bc-update}
     \RETURN $\mathcal{K}(p_q, p_r)$
@@
 %\begin{equation}
-$O(c_r^{8 + \log_2 \zeta} (N + \theta))$
+$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + \theta))$
 %\end{equation}
@@
 2^{s_r^{\max} + 1} + 1$.  Using Lemma \ref{lem:packing} with $\delta =
-2^{s_r^{\max}}$ and $\rho = 2 \omega$ gives $|R^*| = c_r^{3 + \log_2 \omega}$.
+2^{s_r^{\max}}$ and $\rho = 2 \omega$ gives $|R^*| = c_r^{3 + \lceil \log_2
+\omega \rceil}$.
 
@@
 Finally, with $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon)
-\epsilon^{-1}$, we are able to conclude that $|R^*| \le c_r^{3 + \log_2 (2
-\zeta)} = c_r^{4 + \log_2 \zeta}$.  Therefore, the entire dual-tree traversal
-takes $O(c_r^{8 + \log_2 \zeta} (N + \theta))$ time.
+\epsilon^{-1}$, we are able to conclude that $|R^*| \le c_r^{3 + \lceil \log_2
+(2 \zeta ) \rceil} = c_r^{4 + \lceil \log_2 \zeta \rceil}$.  Therefore, the
+entire dual-tree traversal takes $O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N +
+\theta))$ time.
 
@@
 
-Other kernels do not exhibit this effect.  Consider the Laplacian kernel:
+Next, consider the Laplacian kernel:
+$\mathcal{K}_l(d) = \exp(-d / \sigma)$.  For this kernel, $h = 0$ (that is, the
+kernel is always convex), so then $\mathcal{K}'_l(h) = \sigma^{-1}$.  Simple
+algebraic manipulation gives $\mathcal{K}^{-1}_l(\epsilon) = -\sigma \ln
+\epsilon$, resulting in $\zeta = -\mathcal{K}'_l(h) \mathcal{K}^{-1}_l(\epsilon)
+\epsilon^{-1} = \epsilon^{-1} \ln \epsilon$.  So both the Laplacian and Gaussian
+kernels do not exhibit dependence on the bandwidth.
 

### 2014-12-02 bounds.tex
The appendix is embarrassingly wrong.

@@
 %range count/search.
-Numerous machine learning algorithms contain pairwise statistical problems in
+Numerous machine learning algorithms contain pairwise statistical problems at
 their core. Often, tree structures are used to solve these problems efficiently.
@@
 those algorithms.  Each of these bounds is an improvement on the
-state-of-the-art, and in the case of range search, is the first such bound.  We
-then investigate the empirical performance of cover trees for these algorithms
-in Section \ref{sec:emp}, with a focus on understanding the dataset-dependent
-constants that our bounds use.
+state-of-the-art, and in the case of range search, is the first such bound.
+%  We
+%then investigate the empirical performance of cover trees for these algorithms
+%in Section \ref{sec:emp}, with a focus on understanding the dataset-dependent
+%constants that our bounds use.
 
@@
 
-\section{Empirical investigations}
-\label{sec:emp}
+%\section{Empirical investigations}
+%\label{sec:emp}
 
-\input{bounds-files/empirical.tex}
+%\input{bounds-files/empirical.tex}
 
 \section{Conclusion}
+\label{sec:conclusion}
 
@@
 % Appendix doesn't have anything useful or non-straightforward.
-\appendix
-\input{appendix-ec}
+%\appendix
+%\input{appendix-ec}
 

### 2014-12-24 bounds-files/conclusion.tex
Ready for review.

@@
 algorithms that use cover trees and the standard cover tree pruning dual-tree
-traversal (Algorithm \ref{alg:cover-tree-dual}).  Our main result, Theorem
-\ref{thm:ct-runtime}, allows plug-and-play runtime bounding of these algorithms.
-We have shown that Theorem \ref{thm:ct-runtime} is useful for bounding the
-runtime of nearest neighbor search (Theorem \ref{thm:nns}), approximate kernel
-density estimation (Theorem \ref{thm:kde-bound}), exact range count, and exact
-range search (Theorem \ref{thm:rs}).  With our contribution, bounding a cover
-tree dual-tree algorithm is straightforward and only involves bounding the
-maximum size of the reference set, $| R^* |$.
+traversal (Algorithm \ref{alg:cover-tree-dual}).  In order to produce an
+understandable bound, we have introduced the notion of cover tree imbalance; one
+possible interesting direction of future work is to empirically and
+theoretically minimize this quantity by way of modified tree construction
+algorithms; this is likely to provide both tighter runtime bounds and also
+accelerated empirical results.
+
+Our main result, Theorem \ref{thm:ct-runtime}, allows plug-and-play runtime
+bounding of these algorithms.  We have shown that Theorem \ref{thm:ct-runtime}
+is useful for bounding the runtime of nearest neighbor search (Theorem
+\ref{thm:nns}), approximate kernel density estimation (Theorem
+\ref{thm:kde-bound}), exact range count, and exact range search (Theorem
+\ref{thm:rs}).  With our contribution, bounding a cover tree dual-tree algorithm
+is straightforward and only involves bounding the maximum size of the reference
+set, $| R^* |$.

### 2014-12-24 bounds-files/general-runtime-bound.tex
Ready for review.

@@
     \STATE $s^{\max}_r \gets \max_{\mathscr{N}_r \in R} s_r$
+\label{alg:line:ct-dual-srmax}
     \IF{$(s_q < s^{\max}_r)$ {\bf or} $(s^{\max}_r = -\infty)$}
@@
 This dual-tree recursion is a depth-first recursion in the query tree and a
-breadth-first recursion in the reference tree.  It descends either the query
-tree or the reference tree (not both) during a single iteration, and the
-conditional in line \ref{alg:line:ct-dual-ref-recursion-start} ensures that the
-combinations of query and reference nodes that are being visited are close in
-scale.
+breadth-first recursion in the reference tree; to this end, the recursion
+maintains one query node $\mathscr{N}_q$ and a reference set $R$.  The set $R$
+may contain reference nodes with many different scales; the maximum scale in the
+reference set is $s_r^{\max}$ (line \ref{alg:line:ct-dual-srmax}).  Each single
+recursion will descend either the query tree or the reference tree, not both;
+the conditional in line \ref{alg:line:ct-dual-ref-recursion-start}, which
+determines whether the query or reference tree will be recursed, is aimed at
+keeping the relative scales of query nodes and reference nodes equal.
 
-Due to the favorable theoretical properties of the cover tree, we can prove
-general runtime bounds when this traversal is used.
+A query recursion (lines
+\ref{alg:line:ct-dual-query-recursion-start}--\ref{alg:line:ct-dual-query-recursion-end})
+is straightforward: for each child $\mathscr{N}_{qc}$ of $\mathscr{N}_q$, the
+node combinations $(\mathscr{N}_{qc}, \mathscr{N}_r)$ are scored for each
+$\mathscr{N}_r$ in the reference set $R$.  If possible, these combinations are
+pruned to form the set $R'$ (line \ref{alg:line:ct-dual-query-recursion}), and
+then the algorithm recurses with $\mathscr{N}_{qc}$ and $R'$.
+
+A reference recursion (lines
+\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
+is similar to a query recursion, but the pruning strategy is significantly more
+complicated.  We score each node combination $(\mathscr{N}_q, \mathscr{N}_r)$
+for all $\mathscr{N}_r$ in $R_{r - 1}$, which is the set of children of all
+nodes in $R$ that have scale $s_r^{\max}$ (that set is $R_r$).  The set of
+reference nodes in $R_{r - 1}$ that were not pruned create the set $R'_{r - 1}$.
+Then, this set is combined with $R \setminus R_r$---that is, each of the nodes
+in $R$ that was {\em not} recursed into---to produce $R'$, and the algorithm
+recurses with $\mathscr{N}_q$ and the reference set $R'$.
+
+The reference recursion only recurses into the top-level subset of the reference
+nodes in order to preserve the separation invariant.  It is easy to show that
+every pair of points held in nodes in $R$ is separated by at least
+$2^{s_r^{\max}}$:
+
+\begin{lemma}
+For all nodes $\mathscr{N}_i, \mathscr{N}_j \in R$ (in the context of Algorithm
+\ref{alg:cover-tree-dual}) which contain points $p_i$ and $p_j$, respectively,
+$d(p_i, p_j) > 2^{s_r^{\max}}$, with $s_r^{\max}$ defined as in line
+\ref{alg:line:ct-dual-srmax}.
+\end{lemma}
+
+\begin{proof}
+This proof is by induction.  If $|R| = 1$, such as during the first reference
+recursion, the result obviously holds.  Now consider any reference set $R$ and
+assume the statement of the lemma holds for this set $R$, and define
+$s_r^{\max}$ as the maximum scale of any node in $R$.  Construct the set
+$R_{r - 1}$ as in line \ref{alg:line:ct-dual-ref-children} of Algorithm
+\ref{alg:cover-tree-dual}.
+
+Now, take any $\mathscr{N}_i, \mathscr{N}_j$ in $R_{r - 1}$, with points $p_i$
+and $p_j$, respectively, and scales $s_i$ and $s_j$, respectively.  Clearly, if
+$s_i = s_j = s_r^{\max} - 1$, then by the separation invariant $d(p_i, p_j) >
+2^{s_r^{\max} - 1}$.
+
+Now suppose that $s_i < s_r^{\max} - 1$.  This implies that there exists some
+implicit cover tree node with point $p_i$ and scale $s_r^{\max} - 1$ (as well an
+implicit child of this node $p_i$ with scale $s_r^{\max} - 2$ and so forth until
+one of these implicit nodes has child $p_i$ with scale $s_i$).  Because the
+separation invariant applies to both implicit and explicit representations of
+the tree, we conclude that $d(p_i, p_r) > 2^{s_r^{\max}} - 1$.  The same
+argument may be made for the case where $s_j < s_r^{\max} - 1$, with the same
+conclusion.
+
+We may therefore conclude that each point of each node in $R_{r - 1}$ is
+separated by $2^{s_r^{\max} - 1}$.  Note that $R'_{r - 1} \subseteq R_{r - 1}$
+and that $R \setminus R_{r - 1} \subseteq R$, and that each point of each node
+is separated by $2^{s_r^{\max} - 1}$, to see that this condition holds for all
+nodes in $R' = R'_{r - 1} \cup (R \setminus R_{r - 1})$.
+
+Because we have shown that the condition holds for the initial reference set and
+for any reference set produced by a reference recursion, we have shown that the
+statement of the lemma is true.
+\end{proof}
+
+This observation means that the set of points $P$ held by all nodes in $R$ is
+always a subset of $C_{s_r^{\max}}$.  This fact will be useful in our later
+runtime proofs.
+
+Next, we develop notions with which to understand the behavior of the cover tree
+dual-tree traversla when the datasets are of significantly different scale
+distributions.
+
+If the datasets are similar in scale distribution (that is, inter-point distances
+tend to follow the same distribution), then the recursion will alternate between
+query recursions and reference recursions.  But if the query set contains points
+which are, in general, much further apart than the reference set, then the
+recursion will start with many query recursions before reaching a reference
+recursion.  The converse case also holds.  We are interested in formalizing
+this notion of scale distribution; therefore, define the following
+dataset-dependent constants for the query set $S_q$ and the reference set $S_r$:
+
+\begin{itemize}
+  \item $\eta_q$: the largest pairwise distance in $S_q$
+  \item $\delta_q$: the smallest nonzero pairwise distance in $S_q$
+  \item $\eta_r$: the largest pairwise distance in $S_r$
+  \item $\delta_r$: the smallest nonzero pairwise distance in $S_r$
+\end{itemize}
+
+These constants are directly related to the aspect ratio of the datasets;
+indeed, $\eta_q / \delta_q$ is exactly the aspect ratio of $S_q$.  Further,
+let us define the top and bottom levels of each tree:
+
+\begin{itemize}
+  \item The {\it top scale} of the query tree $\mathscr{T}_q$ is defined as
+$s_q^T = \lceil \log_2(\eta_q) \rceil$.
+  \item The {\it minimum scale} of the query tree $\mathscr{T}_r$ is defined as
+$s_q^{\min} = \lceil \log_2(\delta_q) \rceil$.
+  \item The top scale of the reference tree $\mathscr{T}_r$ is defined as $s_r^T
+= \lceil \log_2(\eta_r) \rceil$.
+  \item The minimum scale of the reference tree $\mathscr{T}_r$ is defined as
+$s_r^{\min} = \lceil \log_2(\delta_r) \rceil$.
+\end{itemize}
+
+Note that the minimum scale is not the minimum scale of {\it any} cover tree
+node (that would be $-\infty$), but the minimum scale of any non-leaf node in
+the tree.
+
+Suppose that our datasets are of a similar scale distribution: $s_q^T = s_r^T$,
+and $s_q^{\min} = s_r^{\min}$.  In this setting we will have alternating query
+and reference recursions.  But if this is not the case, then we have extra
+reference recursions before the first query recursion or after the last query
+recursion (situations where both these cases happen are possible).  Motivated by
+this observation, let us quantify these extra reference recursions:
+
+\begin{lemma}
+\label{lem:extcase1}
+For a dual-tree algorithm with $S_q \sim S_r \sim O(N)$ using cover trees and
+the traversal given in Algorithm \ref{alg:cover-tree-dual}, the number of extra
+reference recursions that happen before the first query recursion is bounded by
+
+\begin{equation}
+\min\left(O(N), \lceil \log_2(\eta_r / \eta_q) \rceil\right).
+\end{equation}
+\end{lemma}
+
+\begin{proof}
+The first query recursion happens once $s_q > s_r^{\max}$.  The number of
+reference recursions before the first query recursion is then bounded as the
+number of levels in the reference tree between $s_r^T$ and $s_q^T$ that have at
+least one explicit node.  Because there are $O(N)$ nodes in the reference tree,
+the number of levels cannot be greater than $O(N)$ and thus the result holds.
+
+The second bound holds by applying the definitions of $s_r^T$ and $s_q^T$ to the
+expression $s_r^T - s_q^T$.
+\end{proof}
+
+Note that the $O(N)$ bound may be somewhat loose, but it suffices for our later
+purposes.  Now let us consider the other case:
+
+\begin{lemma}
+\label{lem:extcase3}
+For a dual-tree algorithm with $S_q \sim S_r \sim O(N)$ using cover trees and
+the traversal given in Algorithm \ref{alg:cover-tree-dual}, the number of extra
+reference recursions that happen after the last query recursion is bounded by
+
+\begin{equation}
+\theta = \max\left(\min\left(O(N \lceil \log_2(\delta_q / \delta_r) \rceil),
+O(N^2)\right), \; 0\right).
+\end{equation}
+\end{lemma}
+
+\begin{proof}
+Our goal here is to count the number of reference recursions after the final
+query recursion at level $s_q^{\min}$.  Because query nodes are not pruned in
+this traversal, each reference recursion we are counting will be duplicated over
+the whole set of $O(N)$ query nodes.  The first part of the bound follows by
+observing that $s_q^{\min} - s_r^{\min} = \lceil \log_2(\delta_q) \rceil -
+\lceil \log_2(\delta_r) \rceil \le \lceil \log_2(\delta_q / \delta_r) \rceil$.
+The second part follows by simply observing that there are $O(N)$ reference
+nodes.
+\end{proof}
+
+These two previous lemmas allow us a better understanding of what happens as the
+reference set and query set become different.  Using them, we will prove general
+runtime bounds for the cover tree traversal.
 
@@
 \begin{equation}
-O\left(c_r^4 | R^* | \chi \psi (N + \theta)\right)
+O\left(c_r^4 | R^* | \chi \psi (N + i_t(\mathscr{T}_q) + \theta)\right)
 \end{equation}
@@
 the maximum possible runtime of \texttt{BaseCase()}, $\psi$ is the maximum
-possible runtime of \texttt{Score()}, and $\theta$ is a measure of imbalance
-defined by
-
-\begin{equation}
-\theta = \max(\log_2(\eta_r / \eta_q), 0) + i_t(\mathscr{T}_q) + N
-\max(\log_2(\delta_q / \delta_r), 0)
-\end{equation}
-
-\noindent with $\eta_q$ as the maximum pairwise distance between points in
-$S_q$, $\delta_q$ as the minimum nonzero pairwise distance between points in
-$S_q$, $\eta_r$ as the maximum pairwise distance between points in $S_r$, and
-$\delta_r$ as the minimum nonzero pairwise distance between points in $S_r$.
+possible runtime of \texttt{Score()}, and $\theta$ is defined as in Lemma
+\ref{lem:extcase3}.
 \end{thm}
@@
 
-First, consider case \textit{(1)}.  If the scale of the root of the query tree
-is $s_q^T$ and the scale of the root of the reference tree is $s_r^T$, then
-$s_r^T - s_q^T$ recursions of this sort occur before the first query recursion.
+First, consider case \textit{(1)}.  Lemma \ref{lem:extcase1} shows that the
+number of reference recursions of this type is bounded by $O(N)$.  Although
+there is also a bound that depends on the sizes of the datasets, we only aim to
+show a linear runtime bound, so the $O(N)$ bound is sufficient here.
 
@@
 implies at least one reference recursion before another query recursion.  For
-some query node $\mathscr{N}_q$, the
-exact number of reference recursions before the children of $\mathscr{N}_q$ are
-recursed into is bounded above by $i_t(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has
-imbalance $0$, then it is exactly one level below its parent, and thus there is
-only one reference recursion.  On the other hand, if $\mathscr{N}_q$ is many
-levels below its parent, then it is possible that a reference recursion may
-occur for each level in between; this is a maximum of $i_t(\mathscr{N}_q) + 1$.
+some query node $\mathscr{N}_q$, the exact number of reference recursions before
+the children of $\mathscr{N}_q$ are recursed into is bounded above by
+$i_t(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has imbalance $0$, then it is
+exactly one level below its parent, and thus there is only one reference
+recursion.  On the other hand, if $\mathscr{N}_q$ is many levels below its
+parent, then it is possible that a reference recursion may occur for each level
+in between; this is a maximum of $i_t(\mathscr{N}_q) + 1$.
 
@@
 
-Lastly, case \textit{(3)} occurs when the reference scale is less than the
-minimum query scale $s_q^{\min}$; that is, all of the query recursions are done.
-Define the minimum reference scale as $s_r^{\min}$.  If $s_q^{\min} - s_r^{\min}
-> 0$, each query node may have up to $s_q^{\min} - s_r^{\min}$ reference
-recursions of type \textit{(3)}, and there are $O(N)$ query nodes.  Thus, the
-total number of reference recursions of type \textit{(3)} is bounded by $O(N)
-\max(s_q^{\min} - s_r^{\min}, 0)$.
+Lastly, for case \textit{(3)}, we may refer to Lemma \ref{lem:extcase3}, giving
+a bound of $\theta$ reference recursions in this case.
 
-Combining these three cases, we get the following expression for an upper bound
-$\theta$ on the total number of reference recursions:
+We may now combine these results for the runtime of a query recursions with the
+total number of reference recursions in order to give the result of the theorem:
 
 \begin{equation}
-\theta = \max(s_r^T - s_q^T, 0) + b(\mathscr{T}_q) + O(N) + O(N) (s_q^{\min} -
-s_r^{\min}).
-\end{equation}
-
-Now, we can express these in terms of more common constants which are related to
-the aspect ratio of the data.  Define $\delta_q$ as the minimum nonzero pairwise
-distance in $S_q$ and $\eta_q$ as the furthest pairwise distance in $S_q$ (and
-define $\delta_r$ and $\eta_r$ accordingly for the set $S_r$).  Then, $s_r^T =
-\lceil \log_2 \eta_r \rceil$, $s_q^T = \lceil \log_2 \eta_q \rceil$, $s_r^{\min}
-= \lceil \log_2 \delta_r \rceil$, and $s_q^{\min} = \lceil \log_2 \delta_q
-\rceil$.  This gives a more reasonable expression:
-
-\begin{equation}
-\theta = \max\left(\log_2\left(\frac{\eta_r}{\eta_q}\right), 0\right) +
-b(\mathscr{T}_q) + O(N) + O(N)
-\max\left(\log_2\left(\frac{\delta_q}{\delta_r}\right), 0\right).
-\end{equation}
-
-Returning to our original line of reasoning, we can combine these results to
-give the result of the theorem:
-
-\begin{equation}
-O(c_r^4 \psi \chi | R^* | \theta) + O(c_r^4 \psi | R^* | N) \sim O(c_r^4 \psi
-\chi |R^*| (N + \theta)).
+O(c_r^4 \psi \chi | R^* | (N + i_t(\mathscr{T}_q) + \theta)) + O(c_r^4 \psi | R^* | N) \sim O(c_r^4 \psi
+\chi |R^*| (N + i_t(\mathscr{T}_q + \theta)).
 \end{equation}
@@
 deserves a bit more attention.  The intuition for all terms except $\theta$
-remain virtually the same, although it is worth pointing out that there is
-no dependence on the reference set size.
+remain virtually the same.
 
 The term $\theta$ captures the effect of query and reference datasets with
-different widths, and has two corner cases: all extra reference recursions are of
-type \textit{(1)}, and all extra reference recursions are of type \textit{(3)}.
-The first case presents the best situation, because all of the reference
-recursions are done before the first query node is recursed into\footnote{In
-fact, re-derivation for this specific case can remove the dependence on
-$i_t(\mathscr{T}_q)$ entirely.  But our bound, as derived, does not bound this
-behavior tightly.  To do so would introduce a number of complex constants, and
-thus in that respect $i_t(\mathscr{T}_q)$ is a more suitable bounding quantity}.
-Therefore the bulk of the reference recursion work is done by with the query
-tree root as the query node; this is the best case.  The worst case is the
-latter, where recursion in the reference tree cannot begin until reaching the
-bottom of the query tree.  Thus, no pruning can occur until reaching the leaves
-of the query tree, and the recursion is essentially equivalent to a single-tree
-recursion for each query point plus the added overhead of the recursion of the
-query tree with one reference node.
-
-{\bf TODO: what does the second case degenerate to in terms of running time?}
+different widths, and has one unfortunate corner case: when $\delta_q > \eta_r$,
+then the query tree must be entirely descended before any reference recursion.
+This results in a bound of the form $O(N \log (\eta_r / \delta_r))$, or
+$O(N^2)$.  This is because the reference tree must be descended individually for
+each query point.
 
@@
 given as $c_q^{4 \kappa}$, which is not a good bound, especially because
-$\kappa$ may easily be very large.  Further, it is incorrectly claimed that
-$\kappa = 1$ in the monochromatic case.
+$\kappa$ may easily be very large.
 
@@
 $i_t(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree
-(assuming the tree is already built).  Thus, not only is our bound tighter, it
-more closely reflects the actual behavior of dual-tree algorithms, and the
-constants which it depends upon are straightforward to calculate.
+(assuming the tree is already built).  Thus, not only is our bound tighter when
+the cover tree imbalance is sublinear in $N$, it more closely reflects the
+actual behavior of dual-tree algorithms, and the constants which it depends upon
+are straightforward to calculate.
 

### 2014-12-24 bounds-files/imbalance.tex
Ready for review.

--- 2014-11-14 bounds-files/imbalance.tex
@@
 
-An imbalanced cover tree can---and often does---happen in practice, and the
+An imbalanced cover tree can happen in practice, and in the worst cases, the
 imbalance may be far worse than the simple graphs of Figure \ref{fig:imbalance}.
@@
 the outlier as a descendant, and the other child has the rest of the points in
-the dataset as a descendant.  In fact, it is easy to find datasets with multiple
-outliers which give rise to a chain-like structure at the top of the tree: see
-Figure \ref{fig:outliers} for an illustration\footnote{As a side note, this
-behavior is not limited to cover trees, and can happen to mean-split $kd$-trees
-too, especially in higher dimensions.}.
+the dataset as a descendant.  In fact, it is easy to find datasets with a
+handful of outliers that give rise to a chain-like structure at the top of the
+tree: see Figure \ref{fig:outliers} for an illustration\footnote{As a side note,
+this behavior is not limited to cover trees, and can happen to mean-split
+$kd$-trees too, especially in higher dimensions.}.
 
@@
 \begin{defn}
+\label{def:imbalance}
 The {\it cover tree imbalance} $i_t(\mathscr{T})$ for a cover tree $\mathscr{T}$
@@
 and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
-that the imbalance is actually quadratic in $N$!
+that in the worst case the imbalance is quadratic in $N$.
 
-As we will see later empirically, the notion of cover tree imbalance does a good
-job of capturing the `goodness' of a tree.  We will use this notion in order to
-quantify the worst-case performance of dual-tree algorithms using the cover
-tree.
+However, for most real-world datasets with the cover tree implementation in {\bf
+mlpack} \citep{mlpack2013} and the reference implementation
+\citep{langford2006}, the tree imbalance is near-linear with the number of
+points.  Most of the cover tree imbalance is contributed by leaf nodes whose
+parent has scale greater than $s_{\min}$.  At this time, no cover tree
+construction algorithm specifically aims to minimize imbalance.
 
-{\bf TODO: Can any bound be proven on $i_t(\mathscr{T})$?  I haven't thought of
-anything successfully yet.}
+%To demonstrate this, Table \ref{tab:emp_imbalance} shows empirically
+%calculated cover tree imbalance for cover trees on a number of datasets built
+%with both the cover tree implementation in {\bf mlpack} \citep{mlpack2013} and
+%the original reference implementation \citep{langford2006}.
+
+%\begin{table}[tb]
+%\begin{center}
+%\begin{tabular}{|c|c|c|r|r|}
+%\hline
+%Dataset & $N$ & $d$ & reference & mlpack \\
+%\hline
+%wine & & & & \\
+%isolet & & & & \\
+%corel & & & & \\
+%mnist & 70000 & 784 & & \\
+%bio & & & & \\
+%phy & & & & \\
+%covertype & & & & \\
+%LCDM & & & & \\
+%\hline
+%\end{tabular}
+%\end{center}
+%\caption{Empirical tree imbalance calculations.}
+%\label{tab:emp_imbalance}
+%\end{table}
+
+%{\bf TODO: Can any bound be proven on $i_t(\mathscr{T})$?  I haven't thought of
+%anything successfully yet.}

### 2014-12-24 bounds-files/sec01-supporting-lemmas.tex
Ready for review.

@@
 representation in which all nodes that have only a self-child are coalesced.
-Figure \ref{fig:implicit-explicit} shows an example cover tree and highlights
-implicit and explicit nodes.
 
@@
 \le 1$, $f(x)$ is a constant; for $|x| > 1$, $f(x) = 0$.  It is easy to see that
-$c_f$ in this situation is $2^d$.  To demonstrate this convergence, Table
-\ref{tab:ec_scaling} shows results for 10 trials of empirically calculated
-expansion constants for datasets drawn from this distribution, for varying $d$.
+$c_f$ in this situation is $2^d$. % To demonstrate this convergence, Table
+%\ref{tab:ec_scaling} shows results for 10 trials of empirically calculated
+%expansion constants for datasets drawn from this distribution, for varying $d$.
 
-\begin{table}[htb]
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|}
-\hline
-$|S|$ & $d = 2$ & $d = 5$ & $d = 10$ & $d = 20$ \\
-\hline
-10      & 6     & 10    & 10    &         \\
-31      & 12    & 31    & 31    &         \\
-100     & 13    & 64    & 100   &         \\
-316     & 23    & 96    & 286   &         \\
-1000    & 26    & 126   & 673   &         \\
-3162    & 29    & 189   & 1348  &         \\
-10000   & 30    & 259   & 2646  &         \\
-31622   & 39    & 309   & 3858  &         \\
-100000  & 40    & 360   & 7162  &         \\
-316227  & 46    & 385   & 8297  &         \\
-1000000 &       &       &       &         \\
-$\infty$ ($c_f$)   & 4     & 32    & 1024  & 1048576 \\
-\hline
-\end{tabular}
-\end{center}
-\caption{Calculated $c$ for uniform hypersphere distribution as
-$|S|$ scales.}
-\label{tab:ec_scaling}
-\end{table}
+%\begin{table}[htb]
+%\begin{center}
+%\begin{tabular}{|l|c|c|c|c|}
+%\hline
+%$|S|$ & $d = 2$ & $d = 5$ & $d = 10$ & $d = 20$ \\
+%\hline
+%10      & 6     & 10    & 10    &         \\
+%31      & 12    & 31    & 31    &         \\
+%100     & 13    & 64    & 100   &         \\
+%316     & 23    & 96    & 286   &         \\
+%1000    & 26    & 126   & 673   &         \\
+%3162    & 29    & 189   & 1348  &         \\
+%10000   & 30    & 259   & 2646  &         \\
+%31622   & 39    & 309   & 3858  &         \\
+%100000  & 40    & 360   & 7162  &         \\
+%316227  & 46    & 385   & 8297  &         \\
+%1000000 &       &       &       &         \\
+%$\infty$ ($c_f$)   & 4     & 32    & 1024  & 1048576 \\
+%\hline
+%\end{tabular}
+%\end{center}
+%\caption{Calculated $c$ for uniform hypersphere distribution as
+%$|S|$ scales.}
+%\label{tab:ec_scaling}
+%\end{table}
 
-We are somewhat limited in the sizes of $|S|$ we can scale to, because
-calculating the expansion constant of a dataset is a computationally intensive
-task.  Nonetheless, we can see that the expansion constant moves towards the
-final value of $c_f$, though potentially from above ({\it The results are not
-done yet.  This paragraph may need rewriting.  But I strongly suspect the
-results will show what I've said.}).  We postulate that the behavior of more
-complex real-world datasets is similar.
+%We are somewhat limited in the sizes of $|S|$ we can scale to, because
+%calculating the expansion constant of a dataset is a computationally intensive
+%task.  Nonetheless, we can see that the expansion constant moves towards the
+%final value of $c_f$, though potentially from above ({\it The results are not
+%done yet.  This paragraph may need rewriting.  But I strongly suspect the
+%results will show what I've said.}).  We postulate that the behavior of more
+%complex real-world datasets is similar.
 

### 2014-12-24 bounds-files/sec02-nearest-neighbor-search.tex
Refactor KDE a bit; we can write about both approximation schemes.
Ready for review.

--- 2014-11-14 bounds-files/sec02-nearest-neighbor-search.tex
@@
 given a reference set $S_r$ with expansion constant $c_r$, a query set $S_q$,
-and an inverse degree of bichromaticity $\nu$, the running time of the
-algorithm is bounded by $O(c_r^4 c_{qr}^5 \nu N)$.
+the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N +
+i_t(\mathscr{T}_q) + \theta))$ with $i_t(\mathscr{T}_q$ and $\theta$ defined as
+in Definition \ref{def:imbalance} and Lemma \ref{lem:extcase3}, respectively.
 \label{thm:nns}
@@
 Due to Theorem \ref{thm:ct-runtime}, we therefore know that the runtime of the
-algorithm is bounded by $O(c_r^4 \nu |R^*| N)$.  Thus, the only thing that
-remains is to bound the maximum size of the reference set, $|R^*|$.
+algorithm is bounded by $O(c_r^4 |R^*| (N + i_t(\mathscr{T}_q) + \theta))$.
+Thus, the only thing that remains is to bound the maximum size of the reference
+set, $|R^*|$.
 
@@
 
-In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 \nu N)$ because
-$c = c_r = c_{qr}$.
+In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 (N +
+i_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For
+well-behaved trees where $i_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
+this represents the current best worse-case runtime bound for nearest neighbor
+search.

### 2014-12-24 bounds-files/sec03-approx-kde.tex
Ready for review.

@@
 \begin{equation}
+\label{eqn:ava}
 | f(p_q) - f^*(p_q) | < N \epsilon.
@@
 \begin{equation}
+\label{eqn:rva}
 \frac{| f(p_q) - f^*(p_q) |}{| f^*(p_q) |} < N \epsilon.
@@
 
-\begin{algorithm}[tb]
-  \begin{algorithmic}[1]
-    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, reference node
-$\mathscr{N}_r$, list of node kernel estimates $\hat{f}_n$
-    \STATE {\bfseries Output:} a score for the node combination $(\mathscr{N}_q,
-\mathscr{N}_r)$, or $\infty$ if the combination should be pruned
-
-    \medskip
-
-    \IF{$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
-\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) /
-\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) < \epsilon$}
-      \STATE $f_n(\mathscr{N}_q) \gets f_n(\mathscr{N}_q) + |
-\mathscr{D}^p(\mathscr{N}_r) | \left(\mathcal{K}(d_{\min}(\mathscr{N}_q,
-\mathscr{N}_r)) + \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))\right) /\;2$
-      \RETURN $\infty$
-    \ENDIF
-
-    \RETURN $\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
-\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))$
-  \end{algorithmic}
-
-  \caption{Relative-value approximate kernel density estimation
-\texttt{Score()}}
-  \label{alg:kde_rva_score}
-\end{algorithm}
-
 \begin{thm}
@@
 %\begin{equation}
-$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + \theta))$
+$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + i_t(\mathscr{T}_q) + \theta))$
 %\end{equation}
-where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$ and
-$\theta$ as in Theorem \ref{thm:ct-runtime}.
+where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$,
+$i_t(\mathscr{T}_q)$ defined as in Definition \ref{def:imbalance}, and $\theta$
+defined as in Lemma \ref{lem:extcase3}.
 
@@
 so Theorem \ref{thm:ct-runtime} implies the total runtime of the dual-tree
-algorithm is bounded by $O(c_r^4 |R^*| (N + \theta))$.  Thus, we will bound
-$|R^*|$ using techniques related to those used by \citet{ram2009}.  The bounding
-of $|R^*|$ is split into two sections: first, show that when the scale
-$s_r^{\max}$ is small enough, $R^*$ is empty.  Second, bound $R^*$ when
-$s_r^{\max}$ is larger.
+algorithm is bounded by $O(c_r^4 |R^*| (N + i_t(\mathscr{T}_q) + \theta))$.
+Thus, we will bound $|R^*|$ using techniques related to those used by
+\citet{ram2009}.  The bounding of $|R^*|$ is split into two sections: first,
+show that when the scale $s_r^{\max}$ is small enough, $R^*$ is empty.  Second,
+bound $R^*$ when $s_r^{\max}$ is larger.
 
@@
 Approximate kernel density estimation using relative-value approximation may be
-bounded in a similar manner.  A \texttt{Score()} function is given in Algorithm
-\ref{alg:kde_rva_score}.
+bounded by reducing the absolute-value approximation algorithm (in linear time
+or less) to relative-value approximation.  This is the same strategy as
+performed by \citet{ram2009}.
+
+First, we must establish a \texttt{Score()} function for relative value
+approximation.  The difference between Equations \ref{eqn:ava} and \ref{eqn:rva}
+is the division by the term $|f^*(p_q)|$.  But we can quickly bound
+$|f^*(p_q)|$:
+
+\begin{equation}
+f^*(p_q) \ge N \mathcal{K}\left(\max_{p_r \in S_r} d(p_q, p_r)\right).
+\end{equation}
+
+This is clearly true, because each point in $S_r$ must contribute more than
+$\mathcal{K}(\max_{p_r \in S_r} d(p_q, p_r))$ to $f^*(p_q)$.  Now, we may revise
+the relative approximation condition in Equation \ref{eqn:rva}:
+
+\begin{equation}
+| f(p_q) - f^*(p_q) | \le \epsilon \mathcal{K}^{\max}
+\end{equation}
+
+\noindent where $\mathcal{K}^{\max}$ is lower bounded by $\mathcal{K}(\max_{p_r
+\in S_r} d(p_q, p_r))$.  This allows us to create a \texttt{Score()} algorithm,
+given in Algorithm \ref{alg:kde_rva_score}.
+
+\begin{algorithm}[tb]
+  \begin{algorithmic}[1]
+    \STATE {\bfseries Input:} query node $\mathscr{N}_q$, reference node
+$\mathscr{N}_r$, list of node kernel estimates $\hat{f}_n$
+    \STATE {\bfseries Output:} a score for the node combination $(\mathscr{N}_q,
+\mathscr{N}_r)$, or $\infty$ if the combination should be pruned
+
+    \medskip
+
+    \IF{$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
+\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) < \epsilon
+\mathcal{K}^{\max}$}
+\label{alg:rva-kde-prune}
+      \STATE $f_n(\mathscr{N}_q) \gets f_n(\mathscr{N}_q) + |
+\mathscr{D}^p(\mathscr{N}_r) | \left(\mathcal{K}(d_{\min}(\mathscr{N}_q,
+\mathscr{N}_r)) + \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))\right)
+/\;2$ \label{alg:rva-kde-update}
+      \RETURN $\infty$
+    \ENDIF
+
+    \RETURN $\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
+\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))$
+  \end{algorithmic}
+
+  \caption{Relative-value approximate kernel density estimation
+\texttt{Score()}}
+  \label{alg:kde_rva_score}
+\end{algorithm}
+
+Using this, we may prove linear runtime bounds for relative value approximate
+kernel density estimation.
+
+\begin{thm}
+Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
+of the previous section.  Then, given a query set $S_q$ and a reference set
+$S_r$ both of size $O(N)$, it is possible to perform relative value
+approximate kernel density estimation (satisfying the condition of Equation
+\ref{eqn:rva}) in $O(N)$ time, assuming that the expansion constant $c_r$ of
+$S_r$ is not dependent on $N$.
+\end{thm}
+
+\begin{proof}
+It is easy to see that Theorem \ref{thm:kde-bound} may be adapted to the very
+slightly different \texttt{Score()} rule of Algorithm \ref{alg:kde_rva_score}
+while still providing an $O(N)$ bound.  With that \texttt{Score()} function, the
+dual-tree algorithm will return relative-value approximate kernel density
+estimates satisfying Equation \ref{eqn:rva}.
+
+We now turn to the calculation of $\mathcal{K}^{\max}$.  Given the cover trees
+$\mathscr{T}_q$ and $\mathscr{T}_r$ with root nodes $\mathscr{N}_{r}^{R}$ and
+$\mathscr{N}_{r}^{R}$, respectively, we may calculate a suitable
+$\mathcal{K}^{\max}$ value in constant time:
+
+\begin{equation}
+\mathcal{K}^{\max} = d_{\max}(\mathscr{N}_q^R, \mathscr{N}_r^R) = d(p_q^R,
+p_r^R) + 2^{s_q^{\max} + 1} + 2^{s_r^{\max} + 1}.
+\end{equation}
+
+This proves the statement of the theorem.
+\end{proof}
+
+In this case, we have not shown tighter bounds because the algorithm we have
+proposed is not useful in practice.  For an example of a better relative-value
+approximate kernel density estimation dual-tree algorithm, see the work of
+\citet{gray2003nonparametric}.

### 2014-12-24 bounds-files/sec04-range-search.tex
Propagate changed Lemma 1 to the document.
Ready for review.

@@
 \begin{equation}
-O\left(c_r^{4} \max\left(c_r^{4 + \beta}, |S_{\max}| + C\right) (N + \theta)
+O\left(c_r^{4} \max\left(c_r^{4 + \beta}, |S_{\max}| + C\right) (N +
+i_t(\mathscr{N}_q) + \theta)
 \right)
@@
 
-with $\theta$ defined as in Theorem
-\ref{thm:ct-runtime},
-$\beta = \log_2 (1 + \alpha^{-1})$, and $S_{\max}$ as defined in Equation
-\ref{eqn:smax}.
+with $\theta$ defined as in Lemma \ref{lem:extcase3},
+$\beta = \lceil \log_2 (1 + \alpha^{-1}) \rceil$, and $S_{\max}$ as defined in
+Equation \ref{eqn:smax}.
 %\end{equation}
@@
 \ref{thm:ct-runtime} we know that the runtime of the algorithm is bounded by
-$O(c_r^4 |R^*| (N + \theta))$.  As with the previous proofs, then, our only task
-is to bound the maximum size of the reference set, $|R^*|$.
+$O(c_r^4 |R^*| (N + i_t(\mathscr{N}_q) + \theta))$.  As with the previous
+proofs, then, our only task is to bound the maximum size of the reference set,
+$|R^*|$.
 
@@
 \begin{equation}
-|R^*| \le c_r^{4 + \log_2(1 + \alpha^{-1})}.
+|R^*| \le c_r^{4 + \lceil \log_2(1 + \alpha^{-1}) \rceil}.
 \end{equation}
@@
 \begin{equation}
-O(c_r^{8 + \beta} (N + \theta)).
+O(c_r^{8 + \beta} (N + i_t(\mathscr{N}_q) + \theta)).
 \end{equation}
@@
 $\alpha$.  Consider $\alpha = (1 / 3)$; this yields a running time of $O(c^8 (N
-+ \theta))$.  $\alpha = (1 / 7)$ yields $O(c^9 (N + \theta))$, $\alpha = (1 /
-15)$ yields $O(c^{10} (N + \theta))$, and so forth.  As $\alpha$ gets smaller,
-the exponent on $c$ gets larger, and diverges as $\alpha \to 0$.
++ \theta))$.  $\alpha = (1 / 7)$ yields $O(c^9 (N + i_t(\mathscr{N}_q +
+\theta))$, $\alpha = (1 / 15)$ yields $O(c^{10} (N + i_t(\mathscr{N}_q) +
+\theta))$, and so forth.  As $\alpha$ gets smaller, the exponent on $c$ gets
+larger, and diverges as $\alpha \to 0$.
 

### 2014-12-29 bounds-files/alternative_imbalance.tex
updated abstract and intro, typo fixes, added alternative imbalance notes

The purpose of this section is to explore an alternative to the current 
definition of the tree imbalance. I'll start with some intuition, then provide
an alternative definition. Finally, I'll try to re-write the proof of Theorem
1 using this definition.

I started by wondering if the tree imbalance Equation~\ref{eqn_node_imbalance}
can be bounded in terms of the expansion constant. I tried to construct 
examples where the expansion constant is small, but the tree imbalance is 
large.

The only one I could come up with is this: construct a set of $N/2$ points with
a small expansion constant (i.e. points on a lattice in low dimensions). Make a 
copy of this set, then place it arbitrarily far away from the first set.  The 
expansion constant of the combined set is no larger than for one of the small
sets. However, the gap in level between the root and its children is as 
large as we want.

The problem is that a dual-tree algorithm on this set will generally still be 
efficient. The key is that all of the levels between the root and second level 
of the tree don't matter. The algorithm can skip over all of them in one 
recursion.

So, the idea is to see if we can bound the runtime in terms of the number of 
missing levels \emph{which actually exist} somewhere in the tree. In the 
bichromatic case, we will need to take into account the levels of the other
tree.

Let $T_q$ and $T_r$ be a query and reference tree. Let $s_1$ and $s_2$ be two 
scales such that $s_1 > s_2$. 
Define $\mathcal{M}(s_1, s_2, T_q, T_r)$ to be the number of levels between 
$s_1$ and $s_2$ (inclusive) which one (or both) trees have a node at.

We provide the following definition of the imbalance of a node: 
\begin{definition}
  Let $T_r$ and $T_q$ be two cover trees. 
  Let $N$ be a node in either tree at scale $s$ with parent at scale $s_p$. 
  Then, define the imbalance of the node to be
  \begin{equation}
    \mathcal{I}(N) = \mathcal{M}(s_p, s, T_r, T_q)
  \end{equation}
  The imbalance of the root is defined to be $0$ and the level of a leaf is 
  defined to be $s_{\textrm{min}} - 1$ where $s_{\textrm{min}}$ is the minimum
  level of any non-leaf in either tree.

  The imbalance of a tree is the sum of the imbalances of all nodes.
\end{definition}

Note that the imbalance of a tree is $\bigO(N)$, since in the worst case 
each node has a different level from all of the others.

Is this really true, though? If we're counting for query nodes, then we might
count each reference node many times. This is where my thinking starts to break 
down. . .

### 2014-12-29 bounds-files/general-runtime-bound.tex
updated abstract and intro, typo fixes, added alternative imbalance notes

@@
 Next, we develop notions with which to understand the behavior of the cover tree
-dual-tree traversla when the datasets are of significantly different scale
+dual-tree traversal when the datasets are of significantly different scale
 distributions.
@@
 query recursions and reference recursions.  But if the query set contains points
-which are, in general, much further apart than the reference set, then the
+which are, in general, much farther apart than the reference set, then the
 recursion will start with many query recursions before reaching a reference
@@
 \begin{equation}
-\theta = \max\left(\min\left(O(N \lceil \log_2(\delta_q / \delta_r) \rceil),
-O(N^2)\right), \; 0\right).
+\theta = \max\left\{\min\left[O(N \lceil \log_2(\delta_q / \delta_r) \rceil),
+O(N^2)\right], \; 0\right\}.
 \end{equation}

### 2014-12-29 bounds.tex
updated abstract and intro, typo fixes, added alternative imbalance notes

@@
 %range count/search.
+%Numerous machine learning algorithms contain pairwise statistical problems at
+%their core. Often, tree structures are used to solve these problems efficiently.
+%Dual-tree algorithms using cover trees have rigorous worst-case runtime
+%guarantees for some of these problems. In this paper, we present a {\em
+%problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
+%the cover tree, separating out the problem-dependent and the problem-independent
+%elements. This allows us to just plug in bounds for the problem-dependent
+%elements to get runtime guarantees for dual-tree algorithms for any pairwise
+%statistical problem. We demonstrate this plug-and-play procedure for
+%nearest-neighbor search and approximate kernel density estimation to get
+%improved runtime guarantees.  Under some assumptions, we also present the first
+%linear runtime guarantee for dual-tree based range search.
 Numerous machine learning algorithms contain pairwise statistical problems at
-their core. Often, tree structures are used to solve these problems efficiently.
-Dual-tree algorithms using cover trees have rigorous worst-case runtime
-guarantees for some of these problems. In this paper, we present a {\em
+their core -- i.e. tasks that require computations over all pairs of input points if implemented directly. 
+Often, tree structures are used to solve these problems efficiently. Dual-tree algorithms 
+can efficiently solve or approximate many of these problems.
+Using cover trees, rigorous worst-case runtime
+guarantees have been proven for some of these algorithms. In this paper, we present a {\em
 problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
 the cover tree, separating out the problem-dependent and the problem-independent
-elements. This allows us to just plug in bounds for the problem-dependent
+elements. This allows us to plug in bounds for the problem-dependent
 elements to get runtime guarantees for dual-tree algorithms for any pairwise
-statistical problem. We demonstrate this plug-and-play procedure for
+statistical problem without re-deriving the entire proof. We demonstrate this plug-and-play procedure for
 nearest-neighbor search and approximate kernel density estimation to get
-improved runtime guarantees.  Under some assumptions, we also present the first
+improved runtime guarantees.  Under mild assumptions, we also present the first
 linear runtime guarantee for dual-tree based range search.
@@
 bottlenecks that can be expressed as pairwise statistical problems. 
-Nearest neighbor search is one such problem, naively requiring $O(N)$ time
+By this, we mean computational tasks that can be evaluated directly by iterating over all 
+pairs of input points. 
+Nearest neighbor search is one such problem, since for every query point, we can 
+evaluate its distance to every reference point and keep the closest one. 
+This naively requires $O(N)$ time
 to answer in single query in a reference set of size $N$; answering $O(N)$
 queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
-estimation is also a pairwise statistical problem, naively requiring $O(N^2)$
-time to answer $O(N)$ queries. The reference set is typically indexed with
+estimation is also a pairwise statistical problem, since we compute a sum over all 
+reference points for each query point.
+This again requires $O(N^2)$
+time to answer $O(N)$ queries if done directly. 
+The reference set is typically indexed with
 spatial data structures to accelerate this type of computation
@@
 significant similarity to the more common single-tree branch-and-bound
-algorithms, except that the algroithm must recurse into child nodes of {\it
+algorithms, except that the algorithm must recurse into child nodes of {\it
 both} the query tree and reference tree.
@@
 \citep[especially][]{curtin2013tree} may find that section to be review.
-Following that, we introduce an intuitive measure of cover tree imbalance in
+Following that, we introduce an intuitive measure of cover tree imbalance, an 
+important property for understanding the runtime of dual-tree algorithms, in
 Section \ref{sec:imbalance}.  This measure of imbalance is then used to prove

### 2015-01-08 bounds-files/general-runtime-bound.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

@@
 are actually used in dual-tree algorithms.  Although cover trees were originally
-intended for nearest neighbor search \citep*[see Algorithm
+intended for nearest neighbor search \citep[See Algorithm
 \texttt{Find-All-Nearest},][]{langford2006}, they can be adapted to a wide
@@
 search \citep{ram2009rank}, Gaussian processes posterior calculation
-\citep{moore2014fast}, max-kernel search \citep{curtin2014dual} are some
+\citep{moore2014fast}, and max-kernel search \citep{curtin2014dual} are some
 examples.  Further, through the tree-independent dual-tree algorithm abstraction
@@
 \label{alg:line:ct-dual-srmax}
-    \IF{$(s_q < s^{\max}_r)$ {\bf or} $(s^{\max}_r = -\infty)$}
+    \IF{$(s_q < s^{\max}_r)$}
 \label{alg:line:ct-dual-ref-recursion-start}
@@
 both the cover tree reference implementation and in the more flexible {\bf
-mlpack} library \citep{mlpack2013}.  The algorithm is given in Algorithm
-\ref{alg:cover-tree-dual} and was originally presented by
+mlpack} library \citep{mlpack2013}.  The problem-independent traversal is given
+in Algorithm \ref{alg:cover-tree-dual} and was originally presented by
 \citet{curtin2014dual}.  Initially, it is called with the root of the query tree
-and a reference set containing only the root of the reference tree.
+and a reference set $R$ containing only the root of the reference tree.
 
@@
 determines whether the query or reference tree will be recursed, is aimed at
-keeping the relative scales of query nodes and reference nodes equal.
+keeping the relative scales of query nodes and reference nodes close.
 
@@
 $\mathscr{N}_r$ in the reference set $R$.  If possible, these combinations are
-pruned to form the set $R'$ (line \ref{alg:line:ct-dual-query-recursion}), and
-then the algorithm recurses with $\mathscr{N}_{qc}$ and $R'$.
+pruned to form the set $R'$ (line \ref{alg:line:ct-dual-query-recursion}) by
+checking the output of the \texttt{Score()} function, and then the algorithm
+recurses with $\mathscr{N}_{qc}$ and $R'$.
 
@@
 is similar to a query recursion, but the pruning strategy is significantly more
-complicated.  We score each node combination $(\mathscr{N}_q, \mathscr{N}_r)$
-for all $\mathscr{N}_r$ in $R_{r - 1}$, which is the set of children of all
-nodes in $R$ that have scale $s_r^{\max}$ (that set is $R_r$).  The set of
-reference nodes in $R_{r - 1}$ that were not pruned create the set $R'_{r - 1}$.
-Then, this set is combined with $R \setminus R_r$---that is, each of the nodes
-in $R$ that was {\em not} recursed into---to produce $R'$, and the algorithm
-recurses with $\mathscr{N}_q$ and the reference set $R'$.
+complicated.  Given $R$, we calculate $R_r$, which is the set of nodes in $R$
+that have scale $s_r^{\max}$.  Then, for each node $\mathscr{N}_r$ in the set of
+children of nodes in $R_r$, the node combinations $(\mathscr{N}_q,
+\mathscr{N}_r)$ are scored and pruned if possible.  Those reference nodes that
+were not pruned form the set $R'_{r - 1}$.  Then, this set is combined with $R
+\setminus R_r$---that is, each of the nodes in $R$ that was {\em not} recursed
+into---to produce $R'$, and the algorithm recurses with $\mathscr{N}_q$ and the
+reference set $R'$.
 
@@
 $R_{r - 1}$ as in line \ref{alg:line:ct-dual-ref-children} of Algorithm
-\ref{alg:cover-tree-dual}.
+\ref{alg:cover-tree-dual}; if $| R_{r - 1} | \le 1$, then $R_{r - 1}$ satisfies
+the desired property.
 
-Now, take any $\mathscr{N}_i, \mathscr{N}_j$ in $R_{r - 1}$, with points $p_i$
-and $p_j$, respectively, and scales $s_i$ and $s_j$, respectively.  Clearly, if
-$s_i = s_j = s_r^{\max} - 1$, then by the separation invariant $d(p_i, p_j) >
-2^{s_r^{\max} - 1}$.
+Otherwise, take any $\mathscr{N}_i, \mathscr{N}_j$ in $R_{r - 1}$, with points
+$p_i$ and $p_j$, respectively, and scales $s_i$ and $s_j$, respectively.
+Clearly, if $s_i = s_j = s_r^{\max} - 1$, then by the separation invariant
+$d(p_i, p_j) > 2^{s_r^{\max} - 1}$.
 
@@
 separated by $2^{s_r^{\max} - 1}$.  Note that $R'_{r - 1} \subseteq R_{r - 1}$
-and that $R \setminus R_{r - 1} \subseteq R$, and that each point of each node
-is separated by $2^{s_r^{\max} - 1}$, to see that this condition holds for all
-nodes in $R' = R'_{r - 1} \cup (R \setminus R_{r - 1})$.
+and that $R \setminus R_{r - 1} \subseteq R$ in order to see that this condition
+holds for all nodes in $R' = R'_{r - 1} \cup (R \setminus R_{r - 1})$.
 
@@
 indeed, $\eta_q / \delta_q$ is exactly the aspect ratio of $S_q$.  Further,
-let us define the top and bottom levels of each tree:
+let us define and bound the top and bottom levels of each tree:
 
 \begin{itemize}
-  \item The {\it top scale} of the query tree $\mathscr{T}_q$ is defined as
-$s_q^T = \lceil \log_2(\eta_q) \rceil$.
+  \item The {\it top scale} $s_q^T$ of the query tree $\mathscr{T}_q$ is such
+that as $\lceil \log_2(\eta_q) \rceil - 1 \le s_q^T \le \lceil \log_2(\eta_q)
+\rceil$.
   \item The {\it minimum scale} of the query tree $\mathscr{T}_r$ is defined as
 $s_q^{\min} = \lceil \log_2(\delta_q) \rceil$.
-  \item The top scale of the reference tree $\mathscr{T}_r$ is defined as $s_r^T
-= \lceil \log_2(\eta_r) \rceil$.
+  \item The top scale $s_r^T$ of the reference tree $\mathscr{T}_r$ is such that
+as $\lceil \log_2(\eta_r) \rceil - 1 \le s_r^T \le \lceil \log_2(\eta_r)
+\rceil$.
   \item The minimum scale of the reference tree $\mathscr{T}_r$ is defined as
@@
 \begin{equation}
-\min\left(O(N), \lceil \log_2(\eta_r / \eta_q) \rceil\right).
+\min\left(O(N), \log_2(\eta_r / \eta_q) - 1\right).
 \end{equation}
@@
 \begin{proof}
-The first query recursion happens once $s_q > s_r^{\max}$.  The number of
+The first query recursion happens once $s_q \ge s_r^{\max}$.  The number of
 reference recursions before the first query recursion is then bounded as the
@@
 The second bound holds by applying the definitions of $s_r^T$ and $s_q^T$ to the
-expression $s_r^T - s_q^T$.
+expression $s_r^T - s_q^T - 1$:
+
+\begin{eqnarray}
+s_r^T - s_q^T - 1 &\le& \lceil \log_2(\eta_r) \rceil - (\lceil \log_2(\eta_q)
+\rceil - 1) - 1 \\
+&\le& \log_2(\eta_r) + 1 - \log_2(\eta_q)
+\end{eqnarray}
+
+\noindent which gives the statement of the lemma after applying logarithmic
+identities.
 \end{proof}
@@
 \begin{equation}
-\theta = \max\left\{\min\left[O(N \lceil \log_2(\delta_q / \delta_r) \rceil),
+\theta = \max\left\{\min\left[O(N \log_2(\delta_q / \delta_r)),
 O(N^2)\right], \; 0\right\}.
@@
 Our goal here is to count the number of reference recursions after the final
-query recursion at level $s_q^{\min}$.  Because query nodes are not pruned in
+query recursion at level $s_q^{\min}$; the first of these reference recursions
+is at scale $s_r^{\max} = s_q^{\min}$.  Because query nodes are not pruned in
 this traversal, each reference recursion we are counting will be duplicated over
 the whole set of $O(N)$ query nodes.  The first part of the bound follows by
-observing that $s_q^{\min} - s_r^{\min} = \lceil \log_2(\delta_q) \rceil -
-\lceil \log_2(\delta_r) \rceil \le \lceil \log_2(\delta_q / \delta_r) \rceil$.
+observing that
+$s_q^{\min} - s_r^{\min} \le \lceil \log_2(\delta_q) \rceil -
+\lceil \log_2(\delta_r) \rceil - 1 \le \log_2(\delta_q / \delta_r)$.
+
 The second part follows by simply observing that there are $O(N)$ reference
@@
 These two previous lemmas allow us a better understanding of what happens as the
-reference set and query set become different.  Using them, we will prove general
-runtime bounds for the cover tree traversal.
+reference set and query set become different.  Lemma \ref{lem:extcase1} shows
+that the number of extra recursions caused by a reference set with larger
+pairwise distances than the query set ($\eta_r$ larger than $\eta_q$) is modest;
+on the other hand, Lemma \ref{lem:extcase3} shows that for each extra level in
+the reference tree below $s_q^{\min}$, $O(N)$ extra recursions are required.
+Using these lemmas and this intuition, we will prove general runtime bounds for
+the cover tree traversal.
 
@@
 \ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end}).
-The work done in the base case loop from lines
+Define $R^*$ to be the largest set $R$ for any scale
+$s_r^{\max}$ and any query node $\mathscr{N}_q$ during the course of the
+algorithm; then, it is true that $| R | \le | R^* |$.  The work done in the base
+case loop from lines
 \ref{alg:line:ct-dual-base-case-start}--\ref{alg:line:ct-dual-base-case-end} is
-$O(\chi | R |)$.  Define $| R^* |$ to be the largest set $|R|$ for any scale
-$s_r^{\max}$ and any query node $\mathscr{N}_q$ during the course of the
-algorithm; then, it is true that $| R | \le | R^* |$.
+thus $O(\chi | R |) \le O(\chi | R^*|)$.
 %
@@
 \ref{alg:line:ct-dual-ref-score} take $O(c_r^4 \psi | R |) \le O(c_r^4 \psi |
-R^* |)$ time, because each reference node has $c_r^4$ children.  So, one
+R^* |)$ time, because each reference node has up to $c_r^4$ children.  So, one
 full reference recursion takes $O(c_r^4 \psi \chi | R^* |)$ time.
@@
 work in line \ref{alg:line:ct-dual-query-pruning}, like in the reference
-recursion, is bounded as $O(c_r^4 \psi | R^* |)$. Therefore, the total
-runtime of all query recursions is $O(c_r^4 \psi | R^* | N)$.
+recursion, is bounded as $O(c_r^4 \psi | R^* |)$. Therefore, the total runtime
+of all query recursions is $O(c_r^4 \psi | R^* | N)$.
 
@@
 the children of $\mathscr{N}_q$ are recursed into is bounded above by
-$i_t(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has imbalance $0$, then it is
+$i_n(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has imbalance $0$, then it is
 exactly one level below its parent, and thus there is only one reference
@@
 parent, then it is possible that a reference recursion may occur for each level
-in between; this is a maximum of $i_t(\mathscr{N}_q) + 1$.
+in between; this is a maximum of $i_n(\mathscr{N}_q) + 1$.
 
@@
 \begin{equation}
-\sum_{\mathscr{N}_q \in \mathscr{T}_q} i_t(\mathscr{N}_q) + 1 = b(\mathscr{T}_q) +
+\sum_{\mathscr{N}_q \in \mathscr{T}_q} i_n(\mathscr{N}_q) + 1 = i_t(\mathscr{T}_q) +
 O(N)
@@
 \begin{equation}
-O(c_r^4 \psi \chi | R^* | (N + i_t(\mathscr{T}_q) + \theta)) + O(c_r^4 \psi | R^* | N) \sim O(c_r^4 \psi
-\chi |R^*| (N + i_t(\mathscr{T}_q + \theta)).
+O\left(c_r^4 |R^*| \psi \chi \left(N + i_t(\mathscr{T}_q) + \theta\right)\right)
++ O\left(c_r^4 |R^*| \psi N\right) \sim O\left(c_r^4 |R^*| \psi
+\chi \left(N + i_t(\mathscr{T}_q) + \theta\right)\right).
 \end{equation}
@@
 that $c_q = c_r = c$ and $\mathscr{T}_q = \mathscr{T}_r = \mathscr{T}$, a
-standard cover tree based dual-tree algorithm (Algorithm
+dual-tree algorithm using the standard cover tree traversal (Algorithm
 \ref{alg:cover-tree-dual}) takes
@@
 \begin{equation}
-O\left(c^4 |R^*| \chi \psi (N + i_t(\mathscr{T}))\right)
+O\left(c^4 |R^*| \chi \psi \left(N + i_t(\mathscr{T})\right)\right)
 \end{equation}
@@
 also have a high cover tree imbalance $i_t(\mathscr{T})$; the linear dependence of
-runtime on imbalance is thus sensible.
+runtime on imbalance is thus sensible for well-behaved datasets.
 
@@
 This results in a bound of the form $O(N \log (\eta_r / \delta_r))$, or
-$O(N^2)$.  This is because the reference tree must be descended individually for
-each query point.
+$O(N^2)$ (See Lemma \ref{lem:extcase3}).  This is because the reference tree
+must be descended individually for each query point.
 
@@
 small value of $|R^*|$ will imply that the algorithm is extremely successful in
-pruning away work at Line \ref{alg:line:ct-dual-ref-score} in Algorithm
-\ref{alg:cover-tree-dual}. An (upper) bound on $|R^*|$ (and the algorithm's
+pruning away work.  An (upper) bound on $|R^*|$ (and the algorithm's
 success in pruning work) will depend on the problem and the data.  As we will
@@
 problem-independent elements of the runtime bound, which allows us to simply
-plug in the problem-dependent bounds to get runtime bounds for any dual-tree
-algorithm without requiring an analysis from scratch.
+plug in the problem-dependent bounds in order to get runtime bounds for any
+dual-tree algorithm without requiring an analysis from scratch.
 
@@
 given as $c_q^{4 \kappa}$, which is not a good bound, especially because
-$\kappa$ may easily be very large.
+$\kappa$ may be much greater than $1$ in the bichromatic case (where $S_q \ne
+S_r$).
 

### 2015-01-08 bounds-files/imbalance.tex
updated abstract and intro, typo fixes, added alternative imbalance notes
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

@@
 It is well-known that imbalance in trees leads to degradation in performance;
-for instance, a $kd$-tree node with every descendant except one in its left
-child is effectively useless.  A $kd$-tree full of nodes like this will perform
+for instance, a $kd$-tree node with every descendant in its left child except
+one is effectively useless.  A $kd$-tree full of nodes like this will perform
 abysmally for nearest neighbor search, and it is not hard to generate a
@@
 other points\footnote{Note also that for an outlier sufficiently far away, the
-expansion constant is $N - 1$.}.  Figure \ref{fig:outlier} shows what happens in
+expansion constant is $N - 1$, so we should expect poor performance with the
+cover tree anyway.}.  Figure \ref{fig:outlier} shows what happens in
 this situation: the root node has two children; one of these children has only
@@
 
-A tree that has this chain-like structure all the way down is going to perform
+A tree that has this chain-like structure all the way down, which is similar to
+the $kd$-tree example at the beginning of this section, is going to perform
 horrendously; motivated by this observation, we define a measure of tree
@@
   \end{dcases*}
+  \label{eqn_node_imbalance}
 \end{equation}
@@
 and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
-that in the worst case the imbalance is quadratic in $N$.
+that in the worst case the imbalance is quadratic in $N$\footnote{Note that in
+this situation, $c \sim N$ also.}.
 
@@
 \citep{langford2006}, the tree imbalance is near-linear with the number of
-points.  Most of the cover tree imbalance is contributed by leaf nodes whose
-parent has scale greater than $s_{\min}$.  At this time, no cover tree
-construction algorithm specifically aims to minimize imbalance.
+points.  Generally, most of the cover tree imbalance is contributed by leaf
+nodes whose parent has scale greater than $s_{\min}$.  At this time, no cover
+tree construction algorithm specifically aims to minimize imbalance.
 

### 2015-01-08 bounds-files/sec01-supporting-lemmas.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

@@
 of traversal we will consider is the cover tree pruning dual-tree traversal,
-which we will describe below.
+which we will describe later.
 
@@
 
-A consequence of this definition is if there exists a node $\mathscr{N}_i$,
+As a consequence of this definition, if there exists a node $\mathscr{N}_i$,
 containing the point $p_i$ at some scale $s_i$, then there will also exist a
@@
 $C_{\infty}$ containing only the root point, $C_{-\infty} = S$, and all levels
-between defined according to the definitions above.  \citep{langford2006} find
+between defined as above.  \citet{langford2006} find
 this representation (which they call the {\it implicit} representation) easier
-for description of their algorithms and some of their proofs.  But naturally,
+for description of their algorithms and some of their proofs.  But clearly,
 this is not suitable for implementation; hence, there is an {\it explicit}
-representation in which all nodes that have only a self-child are coalesced.
+representation in which all nodes that have only a self-child are coalesced
+upwards (that is, the node's self-child is removed, and the children of that
+self-child are taken to be the children of the node).  Figure
+\ref{fig:implicit-explicit} shows an example cover tree and highlights implicit
+and explicit nodes.
 
-In our work here, we consider only the explicit representation of a cover tree,
-and do not concern ourselves with the construction of the tree\footnote{A batch
+In this work, we consider only the explicit representation of a cover tree, and
+do not concern ourselves with the details of tree construction\footnote{A batch
 construction algorithm is given by \citet{langford2006}, called
@@
 
-Take $f(x)$ as a uniform spherical distribution in $\mathcal{R}^d$: for any $|x|
-\le 1$, $f(x)$ is a constant; for $|x| > 1$, $f(x) = 0$.  It is easy to see that
-$c_f$ in this situation is $2^d$. % To demonstrate this convergence, Table
+As a simple example, take $f(x)$ as a uniform spherical distribution in
+$\mathcal{R}^d$: for any $|x| \le 1$, $f(x)$ is a constant; for $|x| > 1$, $f(x)
+= 0$.  It is easy to see that $c_f$ in this situation is $2^d$, and thus for
+some dataset $S$, $c$ must converge to that value as more and more points are
+added to $S$.  Closed-form solutions for $c_f$ for more complex distributions
+are less easy to derive; however, empirical speedup results from
+\citet{langford2006} suggest the existence of datasets where $c$ is not strongly
+dependent on $d$.  For instance, the \texttt{covtype} dataset has 54 dimensions
+but the expansion constant is much smaller than other, lower-dimensional
+datasets.
+% To demonstrate this convergence, Table
 %\ref{tab:ec_scaling} shows results for 10 trials of empirically calculated
@@
 decrease $c$ significantly.  Suppose one adds a point arbitrarily close to the
-origin to $S'$; now, the expansion constant will be $|S'| / 2$.  So, although we
-can bound the behavior of $c$ as $|S| \to \infty$ for $S$ from a stationary
-distribution, we are not able to easily say much about its convergence behavior.
+origin to $S'$; now, the expansion constant will be $|S'| / 2$.  Both of these
+situations are degenerate cases not commonly encountered in real-world behavior;
+we discuss them in order to point out that although we can bound the behavior of
+$c$ as $|S| \to \infty$ for $S$ from a stationary distribution, we are not able
+to easily say much about its convergence behavior.
 
@@
 such that every point in $C$ is separated by $\delta$.  Then, for any point
-$p \not\in S$ and any radius $\rho \delta > 0$:
+$p$ (which may or may not be in $S$), and any radius $\rho \delta > 0$:
 \begin{equation}

### 2015-01-08 bounds-files/sec02-nearest-neighbor-search.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

@@
 and \texttt{Score()} function; these are given in Algorithms
-\ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively.
+\ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively.  The
+point-to-point \texttt{BaseCase()} function compares a query point $p_q$ and a
+reference point $p_r$, updating the list of candidate neighbors for $p_q$ if
+necessary.
+
+The node-to-node \texttt{Score()} function determines if the entire
+subtree of nodes under the reference node $\mathscr{N}_r$ can improve the
+candidate neighbors for all descendant points of the query node $\mathscr{N}_q$;
+if not, the node combination is pruned.  The \texttt{Score()} function depends
+on the function $d_{\min}(\cdot, \cdot)$, which represents the minimum possible
+distance between any two descendants of two nodes.  Its definition for cover
+tree nodes is
+
+\begin{equation}
+d_{\min}(\mathscr{N}_q, \mathscr{N}_r) = d(p_q, p_r) - 2^{s_q + 1} - 2^{s_r +
+1}.
+\end{equation}
 
@@
 nearest neighbor candidates in the array $N$ and their distances in the array
-$D$. \citep[see][for a more complete
+$D$. \citep[See][for a more complete
 discussion of how this algorithm works and a proof of
@@
 candidate for any descendant point of the query node $\mathscr{N}_q$.  The
-standard bound function $B(\mathscr{N}_q)$ used for cover trees is
-%\begin{equation}
-$B(\mathscr{N}_q) \le D[p_q] + 2^{s_q + 1}$ \citep[this is adapted from][]{langford2006}.
-%\end{equation}
+standard bound function $B(\mathscr{N}_q)$ used for cover trees is adapted from
+\citep{langford2006}:
+
+\begin{equation}
+B(\mathscr{N}_q) := D[p_q] + 2^{s_q + 1}
+\end{equation}
 
@@
 Algorithms \ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively, and also
-given a reference set $S_r$ with expansion constant $c_r$, a query set $S_q$,
+given a reference set $S_r$ with expansion constant $c_r$, and a query set $S_q$,
 the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N +
@@
 current nearest neighbor candidate for $p_q$; that is, $D[p_q] = d(p_q,
-\hat{p}_r)$.  Then,
+\hat{p}_r)$ and $N[p_q] = \hat{p}_r$.  Then,
 
@@
 %d(p_q, p_r) - 2^{s_q + 1} - 2^{s_r + 1} &\le& D[p_q] + 2^{s_q + 1} \\
+d_{\min}(\mathscr{N}_q, \mathscr{N}_r) &\le& B(\mathscr{N}_q) \\
 d(p_q, p_r) &\le& d(p_q, \hat{p}_r) + 2^{s_q + 1} + 2^{s_r + 1} + 2^{s_q + 1}
@@
 \begin{equation}
-P \subseteq B_{S_r}(p_q, d(p_q, \hat{p}_r) + 2(2^{s_r^{\max} + 1}))
+P \subseteq B_{S_r}(p_q, d(p_q, \hat{p}_r) + 2(2^{s_r^{\max} + 1})).
 \end{equation}
@@
 % &\le& (c'_r)^3 | B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2) | \\
- &\le& c_{qr}^3 | B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2 |
+ &\le& c_{qr}^3 | B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2) |
 \end{eqnarray}
@@
 well-behaved trees where $i_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
-this represents the current best worse-case runtime bound for nearest neighbor
-search.
+this represents the current tightest worst-case runtime bound for nearest
+neighbor search.

### 2015-01-08 bounds-files/sec03-approx-kde.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

@@
 In the case of an infinite-tailed kernel $\mathcal{K}(\cdot, \cdot)$, the
-computation cannot be accelerated; thus, attention has turned towards tractable
-approximation schemes.  Two simple schemes for the approximation of $f^*(p_q)$
-are well-known: {\it absolute value approximation} and {\it relative value
-approximation}.  Absolute value approximation requires that each density
+exact computation cannot be accelerated; thus, attention has turned towards
+tractable approximation schemes.  Two simple schemes for the approximation of
+$f^*(p_q)$ are well-known: {\it absolute value approximation} and {\it relative
+value approximation}.  Absolute value approximation requires that each density
 estimate $f(p_q)$ is within $\epsilon$ of the true estimate $f^*(p_q)$:
@@
 \label{eqn:ava}
-| f(p_q) - f^*(p_q) | < N \epsilon.
+| f(p_q) - f^*(p_q) | < \epsilon \; \; \forall p_q \in S_q.
 \end{equation}
@@
 \label{eqn:rva}
-\frac{| f(p_q) - f^*(p_q) |}{| f^*(p_q) |} < N \epsilon.
+\frac{| f(p_q) - f^*(p_q) |}{| f^*(p_q) |} < \epsilon \; \; \forall p_q \in S_q.
 \end{equation}
@@
 A tree-independent algorithm for solving approximate kernel density estimation
-with absolute value approximation under the previous assumptions on th kernel
+with absolute value approximation under the previous assumptions on the kernel
 is given as a \texttt{BaseCase()} function in Algorithm \ref{alg:kde_base_case}
@@
 have $p_q$ as a descendant.  %It is possible to implement this in a way where
-Each query node needs to be visited only once.
+Each query node needs to be visited only once to perform this calculation; it
+may therefore be accomplished in $O(N)$ time.
 
@@
 Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
-of the previous section.  Then, given a query set $S_q$ and a reference set
+of the previous subsection.  Then, given a query set $S_q$ and a reference set
 $S_r$ with expansion constant $c_r$, and using the approximate kernel density
@@
 traversal given in Algorithm \ref{alg:cover-tree-dual}, the running time of
-approximate kernel density estimation for some $\epsilon$ error parameter is
+approximate kernel density estimation for some error parameter $\epsilon$ is
 bounded by
@@
 %\end{equation}
-where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$,
+with $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$,
 $i_t(\mathscr{T}_q)$ defined as in Definition \ref{def:imbalance}, and $\theta$
@@
 \citet{ram2009}.  The bounding of $|R^*|$ is split into two sections: first,
-show that when the scale $s_r^{\max}$ is small enough, $R^*$ is empty.  Second,
-bound $R^*$ when $s_r^{\max}$ is larger.
+we show that when the scale $s_r^{\max}$ is small enough, $R^*$ is empty.
+Second, we bound $R^*$ when $s_r^{\max}$ is larger.
 
-The \texttt{Score()} function is such that any node in $|R^*|$ for a given query
+The \texttt{Score()} function is such that any node in $R^*$ for a given query
 node $\mathscr{N}_q$ obeys
@@
 $\mathcal{K}'(h)$; therefore, the maximum possible value of $\mathcal{K}(a) -
-\mathcal{K}(b)$ is when the interval $[a, b]$ is centered on $h$.  Further,
-$\mathcal{K}(a) - \mathcal{K}(b) \le \epsilon$ when $(b - a) \le (-\epsilon /
-\mathcal{K}'(h))$.  Note that
+\mathcal{K}(b)$ is when the interval $[a, b]$ is centered on $h$.  This allows
+us to say that $\mathcal{K}(a) - \mathcal{K}(b) \le \epsilon$ when $(b - a) \le
+(-\epsilon / \mathcal{K}'(h))$.  Note that
 
@@
 Consider, then, the case when $s_r^{\max} > \log_2( -\epsilon /
-\mathcal{K}'(h) ) - 2$.  It is trivial to see that for any $\mathscr{N}_r \in
+\mathcal{K}'(h) ) - 2$.  Because of the pruning rule, for any $\mathscr{N}_r \in
 R^*$, $\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) > \epsilon$;
-alternately stated, this means that
+we may refactor this by applying definitions to show
 %\begin{equation}
@@
 %\end{equation}
-is sufficient to bound $|R^*|$.  For notational convenience, define $\omega = \mathcal{K}^{-1}(\epsilon) /
-2^{s_r^{\max} + 1} + 1$.  Using Lemma \ref{lem:packing} with $\delta =
-2^{s_r^{\max}}$ and $\rho = 2 \omega$ gives $|R^*| = c_r^{3 + \lceil \log_2
-\omega \rceil}$.
+is sufficient to bound $|R^*|$.  For notational convenience, define $\omega =
+(\mathcal{K}^{-1}(\epsilon) / 2^{s_r^{\max} + 1}) + 1$, and the statement may be
+more concisely written as $B_{S_r}(p_q, \omega 2^{s_r^{\max} + 1}) \cap
+C_{s_r^{\max}}$.
+
+Using Lemma \ref{lem:packing} with $\delta = 2^{s_r^{\max}}$ and $\rho = 2
+\omega$ gives $|R^*| = c_r^{3 + \lceil \log_2 \omega \rceil}$.
 
@@
 runtime diverges; however, note that there is no dependence on the kernel
-bandwidth $\sigma$.  For a reasonably chosen $\epsilon = 0.05$, the runtime is
-approximately $O(c_r^{8.89} (N + \theta))$; for $\epsilon = 0.01$, the runtime
-is approximately $O(c_r^{11.52} (N + \theta))$.  For very small $\epsilon =
+bandwidth $\sigma$.  To demonstrate the relationship of runtime to $\epsilon$,
+see that for a reasonably chosen $\epsilon = 0.05$, the runtime is approximately
+$O(c_r^{8.89} (N + \theta))$; for $\epsilon = 0.01$, the runtime is
+approximately $O(c_r^{11.52} (N + \theta))$.  For very small $\epsilon =
 0.00001$, the runtime is approximately $O(c_r^{22.15} (N + \theta))$.
-
-To understand the lack of dependence on kernel bandwidth more intuitively,
-consider that as the kernel bandwidth increases, two things happen: {\it (a)}
-the reference set $R$ becomes empty at larger scales, and {\it (b)}
-$\mathcal{K}^{-1}(\epsilon)$ grows, allowing less pruning at higher levels.
-These effects are opposite, and for the Gaussian kernel they cancel each other
-out, giving the same bound regardless of bandwidth.
 
@@
 kernels do not exhibit dependence on the bandwidth.
+
+To understand the lack of dependence on kernel bandwidth more intuitively,
+consider that as the kernel bandwidth increases, two things happen: {\it (a)}
+the reference set $R$ becomes empty at larger scales, and {\it (b)}
+$\mathcal{K}^{-1}(\epsilon)$ grows, allowing less pruning at higher levels.
+These effects are opposite, and for the Gaussian and Laplacian kernels they
+cancel each other out, giving the same bound regardless of bandwidth.
 
@@
 \begin{equation}
-f^*(p_q) \ge N \mathcal{K}\left(\max_{p_r \in S_r} d(p_q, p_r)\right).
+|f^*(p_q)| \ge N \mathcal{K}\left(\max_{p_r \in S_r} d(p_q, p_r)\right).
 \end{equation}
 
-This is clearly true, because each point in $S_r$ must contribute more than
+This is clearly true: each point in $S_r$ must contribute more than
 $\mathcal{K}(\max_{p_r \in S_r} d(p_q, p_r))$ to $f^*(p_q)$.  Now, we may revise
@@
 \noindent where $\mathcal{K}^{\max}$ is lower bounded by $\mathcal{K}(\max_{p_r
-\in S_r} d(p_q, p_r))$.  This allows us to create a \texttt{Score()} algorithm,
-given in Algorithm \ref{alg:kde_rva_score}.
+\in S_r} d(p_q, p_r))$.  Assuming we have some estimate $\mathcal{K}^{\max}$,
+this allows us to create a \texttt{Score()} algorithm, given in Algorithm
+\ref{alg:kde_rva_score}.
 
@@
 \begin{thm}
-Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
-of the previous section.  Then, given a query set $S_q$ and a reference set
-$S_r$ both of size $O(N)$, it is possible to perform relative value
-approximate kernel density estimation (satisfying the condition of Equation
-\ref{eqn:rva}) in $O(N)$ time, assuming that the expansion constant $c_r$ of
-$S_r$ is not dependent on $N$.
+Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the same
+assumptions as Theorem \ref{thm:kde-bound}.  Then, given a query set $S_q$ and a
+reference set $S_r$ both of size $O(N)$, it is possible to perform relative
+value approximate kernel density estimation (satisfying the condition of
+Equation \ref{eqn:rva}) in $O(N)$ time, assuming that the expansion constant
+$c_r$ of $S_r$ is not dependent on $N$.
 \end{thm}

### 2015-01-08 bounds-files/sec04-range-search.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

@@
 In the range search problem, the task is to find the set of reference points
-%\begin{equation}
+
+\begin{equation}
 $S[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}$
-%\end{equation}
+\end{equation}
+
 for each query point $p_q$, where $[l, u]$ is the given
@@
 Our proof works for both of these algorithms
-identically, but we will focus on range search.  A \texttt{BaseCase()} and
+similarly, but we will focus on range search.  A \texttt{BaseCase()} and
 \texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
@@
 difficult problem.  For bounding the running time of range search, we require
-one more definition.
+one more notion of difficulty, related to how $| S_{\max} |$ changes due to
+changes in the range $[l, u]$.
 
@@
 
-Using these notions, we may now bound the running time of range search.
+When the $\alpha$-expansion of the set $S_{\max}$ is approximately the same size
+as $S_{\max}$, then the problem would not be significantly more difficult if the
+range $[l, u]$ was increased slightly.  Using these notions, then, we may now
+bound the running time of range search.
 
@@
 (Algorithm \ref{alg:rs_sc}) take $O(1)$ time.  Therefore, using Lemma
-\ref{thm:ct-runtime} we know that the runtime of the algorithm is bounded by
+\ref{thm:ct-runtime}, we know that the runtime of the algorithm is bounded by
 $O(c_r^4 |R^*| (N + i_t(\mathscr{N}_q) + \theta))$.  As with the previous
@@
 This set is necessarily a subset of $S^{\alpha}[p_q]$; by
-definition, the number of points in this set is bounded above by $|S_{\max}| +
+assumption, the number of points in this set is bounded above by $|S_{\max}| +
 C$.  We may then conclude that $|R^*| \le |S_{\max}| + C$.  By taking the

### 2015-01-08 bounds.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

@@
 the cover tree, separating out the problem-dependent and the problem-independent
-elements. This allows us to plug in bounds for the problem-dependent
+elements.  This allows us to just plug in bounds for the problem-dependent
 elements to get runtime guarantees for dual-tree algorithms for any pairwise
@@
 
-\citet{nbody} generalized the fast multipole method from
-computational physics to obtain dual-tree algorithms.  These are extremely
-useful when there are large query sets, not just a few query points.  Instead of
-building a tree on the reference set and searching with each query point
-separately, Gray and Moore suggest also building a query tree and traversing
-both the query and reference trees simultaneously (a {\it dual-tree traversal},
-from which the class of algorithms takes its name).
+Building upon this intuition, \citet{nbody} generalized the fast multipole
+method from computational physics to obtain dual-tree algorithms.  These are
+extremely useful when there are large query sets, not just a few query points.
+Instead of building a tree on the reference set and searching with each query
+point separately, Gray and Moore suggest also building a query tree and
+traversing both the query and reference trees simultaneously (a {\it dual-tree
+traversal}, from which the class of algorithms takes its name).
 
@@
 \citep{curtin2013fast}, particle smoothing \citep{klaas2006fast}, variational
-inference \citep{amizadeh2012variational}, and range search \citep{nbody}, to name
-a few.
+inference \citep{amizadeh2012variational}, range search \citep{nbody}, and
+embedding techniques \cite{maaten2014accelerating}, to name a few.
 
 Some of these algorithms are derived using the cover tree \citep{langford2006}, a
-data structure with beneficial theoretical qualities.  Dual-tree all
-nearest-neighbor search and approximate kernel density estimation with cover
-trees have $O(N)$ runtime guarantees for $O(N)$ queries \citep{ram2009}; minimum
-spanning tree calculation scales as $O(N \log N)$ \citep{march2010euclidean}.
-Other problems have similar worst-case guarantees \citep{curtin2014dual,
-march2013multi}.
+data structure with compelling theoretical qualities.  When cover trees are
+used, Dual-tree all-nearest-neighbor search and approximate kernel density
+estimation have $O(N)$ runtime guarantees for $O(N)$ queries \citep{ram2009};
+minimum spanning tree calculation scales as $O(N \log N)$
+\citep{march2010euclidean}.  Other problems have similar worst-case guarantees
+\citep{curtin2014dual, march2013multi}.
 

### 2015-01-08 bounds-files/general-runtime-bound.tex
Add a quick discussion.

@@
 \label{alg:line:ct-dual-srmax}
-    \IF{$(s_q < s^{\max}_r)$}
-\label{alg:line:ct-dual-ref-recursion-start}
+    \IF{$(s_q < s^{\max}_r)$} \label{alg:line:ct-dual-ref-recursion-start}
+      \STATE \COMMENT{Perform a reference recursion.}
       \FORALL{$\mathscr{N}_r \in R$} \label{alg:line:ct-dual-base-case-start}
@@
     \ELSE \label{alg:line:ct-dual-query-recursion-start}
+      \STATE \COMMENT{Perform a query recursion.}
       \FORALL{$\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q)$}
@@
 
-Both $\nu$ and $\kappa$ are difficult to empirically calculate and require an
-entire run of the dual-tree algorithm.  On the other hand, bounding
+$\nu$ corresponds to the maximum number of reference recursions between a single
+query recursion, and $\kappa$ corresponds to the maximum number of query
+recursions between a single reference recursion.  The respective proofs that use
+these constants then apply them as a worst-case measure for the whole algorithm:
+when using $\kappa$, \citet{ram2009} assume that {\it every} reference recursion
+may be followed by $\kappa$ query recursions; similarly, \citet{curtin2014dual}
+assume that {\it every} query recursion may be followed by $\nu$ reference
+recursions.  Here, we have simply used $i_t(\mathscr{T}_q)$ and $\theta$ as an
+exact summation of the total extra reference recursions, which gives us a much
+tighter bound than $\nu$ or $\kappa$ on the running time of the whole algorithm.
+
+Further, both $\nu$ and $\kappa$ are difficult to empirically calculate and
+require an entire run of the dual-tree algorithm.  On the other hand, bounding
 $i_t(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree

### 2015-01-14 bounds-files/sec03-approx-kde.tex
Some changes.

@@
 assumptions; most standard kernels fall into this class, including the Gaussian,
-Laplacian and Epanechnikov kernels.
+exponential, and Epanechnikov kernels.
 
@@
 
-Next, consider the Laplacian kernel:
+Next, consider the exponential kernel:
 $\mathcal{K}_l(d) = \exp(-d / \sigma)$.  For this kernel, $h = 0$ (that is, the
@@
 \epsilon$, resulting in $\zeta = -\mathcal{K}'_l(h) \mathcal{K}^{-1}_l(\epsilon)
-\epsilon^{-1} = \epsilon^{-1} \ln \epsilon$.  So both the Laplacian and Gaussian
+\epsilon^{-1} = \epsilon^{-1} \ln \epsilon$.  So both the exponential and Gaussian
 kernels do not exhibit dependence on the bandwidth.
@@
 $\mathcal{K}^{-1}(\epsilon)$ grows, allowing less pruning at higher levels.
-These effects are opposite, and for the Gaussian and Laplacian kernels they
+These effects are opposite, and for the Gaussian and exponential kernels they
 cancel each other out, giving the same bound regardless of bandwidth.

### 2015-01-21 bounds.tex
Some changes.
Some minor last-minute changes.

@@
   \name Dongryeol Lee \email drselee@gmail.com\\
-  \addr GE Research\\
-  Schenedectadadaty, NY\\
+  \addr GE Global Research Center\\
+  Schenectady, NY 12309
   \AND
   \name William B. March \email march@ices.utexas.edu\\
-  \addr Austin, TX, USA\\
+  \addr Institute for Computational Engineering and Sciences\\
+  University of Texas, Austin\\
+  Austin, TX 78712-1229
   \AND
   \name Parikshit Ram \email p.ram@gatech.edu\\
-  \addr Skytree
+  \addr Skytree, Inc.\\
+  Atlanta, GA 30332
 }
@@
 Numerous machine learning algorithms contain pairwise statistical problems at
-their core -- i.e. tasks that require computations over all pairs of input points if implemented directly. 
-Often, tree structures are used to solve these problems efficiently. Dual-tree algorithms 
-can efficiently solve or approximate many of these problems.
-Using cover trees, rigorous worst-case runtime
-guarantees have been proven for some of these algorithms. In this paper, we present a {\em
-problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
-the cover tree, separating out the problem-dependent and the problem-independent
-elements.  This allows us to just plug in bounds for the problem-dependent
-elements to get runtime guarantees for dual-tree algorithms for any pairwise
-statistical problem without re-deriving the entire proof. We demonstrate this plug-and-play procedure for
-nearest-neighbor search and approximate kernel density estimation to get
-improved runtime guarantees.  Under mild assumptions, we also present the first
-linear runtime guarantee for dual-tree based range search.
+their core---that is, tasks that require computations over all pairs of input
+points if implemented naively.  Often, tree structures are used to solve these
+problems efficiently. Dual-tree algorithms can efficiently solve or approximate
+many of these problems.  Using cover trees, rigorous worst-case runtime
+guarantees have been proven for some of these algorithms. In this paper, we
+present a {\em problem-independent} runtime guarantee for {\em any} dual-tree
+algorithm using the cover tree, separating out the problem-dependent and the
+problem-independent elements.  This allows us to just plug in bounds for the
+problem-dependent elements to get runtime guarantees for dual-tree algorithms
+for any pairwise statistical problem without re-deriving the entire proof. We
+demonstrate this plug-and-play procedure for nearest-neighbor search and
+approximate kernel density estimation to get improved runtime guarantees.  Under
+mild assumptions, we also present the first linear runtime guarantee for
+dual-tree based range search.
 \end{abstract}

### 2015-01-21 cover-letter.tex
Update cover letter.

\documentclass[twoside,11pt]{article} % For LaTeX2e
\usepackage{jmlr2e}
\usepackage{hyperref}
\hypersetup{colorlinks,
            urlcolor=blue,
            citecolor=green,
            linkcolor=red}
\usepackage{url}
%\documentstyle[nips13submit_09,times,art10]{article} % For LaTeX 2.09
\usepackage{mathrsfs}
%\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{color}
\usepackage{mathtools}

\usepackage{pstricks}
\usepackage{tikz}
\usepackage{subcaption}

\begin{document}

\noindent Dear Drs. Murphy and Schölkopf: \\

\medskip

My co-authors (William March, Parikshit Ram, and Dongryeol Lee) would like to
submit our manuscript "Plug-and-play dual-tree algorithm runtime analysis" for
publication in the Journal of Machine Learning Research.  This paper establishes
a generalized technique for proving runtime bounds on the class of algorithms
known as dual-tree algorithms.  These algorithms are commonly used for
computations such as nearest neighbor search, range search, kernel density
estimation, and other applications.  For each of the three problems just listed,
we have shown linear runtime bounds using our generalized bounds (though it must
be noted that these bounds depend on the intrinsic dimensionality of the
dataset). \\

\medskip

Each co-author is aware of this submission and consent to its review, and the
paper has not been submitted to any other venues.  There are no conflicts of
interest, since no co-author has recently collaborated with any action editor.
\\
\medskip

Our suggestions for action editors are those who have published papers relating
to dual-tree algorithms in the past or are familiar with the approach and
existing theory.  These include Sanjoy Dasgupta, Nando de Freitas, and possibly
Guy Lebanon. \\
\medskip

If there is anything we have neglected to do in the submission process, please
let us know and we will rectify the issue immediately. \\

\medskip

\noindent Thank you for your consideration, \\

\medskip

\noindent Ryan Curtin

\end{document}

### 2015-06-18 bounds-files/general-runtime-bound-mod.tex
Last modifications of plug-and-play bounds paper.

\begin{thm}
\label{thm:ct-runtime}
%For any dual-tree algorithm using cover trees and the standard cover tree
%dual-tree pruning traversal (Algorithm \ref{alg:cover-tree-dual}) with reference
%set $S_r$ that has expansion constant $c_r$ and query set $S_q$, the runtime is
%bounded above by $O(c_r^4 | R^* | \nu \chi \psi N)$, where $ | R^* | $ is
%the maximum size of the reference set $R$ (line
%\ref{alg:line:ct-dual-input}) at any point in the dual-tree pruning traversal,
%$\chi$ is the longest possible running time of \texttt{BaseCase()}, and $\psi$
%is the longest possible running time of \texttt{Score()}.
Given a reference set $S_r$ of size $N$ with an expansion constant $c_r$ and a
set of queries $S_q$ of size $O(N)$, a standard cover tree based dual-tree
algorithm (Algorithm \ref{alg:cover-tree-dual}) takes $O(c_r^4 | R^* | \chi
\psi (N + \theta))$, where $ | R^* | $ is the maximum size of the reference set $R$ (line
\ref{alg:line:ct-dual-input}) during the dual-tree recursion, $\chi$ is
the maximum possible runtime of \texttt{BaseCase()}, $\psi$ is the maximum
possible runtime of \texttt{Score()}, and $\theta$ is

\begin{equation}
\theta = \max(\log_2(\eta_r / \eta_q), 0) + b + N \max(\log_2(\delta_q / \delta_r), 0)
\end{equation}

\noindent with $\eta_q$ as the maximum pairwise distance between points in
$S_q$, $\delta_q$ as the minimum nonzero pairwise distance between points in
$S_q$, $\eta_r$ as the maximum pairwise distance between points in $S_r$,
$\delta_r$ as the minimum nonzero pairwise distance between points in $S_r$, and
$b$ as some type of measure of the imbalance of the tree which is defined as

\begin{equation}
b = \sum_{\mathscr{N}_q \in \mathscr{T}_q} \left(\sum_{\mathscr{N}_{qc} \in
\mathscr{C}(\mathscr{N}_q), s_{qc} \ne -\infty} s_q - s_{qc} - 1 \right) +
\left(\sum_{\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q), s_{qc} = -\infty}
\max(s_q - s_q^{\min} - 1, 0) \right).
\end{equation}
\end{thm}

\vspace*{-1.23em}
\begin{proof}
First, split the algorithm into two parts: reference recursions (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
and query recursions (lines
\ref{alg:line:ct-dual-query-recursion-start}--\ref{alg:line:ct-dual-query-recursion-end}).
The runtime of the algorithm is bounded as the runtime of a reference recursion
times the total number of reference recursions plus the total runtime of all
query recursions.

Consider a reference recursion (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end}).
The work done in the base case loop from lines
\ref{alg:line:ct-dual-base-case-start}--\ref{alg:line:ct-dual-base-case-end} is
$O(\chi | R |)$.  Define $| R^* |$ to be the largest set $|R|$ for any scale
$s_r^{\max}$ and any query node $\mathscr{N}_q$ during the course of the
algorithm; then, it is true that $| R | \le | R^* |$.
%
Then, lines \ref{alg:line:ct-dual-ref-children} and
\ref{alg:line:ct-dual-ref-score} take $O(c_r^4 \psi | R |) \le O(c_r^4 \psi |
R^* |)$ time, because each reference node has $c_r^4$ children.  So, one
full reference recursion takes $O(c_r^4 \psi \chi | R^* |)$ time.

Now, note that there are $O(N)$ nodes in $\mathscr{T}_q$.  Thus, line
\ref{alg:line:ct-dual-query-recursion} is visited $O(N)$ times.  The amount of
work in line \ref{alg:line:ct-dual-query-pruning}, like in the reference
recursion, is bounded as $O(c_r^4 \psi | R^* |)$. Therefore, the total
runtime of all query recursions is $O(c_r^4 \psi | R^* | N)$.

Lastly, we must bound the total number of reference recursions.  Reference
recursions happen in three cases: \textit{(1)} $s_r^{\max}$ is greater than the
scale of the root of the query tree (no query recursions have happened yet);
\textit{(2)} $s_r^{\max}$ is less than or equal to the scale of the root of the
query tree, but is greater than the minimum scale of the query tree that is not
$-\infty$; \textit{(3)} $s_r^{\max}$ is less than the minimum scale of the query
tree that is not $-\infty$.

First, consider case \textit{(1)}.  If the scale of the root of the query tree
is $s_q^T$ and the scale of the root of the reference tree is $s_r^T$, then
$s_r^T - s_q^T$ recursions of this sort occur before the first query recursion.

Next, consider case \textit{(2)}.  Each query recursion implies at least one
reference recursion; thus there will be at least $O(N)$ recursions.  Any
extra recursions above and beyond this occur because the query node does not
have scale which is exactly one less than its parent.  When the query node is
not a leaf, the number of extra reference recursions is bounded by the
difference between the scales of the query node and its parent minus one (one is
subtracted to account for the one reference recursion that must happen).  Each
query node is visited once, and therefore the total number of extra recursions
when the query node is not a leaf is bounded by

\begin{equation}
\sum_{\mathscr{N}_q \in \mathscr{T}_q} \sum_{\mathscr{N}_{qc} \in
\mathscr{C}(\mathscr{N}_q), s_{qc} \ne -\infty} s_q - s_{qc} - 1.
\end{equation}

When the query node is a leaf, then the number of recursions in case
\textit{(2)} for that query node is bounded by the difference between the scale
of the query node's parent and the smallest scale of a non-leaf node in the
query tree (call this $s_q^{\min}$) minus one:

\begin{equation}
\sum_{\mathscr{N}_q \in \mathscr{T}_q} \sum_{\mathscr{N}_{qc} \in
\mathscr{C}(\mathscr{N}_q), s_{qc} = -\infty} \max(s_q - s_q^{\min} - 1, 0).
\end{equation}

The $\max$ is necessary to account for the case where the query node's parent
has scale $s_q^{\min}$; no reference recursions of type \textit{(2)} happen at
all in this case.  Therefore the total number of extra reference recursions of
type \textit{(2)} is bounded by $b$, where

\begin{equation}
b = \sum_{\mathscr{N}_q \in \mathscr{T}_q} \left(\sum_{\mathscr{N}_{qc} \in
\mathscr{C}(\mathscr{N}_q), s_{qc} \ne -\infty} s_q - s_{qc} - 1 \right) +
\left(\sum_{\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q), s_{qc} = -\infty}
\max(s_q - s_q^{\min} - 1, 0) \right).
\end{equation}

(Note that $b$ can be calculated in a single $O(N)$ traversal of the tree.)

Lastly, case \textit{(3)} occurs when the reference scale is less than the
minimum query scale $s_q^{\min}$.  Define the minimum reference scale as
$s_r^{\min}$.  If $s_q^{\min} - s_r^{\min} > 0$, each query node may have up to
$s_q^{\min} - s_r^{\min}$ reference recursions of type \textit{(3)}, and there
are $N$ query nodes.  Thus, the total number of reference recursions of type
\textit{(3)} is bounded by $N \max(s_q^{\min} - s_r^{\min}, 0)$.

Combining these three cases, we get the following expression for an upper bound
$\theta$ on the total number of reference recursions:

\begin{equation}
\theta = \max(s_r^T - s_q^T, 0) + b + N (s_q^{\min} - s_r^{\min}).
\end{equation}

Now, we can express these in terms of more common constants which are related to
the aspect ratio of the data.  Define $\delta_q$ as the minimum nonzero pairwise
distance in $S_q$ and $\eta_q$ as the furthest pairwise distance in $S_q$ (and
define $\delta_r$ and $\eta_r$ accordingly for the set $S_r$).  Then, $s_r^T =
\lceil \log_2 \eta_r \rceil$, $s_q^T = \lceil \log_2 \eta_q$, $s_r^{\min} =
\lceil \log_2 \delta_r \rceil$, and $s_q^{\min} = \lceil \log_2 \delta_q
\rceil$.  This gives a more reasonable expression:

\begin{equation}
\theta = \max(\log_2(\eta_r / \eta_q), 0) + b + N \max(\log_2(\delta_q / \delta_r), 0).
\end{equation}

Returning to our original line of reasoning, we can combine these results to
give the result of the theorem:

\begin{equation}
O(c_r^4 \psi \chi | R^* | \theta) + O(c_r^4 \psi | R^* | N) \sim O(c_r^4 \psi
\chi |R^*| (N + \theta).
\end{equation}
\end{proof}
\vspace*{-0.8em}

A quick discussion of the numerous constants we have introduced and what they
mean.  In the monochromatic case, the terms of $\theta$ depending on $\eta_q$,
$\eta_r$, $\delta_q$, and $\delta_r$ disappear entirely.  However we are left
with the $b$ term, which is kind of a measure of imbalance of the tree.  The
best case is a case where the tree is perfectly balanced and there are no
`missing levels' where a child of a node is more than one level away from its
parent.  In practice, this quantity is probably much less than $N$ and should
not scale significantly with $N$ (those points which contribute to $b$ the most
will probably be outliers).  In the worst case, $b$ is bounded by the maximum
depth of the tree multiplied by the number of points, which is trivially $O(N
\log (\eta_q / \eta_r)$.  It is likely that a tighter bound of $O(N \log N)$
could be established.

There are two corner cases for the non-monochromatic case: all extra reference
recursions are of type \textit{(1)}, and all extra reference recursions are of
type \textit{(3)}.  The first case presents the best situation, because all of
the reference recursions are done before the first query node is recursed in.
Therefore the bulk of the reference recursion work is done by with the query
tree root as the query node; this is the best case.  The worst case is the
latter, where recursion in the reference tree cannot begin until reaching the
bottom of the query tree.  Thus, no pruning can occur until reaching the leaves
of the query tree, and the recursion is essentially equivalent to a single-tree
recursion for each query point plus the added overhead of the recursion of the
query tree with one reference node.  This second case degenerates to something
like $O(N \log N)$ or $O(N^2)$ time depending on how you look at it.

\subsection{Back to your regularly scheduled paper.}

This result holds for any dual-tree algorithm regardless of the problem. Hence,
%for any pairwise statistical problem, the runtime of the dual-tree algorithm (of
the runtime of any dual-tree algorithm
%that problem)
would be at least $O(N)$ using our bound, which matches the intuition that
answering $O(N)$ queries would take at least $O(N)$ time. For a particular
problem and data, if $c_r$, $|R^*|$, $\nu$, $\chi$ and
$\psi$ are bounded by constants independent of $N$ (for large enough $N$), then
the dual-tree algorithm for that problem has a runtime linear in $N$. Our
theoretical result separates out the problem-dependent and the
problem-independent elements of the runtime bound, which allows us to simply
plug in the problem-dependent bounds to get runtime bounds for any dual-tree
algorithm without requiring an analysis from scratch.
% This discussion already happened earlier... sort of.
%The expansion constant is related to a notion of intrinsic dimensionality of a
%dataset (as discussed earlier). A relatively small $c_r$ corresponds to
%a dataset with low intrinsic dimensionality, while a value of $c_r$ independent
%of $N$ implies that the intrinsic dimensionality of the data is independent of
%the number of points in the dataset. The latter is a reasonable assumption in
%many scenarios.
%Our bound has a linear dependence on the inverse constant of bichromaticity
%$\nu$; this results from the worst-case situation where the reference tree is
%recursed entirely after all query tree recursions
% this makes sense, given that $\nu$ represents the maximum number of times
%any set $R$ is descended for any query node $\mathscr{N}_q$ before
%$\mathscr{N}_q$ is recursed.
If $\nu \sim \kappa$, which is a reasonable assumption, this is a better result
than the results of Ram et.~al.  \cite{ram2009} who found a dependence of
$c_q^{4 \kappa}$ in their proofs.

The quantity $|R^*|$ bounds the amount of work that needs to be done for each
recursion. In the worst case, $|R^*|$ can be $N$. However,
dual-tree algorithms rely on branch-and-bound techniques to prune away
work (Lines \ref{alg:line:ct-dual-ref-score} and
\ref{alg:line:ct-dual-query-pruning} in Algorithm \ref{alg:cover-tree-dual}). A
small value of $|R^*|$ will imply that the algorithm is extremely successful in
pruning away work at Line \ref{alg:line:ct-dual-ref-score} in Algorithm
\ref{alg:cover-tree-dual}. An (upper) bound on $|R^*|$ (and the algorithm's
success in pruning work) will depend on the problem and the data.  As we will
show, bounding $|R^*|$ is often possible. % In addition, we will also show that
For many dual-tree algorithms, $\chi \sim \psi \sim O(1)$; often, cached
sufficient statistics~\cite{moore2000anchors} can enable $O(1)$ runtime
implementations of \texttt{BaseCase()} and \texttt{Score()}.

% Space constraints :(

%The runtime bound $\chi$ for the \texttt{BaseCase($p_q$, $p_r$)} function
%corresponds to the time required to process a pair of points. For example, in
%range search, \texttt{BaseCase()} corresponds to the
%time taken to compute $d(p_q, p_r)$ and determine if that falls into the desired
%range.  For general kernel summations, this corresponds to computing the
%pairwise kernel function between $p_q$ and $p_r$ and then summing it to the
%output for $p_q$.  In general, the runtime of \texttt{BaseCase()} is dominated
%by the pairwise function computation which is independent of $N$ and is usually
%$O(1)$. The \texttt{Score()} function is an operation on a pair of nodes and
%usually involves computing a bound which requires a pairwise function
%computation between the node centers (similar to the \texttt{BaseCase()}
%function except that it is for node centres) and certain cached node statistics
%(such as node radius in the metric space). Hence its runtime bound $\psi$ is
%usually a constant independent of $N$ or simply $O(1)$.

### 2015-08-14 bounds-files/imbalance.tex
Some minor last-minute changes.
Add empirical imbalance numbers.

@@
 \end{center}
-\caption{A badly-drawn multiple-outlier cover tree.}
+\caption{A multiple-outlier cover tree.}
 \label{fig:outliers}
@@
 \citep{langford2006}, the tree imbalance is near-linear with the number of
-points.  Generally, most of the cover tree imbalance is contributed by leaf
-nodes whose parent has scale greater than $s_{\min}$.  At this time, no cover
-tree construction algorithm specifically aims to minimize imbalance.
+points; see Table \ref{tab:imbalance} for empirically-calculated imbalances
+on a number of datasets.  The `power', `susy', `higgs', and
+`covertype' datasets are found in the UCI Machine Learning Repository
+\cite{ucimlrepository}, the `mnist' dataset is from \cite{lecun2000mnist}, and
+the `sdss' dataset is Sloan Digital Sky Survey data \cite{adelman2008sixth}.
+Imbalance is shown for subsets of the datasets of size $5000$, $50000$, and
+$500000$; the actual value of the imbalance is of less importance than the
+scaling properties with $N$---which tend to be near-linear.
+
+Currently, no cover tree construction algorithm specifically aims to minimize
+imbalance.
+
+\begin{table}[tb]
+\begin{center}
+\begin{tabular}{| c | c | c | c | c |}
+\hline
+ & & \multicolumn{3}{|c|}{Imbalance} \\
+\hline
+Dataset & $d$ & $N = 5k$ & $N = 50k$ & $N = 500k$ \\
+\hline
+%lcdm & 3 & 29402 & 313190 & 3400305 \\
+sdss & 4 & 17691 & 191716 & 1483711 \\
+power & 7 & 22152 & 202050 & 1345825 \\
+susy & 18 & 3686 & 64231 & 930969 \\
+higgs & 29 & 404 & 1272 & 73793 \\
+covertype & 55 & 9680 & 100344 & 1319447 \\
+mnist & 784 & 4004 & 79375 & 983554 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Empirically calculated tree imbalances.}
+\label{tab:imbalance}
+\end{table}
 

### 2015-08-14 bounds-files/general-runtime-bound.tex
i_t -> I_t, i_n -> I_n.

@@
 \begin{equation}
-O\left(c_r^4 | R^* | \chi \psi (N + i_t(\mathscr{T}_q) + \theta)\right)
+O\left(c_r^4 | R^* | \chi \psi (N + I_t(\mathscr{T}_q) + \theta)\right)
 \end{equation}
@@
 the children of $\mathscr{N}_q$ are recursed into is bounded above by
-$i_n(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has imbalance $0$, then it is
+$I_n(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has imbalance $0$, then it is
 exactly one level below its parent, and thus there is only one reference
@@
 parent, then it is possible that a reference recursion may occur for each level
-in between; this is a maximum of $i_n(\mathscr{N}_q) + 1$.
+in between; this is a maximum of $I_n(\mathscr{N}_q) + 1$.
 
@@
 \begin{equation}
-\sum_{\mathscr{N}_q \in \mathscr{T}_q} i_n(\mathscr{N}_q) + 1 = i_t(\mathscr{T}_q) +
+\sum_{\mathscr{N}_q \in \mathscr{T}_q} I_n(\mathscr{N}_q) + 1 = I_t(\mathscr{T}_q) +
 O(N)
@@
 \begin{equation}
-O\left(c_r^4 |R^*| \psi \chi \left(N + i_t(\mathscr{T}_q) + \theta\right)\right)
+O\left(c_r^4 |R^*| \psi \chi \left(N + I_t(\mathscr{T}_q) + \theta\right)\right)
 + O\left(c_r^4 |R^*| \psi N\right) \sim O\left(c_r^4 |R^*| \psi
-\chi \left(N + i_t(\mathscr{T}_q) + \theta\right)\right).
+\chi \left(N + I_t(\mathscr{T}_q) + \theta\right)\right).
 \end{equation}
@@
 \begin{equation}
-O\left(c^4 |R^*| \chi \psi \left(N + i_t(\mathscr{T})\right)\right)
+O\left(c^4 |R^*| \chi \psi \left(N + I_t(\mathscr{T})\right)\right)
 \end{equation}
@@
 \sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
-also have a high cover tree imbalance $i_t(\mathscr{T})$; the linear dependence of
+also have a high cover tree imbalance $I_t(\mathscr{T})$; the linear dependence of
 runtime on imbalance is thus sensible for well-behaved datasets.
@@
 assume that {\it every} query recursion may be followed by $\nu$ reference
-recursions.  Here, we have simply used $i_t(\mathscr{T}_q)$ and $\theta$ as an
+recursions.  Here, we have simply used $I_t(\mathscr{T}_q)$ and $\theta$ as an
 exact summation of the total extra reference recursions, which gives us a much
@@
 require an entire run of the dual-tree algorithm.  On the other hand, bounding
-$i_t(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree
+$I_t(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree
 (assuming the tree is already built).  Thus, not only is our bound tighter when

### 2015-08-14 bounds-files/imbalance.tex
i_t -> I_t, i_n -> I_n.

@@
 \begin{defn}
-The {\it cover node imbalance} $i_n(\mathscr{N}_i)$ for a cover tree node
+The {\it cover node imbalance} $I_n(\mathscr{N}_i)$ for a cover tree node
 $\mathscr{N}_i$ with scale $s_i$ in the cover tree $\mathscr{T}$ is defined as
@@
 \begin{equation}
-i_n(\mathscr{N}_i) = \begin{dcases*}
+I_n(\mathscr{N}_i) = \begin{dcases*}
   s_p - s_i - 1 & if $\mathscr{N}_i$ is not a leaf and not the root node \\
@@
 \label{def:imbalance}
-The {\it cover tree imbalance} $i_t(\mathscr{T})$ for a cover tree $\mathscr{T}$
+The {\it cover tree imbalance} $I_t(\mathscr{T})$ for a cover tree $\mathscr{T}$
 is defined as the cumulative number of missing levels in the tree.  This can be
@@
 \begin{equation}
-i_t(\mathscr{T}) = \sum_{\mathscr{N}_i \in \mathscr{T}} i_n(\mathscr{N}_i).
+I_t(\mathscr{T}) = \sum_{\mathscr{N}_i \in \mathscr{T}} I_n(\mathscr{N}_i).
 \end{equation}
@@
 A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
-imbalance $i_t(\mathscr{T}_b) = 0$ (for instance, Figure
+imbalance $I_t(\mathscr{T}_b) = 0$ (for instance, Figure
 \ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
 entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
-$s_{\min}$ will have imbalance $i_t(\mathscr{T}_w) \sim N (s_{\max} -
+$s_{\min}$ will have imbalance $I_t(\mathscr{T}_w) \sim N (s_{\max} -
 s_{\min})$.  Because of this chain-like structure, each level has only one node

### 2015-08-14 bounds-files/sec02-nearest-neighbor-search.tex
i_t -> I_t, i_n -> I_n.

@@
 the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N +
-i_t(\mathscr{T}_q) + \theta))$ with $i_t(\mathscr{T}_q$ and $\theta$ defined as
+I_t(\mathscr{T}_q) + \theta))$ with $I_t(\mathscr{T}_q$ and $\theta$ defined as
 in Definition \ref{def:imbalance} and Lemma \ref{lem:extcase3}, respectively.
@@
 Due to Theorem \ref{thm:ct-runtime}, we therefore know that the runtime of the
-algorithm is bounded by $O(c_r^4 |R^*| (N + i_t(\mathscr{T}_q) + \theta))$.
+algorithm is bounded by $O(c_r^4 |R^*| (N + I_t(\mathscr{T}_q) + \theta))$.
 Thus, the only thing that remains is to bound the maximum size of the reference
@@
 In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 (N +
-i_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For
-well-behaved trees where $i_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
+I_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For
+well-behaved trees where $I_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
 this represents the current tightest worst-case runtime bound for nearest

### 2015-08-14 bounds-files/sec03-approx-kde.tex
Some minor last-minute changes.
i_t -> I_t, i_n -> I_n.

@@
 %\begin{equation}
-$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + i_t(\mathscr{T}_q) + \theta))$
+$O(c_r^{8 + \lceil \log_2 \zeta \rceil} (N + I_t(\mathscr{T}_q) + \theta))$
 %\end{equation}
 with $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$,
-$i_t(\mathscr{T}_q)$ defined as in Definition \ref{def:imbalance}, and $\theta$
+$I_t(\mathscr{T}_q)$ defined as in Definition \ref{def:imbalance}, and $\theta$
 defined as in Lemma \ref{lem:extcase3}.
@@
 so Theorem \ref{thm:ct-runtime} implies the total runtime of the dual-tree
-algorithm is bounded by $O(c_r^4 |R^*| (N + i_t(\mathscr{T}_q) + \theta))$.
+algorithm is bounded by $O(c_r^4 |R^*| (N + I_t(\mathscr{T}_q) + \theta))$.
 Thus, we will bound $|R^*|$ using techniques related to those used by
@@
 
-This is clearly true: each point in $S_r$ must contribute more than
-$\mathcal{K}(\max_{p_r \in S_r} d(p_q, p_r))$ to $f^*(p_q)$.  Now, we may revise
-the relative approximation condition in Equation \ref{eqn:rva}:
-
-\begin{equation}
-| f(p_q) - f^*(p_q) | \le \epsilon \mathcal{K}^{\max}
-\end{equation}
-
-\noindent where $\mathcal{K}^{\max}$ is lower bounded by $\mathcal{K}(\max_{p_r
-\in S_r} d(p_q, p_r))$.  Assuming we have some estimate $\mathcal{K}^{\max}$,
-this allows us to create a \texttt{Score()} algorithm, given in Algorithm
-\ref{alg:kde_rva_score}.
-
 \begin{algorithm}[tb]
@@
 \end{algorithm}
+
+This is clearly true: each point in $S_r$ must contribute more than
+$\mathcal{K}(\max_{p_r \in S_r} d(p_q, p_r))$ to $f^*(p_q)$.  Now, we may revise
+the relative approximation condition in Equation \ref{eqn:rva}:
+
+\begin{equation}
+| f(p_q) - f^*(p_q) | \le \epsilon \mathcal{K}^{\max}
+\end{equation}
+
+\noindent where $\mathcal{K}^{\max}$ is lower bounded by $\mathcal{K}(\max_{p_r
+\in S_r} d(p_q, p_r))$.  Assuming we have some estimate $\mathcal{K}^{\max}$,
+this allows us to create a \texttt{Score()} algorithm, given in Algorithm
+\ref{alg:kde_rva_score}.
 

### 2015-08-14 bounds-files/sec04-range-search.tex
Bibliography changes, I want to stab myself, I hate this shit so much.
Some minor last-minute changes.
i_t -> I_t, i_n -> I_n.

@@
+In the range search problem, the task is to find the set of reference points
+
+\begin{equation}
+S[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}
+\end{equation}
+
+\noindent for each query point $p_q$, where $[l, u]$ is the given
+range.  The range count problem is practically identical, but only the size of
+the set, $|S[p_q]|$, is desired.
+Our proof works for both of these algorithms
+similarly, but we will focus on range search.  A \texttt{BaseCase()} and
+\texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
+\ref{alg:rs_sc}, respectively \citep[a correctness proof can be found
+in][]{curtin2013tree}.  The sets $N[p_q]$ (for each $p_q$) are
+initialized to $\emptyset$ at the beginning of the traversal.
+
+In order to bound the running time of dual-tree range search, we require better
+notions for understanding the difficulty of the problem.  Observe that if the
+range is sufficiently large, then for every query point $p_q$, $S[p_q] = S_r$.
+Clearly, for $S_q \sim S_r \sim O(N)$, this cannot be solved in anything less
+than quadratic time simply due to the time required to fill each output array
+$S[p_q]$.  Define the maximum result size for a given query set $S_q$, reference
+set $S_r$, and range $[l, u]$ as
+
+\begin{equation}
+| S_{\max} | = \max_{p_q \in S_q} | S[p_q] |.
+\label{eqn:smax}
+\end{equation}
+
 \begin{algorithm}[tb]
@@
 \end{algorithm}
-
-In the range search problem, the task is to find the set of reference points
-
-\begin{equation}
-$S[p_q] = \{ p_r \in S_r : d(p_q, p_r) \in [l, u] \}$
-\end{equation}
-
-for each query point $p_q$, where $[l, u]$ is the given
-range.  The range count problem is practically identical, but only the size of
-the set, $|S[p_q]|$, is desired.
-Our proof works for both of these algorithms
-similarly, but we will focus on range search.  A \texttt{BaseCase()} and
-\texttt{Score()} function are given in Algorithms \ref{alg:rs_bc} and
-\ref{alg:rs_sc}, respectively \citep[a correctness proof can be found
-in][]{curtin2013tree}.  The sets $N[p_q]$ (for each $p_q$) are
-initialized to $\emptyset$ at the beginning of the traversal.
-
-In order to bound the running time of dual-tree range search, we require better
-notions for understanding the difficulty of the problem.  Observe that if the
-range is sufficiently large, then for every query point $p_q$, $S[p_q] = S_r$.
-Clearly, for $S_q \sim S_r \sim O(N)$, this cannot be solved in anything less
-than quadratic time simply due to the time required to fill each output array
-$S[p_q]$.  Define the maximum result size for a given query set $S_q$, reference
-set $S_r$, and range $[l, u]$ as
-
-\begin{equation}
-| S_{\max} | = \max_{p_q \in S_q} | S[p_q] |.
-\label{eqn:smax}
-\end{equation}
 
@@
 O\left(c_r^{4} \max\left(c_r^{4 + \beta}, |S_{\max}| + C\right) (N +
-i_t(\mathscr{N}_q) + \theta)
+I_t(\mathscr{N}_q) + \theta)
 \right)
@@
 \ref{thm:ct-runtime}, we know that the runtime of the algorithm is bounded by
-$O(c_r^4 |R^*| (N + i_t(\mathscr{N}_q) + \theta))$.  As with the previous
+$O(c_r^4 |R^*| (N + I_t(\mathscr{N}_q) + \theta))$.  As with the previous
 proofs, then, our only task is to bound the maximum size of the reference set,
@@
 \begin{equation}
-O(c_r^{8 + \beta} (N + i_t(\mathscr{N}_q) + \theta)).
+O(c_r^{8 + \beta} (N + I_t(\mathscr{N}_q) + \theta)).
 \end{equation}
@@
 $\alpha$.  Consider $\alpha = (1 / 3)$; this yields a running time of $O(c^8 (N
-+ \theta))$.  $\alpha = (1 / 7)$ yields $O(c^9 (N + i_t(\mathscr{N}_q +
-\theta))$, $\alpha = (1 / 15)$ yields $O(c^{10} (N + i_t(\mathscr{N}_q) +
++ \theta))$.  $\alpha = (1 / 7)$ yields $O(c^9 (N + I_t(\mathscr{N}_q +
+\theta))$, $\alpha = (1 / 15)$ yields $O(c^{10} (N + I_t(\mathscr{N}_q) +
 \theta))$, and so forth.  As $\alpha$ gets smaller, the exponent on $c$ gets

### 2015-08-14 bounds-files/sec01-supporting-lemmas.tex
Bibliography changes, I want to stab myself, I hate this shit so much.
Add an example cover tree figure.

@@
 self-child are taken to be the children of the node).  Figure
-\ref{fig:implicit-explicit} shows an example cover tree and highlights implicit
-and explicit nodes.
+\ref{fig:example-cover-tree} shows each of the levels of an example cover tree
+(in the explicit representation) on a simple six-point dataset.
 
@@
 \texttt{Construct}.}.
+
+\begin{figure}
+\begin{subfigure}[b]{0.495\textwidth}
+  \begin{center}
+    \begin{tikzpicture}[scale=0.8]
+      \input{figures/cover-tree-points}
+      \input{figures/cover-tree-s1}
+    \end{tikzpicture}
+  \end{center}
+  \vspace*{-0.5em}
+  \caption{Root node at scale $1$.}
+  \label{fig:cover-tree-s1}
+\end{subfigure}
+\begin{subfigure}[b]{0.495\textwidth}
+  \begin{center}
+    \begin{tikzpicture}[scale=0.85]
+      \input{figures/cover-tree-points}
+      \input{figures/cover-tree-s0}
+    \end{tikzpicture}
+  \end{center}
+  \vspace*{-0.5em}
+  \caption{Nodes at scale $0$.}
+  \label{fig:cover-tree-s0}
+\end{subfigure}
+\begin{subfigure}[b]{0.355\textwidth}
+  \begin{center}
+    \begin{tikzpicture}[scale=0.85]
+      \input{figures/cover-tree-points}
+      \input{figures/cover-tree-s-1}
+    \end{tikzpicture}
+  \end{center}
+  \vspace*{-0.5em}
+  \caption{Nodes at scale $-1$.}
+  \label{fig:cover-tree-s-1}
+\end{subfigure}
+\begin{subfigure}[b]{0.64\textwidth}
+  \begin{center}
+    \begin{tikzpicture}[scale=0.6]
+      \input{figures/cover-tree-ex}
+    \end{tikzpicture}
+  \end{center}
+  \vspace*{-0.5em}
+  \caption{Abstract representation.}
+  \label{fig:cover-tree-ex}
+\end{subfigure}
+  \vspace*{-0.5em}
+\caption{Example cover tree on six points.  (a) $\mathscr{N}_a$ is centered at
+$p_0$ with scale $1$.  (b) $\mathscr{N}_b$ and $\mathscr{N}_c$ are centered at
+$p_0$ and $p_1$, respectively, and have scale $0$.  (c) $\mathscr{N}_d$ and
+$\mathscr{N}_e$ are centered at $p_0$ and $p_2$, respectively, and have scale
+$-1$.  The leaves, $\mathscr{N}_0$ through $\mathscr{N}_6$, are centered at each
+of the six points, with scale $-\infty$ (and therefore radius 0).  Note that
+although node $\mathscr{N}_b$ in subfigure (b) overlaps node $\mathscr{N}_c$,
+point $p_1$ only belongs to $\mathscr{N}_c$, not $\mathscr{N}_b$.  Note also
+that this is only one valid cover tree that could be built on the data; other
+configurations are possible; for instance, selecting a different root point
+gives different valid cover trees.}
+\label{fig:example-cover-tree}
+\end{figure}
 

### 2015-08-14 figures/cover-tree-ex.tex
Add an example cover tree figure.

%%
\begin{scope}
  \pgfsetstrokecolor{black}
  \definecolor{strokecol}{rgb}{1.0,1.0,1.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{1.0,1.0,1.0};
  \pgfsetfillcolor{fillcol}
  \filldraw (0bp,0bp) -- (0bp,252bp) -- (414bp,252bp) -- (414bp,0bp) -- cycle;
\end{scope}
\begin{scope}
  \pgfsetstrokecolor{black}
  \definecolor{strokecol}{rgb}{1.0,1.0,1.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{1.0,1.0,1.0};
  \pgfsetfillcolor{fillcol}
  \filldraw (0bp,0bp) -- (0bp,252bp) -- (414bp,252bp) -- (414bp,0bp) -- cycle;
\end{scope}
\begin{scope}
  \pgfsetstrokecolor{black}
  \definecolor{strokecol}{rgb}{1.0,1.0,1.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{1.0,1.0,1.0};
  \pgfsetfillcolor{fillcol}
  \filldraw (0bp,0bp) -- (0bp,252bp) -- (414bp,252bp) -- (414bp,0bp) -- cycle;
\end{scope}
  \pgfsetcolor{black}
  % Edge: ne -> ni
  \draw [->] (215.35bp,72.765bp) .. controls (219.71bp,64.283bp) and (225.15bp,53.714bp)  .. (234.7bp,35.147bp);
  % Edge: nd -> ng
  \draw [->] (99bp,71.697bp) .. controls (99bp,63.983bp) and (99bp,54.712bp)  .. (99bp,36.104bp);
  % Edge: nd -> nf
  \draw [->] (84.43bp,74.834bp) .. controls (74.25bp,64.938bp) and (60.476bp,51.546bp)  .. (41.796bp,33.385bp);
  % Edge: nb -> ne
  \draw [->] (207bp,143.7bp) .. controls (207bp,135.98bp) and (207bp,126.71bp)  .. (207bp,108.1bp);
  % Edge: nc -> nk
  \draw [->] (323.26bp,144.71bp) .. controls (335.67bp,120.23bp) and (359.24bp,73.739bp)  .. (378.62bp,35.537bp);
  % Edge: nb -> nd
  \draw [->] (188.19bp,148.81bp) .. controls (171bp,137.67bp) and (145.38bp,121.06bp)  .. (117.6bp,103.05bp);
  % Edge: na -> nb
  \draw [->] (264.43bp,218.83bp) .. controls (254.25bp,208.94bp) and (240.48bp,195.55bp)  .. (221.8bp,177.38bp);
  % Edge: na -> nc
  \draw [->] (287.35bp,216.76bp) .. controls (291.71bp,208.28bp) and (297.15bp,197.71bp)  .. (306.7bp,179.15bp);
  % Edge: nc -> nj
  \draw [->] (315bp,143.87bp) .. controls (315bp,119.67bp) and (315bp,75.211bp)  .. (315bp,36.189bp);
  % Edge: ne -> nh
  \draw [->] (198.65bp,72.765bp) .. controls (194.29bp,64.283bp) and (188.85bp,53.714bp)  .. (179.3bp,35.147bp);
  % Node: nh
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (171bp,18bp) ellipse (27bp and 18bp);
  \draw (171bp,18bp) node {$\mathscr{N}_2$};
\end{scope}
  % Node: ni
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (243bp,18bp) ellipse (27bp and 18bp);
  \draw (243bp,18bp) node {$\mathscr{N}_5$};
\end{scope}
  % Node: nj
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (315bp,18bp) ellipse (27bp and 18bp);
  \draw (315bp,18bp) node {$\mathscr{N}_1$};
\end{scope}
  % Node: nk
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (387bp,18bp) ellipse (27bp and 18bp);
  \draw (387bp,18bp) node {$\mathscr{N}_3$};
\end{scope}
  % Node: na
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (279bp,234bp) ellipse (27bp and 18bp);
  \draw (279bp,234bp) node {$\mathscr{N}_a$};
\end{scope}
  % Node: nb
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (207bp,162bp) ellipse (27bp and 18bp);
  \draw (207bp,162bp) node {$\mathscr{N}_b$};
\end{scope}
  % Node: nc
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (315bp,162bp) ellipse (27bp and 18bp);
  \draw (315bp,162bp) node {$\mathscr{N}_c$};
\end{scope}
  % Node: nd
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (99bp,90bp) ellipse (27bp and 18bp);
  \draw (99bp,90bp) node {$\mathscr{N}_d$};
\end{scope}
  % Node: ne
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (207bp,90bp) ellipse (27bp and 18bp);
  \draw (207bp,90bp) node {$\mathscr{N}_e$};
\end{scope}
  % Node: nf
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (27bp,18bp) ellipse (27bp and 18bp);
  \draw (27bp,18bp) node {$\mathscr{N}_0$};
\end{scope}
  % Node: ng
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \definecolor{fillcol}{rgb}{0.83,0.83,0.83};
  \pgfsetfillcolor{fillcol}
  \filldraw [opacity=1] (99bp,18bp) ellipse (27bp and 18bp);
  \draw (99bp,18bp) node {$\mathscr{N}_4$};
\end{scope}

\draw[gray,thin,dashed] (0bp,53bp) -- (420bp,53bp);
\draw[gray,thin,dashed] (0bp,125bp) -- (420bp,125bp);
\draw[gray,thin,dashed] (0bp,197bp) -- (420bp,197bp);

\draw (400bp,250bp) node[darkgray] {\scriptsize scale $1$};
\draw (400bp,185bp) node[darkgray] {\scriptsize scale $0$};
\draw (400bp,115bp) node[darkgray] {\scriptsize scale $-1$};
\draw (400bp,45bp) node[darkgray] {\scriptsize scale $-\infty$};

%

### 2015-08-14 figures/cover-tree-points.tex
Add an example cover tree figure.

\coordinate (p0) at (0, 0);
\coordinate (p1) at (1.0, -1.3);
\coordinate (p2) at (-0.3, 0.85);
\coordinate (p3) at (1.5, -1.8);
\coordinate (p4) at (0.4, -0.2);
\coordinate (p5) at (-0.6, 1.3);

\node [draw, circle, inner sep=1pt, fill] at (p0) { };
\node [below] at (p0) { $p_0$ };
\node [draw, circle, inner sep=1pt, fill] at (p1) { };
\node [below] at (p1) { $p_1$ };
\node [draw, circle, inner sep=1pt, fill] at (p2) { };
\node [below] at (p2) { $p_2$ };
\node [draw, circle, inner sep=1pt, fill] at (p3) { };
\node [below] at (p3) { $p_3$ };
\node [draw, circle, inner sep=1pt, fill] at (p4) { };
\node [below] at (p4) { $p_4$ };
\node [draw, circle, inner sep=1pt, fill] at (p5) { };
\node [below] at (p5) { $p_5$ };

### 2015-08-14 bounds-files/sec01-supporting-lemmas.tex
Rewording about the generalizability of the bound.

@@
   \vspace*{-0.5em}
-\caption{Example cover tree on six points.  (a) $\mathscr{N}_a$ is centered at
-$p_0$ with scale $1$.  (b) $\mathscr{N}_b$ and $\mathscr{N}_c$ are centered at
-$p_0$ and $p_1$, respectively, and have scale $0$.  (c) $\mathscr{N}_d$ and
-$\mathscr{N}_e$ are centered at $p_0$ and $p_2$, respectively, and have scale
-$-1$.  The leaves, $\mathscr{N}_0$ through $\mathscr{N}_6$, are centered at each
-of the six points, with scale $-\infty$ (and therefore radius 0).  Note that
-although node $\mathscr{N}_b$ in subfigure (b) overlaps node $\mathscr{N}_c$,
-point $p_1$ only belongs to $\mathscr{N}_c$, not $\mathscr{N}_b$.  Note also
-that this is only one valid cover tree that could be built on the data; other
-configurations are possible; for instance, selecting a different root point
-gives different valid cover trees.}
+\caption{Example cover tree on six points in $\mathcal{R}^2$.  (a)
+$\mathscr{N}_a$ is centered at $p_0$ with scale $1$.  (b) $\mathscr{N}_b$ and
+$\mathscr{N}_c$ are centered at $p_0$ and $p_1$, respectively, and have scale
+$0$.  (c) $\mathscr{N}_d$ and $\mathscr{N}_e$ are centered at $p_0$ and $p_2$,
+respectively, and have scale $-1$.  The leaves, $\mathscr{N}_0$ through
+$\mathscr{N}_6$, are centered at each of the six points, with scale $-\infty$
+(and therefore radius 0).  Note that although node $\mathscr{N}_b$ in subfigure
+(b) overlaps node $\mathscr{N}_c$, point $p_1$ only belongs to $\mathscr{N}_c$,
+not $\mathscr{N}_b$.  Note also that this is only one valid cover tree that
+could be built on the data; other configurations are possible; for instance,
+selecting a different root point gives different valid cover trees.}
 \label{fig:example-cover-tree}
@@
 The expansion constant is used heavily in the cover tree literature.  It is,
-in some sense, a notion of instrinic dimensionality, and previous work has shown
-that there are many scenarios where $c$ is independent of the number of points
-in the dataset \citep{karger2002finding, langford2006,
-krauthgamer2004navigating, ram2009}.  Note also that if points in $S \subset
-\mathcal{H}$ are being drawn according to a stationary distribution $f(x)$, then
-$c$ will converge to some finite value $c_f$ as $|S| \to \infty$.  To see this,
-define $c_f$ as a generalization of the expansion constant for distributions.
-$c_f \ge 2$ is the smallest value such that
+in some sense, a notion of instrinic dimensionality, most useful in scenarios
+where $c$ is independent of the number of points in the dataset
+\citep{karger2002finding, langford2006, krauthgamer2004navigating, ram2009}.
+Note also that if points in $S \subset \mathcal{H}$ are being drawn according to
+a stationary distribution $f(x)$, then $c$ will converge to some finite value
+$c_f$ as $|S| \to \infty$.  To see this, define $c_f$ as a generalization of the
+expansion constant for distributions.  $c_f \ge 2$ is the smallest value such
+that
 

### 2015-08-14 bounds.tex
Add an example cover tree figure.
Rewording about the generalizability of the bound.

@@
 \usepackage{subcaption}
+
+\usetikzlibrary{calc}
 
@@
 state-of-the-art, and in the case of range search, is the first such bound.
+Despite the intuition this provides for the scaling properties of all dual-tree
+algorithms\footnote{Dual-tree algorithms using $kd$-trees and other types of
+trees have been observed to empirically scale linearly for tasks that take
+quadratic time without the use of trees; see the empirical results of
+\citet{nbody, march2010euclidean, vladymyrov2014linear, klaas2006fast,
+gray2003nonparametric}.}, it must be kept in mind that these worst-case bounds
+only apply to dual-tree algorithms that use the cover tree and the standard
+cover tree traversal.
 %  We

### 2015-08-14 bounds-files/sec02-nearest-neighbor-search.tex
Fix typo.

@@
 the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N +
-I_t(\mathscr{T}_q) + \theta))$ with $I_t(\mathscr{T}_q$ and $\theta$ defined as
+I_t(\mathscr{T}_q) + \theta))$ with $I_t(\mathscr{T}_q)$ and $\theta$ defined as
 in Definition \ref{def:imbalance} and Lemma \ref{lem:extcase3}, respectively.

### 2015-08-14 bounds-files/general-runtime-bound.tex
Pull definition of \theta out of Lemma 4.
Clarify what the monochromatic case actually is.

@@
 \begin{equation}
-\theta = \max\left\{\min\left[O(N \log_2(\delta_q / \delta_r)),
+\max\left\{\min\left[O(N \log_2(\delta_q / \delta_r)),
 O(N^2)\right], \; 0\right\}.
@@
 \end{lemma}
+
+For convenience, we define a term that encapsulates this bound.
+
+\begin{defn}
+Define $\theta$ as a bound on the number of extra reference recursions tha
+happen after the last query recursion.  Then,
+
+\begin{equation}
+\theta = \max\left\{\min\left[O(N \log_2(\delta_q / \delta_r)), O(N^2)\right],
+\; 0\right\}.
+\end{equation}
+\end{defn}
 
@@
 
-When we consider the monochromatic case (where $S_q = S_r$), the results
-trivially simplify.
+When we consider the monochromatic case (where $S_q = S_r$)\footnote{In the
+monochromatic case, we do not take a point as its own nearest neighbor, so
+slight modification of {\tt BaseCase()} is necessary.  The runtime bound result
+remains unchanged.}, the results trivially simplify.
 

### 2015-08-14 bounds-files/general-runtime-bound.tex
Refactor for similar notation in theorem/lemma statements.

@@
 \label{lem:extcase1}
-For a dual-tree algorithm with $S_q \sim S_r \sim O(N)$ using cover trees and
+For a dual-tree algorithm with $|S_q| \sim |S_r| \sim O(N)$ using cover trees and
 the traversal given in Algorithm \ref{alg:cover-tree-dual}, the number of extra
@@
 \label{lem:extcase3}
-For a dual-tree algorithm with $S_q \sim S_r \sim O(N)$ using cover trees and
+For a dual-tree algorithm with $|S_q| \sim |S_r| \sim O(N)$ using cover trees and
 the traversal given in Algorithm \ref{alg:cover-tree-dual}, the number of extra
@@
 %is the longest possible running time of \texttt{Score()}.
-Given a reference set $S_r$ of size $N$ with an expansion constant $c_r$ and a
+Given a reference set $S_r$ of size $O(N)$ with an expansion constant $c_r$ and a
 set of queries $S_q$ of size $O(N)$, a standard cover tree based dual-tree

### 2015-08-14 bounds-files/sec03-approx-kde.tex
Refactor for similar notation in theorem/lemma statements.

@@
 Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
-of the previous subsection.  Then, given a query set $S_q$ and a reference set
-$S_r$ with expansion constant $c_r$, and using the approximate kernel density
-estimation \texttt{BaseCase()} and \texttt{Score()} as given in Algorithms
-\ref{alg:kde_base_case} and \ref{alg:kde_score}, respectively, with the
-traversal given in Algorithm \ref{alg:cover-tree-dual}, the running time of
-approximate kernel density estimation for some error parameter $\epsilon$ is
-bounded by
+of the previous subsection.  Then, given a query set $S_q$ of size $O(N)$ and a
+reference set $S_r$ of size $O(N)$ with expansion constant $c_r$, and using the
+approximate kernel density estimation \texttt{BaseCase()} and \texttt{Score()}
+as given in Algorithms \ref{alg:kde_base_case} and \ref{alg:kde_score},
+respectively, with the traversal given in Algorithm \ref{alg:cover-tree-dual},
+the running time of approximate kernel density estimation for some error
+parameter $\epsilon$ is bounded by
 %\begin{equation}

### 2015-08-14 bounds-files/sec04-range-search.tex
Refactor for similar notation in theorem/lemma statements.

@@
 \begin{thm}
-Given a reference set $S_r$ of size $N$ with expansion constant $c_r$, and a
+Given a reference set $S_r$ of size $O(N)$ with expansion constant $c_r$, and a
 query set $S_q$ of size $O(N)$, a search range of $[l, u]$, and using the range

### 2015-08-14 bounds-files/sec02-nearest-neighbor-search.tex
Refactor for consistent use of O(N).

--- 2015-08-14 bounds-files/sec02-nearest-neighbor-search.tex
@@
 Algorithms \ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively, and also
-given a reference set $S_r$ with expansion constant $c_r$, and a query set $S_q$,
-the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N +
-I_t(\mathscr{T}_q) + \theta))$ with $I_t(\mathscr{T}_q$ and $\theta$ defined as
-in Definition \ref{def:imbalance} and Lemma \ref{lem:extcase3}, respectively.
+given a reference set $S_r$ of size $O(N)$ with expansion constant $c_r$, and a
+query set $S_q$ of size $O(N)$, the running time of the algorithm is bounded by
+$O(c_r^4 c_{qr}^5 (N + I_t(\mathscr{T}_q) + \theta))$ with $I_t(\mathscr{T}_q)$
+and $\theta$ defined as in Definition \ref{def:imbalance} and Lemma
+\ref{lem:extcase3}, respectively.
 \label{thm:nns}

### 2015-08-14 bounds-files/general-runtime-bound.tex
Minor fixes and changes for reviewer #1.

@@
 \begin{lemma}
-For all nodes $\mathscr{N}_i, \mathscr{N}_j \in R$ (in the context of Algorithm
-\ref{alg:cover-tree-dual}) which contain points $p_i$ and $p_j$, respectively,
-$d(p_i, p_j) > 2^{s_r^{\max}}$, with $s_r^{\max}$ defined as in line
-\ref{alg:line:ct-dual-srmax}.
+For all distinct nodes $\mathscr{N}_i, \mathscr{N}_j \in R$ (in the context of
+Algorithm \ref{alg:cover-tree-dual}) which contain points $p_i$ and $p_j$,
+respectively, $d(p_i, p_j) > 2^{s_r^{\max}}$, with $s_r^{\max}$ defined as in
+line \ref{alg:line:ct-dual-srmax}.
 \end{lemma}
@@
 \rceil$.
-  \item The {\it minimum scale} of the query tree $\mathscr{T}_r$ is defined as
+  \item The {\it minimum scale} of the query tree $\mathscr{T}_q$ is defined as
 $s_q^{\min} = \lceil \log_2(\delta_q) \rceil$.
@@
 \begin{equation}
-\max\left\{\min\left[O(N \log_2(\delta_q / \delta_r)),
-O(N^2)\right], \; 0\right\}.
+\max\left\{\min\left(O(N \log_2(\delta_q / \delta_r)),
+O(N^2)\right), \; 0\right\}.
 \end{equation}
@@
 \begin{equation}
-\theta = \max\left\{\min\left[O(N \log_2(\delta_q / \delta_r)), O(N^2)\right],
+\theta = \max\left\{\min\left(O(N \log_2(\delta_q / \delta_r)), O(N^2)\right),
 \; 0\right\}.
@@
 by the term $c^4 |R^*|$, with $|R^*|$ being the maximum size of the reference
-set $R$; this value must be derived for each specific problem.  The bad
-performance of poorly-behaved datasets with large $c$ (or, in the worst case, $c
-\sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
-also have a high cover tree imbalance $I_t(\mathscr{T})$; the linear dependence of
-runtime on imbalance is thus sensible for well-behaved datasets.
+set $R$; this value must be derived for each specific problem.  The poor
+performance of trees on datasets with large $c$ (or, in the worst case, $c
+\sim N$) is then captured in both of those terms.  These datasets for which
+trees perform poorly may also have a high cover tree imbalance
+$I_t(\mathscr{T})$; the linear dependence of runtime on imbalance is thus
+sensible for datasets where trees perform well.
 
@@
 %that problem)
-would be at least $O(N)$ using our bound, which matches the intuition that
-answering $O(N)$ queries would take at least $O(N)$ time. For a particular
-problem and data, if $c_r$, $|R^*|$, $\chi$, $\psi$ are bounded by constants
-independent of $N$ and $\theta$ is no more than linear in $N$ (for large enough
-$N$), then the dual-tree algorithm for that problem has a runtime linear in $N$.
-Our theoretical result separates out the problem-dependent and the
-problem-independent elements of the runtime bound, which allows us to simply
+can be bounded no more tightly than $O(N)$ with our bound, which matches the
+intuition that answering $O(N)$ queries will take at least $O(N)$ time. For a
+particular problem and data, if $c_r$, $|R^*|$, $\chi$, $\psi$ are bounded by
+constants independent of $N$ and $\theta$ is no more than linear in $N$ (for
+large enough $N$), then the dual-tree algorithm for that problem has a runtime
+linear in $N$.  Our theoretical result separates out the problem-dependent and
+the problem-independent elements of the runtime bound, which allows us to simply
 plug in the problem-dependent bounds in order to get runtime bounds for any
@@
 
-$\nu$ corresponds to the maximum number of reference recursions between a single
-query recursion, and $\kappa$ corresponds to the maximum number of query
-recursions between a single reference recursion.  The respective proofs that use
-these constants then apply them as a worst-case measure for the whole algorithm:
-when using $\kappa$, \citet{ram2009} assume that {\it every} reference recursion
-may be followed by $\kappa$ query recursions; similarly, \citet{curtin2014dual}
-assume that {\it every} query recursion may be followed by $\nu$ reference
-recursions.  Here, we have simply used $I_t(\mathscr{T}_q)$ and $\theta$ as an
-exact summation of the total extra reference recursions, which gives us a much
-tighter bound than $\nu$ or $\kappa$ on the running time of the whole algorithm.
+The quantity $\nu$ corresponds to the maximum number of reference recursions
+between a single query recursion, and $\kappa$ corresponds to the maximum number
+of query recursions between a single reference recursion.  The respective proofs
+that use these constants then apply them as a worst-case measure for the whole
+algorithm: when using $\kappa$, \citet{ram2009} assume that {\it every}
+reference recursion may be followed by $\kappa$ query recursions; similarly,
+\citet{curtin2014dual} assume that {\it every} query recursion may be followed
+by $\nu$ reference recursions.  Here, we have simply used $I_t(\mathscr{T}_q)$
+and $\theta$ as an exact summation of the total extra reference recursions,
+which gives us a much tighter bound than $\nu$ or $\kappa$ on the running time
+of the whole algorithm.
 

### 2015-08-14 bounds.tex
Minor fixes and changes for reviewer #1.

@@
 This naively requires $O(N)$ time
-to answer in single query in a reference set of size $N$; answering $O(N)$
+to answer a single query in a reference set of size $N$; answering $O(N)$
 queries subsequently requires prohibitive $O(N^2)$ time. Kernel density

### 2015-08-14 responses_to_reviewers.tex
Add some responses to the reviewers.

\documentclass[10pt]{article} % For LaTeX2e
\usepackage{url}
\usepackage{amsmath}
\usepackage{mathrsfs}

\begin{document}

Dear reviewers:

Thank you for your time in reviewing our manuscript.  We appreciate the detailed
comments and have incorporated several important changes:

\begin{itemize}
  \item The imbalance has been renamed $I_t(\cdot)$ (from $i_t(\cdot)$), to
avoid confusion with the index $i$ which used in other contexts.

  \item The spelling and grammatical errors have been fixed.

  \item The quantity $\theta$ is now defined outside Lemma 4.

  \item We have added a table of empirical results for the imbalance of cover
trees on a handful of real-world datasets (Table 1).

  \item We have added Figure 1, an example cover tree, which should help readers
develop intuition for cover trees.

  \item We have emphasized in the introduction that these bounds only apply to
cover trees, not other types of trees.  Unfortunately, other types of trees have
not (at this time) had any comparable adaptive analysis performed, so we cannot
compare our results with results for other tree types.
\end{itemize}

Now, we answer a few questions that came up during the reviews:

\begin{quote}
{\bf (Reviewer 1)} In particular, it seems non-trivial to derive similar bounds
for other efficient search structures like kd-trees, ball trees, or vantage
point trees, as these are not amenable to the notion of imbalancedness that the
authors use in their bounds. Are existing bounds for these tree structures
better or worse (or similar) than the bounds derived here? And more importantly,
what is the performance of cover trees compared to these other structures in
practice?
\end{quote}

We are not aware of existing worst-case bounds for other tree structures that
are non-trivial and very loose (i.e. $O(N^2)$ for nearest neighbor search), so
unfortunately we cannot compare our theoretical results in any meaningful way.
The empirical performance of cover trees is often comparable with that of
$kd$-trees, metric trees, and other types of trees.  Here is an interesting
empirical survey that supports this claim:

A.M. Kibriya, E. Frank, "An empirical comparison of exact nearest neighbour
algorithms", PKDD 2007.  (Table 1 is of particular interest, though it is
restricted to single-tree search.)
\url{http://www.cs.waikato.ac.nz/~ml/publications/2007/KibriyaAndFrankPKDD07.pdf}

In addition, our experiences with mlpack development seem to indicate that
$kd$-trees and cover trees can perform comparably.  Of course, the actual
performance is dependent on the data; sometimes cover trees are faster (usually
in medium dimensions), sometimes $kd$-trees are faster (usually in lower
dimensions).

We've added some references in the introduction to highlight the
already-documented empirical performance of dual-tree algorithms, too.

\begin{quote}
{\bf (Reviewer 1)} The authors state that "tree imbalance is near-linear with
the number of points", which would imply the run-time bounds behave more like
O(N*N) in practice?
\end{quote}

We think there has been a slight misunderstanding here.  The bound is of the
form O(N + imbalance), so linear imbalance scaling still gives O(N) scaling.

\begin{quote}
{\bf (Reviewer 2)} Does the result extend to k-nearest neighbor search?
\end{quote}

Yes; you get a bound of the form $O(k(N + I_t(\mathscr{T}_q) + \theta))$.
Essentially the only thing that needs to change is Equation 24, where the ball
$B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2)$ now contains a maximum of $k$
points instead of $1$ point.

\begin{quote}
{\bf (Reviewer 3)} Lemma 1 proof: the definition of the expansion constant is an
inequality.  How can you assert the equality?
\end{quote}

This was a typo.  We have replaced the $=$ with $\le$ as it should be.

\begin{quote}
{\bf (Reviewer 3)} I find it strange that you are defining the "top" and
"minimum"... why not "top" and "bottom" or "maximum" and "minimum"?
\end{quote}

Unfortunately we are forced to use these terms here.  In our view, ``bottom
scale'' would imply the bottom of the tree, and the minimum scale is not the
scale of the bottom of the tree (that scale is $-\infty$).  We also can't use
the term ``maximum scale'', because that term is used heavily in the proof of
Lemma 2 to refer to the maximum scale present in the set $R$:

$$s_r^{\max} = \max_{\mathscr{N}_i \in R} s_i.$$

Again, thank you all for your comments!

\end{document}

### 2015-08-25 bounds-files/imbalance.tex
Update some comments about the experiment.

@@
 \citep{langford2006}, the tree imbalance is near-linear with the number of
-points; see Table \ref{tab:imbalance} for empirically-calculated imbalances
-on a number of datasets.  The `power', `susy', `higgs', and
-`covertype' datasets are found in the UCI Machine Learning Repository
-\cite{ucimlrepository}, the `mnist' dataset is from \cite{lecun2000mnist}, and
-the `sdss' dataset is Sloan Digital Sky Survey data \cite{adelman2008sixth}.
-Imbalance is shown for subsets of the datasets of size $5000$, $50000$, and
-$500000$; the actual value of the imbalance is of less importance than the
-scaling properties with $N$---which tend to be near-linear.
+points.  We have constructed cover trees on $N$ uniformly subsampled points
+from a variety of datasets and calculated the imbalance; see Table
+\ref{tab:imbalance} for the results.  These results are normalized with respect
+to $N$, for which the values of $5000$, $50000$, and $500000$ were chosen.
+The `power', `susy', `higgs', and `covertype' datasets are found in the UCI
+Machine Learning Repository \cite{ucimlrepository}, the `mnist' dataset is from
+\cite{lecun2000mnist}, and the `lcdm` and `sdss' datasets are Sloan Digital Sky
+Survey data \cite{adelman2008sixth}.  The imbalances on each of these datasets
+tend to be near-linear.
 
@@
 %lcdm & 3 & 29402 & 313190 & 3400305 \\
-sdss & 4 & 17691 & 191716 & 1483711 \\
-power & 7 & 22152 & 202050 & 1345825 \\
-susy & 18 & 3686 & 64231 & 930969 \\
-higgs & 29 & 404 & 1272 & 73793 \\
-covertype & 55 & 9680 & 100344 & 1319447 \\
-mnist & 784 & 4004 & 79375 & 983554 \\
+lcdm & 3 & 5.88 & 6.26 & 6.80 \\
+%sdss & 4 & 17691 & 191716 & 1483711 \\
+sdss & 4 & 3.54 & 3.83 & 2.97 \\
+% power & 7 & 22152 & 202050 & 1345825 \\
+power & 7 & 4.43 & 4.04 & 2.69 \\
+% susy & 18 & 3686 & 64231 & 930969 \\
+susy & 18 & 0.74 & 1.28 & 1.86 \\
+% higgs & 29 & 404 & 1272 & 73793 \\
+higgs & 29 & 0.08 & 0.03 & 0.15 \\
+% covertype & 55 & 9680 & 100344 & 1319447 \\
+covertype & 55 & 1.94 & 2.01 & 2.64 \\
+% mnist & 784 & 4004 & 79375 & 983554 \\
+mnist & 784 & 0.80 & 1.59 & 1.97 \\
 \hline
@@
 \end{center}
-\caption{Empirically calculated tree imbalances.}
+\caption{Empirically calculated tree imbalances, normalized by $N$.}
 \label{tab:imbalance}

### 2015-08-25 bounds-files/general-runtime-bound.tex
Fix misstatement about reference recursions.

@@
 complicated.  Given $R$, we calculate $R_r$, which is the set of nodes in $R$
-that have scale $s_r^{\max}$.  Then, for each node $\mathscr{N}_r$ in the set of
-children of nodes in $R_r$, the node combinations $(\mathscr{N}_q,
-\mathscr{N}_r)$ are scored and pruned if possible.  Those reference nodes that
-were not pruned form the set $R'_{r - 1}$.  Then, this set is combined with $R
-\setminus R_r$---that is, each of the nodes in $R$ that was {\em not} recursed
-into---to produce $R'$, and the algorithm recurses with $\mathscr{N}_q$ and the
-reference set $R'$.
+that have scale $s_r^{\max}$.  We expand each node in $R_r$ to construct $R_{r -
+1}$: this is the set of children of all nodes in $R_r$.  This set is then
+combined with $R \setminus R_r$ (that is, the set of references nodes not at
+scale $s_r^{\max}$) to produce $R_{r - 1}$.  Each node in $R_{r - 1}$ is then
+scored and pruned if possible, resulting in the pruned reference set $R'_{r -
+1}$.  The algorithm then recurses with $\mathscr{N}_q$ and $R'_{r - 1}$.
 

### 2015-10-07 bounds-files/imbalance.tex
Update imbalance numbers with better experiment. That sure did take a long time to run...

@@
 from a variety of datasets and calculated the imbalance; see Table
-\ref{tab:imbalance} for the results.  These results are normalized with respect
-to $N$, for which the values of $5000$, $50000$, and $500000$ were chosen.
-The `power', `susy', `higgs', and `covertype' datasets are found in the UCI
-Machine Learning Repository \cite{ucimlrepository}, the `mnist' dataset is from
-\cite{lecun2000mnist}, and the `lcdm` and `sdss' datasets are Sloan Digital Sky
-Survey data \cite{adelman2008sixth}.  The imbalances on each of these datasets
-tend to be near-linear.
+\ref{tab:imbalance} for the results.  Ten trials were performed for each dataset
+and each $N$, and the mean imbalance is given.  These results are normalized
+with respect to $N$, for which the values of $5000$, $50000$, and $500000$ were
+chosen.  The `power', `susy', `higgs', and `covertype' datasets are found in the
+UCI Machine Learning Repository \cite{ucimlrepository}, the `mnist' dataset is
+from \cite{lecun2000mnist}, the `lcdm` and `sdss' datasets are Sloan Digital Sky
+Survey data \cite{adelman2008sixth}, and the `randu' dataset is
+randomly-generated uniformly-distributed data in 10 dimensions.  The imbalances
+on each of these datasets tend to be near-linear.
 
@@
 \hline
-%lcdm & 3 & 29402 & 313190 & 3400305 \\
-lcdm & 3 & 5.88 & 6.26 & 6.80 \\
-%sdss & 4 & 17691 & 191716 & 1483711 \\
-sdss & 4 & 3.54 & 3.83 & 2.97 \\
-% power & 7 & 22152 & 202050 & 1345825 \\
-power & 7 & 4.43 & 4.04 & 2.69 \\
-% susy & 18 & 3686 & 64231 & 930969 \\
-susy & 18 & 0.74 & 1.28 & 1.86 \\
-% higgs & 29 & 404 & 1272 & 73793 \\
-higgs & 29 & 0.08 & 0.03 & 0.15 \\
-% covertype & 55 & 9680 & 100344 & 1319447 \\
-covertype & 55 & 1.94 & 2.01 & 2.64 \\
-% mnist & 784 & 4004 & 79375 & 983554 \\
-mnist & 784 & 0.80 & 1.59 & 1.97 \\
+%lcdm & 3 & 22417.9 & 257445.6 & 2621239.6 \\
+lcdm & 3 & 4.48 & 5.15 & 5.24 \\
+% sdss & 4 & 10853.6 & 140431.6 & 1486808.8 \\
+sdss & 4 & 2.17 & 2.81 & 2.97 \\
+% power & 7 & 27042.2 & 323125.6 & 2251317.2
+power & 7 & 5.41 & 6.46 & 4.50 \\
+% susy & 18 & 3701.3 & 37823.7 & 431784.8 \\
+susy & 18 & 0.74 & 0.76 & 0.86 \\
+%randu-10 & 10 & 1156.0 & 10763.4 & 297041.2 \\
+randu & 10 & 0.23 & 0.22 & 0.59 \\
+% higgs & 29 & 4963.6 & 84171.8 &  \\ need one more...
+%covertype & 54 & 6609.6 & 88296.4 & 1247820.7 \\
+covertype & 54 & 1.322 & 1.766 & 2.495 \\
+% mnist & 784 & 4974.1 & 83658.6 & 1044075.0 \\
+mnist & 784 & 0.99 & 1.67 & 2.09 \\
 \hline

### 2015-10-07 bounds-files/sec01-supporting-lemmas.tex
Fix a typo.
Remove reference to undefined p_i.

@@
 $\mathscr{D}_i^n$ \\ \hline
-$\mu_i$ & Center of $\mathscr{N}_i$ (for cover trees, $\mu_i = p_i$) \\ \hline
+$\mu_i$ & Center of $\mathscr{N}_i$ \\ \hline
 $\lambda_i$ & Furthest descendant distance \\ \hline
@@
 Observe that $B_S(p, \rho \delta) \subseteq B_S(p_i, 2 \rho \delta)$.
-Also, $| B_S(p_i, 2 \rho \delta) | = c^{2 + \lceil \log_2 \rho
+Also, $| B_S(p_i, 2 \rho \delta) | \le c^{2 + \lceil \log_2 \rho
 \rceil} | B_S(p_i, \delta / 2) |$ by the definition of the expansion constant.

### 2015-10-07 bounds-files/imbalance.tex
Note that c ~ O(N) in the footnote.

@@
 this behavior is not limited to cover trees, and can happen to mean-split
-$kd$-trees too, especially in higher dimensions.}.
+$kd$-trees too, especially in higher dimensions.  In addition, for this
+scenario to arise with cover trees, it must be true that $c \sim O(N)$.}.
 

### 2015-10-07 bounds-files/general-runtime-bound.tex
Discuss why query and reference scale closeness is useful.

@@
 keeping the relative scales of query nodes and reference nodes close.
+
+Keeping the query and reference scales close is both beneficial for the later
+theory and intuitively reasonable: recursing too quickly in the either the query
+or reference node will unnecessarily duplicate work.  Suppose we recurse many
+levels down the query tree before recursing down the reference tree, giving us a
+set of query nodes we are considering.  For {\it each} of these query nodes, we
+will then need to descend the reference tree.  Because these query nodes are
+close together (with respect to the reference nodes we are considering, which
+are of larger scale and thus further apart), the pruning decisions at each level
+of recursion are likely to be the same for each query node.  Therefore,
+recursing too far in the query tree may cause a large amount of duplicated work.
+The symmetric argument applies for recursing too far in the reference tree
+before recursing in the query tree.  This justifies the approach of keeping the
+query and reference scales approximately equal.
 

### 2015-10-07 bounds-files/conclusion.tex
Word change suggestion from Pari.

@@
 \ref{thm:rs}).  With our contribution, bounding a cover tree dual-tree algorithm
-is straightforward and only involves bounding the maximum size of the reference
+is streamlined and only involves bounding the maximum size of the reference
 set, $| R^* |$.

### 2015-10-22 bounds-all.tex
Last modifications of plug-and-play bounds paper.
Minor fixes + Dong's address update

--- 2015-01-21 bounds.tex
@@
   \name Dongryeol Lee \email drselee@gmail.com\\
-  \addr GE Global Research Center\\
-  Schenectady, NY 12309
+  \addr Yahoo Labs\\
+  Sunnyvale, CA 94089
   \AND
@@
 
+\begin{keywords}
+  dual-tree algorithms, branch and bound, nearest neighbor search, kernel
+density estimation, range search
+\end{keywords}
+
 \section{Dual-tree algorithms}
@@
 data structure with compelling theoretical qualities.  When cover trees are
-used, Dual-tree all-nearest-neighbor search and approximate kernel density
+used, dual-tree all-nearest-neighbor search and approximate kernel density
 estimation have $O(N)$ runtime guarantees for $O(N)$ queries \citep{ram2009};

### 2015-10-22 bounds-files/general-runtime-bound.tex
Fix typo.
Minor fixes + Dong's address update

@@
         \STATE $R' \gets \{ \mathscr{N}_r \in R :
-\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
+\texttt{Score(}\mathscr{N}_{qc}\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
 \label{alg:line:ct-dual-query-pruning}
@@
 separation invariant applies to both implicit and explicit representations of
-the tree, we conclude that $d(p_i, p_r) > 2^{s_r^{\max}} - 1$.  The same
+the tree, we conclude that $d(p_i, p_j) > 2^{s_r^{\max}} - 1$.  The same
 argument may be made for the case where $s_j < s_r^{\max} - 1$, with the same

### 2015-10-22 bounds-files/sec01-supporting-lemmas.tex
Minor fixes + Dong's address update

@@
 $\mu_i$ & Center of $\mathscr{N}_i$ \\ \hline
-$\lambda_i$ & Furthest descendant distance \\ \hline
+$\lambda_i$ & Furthest descendant distance from $\mu_i$\\ \hline
 \end{tabular}

### 2015-10-23 bounds-files/general-runtime-bound.tex
Update explanation.

@@
 statement of the lemma is true.
+
+Note that in this proof, we have considered the child reference set $R_{r - 1}$,
+not the original reference set $R$, and shown that with respect to $s_r^{\max}$
+as defined by $R$ (not $R_{r - 1}$), all nodes are separated by $2^{s_r^{\max} -
+1}$.  Then, in the frame of the next recursion where $R \gets R_{r - 1}$, the
+lemma will hold, as $s_r^{\max}$ will then be the maximum scale present in $R$.
 \end{proof}

### 2015-10-29 bounds.tex
Fix citation type.

@@
 inference \citep{amizadeh2012variational}, range search \citep{nbody}, and
-embedding techniques \cite{maaten2014accelerating}, to name a few.
+embedding techniques \citep{maaten2014accelerating}, to name a few.
 

### 2015-10-30 bounds-files/general-runtime-bound.tex
Particularly trivial revisions.

@@
 Now suppose that $s_i < s_r^{\max} - 1$.  This implies that there exists some
-implicit cover tree node with point $p_i$ and scale $s_r^{\max} - 1$ (as well an
-implicit child of this node $p_i$ with scale $s_r^{\max} - 2$ and so forth until
-one of these implicit nodes has child $p_i$ with scale $s_i$).  Because the
-separation invariant applies to both implicit and explicit representations of
-the tree, we conclude that $d(p_i, p_j) > 2^{s_r^{\max}} - 1$.  The same
+implicit cover tree node with point $p_i$ and scale $s_r^{\max} - 1$ (as well as
+an implicit child of this node $p_i$ with scale $s_r^{\max} - 2$ and so forth
+until one of these implicit nodes has child $p_i$ with scale $s_i$).  Because
+the separation invariant applies to both implicit and explicit representations
+of the tree, we conclude that $d(p_i, p_j) > 2^{s_r^{\max}} - 1$.  The same
 argument may be made for the case where $s_j < s_r^{\max} - 1$, with the same
@@
 and that $R \setminus R_{r - 1} \subseteq R$ in order to see that this condition
-holds for all nodes in $R' = R'_{r - 1} \cup (R \setminus R_{r - 1})$.
+holds for all nodes in $R'_{r - 1}$.
 
 Because we have shown that the condition holds for the initial reference set and
-for any reference set produced by a reference recursion, we have shown that the
-statement of the lemma is true.
+for any reference set produced by a reference recursion (which will be $R$ at
+some other level of recursion), we have shown that the statement of the lemma is
+true.
 
@@
 \begin{defn}
-Define $\theta$ as a bound on the number of extra reference recursions tha
+Define $\theta$ as a bound on the number of extra reference recursions that
 happen after the last query recursion.  Then,
@@
 \ref{alg:line:ct-dual-query-recursion-start}--\ref{alg:line:ct-dual-query-recursion-end}).
-The runtime of the algorithm is bounded as the runtime of a reference recursion
+The runtime of the algorithm is the runtime of a reference recursion
 times the total number of reference recursions plus the total runtime of all
@@
 
-When we consider the monochromatic case (where $S_q = S_r$)\footnote{In the
-monochromatic case, we do not take a point as its own nearest neighbor, so
-slight modification of {\tt BaseCase()} is necessary.  The runtime bound result
-remains unchanged.}, the results trivially simplify.
+When we consider the monochromatic case (where $S_q = S_r$), the results trivially simplify.
 
@@
 This results in a bound of the form $O(N \log (\eta_r / \delta_r))$, or
-$O(N^2)$ (See Lemma \ref{lem:extcase3}).  This is because the reference tree
-must be descended individually for each query point.
+$O(N^2)$ (see Lemma \ref{lem:extcase3}).  This is because the reference tree
+must be descended separately for each query point.
 
@@
 dual-tree algorithms rely on branch-and-bound techniques to prune away
-work (Lines \ref{alg:line:ct-dual-ref-score} and
+work (lines \ref{alg:line:ct-dual-ref-score} and
 \ref{alg:line:ct-dual-query-pruning} in Algorithm \ref{alg:cover-tree-dual}). A
@@
 intuition that answering $O(N)$ queries will take at least $O(N)$ time. For a
-particular problem and data, if $c_r$, $|R^*|$, $\chi$, $\psi$ are bounded by
+particular problem and data, if $c_r$, $|R^*|$, $\chi$, and $\psi$ are bounded by
 constants independent of $N$ and $\theta$ is no more than linear in $N$ (for

### 2015-10-30 bounds-files/imbalance.tex
Add higgs dataset.
Particularly trivial revisions.

@@
 
-\begin{figure}
+\begin{figure}[b!]
 \begin{subfigure}[b]{0.585\textwidth}
@@
 $\mathscr{N}_p$ (which has scale $s_p$).  If
-the node is a leaf child (that is, $s_i = -\infty$), then number of missing
+the node is a leaf (that is, $s_i = -\infty$), then the number of missing
 levels is defined as the difference between $s_p$ and $s_{\min} - 1$ where
@@
 and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
-that in the worst case the imbalance is quadratic in $N$\footnote{Note that in
-this situation, $c \sim N$ also.}.
+that in the worst case the imbalance is quadratic in $N$.\footnote{Note that in
+this situation, $c \sim N$ also.}
 
@@
 chosen.  The `power', `susy', `higgs', and `covertype' datasets are found in the
-UCI Machine Learning Repository \cite{ucimlrepository}, the `mnist' dataset is
+UCI Machine Learning Repository \citep{ucimlrepository}, the `mnist' dataset is
 from \cite{lecun2000mnist}, the `lcdm` and `sdss' datasets are Sloan Digital Sky
-Survey data \cite{adelman2008sixth}, and the `randu' dataset is
+Survey data \citep{adelman2008sixth}, and the `randu' dataset is
 randomly-generated uniformly-distributed data in 10 dimensions.  The imbalances
@@
 randu & 10 & 0.23 & 0.22 & 0.59 \\
-% higgs & 29 & 4963.6 & 84171.8 &  \\ need one more...
+%higgs & 29 & 4963.6 & 84171.8 & 781268.7 \\
+higgs & 29 & 0.99 & 1.68 & 1.56 \\
 %covertype & 54 & 6609.6 & 88296.4 & 1247820.7 \\

### 2015-10-30 bounds-files/sec02-nearest-neighbor-search.tex
Particularly trivial revisions.

@@
 standard bound function $B(\mathscr{N}_q)$ used for cover trees is adapted from
-\citep{langford2006}:
+\citet{langford2006}:
 
@@
 
-In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 (N +
+In the monochromatic case where $S_q = S_r$\footnote{In the monochromatic case,
+we do not take a point as its own nearest neighbor, so slight modification of
+{\tt BaseCase()} is necessary.  The runtime bound result remains unchanged.},
+the bound is $O(c^9 (N +
 I_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For

### 2015-10-31 bounds-files/sec03-approx-kde.tex
Minor fixes + Dong's address update
Fix minor typo.

@@
 been proposed for approximate kernel density estimation \citep[see, for
-instance][]{gray2003nonparametric}; however, this version is sufficient for our
+instance,][]{gray2003nonparametric}; however, this version is sufficient for our
 runtime analysis.  Real-world implementations, such as the one found in
@@
 \begin{thm}
-Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
+Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel with bandwidth $h$ satisfying the assumptions
 of the previous subsection.  Then, given a query set $S_q$ of size $O(N)$ and a

### 2015-10-31 submission-cover-letter.tex
Add submission cover letter.

\documentclass[twoside,11pt]{article}
\usepackage{jmlr2e}
\usepackage{hyperref}
\hypersetup{colorlinks, urlcolor=blue, citecolor=green, linkcolor=red}
\usepackage{url}

\usepackage{mathrsfs}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

\begin{document}

Dear editors and reviewers: \\

\medskip

Thank you for your time and consideration in reviewing our paper,
``Plug-and-play dual-tree algorithm runtime analysis''.  We are glad to see that
it has been accepted, and we have addressed the issues pointed out by each
reviewer.  We apologize for the long delay in our final submission; multiple
authors have changed jobs since the original submission, and one (the
corresponding author) was unfortunately particularly busy defending his thesis.
With those things now firmly in the past, we found the time to complete our
revisions and assemble this document (which is a sort of changelog from the
submission).  We'll address the concerns of each reviewer individually in the
following sections.

\medskip

\section{Reviewer 1}

\begin{quote}{\it
In particular, it seems non-trivial to derive similar bounds for other
efficient search structures like kd-trees, ball trees, or vantage point trees,
as these are not amenable to the notion of imbalancedness that the authors use
in their bounds.
}\end{quote}

This is absolutely true, and we have re-emphasized at the end of Section 1 that
these results are only applicable to cover trees.  To our knowledge nobody has
been able to prove better than worst-case runtime bounds on $kd$-trees or ball
trees or any other type of trees for exact nearest neighbor search.  Part of the
reason can be seen in the complexity of the cover tree: very specific structure
and traversal types are required for bounding the number of reference nodes
considered at any point during the traversal.  With the notion of the expansion
constant, at least, it is unclear how to do this for any type of tree other than
the cover tree (or direct derivatives thereof).

However, these bounds do support the empirical observations that dual-tree
algorithms with other types of trees (like the aforementioned $kd$-tree or ball
tree) do tend to scale linearly.  We've mentioned this in footnote 1 at the
bottom of page 3, providing some useful references.

\begin{quote}{\it
And more importantly, what is the performance of cover trees compared to these
other structures in practice?
}\end{quote}

Performance of the cover tree vs. the $kd$-tree has been shown to be about
comparable in a survey by Kibriya et~al.: ``An Empirical Comparison of Exact
Nearest Neighbour Search Algorithms'', 2007.  However, that paper only considers
the single-tree nearest neighbor search algorithm, which isn't what we are
considering in this paper.  Our own experiences (which are readily verifiable
with mlpack's {\tt allknn} program) show that cover trees perform comparably to
$kd$-trees on most datasets, and for larger dimensionalities tend to outperform
$kd$-trees.

\begin{quote}{\it
Can the authors present some empirical results on the actual tree imbalance
measured on real data sets, to allow the reader to gauge the importance of the
imbalancedness value in the O(N * imbalance) runtimes?
}\end{quote}

To this end, we've added Table 2, on page 9, which has some empirical
simulations.

\begin{quote}{\it
The authors state that "tree imbalance is near-linear with the number of
points", which would imply the run-time bounds behave more like O(N*N) in
practice?
}\end{quote}

The bound is of the form $O(N + i(\mathscr{T}_q))$, not $O(N i(\mathscr{T}_q))$,
so linear (or better) imbalance implies $O(N)$ scaling, not $O(N^2)$.

\section{Reviewer 2}

\begin{quote}{\it
1) It would be nice to have a better understanding of the relationship between
the expansion constant and possible values of these tree-dependent quantities.
Is there a similar amortized run time bound of the form poly(c)n?  Are there
examples where these tree-related quantities can be constant in n, while is c is
not?  Having such a discussion would be helpful for the reader.
}\end{quote}

We agree, but unfortunately it is unclear how to relate the imbalance to the
expansion constant despite quite some time trying.  We have elaborated on the
intuition behind the imbalance somewhat, showing (and describing) some
worst-case situations.

\begin{quote}{\it
The authors consider 1-nearest neighbor search, finding the closest point to
each query point.  For this problem, the monochromatic case (where the reference
set is the query set) is not very meaningful, so the simplified bound after
Theorem 2 is not very sensible.
}\end{quote}

We've added a footnote, number 6 on page 19, which describes the {\tt
BaseCase()} modification necessary to perform meaningful monochromatic nearest
neighbor search.  We apologize for the confusion.

\begin{quote}{\it
Does the result extend to k-nearest neighbor search?
}\end{quote}

Yes, the result can be re-derived for $k$-nearest neighbor search, but it
becomes significantly more complicated.  The proof strategy is basically
identical; the major change will be the bound function $B(\mathscr{N}_q)$.  For
brevity we chose to omit this extension.

\begin{quote}{\it
The statement of Theorem 2 is missing a closing ")" after the imbalance
notation.
}\end{quote}

Thank you for pointing this out; this is now fixed.

\begin{quote}{\it
It's not a good idea to have definitions inside lemmas, so I'd strongly suggest
to define theta outside Lemma 4.  For one thing, Lemma 4 defines theta under the
condition that both $S_q$ and $S_r$ are have $O(N)$ points while Theorem 2 is
stated more generally.  Speaking of Lemma 4, $S_q$ and $S_r$ are defined as
sets, so you should probably have $|S_q| = |S_r| = O(N)$ in the statement of the
lemma.
}\end{quote}

This is a good suggestion and we have rewritten the affected lemma, adding a
definition for $\theta$.

\section{Reviewer 3}

\begin{quote}{\it
This naively requires $O(N)$ time to answer [in/a] single query
}\end{quote}

Thanks; fixed.

\begin{quote}{\it
I feel that an illustration of a cover tree would be very helpful.
}\end{quote}

We have added a detailed illustration in Figure 1.  We hope this is helpful.

\begin{quote}{\it
Lemma 1: "every point in $C$ is separated by" is not clear enough.  Do you mean
that no two points can be closer than $\delta$?  That every point in $C$ is
separated from every other point in $C$ by exactly $\delta$?  Spell out what you
mean.
}\end{quote}

We have clarified this; our meaning was ``every two points in $C$ are separated
by at least $\delta$''.

\begin{quote}{\it
Lemma 1 proof: ... The definition of the expansion constant is an inequality.
How can
you assert the equality?
}\end{quote}

This was a typo, and should have been an inequality.  This has been fixed;
thanks for pointing it out.

\begin{quote}{\it
You are already using "$i$" for the scale "$s_i$", so I would urge you to
use another symbol for imbalance.
}\end{quote}

We switched our notation to use capital $I$ as a result; thank you for the
suggestion.

\begin{quote}{\it
pg 10: "aspect ratio"... please indicate somehow that this is a
technical definition of "aspect ratio", not the colloquial one.
}\end{quote}

We've clarified that by aspect ratio we mean the maximum pairwise distance
divided by the minimum pairwise distance.

\begin{quote}{\it
pg 10:
in "The minimum scale fo the query tree", you have a subscript $T_r$, not $T_t$.
}\end{quote}

Thanks, this is now fixed.

\begin{quote}{\it
pg 10: I find it strange that you are defining the "top" and "minimum"... why
not "top" and "bottom" or "maximum" and "minimum"?
}\end{quote}

The goal here was to avoid reusing the term ``maximum scale'', because that
already refers to the maximum scale in a set, $s_r^{\max}$.  However, the term
``bottom'' would imply the scale of the leaves of the tree, but we actually want
the minimum non-leaf scale, so we believe ``minimum'' is a better term to use
here despite the strangeness of combining ``top'' and ``minimum''.

\begin{quote}{\it
eqn 10: just use normal parens on your min and max functions.  Typeset them
bigger to show grouping.
}\end{quote}

Fixed.

\begin{quote}{\it
I don't like the notion of a "poorly-behaved dataset".  The data are
the data!  It is the performance of the data structures \& algorithms
given that dataset that is poor.  Please rephrase.
}\end{quote}

This is an excellent point.  We have refactored our discussion for
clarification: ``These datasets for which trees perform poorly \ldots''.

\begin{quote}{\it
pg 14:
"Hence, the runtime of any dual-tree algorithm would be at least $O(N)$
using our bound, which matches the intuition that answering $O(N)$
queries would take at least $O(N)$ time."

You can't say anything about the minimum amount of time an algorithm
would take when all you have is a big-O bound.
}\end{quote}

We aren't saying that $O(N)$ is a lower bound on the amount of time necessary to
return results for $N$ queries---instead we are saying that when the various
parameters are plugged into our bound, it yields something linear (or greater)
in $N$, and this matches an intuitive inspection of these dual-tree algorithms.
We hope this clarifies our statement somewhat.

\begin{quote}{\it
pg 14:
Rework the paragraphs so that you don't need to start a paragraph with $\nu$.
}\end{quote}

Fixed.

\section{Final notes}

We have reworked a few other sections, provided some more explanation of the
dual-tree traversal and how it works, qualitatively, and fixed some other
minor grammatical errors that we missed for the original submission.  We have
now checked over our final draft thoroughly, addressed the reviewer comments,
and our final submission is enclosed.

Again, thanks to everyone involved in the review of this paper.  We hope our
revisions are satisfactory and again apologize for our high latency final
revision.  If there are any further issues, please let us know and we will not
hesitate to get them taken care of (this time with less turnaround time).

\end{document}

### 2015-10-31 bounds-files/sec01-supporting-lemmas.tex
Final little changes requested by reviewers.

@@
 Consider a dataset $S$ with expansion constant $c$ and a subset $C \subseteq S$
-such that every point in $C$ is separated by $\delta$.  Then, for any point
-$p$ (which may or may not be in $S$), and any radius $\rho \delta > 0$:
+such that every two distinct points in $C$ are separated by at least $\delta$.
+Then, for any point $p$ (which may or may not be in $S$), and any radius $\rho
+\delta > 0$:
 \begin{equation}

### 2015-12-28 bounds.tex
Add editor.
Add acknowledgements.
Update headings and abstract formatting.
Use titlecase for section headings.

@@
 
-\title{Plug-and-play dual-tree algorithm runtime analysis}
+\title{Plug-and-Play Dual-Tree Algorithm Runtime Analysis}
 
@@
 
-\editor{Unknown}
+\editor{Nando de Freitas}
+
+\jmlrheading{12}{2015}{1-29}{1/15}{12/15}{R.R. Curtin, D. Lee, W.B. March, P. Ram}
+
+\ShortHeadings{Plug-and-Play Dual-Tree Algorithm Runtime Analysis}{Curtin, Lee, March,
+and Ram}
 
@@
 
-\begin{abstract}
+\begin{abstract}%
 %A large collection of machine learning algorithms contain subproblems that can
@@
 
-\section{Dual-tree algorithms}
+\section{Dual-tree Algorithms}
 
@@
 
-\section{Tree imbalance}
+\section{Tree Imbalance}
 \label{sec:imbalance}
@@
 
-\section{General runtime bound}
+\section{General Runtime Bound}
 \label{sec:bound}
@@
 
-\section{Nearest neighbor search}
+\section{Nearest Neighbor Search}
 \label{sec:nns}
@@
 
-\section{Approximate kernel density estimation}
+\section{Approximate Kernel Density Estimation}
 \label{sec:akde}
@@
 
-\section{Range search and range count}
+\section{Range Search and Range Count}
 \label{sec:rs}
@@
 
+\section*{Acknowledgements}
+
+The authors gratefully acknowledge the helpful and insightful comments of the
+anonymous reviewers.
+
 %\bibliographystyle{plain}

### 2015-12-28 bounds-files/imbalance.tex
JMLR style guide: "don't use utilize".

@@
 \citet{colless1982review}, and another is Sackin's index \citep{sackin1972good},
-but we aim to capture a different measure of imbalance that utilizes the leveled
+but we aim to capture a different measure of imbalance that uses the leveled
 structure of the cover tree.

### 2015-12-28 bounds-files/sec01-supporting-lemmas.tex
Use titlecase for section headings.
No parentheses in parenthetical citations.

@@
 
-\subsection{The cover tree}
+\subsection{The Cover Tree}
 
@@
 
-\subsection{Expansion constant}
+\subsection{Expansion Constant}
 
@@
   \item {\bf Width bound:} no cover tree node has more than $c^4$ children
-(Lemma 4.1, \cite{langford2006}).
+(Lemma 4.1, \citealp{langford2006}).
 
   \item {\bf Depth bound:} the maximum depth of any node is $O(c^2 \log N)$
-(Lemma 4.3, \cite{langford2006}).
+(Lemma 4.3, \citealp{langford2006}).
 
   \item {\bf Space bound:} a cover tree has $O(N)$ nodes (Theorem 1,
-\cite{langford2006}).
+\citealp{langford2006}).
 \end{itemize}
